### 2021-10-20 title-page.tex
Simplify template file

\begin{titlepage}
\begin{center}
\sffamily
\null\vspace{2cm}
\huge Abstractions for Type-Level Programming\\[24pt]
\textcolor{gray}{\small THIS IS A TEMPORARY TITLE PAGE\\
It will be replaced for the final print by a version\\[-14pt]
provided by the registrar's office.}
\vfill
\normalsize THÈSE N° XXXX (\the\year)\\[10pt]
PRÉSENTÉE LE \uppercase{\today}\\
À LA FACULTÉ INFORMATIQUE ET COMMUNICATIONS\\
LABORATOIRE DE MÉTHODES DE PROGRAMMATION\\
PROGRAMME DOCTORAL EN INFORMATIQUE ET COMMUNICATIONS\\[10pt]
ÉCOLE POLYTECHNIQUE FÉDÉRALE DE LAUSANNE\\
pour l'obtention du grade de Docteur ès Sciences\\
par\\[20pt]
\Large Olivier Blanvillain\\[20pt]
\small acceptée sur proposition du jury:\\[4pt]
Prof Name Surname, président du jury\\
Prof Name Surname, directeur de thèse\\
Prof Name Surname, rapporteur\\
Prof Name Surname, rapporteur\\
Prof Name Surname, rapporteur\\[12pt]
Lausanne, EPFL, \the\year
\end{center}
\vspace{2cm}
\end{titlepage}

### 2021-10-21 proofs/canonical-forms.tex
Import FM proof

\begin{lemma}[Canonical forms]
  \lbl{lem:canonical-forms}~
  \begin{enumerate}
    \item % 1
    If $\Γ \⊢ \t \: \T$,
    where either $\T$ is a type variable,
    or $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$,
    then $\t$ is not a closed value.

    \item % 2
    If $\v$ is a closed value with $\Γ \⊢ \v \: \T$
    where either $\T \= \T_{1} \→ \T_{2}$,
    or $\T$ is a match type and $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$,
    then $\v$ has the form $\λ \x \: \S_{1}.~ \t_{2}$.

    \item % 3
    If $\v$ is a closed value with $\Γ \⊢ \v \: \T$
    where either $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$,
    or $\T$ is a match type and $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$,
    then $\v$ has the form $\λ \X \<: \U_{1}.~ \t_{2}$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  \begin{enumerate}
    \item % 1
    By induction on a derivation of $\Γ \⊢ \t \: \T$
    \begin{itemize}
      \Case\TVar, \TTApp, \TApp, \TMatch:

      In those cases, $\t$ is not a value and the result is immediate.

      \Case\TAbs, \TTAbs, \TClass:

      In those cases, $\T$ is neither a match type nor a type variable and the result is immediate.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      By \Cref{lem:inversion-of-subtyping}, either $\S$ is a match type with $\Γ \⊢ \S \⇌ \Y$, or $\S$ is a type variable.
      In both cases, we use the IH to show that $\t$ is not a closed value, as required.
    \end{itemize}

    \item % 2
    By induction on a derivation of $\Γ \⊢ \t \: \T$.

    \begin{itemize}
      \Case\TVar, \TTApp, \TApp, \TMatch:

      In those cases, $\t$ is not a value and the result is immediate.

      \Case\TAbs:
      \quad $\t \= \λ \x \: \T_{1} \t_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \quad $\Γ, \x \: \T_{1} \⊢ \t_{2} \: \T_{2}$

      $\t \= \λ \x \: \T_{1} \t_{2}$ and the result is immediate.

      \Case\TTAbs, \TClass:

      $\T$ is neither a function type nor a match type and the result is immediate.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using \Cref{lem:inversion-of-subtyping} we get 4 subcases:
      \begin{enumerate}
        \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \S_{1} \→ \S_{2}$:\\
        The result follows from the IH.

        \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \X$:\\
        This case cannot occure since the 1th part of the lemma would lead to a contradiction on the fact that $\v$ is a closed value.

        \item $\S$ is a type variable:\\
        Similarly, this case cannot occure since the 1st part of the lemma would lead to a contradiction.

        \item $\S$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:\\
        The result follows from the IH.
      \end{enumerate}
    \end{itemize}

    \item % 3
    By induction on a derivation of $\Γ \⊢ \t \: \T$.

    \begin{itemize}
      \Case\TVar, \TTApp, \TApp, \TMatch:

      In those cases, $\t$ is not a value and the result is immediate.

      \Case\TAbs, \TClass:

      $\T$ is neither a universal type nor a match type and the result is immediate.

      \Case\TTAbs:
      \quad $\t \= \λ \Y \<: \T_{1}.~ \t_{2}$
      \quad $\T \= \∀ \Y \<: \T_{1}.~ \T_{2}$
      \quad $\Γ, \Y \<: \T_{1} \⊢ \t_{2} \: \T_{2}$

      $\t \= \λ \Y \<: \T_{1}.~ \t_{2}$ and the result is immediate.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using \Cref{lem:inversion-of-subtyping} we get 4 subcases:
      \begin{enumerate}
        \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$,
        The result follows from the IH.

        \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \X$,
        This case cannot occure since the 1th part of the lemma would lead to a contradiction on the fact that $\v$ is a closed value.

        \item $\S$ is a type variable,
        Similarly, this case cannot occure since the 1st part of the lemma would lead to a contradiction.

        \item $\S$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ and $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$.
        The result follows from the IH.
      \end{enumerate}
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/disjointness-subtyping-exclusivity.tex
Import FM proof

\begin{lemma}[Disjointness/subtyping exclusivity]
  \lbl{lem:disjointness-subtyping-exclusivity}
  ~\\\indent
  The type disjointness and subtyping relations are mutually exclusive.
\end{lemma}

\begin{proof}
  We prove mutual exclusivity of type disjointness and subtyping by first defining $\⟦\· \⟧_Γ$, a mapping from \SystemFm types (in a given context $\Γ$) into non-empty subsets of a newly defined set $\P$. We then show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness relation in $\P$.

  This set-theoretical view of subtyping and disjointness renders the proof trivial.
  Indeed, suppose there exist two types $\S$ and $\T$ with $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \disj(\S, \T)$.
  $\⟦ \S \⟧_Γ$ and $\⟦ \T \⟧_Γ$ are two non-empty sets which are both intersecting and disjoint, a contradiction.

  We first define $\P$ as $\P \= \{ \Λ, \V \} \∪ \C$. Elements of $\P$ can be understood as equivalence classes for \SystemFm values: $\Λ$ corresponds to all abstraction values, $\V$ corresponds to type abstraction value, and elements of $\C$ correspond to their respective constructors.
  The definition of $\⟦\· \⟧_Γ$ makes this correspondence apparent.

  We define $\⟦\· \⟧_Γ$, a mapping of \SystemFm types into subsets of $\P$ in a given context $\Γ$:

  \begin{align*}
    \text{$\⟦ \Top \⟧_Γ$} &=
    \text{$\P$}\\
    %
    \text{$\⟦ \X \⟧_Γ$} &=
    \text{$\⟦ \T \⟧_Γ \quad$ where $\X \<: \T \∈ \Γ$}\\
    %
    \text{$\⟦ \T_{1} \→ \T_{2} \⟧_Γ$} &=
    \text{$\{ \Λ \}$}\\
    %
    \text{$\⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_Γ$} &=
    \text{$\{ \V \}$}\\
    %
    \text{$\⟦ \{ \new \C_{1} \} \⟧_Γ$} &=
    \text{$\{ \C_{1} \}$}\\
    %
    \text{$\⟦ \C_{1} \⟧_Γ$} &=
      \{ \c \∈ \C~|~(\c, \C_1) \∈ \Ψ \}\\
    \⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_\Γ &=
      \begin{cases}
        \text{$\⟦ \T_n \⟧_Γ$\phantom{\P} if $\⟦ \T_s \⟧_Γ \⊂ \⟦ \S_n \⟧_Γ$}\\
        \phantom{\text{$\⟦ \T_n \⟧_Γ$}}\phantom{\P}\text{ and $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$}\\
        \text{$\⟦ \T_d \⟧_Γ$\phantom{\P} if $\∀ \m.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$}\\
        \P\phantom{\text{$\⟦ \T_n \⟧_Γ$}}\text{ otherwise}
      \end{cases}
  \end{align*}

  We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
  In other words, we prove the follows statements:

  \begin{enumerate}
    \item $\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$,
    \item $\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$,
  \end{enumerate}

  Both statements are proved simultaneously by induction on derivations of $\Γ \⊢ \disj(\V, \W)$ and $\Γ \⊢ \S \<: \T$.
  More precisely, the induction is done on the cumulative depth of both derivation tree.
  \begin{enumerate}
    \item ($\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$)
    \begin{itemize}
      \Case\SRefl:
      \quad $\T \= \S$

      $\⟦ \S \⟧_Γ \= \⟦ \T \⟧_Γ$ and the result is immediate.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      By the IH, $\⟦ \S \⟧_Γ \⊂ \⟦ \U \⟧_Γ$ and $\⟦ \U \⟧_Γ \⊂ \⟦ \T \⟧_Γ$. From subset transitivity, $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$, as required.

      \Case\STop:
      \quad $\T \= \Top$

      $\⟦ \Top \⟧_Γ \= \P$ coincides with the codomain of $\⟦\· \⟧_Γ$.
      Therefore, for any type $\T$, $\⟦ \T \⟧_Γ \⊂ \⟦ \Top \⟧_Γ$, as required.

      \Case\SSin:
      \quad $\S \= \{ \new \C_{1} \}$
      \quad $\T \= \C_{1}$

      $\⟦ \{ \new \C_{1} \} \⟧_Γ \= \{ \C_{1} \}$ and
      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$.
      The result follows by reflexivity of $\Ψ$.

      \Case\STvar:
      \quad $\S \= \X$
      \quad $\X \<: \T \∈ \Γ$

      By definition $\⟦ \X \⟧_Γ \= \⟦ \T \⟧_Γ$, and the result is immediate.

      \Case\SArrow:
      \quad $\S \= \S_{1} \→ \S_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \\\phantom{\textit{Case}~\SArrow:}
      \quad $\Γ \⊢ \T_{1} \<: \S_{1}$
      \quad $\Γ \⊢ \S_{2} \<: \T_{2}$

      $\⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \⟦ \T_{1} \→ \T_{2} \⟧_Γ \= \{ \Λ \}$ and the result is immediate.

      \Case\SAll:
      \quad $\S \= \∀ \X \<: \U_{1}.~ \S_{2}$
      \quad $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$
      \quad $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$

      $\⟦ \∀ \X \<: \U_{1}.~ \S_{2} \⟧_Γ \= \⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_Γ \= \{ \V \}$ and the result is immediate.

      \Case\SPsi:
      \quad $\S \= \C_{1}$
      \quad $\T \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ψ$

      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
      By transitivity of $\Ψ$, $\⟦ \C_{1} \⟧_Γ \⊂ \⟦ \C_{2} \⟧_Γ$, as required.

      \Case\SMatch1/2:
      \quad $\T_{1} \= \T_n$
      \quad $\T_{2} \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch1/2:}
      \quad $\Γ \⊢ \T_s \<: \S_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \S_m)$

      \hfuzz=1.5pt
      Using the IH we get $\⟦ \T_s \⟧_Γ \⊂ \⟦ \S_n \⟧_Γ$, $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$ and $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_Γ \= \⟦ \T_n \⟧_Γ$.
      This last equality implies both $\⟦ \T_{1} \⟧_Γ \⊂ \⟦ \T_{2} \⟧_Γ$ and $\⟦ \T_{2} \⟧_Γ \⊂ \⟦ \T_{1} \⟧_Γ$, as required.

      \hfuzz=0pt
      \Case\SMatch3/4:
      \quad $\T_{1} \= \T_d$
      \quad $\T_{2} \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch3/4:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \S_n)$

      By the IH $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_n \⟧_Γ \= \{ \}$.
      By definition, $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_Γ \= \⟦ \T_d \⟧_Γ$.
      This equality implies both $\⟦ \T_{1} \⟧_Γ \⊂ \⟦ \T_{2} \⟧_Γ$ and $\⟦ \T_{2} \⟧_Γ \⊂ \⟦ \T_{1} \⟧_Γ$, the desired result for \SMatch3 and \SMatch4 respectively.

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$
      \quad $\Γ \⊢ \S_d \<: \T_d$

      By case analysis on $\⟦ \T \⟧_Γ$:
      \begin{enumerate}
        \item $\⟦ \T \⟧_Γ \= \⟦ \T_n \⟧_Γ$
        and $\⟦ \T_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$
        and $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$:

        The IH gives $\⟦ \S_s \⟧_Γ \⊂ \⟦ \T_s \⟧_Γ$.
        Using $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$, we get $\∀ \m<n.~ \⟦ \S_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$.
        Using $\⟦ \T_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$, we get $\⟦ \S_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$ (by subset transitivity).
        Therefore, $\⟦ \S \⟧_Γ \= \⟦ \S_n \⟧_Γ$, and the result follows from the IH.

        \item $\⟦ \T \⟧_Γ \= \⟦ \T_d \⟧_Γ$ and
        $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$:

        The IH gives $\⟦ \S_s \⟧_Γ \⊂ \⟦ \T_s \⟧_Γ$.
        Using $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$ we get $\∀ \n.~ \⟦ \S_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$.
        Therefore, $\⟦ \S \⟧_Γ \= \⟦ \S_d \⟧_Γ$, and the result follows from the IH.

        \item $\⟦ \T \⟧_Γ \= \P$:

        $\P$ is the codomain of $\⟦\· \⟧_Γ$, therefore $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$, as required.
      \end{enumerate}
    \end{itemize}

    \item ($\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$)
    \begin{itemize}
      \Case\DXi:
      \quad $\S \= \C_{1}$
      \quad $\T \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ξ$

      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
      By definition of Ξ, there is no class $\c$ such that both $\(\c, \C_{1}) \∈ \Ψ$ and $\(\c, \C_{2}) \∈ \Ψ$, and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \C_{2} \⟧_Γ \= \{ \}$.

      \Case\DPsi:
      \quad $\S \= \{ \new \C_{1} \}$
      \quad $\T \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∉ \Ψ$

      $\⟦ \{ \new \C_{1} \} \⟧_Γ \= \C_{1}$ and
      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
      Since $\(\C_{1}, \C_{2}) \∉ \Ψ$, $\C_{1} \∉ \⟦ \C_{2} \⟧_Γ$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \C_{2} \⟧_Γ \= \{ \}$.

      \Case\DSub:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \disj(\U, \T)$

      By the IH, $\⟦ \S \⟧_Γ \⊂ \⟦ \U \⟧_Γ$ and $\⟦ \U \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$.
      $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$ follows directly.

      \Case\DArrow:
      \quad $\S \= \S_{1} \→ \S_{2}$
      \quad $\T \= \C_{1}$

      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \{ \Λ \}$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \{ \}$, as required.

      \Case\DAll:
      \quad $\S \= \∀ \X \<: \S_{1}.~ \S_{2}$
      \quad $\T \= \C_{1}$

      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_Γ \= \{ \S \}$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_Γ \= \{ \}$, as required.
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/inversion-of-subtyping.tex
Import FM proof

\begin{lemma}[Inversion of subtyping]
  \lbl{lem:inversion-of-subtyping}~
  \begin{enumerate}
    \item % 0
    If $\Γ \⊢ \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d \⇌ \T$, then either:
    \begin{enumerate}
      \item
        $\Γ \⊢ \S_s \<: \U_n$,
        $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$
        and $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \T$,
      \item
        $\Γ \⊢ \S_s \<: \U_n$,
        $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$
        and $\S_n \= \T$,
      \item
        $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$
        and $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \T$,
      \item
        $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$
        and $\S_d \= \T$.
    \end{enumerate}

    \item % 1
    If $\Γ \⊢ \S \<: \X$, or $\Γ \⊢ \S \<: \T$ where $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$, then either
    \begin{enumerate}
      \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \Y$, for some $\Y$,
      \item $\S$ is a type variable.
    \end{enumerate}

    \item % 2
    If $\Γ \⊢ \S \<: \T_{1} \→ \T_{2}$, or $\Γ \⊢ \S \<: \T$ where $\T$ is a match type with $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, then either
    \begin{enumerate}
      \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \S_{1} \→ \S_{2}$, for some $\S_{1}$, $\S_{2}$ such that $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$,
      \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \X$, for some $\X$,
      \item $\S$ is a type variable,
      \item $\S$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$.
    \end{enumerate}

    \item % 3
    If $\Γ \⊢ \S \<: \∀ \X \<: \U_{1}.~ \T_{2}$, or $\Γ \⊢ \S \<: \T$ where $\T$ is a match type with $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, then either
    \begin{enumerate}
      \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$, for some $\S_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$,
      \item $\S$ is a match type with $\Γ \⊢ \S \⇌ \X$, for some $\X$,
      \item $\S$ is a type variable,
      \item $\S$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ and $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$.
    \end{enumerate}
  \end{enumerate}
\end{lemma}

\begin{proof}
  \begin{enumerate}
    \item % 0
    By induction on a derivation on $\Γ \⊢ \S \⇌ \T$.
    \begin{itemize}
      \Case\SMatch1/2:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \S_n$
      \\\phantom{\textit{Case}~\SMatch1/2:}
      \quad $\Γ \⊢ \S_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$

      $\S_n \= \T$ and the result follows directly from the premises.

      \Case\SMatch3/4:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \S_d$
      \quad $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$

      $\S_d \= \T$ and the result follows directly from the premises.

      \Case\STrans:
      \quad $\Γ \⊢ \S \⇌ \U$
      \quad $\Γ \⊢ \U \⇌ \T$

      By definition of the $\⇌$ relation, $\Γ \⊢ \U \⇌ \T$ implies that $\U$ is a match type.
      Using the IH on the left premise we get 4 subcases:
      \begin{enumerate}
        \item
          $\Γ \⊢ \S_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$
          and $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \U$:

        Using \STrans we get $\Γ \⊢ \S_n \⇌ \T$, as requiered.

        \item
          $\Γ \⊢ \S_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$
          and $\S_n \= \U$:

        The result is immediate.

        \item
          $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$
          and $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \U$:

        Using \STrans we get $\Γ \⊢ \S_d \⇌ \T$, as requiered.

        \item
          $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$
          and $\S_d \= \U$:

        The result is immediate.
      \end{enumerate}
    \end{itemize}

    \item % 1
    By induction on a derivation of $\Γ \⊢ \S \<: \T$.
    \begin{itemize}
      \Case\SRefl:
      \quad $\S \= \T$

      If $\T$ is a type variable then so is $\S$ and the result is immediate.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$, $\S$ is a match type and $\Γ \⊢ \S \⇌ \Y$, as required.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      If $\T$ is a type variable we use the IH on the right premise to get that $\U$ is either a match type with $\Γ \⊢ \U \⇌ \X$ or a type variable.
      In either case, the result follows from using the IH on the left premise.

      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$ we can also use the the IH on the right premise to get to the same result.

      \Case\STvar:
      \quad $\S \= \Y$
      \quad $\Y \<: \X \∈ \Γ$

      $\S$ is a type variable and the result is immediate.

      \Case\SMatch1:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_n$
      \\\phantom{\textit{Case}~\SMatch1:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      If $\T \= \X$, we use \SMatch2 to obtain $\Γ \⊢ \S \⇌ \X$, as required.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$, we use \SMatch2 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \X$, as required.

      \Case\SMatch2:
      \quad $\S \= \T_n$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch2:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \X$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \X$, we get 4 subcases:
      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \X$:

        $\S \= \T_n$ is a match type and $\Γ \⊢ \S \⇌ \X$, as required.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \X$:

        $\S \= \T_n \= \X$ is a type variable and the result is immediate.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \X$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \X$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.
      \end{enumerate}

      \Case\SMatch3:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      If $\T \= \X$, we use \SMatch4 to obtain $\Γ \⊢ \S \⇌ \X$, as required.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \X$, we use \SMatch4 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \X$, as required.

      \Case\SMatch4:
      \quad $\S \= \T_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \X$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \X$, we get 4 subcases:
      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \X$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \X$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \X$:

        $\S$ is a match type and $\Γ \⊢ \S \⇌ \X$, as required.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \X$:

        $\S \= \T_d \= \X$ is a type variable and the result is immediate.
      \end{enumerate}

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$
      \quad $\Γ \⊢ \S_d \<: \T_d$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \X$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \X$, we get 4 subcases:
      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \X$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we obtain $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \T_n$ and $\Γ \⊢ \T_n \⇌ \X$, we use the IH to get that either $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \Y$, or $\S_n$ is a type variable.
        If $\S_n$ is a type variable, we already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        If $\Γ \⊢ \S_n \⇌ \Y$, then using \STrans with $\Γ \⊢ \S \⇌ \S_n$ gives $\Γ \⊢ \S \⇌ \Y$, as required.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \X$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we obtain $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \X$, we use the IH to get that either $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \Y$, or $\S_n$ is a type variable.
        If $\S_n$ is a type variable, we already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        If $\Γ \⊢ \S_n \⇌ \Y$, then using \STrans with $\Γ \⊢ \S \⇌ \S_n$ gives $\Γ \⊢ \S \⇌ \Y$, as required.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \X$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \⇌ \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$ and $\Γ \⊢ \T_d \⇌ \X$, we use the IH to get that either $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \Y$, or $\S_d$ is a type variable.
        If $\S_d$ is a type variable, we already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        If $\Γ \⊢ \S_d \⇌ \Y$, then using \STrans with $\Γ \⊢ \S \⇌ \S_d$ gives $\Γ \⊢ \S \⇌ \Y$, as required.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \X$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \<: \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$ and $\Γ \⊢ \T_d \⇌ \X$, we use the IH to get that either $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \Y$, or $\S_d$ is a type variable.
        If $\S_d$ is a type variable, we already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        If $\Γ \⊢ \S_d \⇌ \Y$, then using \STrans with $\Γ \⊢ \S \⇌ \S_d$ gives $\Γ \⊢ \S \⇌ \Y$, as required.
      \end{enumerate}

      \Case\STop, \SSin, \SArrow, \SAll, \SPsi:

      In those cases, $\T$ is neither a type variable nor a match type and the result is immediate.
    \end{itemize}

    \item % 2
    By induction on a derivation of $\Γ \⊢ \S \<: \T$.
    \begin{itemize}
      \Case\SRefl:
      \quad $\T \= \S$

      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$ and the result follows directly from \SRefl.
      If $\T \= \T_{1} \→ \T_{2}$, the result also follows directly from \SRefl.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      Using the IH on the right premise we get 4 subcases:
      \begin{enumerate}
        \item $\U$ is a match type with $\Γ \⊢ \U \⇌ \U_{1} \→ \U_{2}$, for some $\U_{1}$, $\U_{2}$ such that $\Γ \⊢ \T_{1} \<: \U_{1}$ and $\Γ \⊢ \U_{2} \<: \T_{2}$:

        The result follows directly from using the IH on the left premise ($\Γ \⊢ \T_{1} \<: \S_{1}$ is obtained using \STrans with $\Γ \⊢ \T_{1} \<: \U_{1}$ and $\Γ \⊢ \U_{1} \<: \S_{1}$, $\Γ \⊢ \S_{2} \<: \T_{2}$ is obtained analogously).

        \item $\U$ is a match type with $\Γ \⊢ \U \⇌ \X$, for some $\X$:

        Using the 2nd part of the lemma on the left premise leads to the desired result.

        \item $\U$ is a type variable:

        The result follows from using the 2nd part of the lemma on the left premise.

        \item $\U$ has the form $\U_{1} \→ \U_{2}$ with $\Γ \⊢ \T_{1} \<: \U_{1}$ and $\Γ \⊢ \U_{2} \<: \T_{2}$:

        The result follows directly from using the IH on the left premise ($\Γ \⊢ \T_{1} \<: \S_{1}$ is obtained using \STrans with $\Γ \⊢ \T_{1} \<: \U_{1}$ and $\Γ \⊢ \U_{1} \<: \S_{1}$, $\Γ \⊢ \S_{2} \<: \T_{2}$ is obtained analogously).
      \end{enumerate}

      \Case\SMatch1:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_n$
      \\\phantom{\textit{Case}~\SMatch1:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      If $\T \= \T_{1} \→ \T_{2}$, we use \SMatch2 to obtain $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, as required.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, we use \SMatch2 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, and the result follows from \SRefl.

      \Case\SMatch2:
      \quad $\S \= \T_n$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch2:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \T_{1} \→ \T_{2}$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, we get 4 subcases:

      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \T_{1} \→ \T_{2}$:

        $\S \= \T_n$ is a match type and $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, and the result follows from \SRefl.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \T_{1} \→ \T_{2}$:

        $\S \= \T_{1} \→ \T_{2}$, $\S_{1} \= \T_{1}$, $\S_{2} \= \T_{2}$, and the result follows from \SRefl.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \T_{1} \→ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \T_{1} \→ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.
      \end{enumerate}

      \Case\SMatch3:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      If $\T \= \T_{1} \→ \T_{2}$, we use \SMatch2 to obtain $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, and the result follows from \SRefl.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, we use \SMatch2 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, and the result follows from \SRefl.

      \Case\SMatch4:
      \quad $\S \= \T_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \T_{1} \→ \T_{2}$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, we get 4 subcases:

      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \T_{1} \→ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \T_{1} \→ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \T_{1} \→ \T_{2}$:

        $\S \= \T_d$ is a match type and $\Γ \⊢ \S \⇌ \T_{1} \→ \T_{2}$, and the result follows from \SRefl.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \T_{1} \→ \T_{2}$:

        $\S \= \T_{1} \→ \T_{2}$, $\S_{1} \= \T_{1}$, $\S_{2} \= \T_{2}$, and the result follows from \SRefl.
      \end{enumerate}

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$
      \quad $\Γ \⊢ \S_d \<: \T_d$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \T_{1} \→ \T_{2}$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \T_{1} \→ \T_{2}$, we get 4 subcases:
      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \T_{1} \→ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we obtain $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \T_n$ and $\Γ \⊢ \T_n \⇌ \T_{1} \→ \T_{2}$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \S_{1} \→ \S_{2}$ for some $\S_{1}$, $\S_{2}$ such that $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_n$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.

          \item $\S_n$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        \end{enumerate}

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \T_{1} \→ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we obtain $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \T_n$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \S_{1} \→ \S_{2}$ for some $\S_{1}$, $\S_{2}$ such that $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_n$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.

          \item $\S_n$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        \end{enumerate}

        \item
          $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \T_{1} \→ \T_{2}$:

        Using \DSub on $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$ and $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$ we obtain $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \C_d$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \⇌ \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$ and $\Γ \⊢ \T_d \⇌ \T_{1} \→ \T_{2}$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \S_{1} \→ \S_{2}$ for some $\S_{1}$, $\S_{2}$ such that $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_d$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.

          \item $\S_d$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        \end{enumerate}

        \item
          $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$
          and $\T_d \= \T_{1} \→ \T_{2}$:

        Using \DSub on $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$ and $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$ we obtain $\∀ \n.~ \Γ \⊢ \disj(\S_s, \U_n)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \C_d$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \⇌ \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \S_{1} \→ \S_{2}$ for some $\S_{1}$, $\S_{2}$ such that $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_d$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.

          \item $\S_d$ has the form $\S_{1} \→ \S_{2}$ with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        \end{enumerate}
      \end{enumerate}

      \Case\STvar:
      \quad $\S \= \Y$
      \quad $\Y \<: \T \∈ \Γ$

      $\S$ is a type variable and the result is immediate.

      \Case\SArrow:
      \quad $\S \= \S_{1} \→ \S_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \\\phantom{\textit{Case}~\SArrow:}
      \quad $\Γ \⊢ \T_{1} \<: \S_{1}$
      \quad $\Γ \⊢ \S_{2} \<: \T_{2}$

      $\S$ has the form $\S_{1} \→ \S_{2}$, with $\Γ \⊢ \T_{1} \<: \S_{1}$ and $\Γ \⊢ \S_{2} \<: \T_{2}$, as required.

      \Case\STop, \SSin, \SAll, \SPsi:

      In those cases, $\T$ is neither a type variable nor a function type and the result is immediate.
    \end{itemize}

    \item % 3
    By induction on a derivation of $\Γ \⊢ \S \<: \T$.
    \begin{itemize}
      \Case\SRefl:
      \quad $\T \= \S$

      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$ the result follows directly from \SRefl.
      If $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$, $\S_{2} \= \T_{1}$, and the result also follows from \SRefl.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      Using the IH on the right premise we get 4 subcases:
      \begin{enumerate}
        \item $\U$ is a match type with $\Γ \⊢ \U \⇌ \U_{1} \→ \U_{2}$ for some $\U_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \U_{2} \<: \T_{2}$:

        The result follows directly from using the IH on the left premise ($\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$ is obtained using \STrans with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \U_{2}$ and $\Γ, \X \<: \U_{1} \⊢ \U_{2} \<: \T_{2}$).

        \item $\U$ is a match type with $\Γ \⊢ \U \⇌ \X$:

        Using the 2nd part of the lemma on the left premise leads to the desired result.

        \item $\U$ is a type variable:

        The result follows from using the 2nd part of the lemma on the left premise.

        \item $\U$ has the form $\∀ \X \<: \U_{1}.~ \U_{2}$ with $\Γ, \X \<: \U_{1} \⊢ \U_{2} \<: \T_{2}$:

        The result follows directly from using the IH on the left premise ($\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$ is obtained using \STrans with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \U_{2}$ and $\Γ, \X \<: \U_{1} \⊢ \U_{2} \<: \T_{2}$).
      \end{enumerate}

      \Case\SMatch1:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_n$
      \\\phantom{\textit{Case}~\SMatch1:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      If $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$, we use \SMatch2 to obtain $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, as required, and the result follows from \SRefl.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, we use \SMatch2 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, and the result follows from \SRefl.

      \Case\SMatch2:
      \quad $\S \= \T_n$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch2:}
      \quad $\Γ \⊢ \T_s \<: \U_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \(\∀ \X \<: \U_{1}.~ \T_{2})$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, we get 4 subcases:

      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        $\S \= \T_n$ is a match type and $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, and the result follows from \SRefl.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        $\S \= \∀ \X \<: \U_{1}.~ \T_{2}$, $\S_{2} \= \T_{2}$, and the result follows from \SRefl.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.
      \end{enumerate}

      \Case\SMatch3:
      \quad $\S \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      If $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$, we use \SMatch2 to obtain $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, and the result follows from \SRefl.
      If $\T$ is a match type with $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, we use \SMatch2 to get $\Γ \⊢ \S \⇌ \T$, and \STrans to obtain $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, and the result follows from \SRefl.

      \Case\SMatch4:
      \quad $\S \= \T_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch3:}
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \U_n)$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \(\∀ \X \<: \U_{1}.~ \T_{2})$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, we get 4 subcases:

      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        This case cannot occur since $\Γ \⊢ \disj(\T_s, \U_n)$ would contradict $\Γ \⊢ \T_s \<: \U_n$, by \Cref{lem:disjointness-subtyping-exclusivity}.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        $\S \= \T_d$ is a match type and $\Γ \⊢ \S \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, and the result follows from \SRefl.

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        $\S \= \∀ \X \<: \U_{1}.~ \T_{2}$, $\S_{2} \= \T_{2}$, and the result follows from \SRefl.
      \end{enumerate}

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$
      \quad $\Γ \⊢ \S_d \<: \T_d$

      In this case the first premise of the statement ($\Γ \⊢ \S \<: \(\∀ \X \<: \U_{1}.~ \T_{2})$) does not apply since $\T$ is a match type.
      Using the 1st part of the lemma on $\Γ \⊢ \T \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, we get 4 subcases:

      \begin{enumerate}
        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n$ is a match type with $\Γ \⊢ \T_n \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we get $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \T_n$ and $\Γ \⊢ \T_n \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$ for some $\S_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_n$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.

          \item $\S_n$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        \end{enumerate}

        \item
          $\Γ \⊢ \T_s \<: \U_n$,
          $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_n \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \U_m)$ we get $\∀ \m<n.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \U_n$.
        Using \SMatch1/2 we get $\Γ \⊢ \S \⇌ \S_n$.
        Since $\Γ \⊢ \S_n \<: \T_n$, the IH, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$ for some $\S_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          The result follows directly from \STrans.

          \item $\S_n$ is a match type with $\Γ \⊢ \S_n \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_n$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.

          \item $\S_n$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_n$, as required.
        \end{enumerate}

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d$ is a match type with $\Γ \⊢ \T_d \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \C_d$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \⇌ \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$ and $\Γ \⊢ \T_d \⇌ \∀ \X \<: \U_{1}.~ \T_{2}$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$ for some $\S_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$::

          The result follows directly from \STrans.

          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_d$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.

          \item $\S_d$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        \end{enumerate}

        \item
          $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$
          and $\T_d \= \∀ \X \<: \U_{1}.~ \T_{2}$:

        Using \DSub on $\∀ \m.~ \Γ \⊢ \S_m \<: \T_m$ and $\∀ \m.~ \Γ \⊢ \disj(\T_s, \U_m)$ we obtain $\∀ \m.~ \Γ \⊢ \disj(\S_s, \U_m)$.
        From \STrans we get $\Γ \⊢ \S_s \<: \C_d$.
        Using \SMatch3/4 we obtain $\Γ \⊢ \S \⇌ \S_d$.
        Since $\Γ \⊢ \S_d \<: \T_d$, the IH gives 4 subsubcases:

        \begin{enumerate}
          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \∀ \X \<: \U_{1}.~ \S_{2}$ for some $\S_{2}$ such that $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$::

          The result follows directly from \STrans.

          \item $\S_d$ is a match type with $\Γ \⊢ \S_d \⇌ \X$:

          The result follows directly from \STrans.

          \item $\S_d$ is a type variable:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.

          \item $\S_d$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$ with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$:

          We already proved $\Γ \⊢ \S \⇌ \S_d$, as required.
        \end{enumerate}
      \end{enumerate}

      \Case\STvar:
      \quad $\S \= \Y$
      \quad $\Y \<: \T \∈ \Γ$

      $\S$ is a type variable and the result is immediate.

      \Case\SAll:
      \quad $\S \= \∀ \X \<: \U_{1}.~ \S_{2}$
      \quad $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$
      \\\phantom{\textit{Case}~\SArrow:}
      \quad $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$

      $\S$ has the form $\∀ \X \<: \U_{1}.~ \S_{2}$, with $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$, as required.

      \Case\STop, \SSin, \SArrow, \SPsi:

      In those cases, $\T$ is neither a type variable nor a universal type and the result is immediate.
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/inversion-of-typing.tex
Import FM proof

\begin{lemma}[Inversion of typing]
  \lbl{lem:inversion-of-typing}~
  \begin{enumerate}
    \item % 1
    If $\Γ \⊢ \λ \x \: \S_{1}.~ \s_{2} \: \T$ and $\Γ \⊢ \T \<: \U_{1} \→ \U_{2}$,
    then $\Γ \⊢ \U_{1} \<: \S_{1}$ and there is some $\S_{2}$ such that
    $\Γ, \x \: \S_{1} \⊢ \s_{2} \: \S_{2}$ and $\Γ \⊢ \S_{2} \<: \U_{2}$.

    \item % 2
    If $\Γ \⊢ \λ \X \<: \S_{1}.~ \s_{2} \: \T$ and $\Γ \⊢ \T \<: \(\∀ \X \<: \U_{1}.~ \U_{2})$,
    then $\U_{1} \= \S_{1}$ and there is some $\S_{2}$ such that
    $\Γ, \X \<: \S_{1} \⊢ \s_{2} \: \S_{2}$ and $\Γ, \X \<: \S_{1} \⊢ \S_{2} \<: \U_{2}$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  \begin{enumerate}
    \item % 1
    By induction on a derivation of $\Γ \⊢ \t \: \T$.

    \begin{itemize}
      \Case\TAbs:
      \quad $\t \= \λ \x \: \S_{1} \s_{2}$
      \quad $\T \= \S_{1} \→ \T_{2}$
      \quad $\Γ \, \x \: \S_{1} \⊢ \s_{2} \: \T_{2}$

      Given $\Γ \⊢ \S_{1} \→ \T_{2} \<: \U_{1} \→ \U_{2}$,
      we use \Cref{lem:inversion-of-subtyping} to obtain
      $\Γ \⊢ \U_{1} \<: \S_{1}$ and $\Γ \⊢ \T_{2} \<: \U_{2}$.
      We pick $\S_{2}$ to be $\T_{2}$ to obtain the desired result.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using \STrans with $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \U_{1} \→ \U_{2}$ we get $\Γ \⊢ \S \<: \U_{1} \→ \U_{2}$. The results follows directly from the IH.

      \Case\TVar, \TApp, \TTAbs, \TTApp, \TClass, \TMatch:

      In those cases, $\t$ is not of the form $\λ \x \: \S_{1}.~ \s_{2}$ and the result is immediate.
    \end{itemize}

    \item % 2
    By induction on a derivation of $\Γ \⊢ \t \: \T$.

    \begin{itemize}
      \Case\TTAbs:
      \quad $\t \= \λ \X \<: \S_{1}.~ \s_{2}$
      \quad $\T \= \∀ \X \<: \S_{1}.~ \T_{2}$
      \quad $\Γ, \X \<: \S_{1} \⊢ \s_{2} \: \T_{2}$

      Given $\Γ \⊢ \(\∀ \X \<: \S_{1}.~ \T_{2}) \<: \(\∀ \X \<: \U_{1}.~ \U_{2})$,
      we use \Cref{lem:inversion-of-subtyping} to obtain
      $\U_{1} \= \S_{1}$ and $\Γ, \X \<: \S_{1} \⊢ \T_{2} \<: \U_{2}$.
      We pick $\S_{2}$ to be $\T_{2}$ to obtain the desired result.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using \STrans with $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \(\∀ \X \<: \U_{1}.~ \U_{2})$
      we get $\Γ \⊢ \S \<: \(\∀ \X \<: \U_{1}.~ \U_{2})$.
      The results follows directly from the IH.

      \Case\TVar, \TAbs, \TApp, \TTApp, \TClass, \TMatch:

      In those cases, $\t$ is not of the form $\λ \X \<: \S_{1}.~ \s_{2} \: \T$ and the result is immediate.

    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/minimum-types.tex
Import FM proof

\begin{lemma}[Minimum types]
  \lbl{lem:minimum-types}~
  \begin{enumerate}
    \item % 1
    If $\Γ \⊢ \new \C \: \T$ then $\Γ \⊢ \{ \new \C \} \<: \T$.

    \item % 2
    If $\Γ \⊢ \λ \x \: \T_{1}.~ \t_{2} \: \T$
    then there is some $\T_{2}$ such that
    $\Γ \⊢ \T_{1} \→ \T_{2} \<: \T$.

    \item % 3
    If $\Γ \⊢ \λ \X \<: \U_{1}.~ \t_{2} \: \T$
    then there is some $\T_{2}$ such that $\Γ \⊢ \∀ \X \<: \U_{1}.~ \T_{2} \<: \T$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  \begin{enumerate}
    \item % 1
    By induction on a derivation of $\Γ \⊢ \t \: \T$.
    \begin{itemize}
      \Case\TVar, \TAbs, \TApp, \TTAbs, \TTApp, \TMatch:

      In those cases, $\t$ is not a constructor call and the result is immediate.

      \Case\TClass:
      \quad $\t \= \new \C$
      \quad $\T \= \{ \new \C \}$

      The result follows directly from \SRefl.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      By the IH, $\Γ \⊢ \{ \new \C \} \<: \S$.
      The result follows from \STrans.
    \end{itemize}

    \item % 2
    By induction on a derivation of $\Γ \⊢ \t \: \T$.
    \begin{itemize}
      \Case\TVar, \TApp, \TTAbs, \TTApp, \TClass, \TMatch:

      In those cases, $\t$ is not an abstraction and the result is immediate

      \Case\TAbs:
      \quad $\T \= \T_{1} \→ \S_{2}$
      \quad $\Γ, \x \: \T_{1} \⊢ \t_{2} \: \S_{2}$

      $\T_{2} \= \S_{2}$ and the result is immediate using \SRefl.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using the IH, there is some $\T_{2}$ such that $\Γ \⊢ \T_{1} \→ \T_{2} \<: \S$.
      The result follows from \STrans.
    \end{itemize}

    \item % 3
    By induction on a derivation of $\Γ \⊢ \t \: \T$.
    \begin{itemize}
      \Case\TVar, \TAbs, \TApp, \TTApp, \TClass, \TMatch:

      In those cases, $\t$ is not a type abstraction and the result is immediate

      \Case\TTAbs:
      \quad $\T \= \∀ \X \<: \U_{1}.~ \S_{2}$
      \quad $\Γ, \X \<: \U_{1} \⊢ \t_{2} \: \S_{2}$

      $\T_{2} \= \S_{2}$ and the result is immediate using \SRefl.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using the IH, there is some $\T_{2}$ such that $\Γ \⊢ \∀ \X \<: \U_{1}.~ \T_{2} \<: \S$.
      The result follows from \STrans.
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/permutation.tex
Import FM proof

\begin{lemma}[Permutation]
  \lbl{lem:permutation}~

  If $\Γ$ and $\Δ$ are well-formed and $\Δ$ is a permutation of $\Γ$, then:
  \begin{enumerate}
    \item % 0
    If   $\Γ \⊢ \disj(\S, \T)$,
    then $\Δ \⊢ \disj(\S, \T)$.

    \item % 1
    If   $\Γ \⊢ \S \<: \T$,
    then $\Δ \⊢ \S \<: \T$.

    \item % 2
    If   $\Γ \⊢ \t \: \T$,
    then $\Δ \⊢ \t \: \T$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  We prove 1. and 2. simultaneously by induction on two derivations of $\Γ \⊢ \disj(\V, \W)$ and $\Γ \⊢ \S \<: \T$.
  More precisely, the induction is done on the cumulative depth of both derivation tree.
  \begin{enumerate}
    \item % 0
    $\Γ \⊢ \disj(\V, \W)$
    \begin{itemize}
      \Case\DXi:
      \quad $\V \= \C_{1}$
      \quad $\W \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ξ$

      Using \DXi with context $\Δ$ directly leads to the desired result.

      \Case\DPsi:
      \quad $\V \= \{ \new \C_{1} \}$
      \quad $\W \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∉ \Ψ$

      Using \DPsi with context $\Δ$ directly leads to the desired result.

      \Case\DSub:
      \quad $\Γ \⊢ \V \<: \U$
      \quad $\Γ \⊢ \disj(\U, \W)$

      By the IH we get $\Δ \⊢ \disj(\U, \W)$.
      Using the 2nd part of the lemma we obtain $\Δ \⊢ \V \<: \U$.
      The result follows from \DSub.

      \Case\DArrow:
      \quad $\V \= \V_{1} \→ \V_{2}$
      \quad $\W \= \C$

      Using \DArrow with context $\Δ$ directly leads to the desired result.

      \Case\DAll:
      \quad $\V \= \∀ \X \<: \V_{1}.~ \V_{2}$
      \quad $\W \= \C$

      Using \DAll with context $\Δ$ directly leads to the desired result.
    \end{itemize}

    \item % 1
    $\Γ \⊢ \S \<: \T$.
    \begin{itemize}
      \Case\SRefl:
      \quad $\T \= \S$

      Using \SRefl with context $\Δ$ directly leads to the desired result.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      The result follows directly from the IH and \STrans.

      \Case\STop:
      \quad $\T \= \Top$

      Using \STop with context $\Δ$ directly leads to the desired result.

      \Case\SSin:
      \quad $\S \= \{ \new \C \}$
      \quad $\T \= \C$

      Using \SSin with context $\Δ$ directly leads to the desired result.

      \Case\STvar:
      \quad $\S \= \X$
      \quad $\X \<: \T \∈ \Γ$
      Since $\Δ$ is a permutation of $\Γ$, $\X \<: \T \∈ \Δ$, and the result follows from \STvar.

      \Case\SArrow:
      \quad $\S \= \S_{1} \→ \S_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \\\phantom{\textit{Case}~\SArrow:}
      \quad $\Γ \⊢ \T_{1} \<: \S_{1}$
      \quad $\Γ \⊢ \S_{2} \<: \T_{2}$

      The result follows directly from the IH and \SArrow.

      \Case\SAll:
      \quad $\S \= \∀ \X \<: \U_{1}.~ \S_{2}$
      \quad $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$
      \quad $\Γ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$

      If $\Δ$ is a permutation of $\Γ$, then $\Δ, \X \<: \U_{1}$ is a permutation of $\Γ, \X \<: \U_{1}$.
      Therefore, we can use the IH to get $\Δ, \X \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$.
      The result follows from \SAll.

      \Case\SPsi:
      \quad $\S \= \C_{1}$
      \quad $\T \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ψ$

      Using \SPsi with context $\Δ$ directly leads to the desired result.

      \Case\SMatch1/2:
      \quad $\T_{1} \= \T_n$
      \quad $\T_{2} \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch1/2:}
      \quad $\Γ \⊢ \T_s \<: \S_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \S_m)$

      By the 1st part of the lemma we get $\∀ \m<n.~ \Δ \⊢ \disj(\T_s, \S_m)$.
      From the IH we obtain $\Δ \⊢ \T_s \<: \S_n$.
      The result follows from \SMatch1/2.

      \Case\SMatch3/4:
      \quad $\S \= \T_d$
      \quad $\T \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \S_n)$

      The result follows from the 1st part of the lemma and \SMatch3/4.

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \U_i \⇒ \S_i \} \otherwise \S_d$
      \quad $\T \= \T_s \match \{ \U_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \n.~ \Γ \⊢ \S_n \<: \T_n$
      \quad $\Γ \⊢ \S_d \<: \T_d$

      The result follows directly from the IH and \SMatch5.
    \end{itemize}

    \item % 2
    By induction on a derivation of $\Γ \⊢ \t \: \T$
    \begin{itemize}
      \Case\TVar:
      \quad $\t \= \x$
      \quad $\x \: \T \∈ \Γ$

      Since $\Δ$ is a permutation of $\Γ$, $\x \: \T \∈ \Δ$, and the result follows from \TVar.

      \Case\TAbs:
      \quad $\t \= \λ \x \: \T_{1} \t_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \quad $\Γ, \x \: \T_{1} \⊢ \t_{2} \: \T_{2}$

      If $\Δ$ is a permutation of $\Γ$, then $\Δ, \x \: \T_{1}$ is a permutation of $\Γ, \x \: \T_{1}$.
      Therefore, we can use the IH to get $\Δ, \x \: \T_{1} \⊢ \t_{2} \: \T_{2}$.
      The result follows from \TAbs.

      \Case\TApp:
      \quad $\t \= \t_{1} \t_{2}$
      \quad $\T \= \T_{12}$
      \quad $\Γ \⊢ \t_{1} \: \T_{11} \→ \T_{12}$
      \quad $\Γ \⊢ \t_{2} \: \T_{11}$

      The result follows directly from the IH and \TApp.

      \Case\TTAbs:
      \quad $\t \= \λ \X \<: \U_{1}.~ \t_{2}$
      \quad $\T \= \∀ \X \<: \U_{1}.~ \T_{2}$
      \quad $\Γ, \X \<: \U_{1} \⊢ \t_{2} \: \T_{2}$

      If $\Δ$ is a permutation of $\Γ$, then $\Δ, \X \<: \U_{1}$ is a permutation of $\Γ, \X \<: \U_{1}$.
      Therefore, we can use the IH to get $\Δ, \X \<: \U_{1} \⊢ \t_{2} \: \T_{2}$.
      The result follows from \TTAbs.

      \Case\TTApp:
      \quad $\t \= \t_{1} \T_{2}$
      \quad $\T \= \[\X \↦ \T_{2}] \T_{12}$
      \\\phantom{\textit{Case}~\TTApp:}
      \quad $\Γ \⊢ \t_{1} \: \(\∀ \X \<: \U_{1}.~ \T_{12})$
      \quad $\Γ \⊢ \T_{2} \<: \U_{1}$

      By the IH we get $\Δ \⊢ \t_{1} \: \(\∀ \X \<: \U_{1}.~ \T_{12})$.
      Using the 2nd part of the lemma we obtain $\Δ \⊢ \T_{2} \<: \U_{1}$.
      The result follows from \TTApp.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      By the IH we get $\Δ \⊢ \t \: \S$.
      Using the 2nd part of the lemma we obtain $\Δ \⊢ \S \<: \T$.
      The result follows from \TSub.

      \Case\TClass:
      \quad $\t \= \new \C$
      \quad $\T \= \{ \new \C \}$

      Using \TClass with context $\Δ$ directly leads to the desired result.

      \Case\TMatch:
      \quad $\t \= \t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$
      \quad $\T \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\TMatch:}
      \quad $\Γ \⊢ \t_s \: \T_s$
      \quad $\Γ, \x_i \: \C_i \⊢ \t_i \: \T_i$
      \quad $\Γ \⊢ \t_d \: \T_d$

      If $\Δ$ is a permutation of $\Γ$, then $\Δ, \x_i \: \C_i$ is a permutation of $\Γ, \x_i \: \C_i$. Therefore the result follows directly from the IH and \TMatch.
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 proofs/preservation.tex
Import FM proof

\begin{theorem}[Preservation]
  \lbl{thm:preservation}
  ~\\\indent
  If $\Γ \⊢ \t \: \T$ and $\t \⟶ \t'$ then $\Γ \⊢ \t' \: \T$.
\end{theorem}

\begin{proof}
  By induction on a derivation of $\Γ \⊢ \t \: \T$.
  \begin{itemize}
    \Case\TVar, \TAbs, \TTAbs, \TClass:

    These cases cannot arise since we assume $\t \⟶ \t'$ but there are no evaluation rules for variables, abstractions, type abstractions and class instantiations.

    \Case\TApp:
    \quad $\t \= \t_{1} \t_{2}$
    \quad $\T \= \T_{12}$
    \quad $\Γ \⊢ \t_{1} \: \T_{11} \→ \T_{12}$
    \quad $\Γ \⊢ \t_{2} \: \T_{11}$

    By definition of the evaluation relation, there are 3 subcases:
    \begin{itemize}
      \Subcase\EAppA:
      \quad $\t_{1} \⟶ \t_{1}'$
      \quad $\t' \= \t_{1}' \t_{2}$

      By the IH and the 1st premise we get $\Γ \⊢ \t_{1}' \: \T_{11} \→ \T_{12}$.
      We use \TApp with that result and the 2nd premise to obtain $\Γ \⊢ \t_{1}' \t_{2} \: \T_{12}$, as required.

      \Subcase\EAppB:
      \quad $\t_{2} \⟶ \t_{2}'$
      \quad $\t_{1} \= \v_{1}$
      \quad $\t' \= \v_{1} \t_{2}'$

      Analogously, the IH gives $\Γ \⊢ \t_{2}' \: \T_{12}$ and the result follows from \TApp.

      \Subcase\EAppAbs:
      \quad $\t_{1} \= \λ \x \: \U_{11}.~ \u_{12}$
      \quad $\t' \= \[\x \↦ \t_{2}] \u_{12}$

      By \Cref{lem:inversion-of-typing} (with $\S_{1} \= \U_{11}$, $\s_{2} \= \u_{12}$, $\U_{1} \= \T_{11}$ and $\U_{2} \= \T_{12}$),
      there is some $\S_{2}$ such that $\Γ, \x \: \U_{11} \⊢ \u_{12} \: \S_{2}$, $\Γ \⊢ \S_{2} \<: \T_{12}$ and $\Γ \⊢ \T_{11} \<: \U_{11}$.
      Using \TSub with $\Γ \⊢ \t_{2} \: \T_{11}$ and $\Γ \⊢ \T_{11} \<: \U_{11}$ we obtain $\Γ \⊢ \t_{2} \: \U_{11}$.
      By \Cref{lem:substitution} we get $\Γ \⊢ \[\x \↦ \t_{2}] \u_{12} \: \S_{2}$.
      Using \TSub we obtain $\Γ \⊢ \[\x \↦ \t_{2}] \u_{12} \: \T_{12}$, as required.
    \end{itemize}

    \Case\TTApp:
    \quad $\t \= \t_{1} \T_{2}$
    \quad $\T \= \[\X \↦ \T_{2}] \T_{12}$
    \\\phantom{\textit{Case}~\TTApp:}
    \quad $\Γ \⊢ \t_{1} \: \(\∀ \X \<: \U_{1}.~ \T_{12})$
    \quad $\Γ \⊢ \T_{2} \<: \U_{1}$

    The proof for case \TTApp is analogous to the one for \TApp.
    By definition of the evaluation relation, there are 2 subcases:
    \begin{itemize}
      \Subcase\ETApp:
      \quad $\t_{1} \⟶ \t_{1}'$
      \quad $\t' \= \t_{1}' \T_{2}$

      By the IH and the 1st premise we get $\Γ \⊢ \t_{1}' \: \(\∀ \X \<: \U_{1}.~ \T_{12})$.
      We use \TTApp with that result and the 2nd premise to obtain $\Γ \⊢ \t_{1}' \T_{2} \: \[\X \↦ \T_{2}] \T_{12}$, as required.

      \Subcase\ETAppTAbs:
      \quad $\t_{1} \= \λ \X \<: \U_{11}.~ \u_{12}$
      \quad $\t' \= \[\X \↦ \T_{2}] \u_{12}$

      By \Cref{lem:inversion-of-typing} (with $\S_{1} \= \U_{11}$, $\s_{2} \= \u_{12}$ and $\U_{2} \= \T_{12}$),
      there is some $\S_{2}$ such that
      $\Γ, \X \<: \U_{11} \⊢ \u_{12} \: \S_{2}$,
      $\U_{1} \= \U_{11}$,
      and $\Γ, \X \<: \U_{11} \⊢ \S_{2} \<: \T_{12}$.
      Since $\Γ \⊢ \T_{2} \<: \U_{11}$, we use \Cref{lem:substitution} twice to
      get $\Γ \⊢ \[\X \↦ \T_{2}] \u_{12} \: \[\X \↦ \T_{2}] \S_{2}$
      and $\Γ \⊢ \[\X \↦ \T_{2}] \S_{2} \<: \[\X \↦ \T_{2}] \T_{12}$.
      By \TSub $\Γ \⊢ \[\X \↦ \T_{2}] \u_{12} \: \[\X \↦ \T_{2}] \T_{12}$, as required.
    \end{itemize}

    \Case\TSub:
    \quad $\Γ \⊢ \t \: \S$
    \quad $\Γ \⊢ \S \<: \T$

    By the IH, $\Γ \⊢ \t' \: \S$. The result follows directly from \TSub.

    \Case\TMatch:
    \quad $\t \= \t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$
    \quad $\T \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$
    \\\phantom{\textit{Case}~\TMatch:}
    \quad $\Γ \⊢ \t_s \: \T_s$
    \quad $\Γ, \x_i \: \C_i \⊢ \t_i \: \T_i$
    \quad $\Γ \⊢ \t_d \: \T_d$

    By definition of the evaluation relation, there are 5 subcases:
    \begin{itemize}
      \Subcase\EMatch1:
      \quad $\t_s \⟶ \t_s'$
      \quad $\t' \= \t_s' \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$

      By the IH $\Γ \⊢ \t_s' \: \T_s$. The result follows directly from \TMatch.

      \Subcase\EMatch2:
      \quad $\t_s \= \new \C$
      \quad $\(\C, \C_n) \∈ \Ψ$
      \\\phantom{\textit{Subcase}~\EMatch2:}
      \quad $\∀ \m<n.~ \(\C, \C_m) \∉ \Ψ$
      \quad $\t' \= \[\x_n \↦ \new \C] \t_n$

      By \SPsi, \SSin and \STrans, $\Γ \⊢ \{ \new \C \} \<: \C_n$.
      From \DPsi, we obtain $\∀ \m<n.~ \Γ \⊢ \disj(\{ \new \C \}, \C_m)$.
      By \Cref{lem:minimum-types}, $\Γ \⊢ \{ \new \C \} \<: \T_s$.
      Let $\T_{1}$ be $\{ \new \C \} \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$.
      From \SMatch1, $\Γ \⊢ \T_n \<: \T_{1}$.
      Using \SMatch5, $\Γ \⊢ \T_{1} \<: \T$.
      By \STrans, $\Γ \⊢ \T_n \<: \T$.

      Using \TClass, \TSub, \SSin and \SPsi (with $\(\C, \C_n) \∈ \Ψ$) we get $\Γ \⊢ \new \C \: \C_n$.
      From the case premises, we have $\Γ, \x_n \: \C_n \⊢ \t_n \: \T_n$.
      By \Cref{lem:substitution}, we get $\Γ \⊢ \[\x_n \↦ \new \C] \t_n \: \T_n$.
      Finally, using \TSub we get , $\Γ \⊢ \[\x_n \↦ \new \C] \t_n \: \T$, as required.

      \Subcase\EMatch3:
      \quad $\t_s \= \new \C$
      \quad $\∀ \n.~ \(\C, \C_n) \∉ \Ψ$
      \quad $\t' \= \t_d$

      The proof for subcase \EMatch3 is analogous to the one for \EMatch2.
      From \DPsi, $\∀ \n.~ \Γ \⊢ \disj(\{ \new \C \}, \C_n)$.
      By \Cref{lem:minimum-types} we get $\Γ \⊢ \{ \new \C \} \<: \T_s$.
      Let $\T_{1}$ be $\{ \new \C \} \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$.
      From \SMatch2, $\Γ \⊢ \T_d \<: \T_{1}$.
      Using \SMatch5, $\Γ \⊢ \T_{1} \<: \T$.
      By \STrans $\Γ \⊢ \T_d \<: \T$.
      Using \TSub, $\Γ \⊢ \t_d \: \T$, as required.

      \Subcase\EMatch4:
      \quad $\t_s \= \λ \x \: \U.~ \u$
      \quad $\t' \= \t_d$

      By \Cref{lem:minimum-types}, there exists a $\V$ such that
      $\Γ \⊢ \U \→ \V \<: \T_s$.
      Using \DArrow, $\∀ \n.~ \Γ \⊢ \disj(\U \→ \V, \C_n)$.
      Let $\T_{1}$ be $\U \→ \V \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$.
      From \SMatch2, $\Γ \⊢ \T_d \<: \T_{1}$.
      Using \SMatch5, $\Γ \⊢ \T_{1} \<: \T$.
      By \STrans $\Γ \⊢ \T_d \<: \T$.
      Using \TSub, $\Γ \⊢ \t_d \: \T$, as required.

      \Subcase\EMatch5:
      \quad $\t_s \= \∀ \X \<: \U.~ \u$
      \quad $\t' \= \t_d$

      The proof for subcase \EMatch5 is analogous to the one for \EMatch4.
      By \Cref{lem:minimum-types}, there exists a $\V$ such that
      $\Γ \⊢ \(\∀ \X \<: \U.~ \V) \<: \T_s$.
      Using \DArrow, we get $\∀ \n.~ \Γ \⊢ \disj(\∀ \X \<: \U.~ \V, \C_n)$.
      Let $\T_{1}$ be $\(\∀ \X \<: \U.~ \V) \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$.
      From \SMatch2, $\Γ \⊢ \T_d \<: \T_{1}$.
      Using \SMatch5, $\Γ \⊢ \T_{1} \<: \T$.
      By \STrans $\Γ \⊢ \T_d \<: \T$.
      Using \TSub, $\Γ \⊢ \t_d \: \T$, as required.
    \end{itemize}
  \end{itemize}
\end{proof}

### 2021-10-21 proofs/progress.tex
Import FM proof

\begin{theorem}[Progress]
  \lbl{thm:progress}
  ~\\\indent
  If $\t$ is a closed, well-typed term,
  then either $\t$ is a value
  or there is some $\t'$ such that $\t \⟶ \t'$.
\end{theorem}

\begin{proof}
  By induction on a derivation of $\Γ \⊢ \t \: \T$.
  \begin{itemize}
    \Case\TVar:
    \quad $\t \= \x$
    \quad $\x \: \T \∈ \Γ$

    This case cannot occure because $\t$ is closed.

    \Case\TAbs, \TTAbs, \TClass:

    In those cases, $\t$ is a value and the result is immediate.

    \Case\TApp:
    \quad $\t \= \t_{1} \t_{2}$
    \quad $\T \= \T_{12}$
    \quad $\Γ \⊢ \t_{1} \: \T_{11} \→ \T_{12}$
    \quad $\Γ \⊢ \t_{2} \: \T_{11}$

    By the IH, either $\t_{1}$ is a value or $\t_{1}$ can take a step
    (there is some $\t_{1}'$ such that $\t_{1} \⟶ \t_{1}'$).
    If $\t_{1}$ can take a step, then \EAppA applies to $\t$.
    If $\t_{1}$ is a value, we use \Cref{lem:canonical-forms}
    to get that $\t_{1}$ has the form $\λ \x \: \S_{1}.~ \t_{2}$.
    Therefore, \EAppAbs applies to $\t$, as required.

    \Case\TTApp:
    \quad $\t \= \t_{1} \T_{2}$
    \quad $\T \= \[\X \↦ \T_{2}] \T_{12}$
    \\\phantom{\textit{Case}~\TTApp:}
    \quad $\Γ \⊢ \t_{1} \: \(\∀ \X \<: \U_{1}.~ \T_{12})$
    \quad $\Γ \⊢ \T_{2} \<: \U_{1}$

    The proof for case \TTApp is analogous to the one for \TApp.

    By the IH, either $\t_{1}$ is a value or $\t_{1}$ can take a step.
    If $\t_{1}$ can take a step, then \ETApp applies to $\t$.
    If $\t_{1}$ is a value, we use \Cref{lem:canonical-forms}
    to get that $\t_{1}$ has the form $\λ \X \<: \T_{1}.~ \t_{2}$.
    Therefore, \ETAppTAbs applies to $\t$, as required.

    \Case\TSub:
    \quad $\Γ \⊢ \t \: \S$
    \quad $\Γ \⊢ \S \<: \T$

    The result follows directly from the IH.

    \Case\TMatch:
    \quad $\t \= \t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$
    \quad $\T \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$
    \\\phantom{\textit{Case}~\TMatch:}
    \quad $\Γ \⊢ \t_s \: \T_s$
    \quad $\Γ, \x_i \: \C_i \⊢ \t_i \: \T_i$
    \quad $\Γ \⊢ \t_d \: \T_d$

    By the IH, either $\t_s$ is a value or $\t_s$ can take a step.
    If $\t_s$ can take a step, then \EMatch1 applies to $\t$, as required.
    If $\t_s$ is a value, $\t_s$ can take 3 different forms:
    \begin{itemize}
      \Subcase $\t_s$ is of the form $\new \C$:\\
      If $\∀m.~ \(\C, \C_m) \∉ \Ψ$, then \EMatch3 applies to $\t$.
      Otherwise, let $\C_k$ be the first class such that $\(\C, \C_k) \∈ \Ψ$.
      By construction we know that $\∀m<k.~ \(\C, \C_k) \∉ \Ψ$.
      Therefore \EMatch2 applies to $\t$, as required.

      \Subcase $\t_s$ is of the form $\λ \x \: \T_{1} \t_{2}$:\\
      \EMatch4 applies to $\t$ and the result is immediate.

      \Subcase $\t_s$ is of the form $\∀ \X \<: \U_{1}.~ \T_{2}$:\\
      \EMatch5 applies to $\t$ and the result is immediate.
    \end{itemize}
  \end{itemize}
\end{proof}

### 2021-10-21 proofs/weakening.tex
Import FM proof

\begin{lemma}[Weakening]
  \lbl{lem:weakening}~
  \begin{enumerate}
    \item % 0
    If $\Γ \⊢ \disj(\S, \T)$ and $\Γ, \X \<: \U$ is well formed,
    then $\Γ, \X \<: \U \⊢ \disj(\S, \T)$.

    \item % 1
    If $\Γ \⊢ \S \<: \T$ and $\Γ, \X \<: \U$ is well formed,
    then $\Γ, \X \<: \U \⊢ \S \<: \T$.

    \item % 2
    If $\Γ \⊢ \S \<: \T$ and $\Γ, \x \: \U$ is well formed,
    then $\Γ, \x \: \U \⊢ \S \<: \T$.

    \item % 3
    If $\Γ \⊢ \t \: \T$ and $\Γ, \x \: \U$ is well formed,
    then $\Γ, \x \: \U \⊢ \t \: \T$.

    \item % 4
    If $\Γ \⊢ \t \: \T$ and $\Γ, \X \<: \U$ is well formed,
    then $\Γ, \X \<: \U \⊢ \t \: \T$.
  \end{enumerate}
\end{lemma}

\begin{proof}
  We prove 1. and 2. simultaneously by induction on two derivations of $\Γ \⊢ \disj(\V, \W)$ and $\Γ \⊢ \S \<: \T$.
  More precisely, the induction is done on the cumulative depth of both derivation tree.
  \begin{enumerate}
    \item % 0
    $\Γ \⊢ \disj(\V, \W)$
    \begin{itemize}
      \Case\DXi:
      \quad $\V \= \C_{1}$
      \quad $\W \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ξ$

      Using \DXi with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\DPsi:
      \quad $\V \= \{ \new \C_{1} \}$
      \quad $\W \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∉ \Ψ$

      Using \DPsi with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\DSub:
      \quad $\Γ \⊢ \V \<: \U$
      \quad $\Γ \⊢ \disj(\U, \W)$

      By the IH we get $\Γ \⊢ \disj(\U, \W)$.
      Using the 2nd part of the lemma we obtain $\Γ, \X \<: \U \⊢ \V \<: \U$.
      The result follows from \DSub.

      \Case\DArrow:
      \quad $\V \= \V_{1} \→ \V_{2}$
      \quad $\W \= \C$

      Using \DArrow with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\DAll:
      \quad $\V \= \∀ \X \<: \V_{1}.~ \V_{2}$
      \quad $\W \= \C$

      Using \DAll with context $\Γ, \X \<: \U$ directly leads to the desired result.
    \end{itemize}

    \item % 1
    $\Γ \⊢ \S \<: \T$.
    \begin{itemize}
      \Case\SRefl:
      \quad $\T \= \S$

      Using \SRefl with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\STrans:
      \quad $\Γ \⊢ \S \<: \U$
      \quad $\Γ \⊢ \U \<: \T$

      The result follows from the IH and \STrans.

      \Case\STop:
      \quad $\T \= \Top$

      Using \STop with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\SSin:
      \quad $\S \= \{ \new \C \}$
      \quad $\T \= \C$

      Using \SSin with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\STvar:
      \quad $\S \= \Y$
      \quad $\Y \<: \T \∈ \Γ$

      If $\Y \<: \T \∈ \Γ$, then $\Y \<: \T \∈ \Γ, \X \<: \U$ and the result follows from \STvar.

      \Case\SArrow:
      \quad $\S \= \S_{1} \→ \S_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \\\phantom{\textit{Case}~\SArrow:}
      \quad $\Γ \⊢ \T_{1} \<: \S_{1}$
      \quad $\Γ \⊢ \S_{2} \<: \T_{2}$

      The result follows from the IH and \SArrow.

      \Case\SAll:
      \quad $\S \= \∀ \Y \<: \U_{1}.~ \S_{2}$
      \quad $\T \= \∀ \Y \<: \U_{1}.~ \T_{2}$
      \quad $\Γ, \Y \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$

      Using the IH with context $\Γ_{IH} \= \Γ, \Y \<: \U_{1}$, we get $\Γ, \Y \<: \U_{1}, \X \<: \U \⊢ \S_{2} \<: \T_{2}$.
      From \Cref{lem:permutation}, $\Γ, \X \<: \U, \Y \<: \U_{1} \⊢ \S_{2} \<: \T_{2}$
      The result follows from \SAll.

      \Case\SPsi:
      \quad $\S \= \C_{1}$
      \quad $\T \= \C_{2}$
      \quad $\(\C_{1}, \C_{2}) \∈ \Ψ$

      Using \SPsi with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\SMatch1/2:
      \quad $\T_{1} \= \T_n$
      \quad $\T_{2} \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\SMatch1/2:}
      \quad $\Γ \⊢ \T_s \<: \S_n$
      \quad $\∀ \m<n.~ \Γ \⊢ \disj(\T_s, \S_m)$

      From the IH we get $\Γ, \X \<: \U \⊢ \T_s \<: \S_n$.
      Using the 1st part of the lemma we obtain $\∀ \m< \n.~\ \Γ, \X \<: \U \⊢ \disj(\T_s, \S_m)$.
      The result follows from \SMatch1/2.

      \Case\SMatch3/4:
      \quad $\T_{1} \= \T_d$
      \quad $\T_{2} \= \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\∀ \n.~ \Γ \⊢ \disj(\T_s, \S_n)$

      Using the 1st part of the lemma we get $\∀ \n.~\ \Γ, \X \<: \U \⊢ \disj(\T_s, \S_n)$.
      The result follows from \SMatch3/4.

      \Case\SMatch5:
      \quad $\S \= \S_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d$
      \quad $\T \= \T_s \match \{ \S_i \⇒ \U_i \} \otherwise \U_d$
      \\\phantom{\textit{Case}~\SMatch5:}
      \quad $\Γ \⊢ \S_s \<: \T_s$
      \quad $\∀ \n.~ \Γ \⊢ \T_n \<: \U_n$
      \quad $\Γ \⊢ \T_d \<: \U_d$

      We use the IH on each premise and the result follows directly from \SMatch5.
    \end{itemize}

    \item % 2
    By inspection of the subtyping rules, it is clear that typing assumbtions play no role in subtyping derivations.

    \item % 3
    By induction on a derivation of $\Γ \⊢ \t \: \T$.
    \begin{itemize}
      \Case\TVar:
      \quad $\t \= \y$
      \quad $\y \: \T \∈ \Γ$

      If $\y \: \T \∈ \Γ$ then $\y \: \T \∈ \Γ, \x \: \U$ and the result follows from \TVar.

      \Case\TAbs:
      \quad $\t \= \λ \y \: \T_{1} \t_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \quad $\Γ, \y \: \T_{1} \⊢ \t_{2} \: \T_{2}$

      Using the IH with $\Γ_{IH} \= \Γ, \y \: \T_{1}$ we get $\Γ, \y \: \T_{1}, \x \: \U \⊢ \t_{2} \: \T_{2}$.
      From \Cref{lem:permutation}, $\Γ, \x \: \U, \y \: \T_{1} \⊢ \t_{2} \: \T_{2}$.
      The result follows from \TAbs.

      \Case\TApp:
      \quad $\t \= \t_{1} \t_{2}$
      \quad $\T \= \T_{12}$
      \quad $\Γ \⊢ \t_{1} \: \T_{11} \→ \T_{12}$
      \quad $\Γ \⊢ \t_{2} \: \T_{12}$

      We use the IH on each premise and the result follows from \TApp.

      \Case\TTAbs:
      \quad $\t \= \λ \X \<: \T_{1}.~ \t_{2}$
      \quad $\T \= \∀ \X \<: \T_{1}.~ \T_{2}$
      \quad $\Γ, \X \<: \T_{1} \⊢ \t_{2} \: \T_{2}$

      Using the IH with $\Γ_{IH} \= \Γ, \X \<: \T_{1}$ we get $\Γ, \X \<: \T_{1}, \x \: \U \⊢ \t_{2} \: \T_{2}$.
      From \Cref{lem:permutation}, $\Γ, \x \: \U, \X \<: \T_{1} \⊢ \t_{2} \: \T_{2}$
      The result follows from \TTAbs.

      \Case\TTApp:
      \quad $\t \= \t_{1} \T_{2}$
      \quad $\T \= \[\X \↦ \T_{2}] \T_{12}$
      \\\phantom{\textit{Case}~\TTApp:}
      \quad $\Γ \⊢ \t_{1} \: \(\∀ \X \<: \T_{11}.~ \T_{12})$
      \quad $\Γ \⊢ \T_{2} \<: \T_{11}$

      Using the IH on the left premise we get $\Γ, \x \: \U \⊢ \t_{1} \: \(\∀ \X \<: \T_{11}.~ \T_{12})$.
      Using the 3rd part of the lemma on the right premise we obtain $\Γ, \x \: \U \⊢ \T_{2} \<: \T_{11}$.
      The result follows from \TTApp.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using the IH on the left premise we get $\Γ, \x \: \U \⊢ \t \: \S$.
      Using the 3rd part of the lemma on the right premise we obtain $\Γ, \x \: \U \⊢ \S \<: \T$.
      The result follows from \TSub.

      \Case\TClass:
      \quad $\t \= \new \C$
      \quad $\T \= \{ \new \C \}$

      Using \TClass with context $\Γ, \x \: \U$ directly leads to the desired result.

      \Case\TMatch:
      \quad $\t \= \t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$
      \quad $\T \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\TMatch:}
      \quad $\Γ \⊢ \t_s \: \T_s$
      \quad $\Γ, \x_i \: \C_i \⊢ \t_i \: \T_i$
      \quad $\Γ \⊢ \t_d \: \T_d$

      We use the IH on each premise and the result follows directly from \Cref{lem:permutation} and \TMatch.
    \end{itemize}

    \item % 4
    By induction on a derivation of $\Γ \⊢ \t \: \T$.
    \begin{itemize}
      \Case\TVar:
      \quad $\t \= \x$
      \quad $\x \: \T \∈ \Γ$

      If $\y \: \T \∈ \Γ$ then $\y \: \T \∈ \Γ, \X \<: \U$ and the result follows from \TVar.

      \Case\TAbs:
      \quad $\t \= \λ \x \: \T_{1} \t_{2}$
      \quad $\T \= \T_{1} \→ \T_{2}$
      \quad $\Γ, \x \: \T_{1} \⊢ \t_{2} \: \T_{2}$

      Using the IH with $\Γ_{IH} \= \Γ, \x \: \T_{1}$ we get $\Γ, \x \: \T_{1}, \X \<: \U \⊢ \t_{2} \: \T_{2}$.
      From \Cref{lem:permutation}, $\Γ, \X \<: \U, \x \: \T_{1} \⊢ \t_{2} \: \T_{2}$.
      The result follows from \TAbs.

      \Case\TApp:
      \quad $\t \= \t_{1} \t_{2}$
      \quad $\T \= \T_{12}$
      \quad $\Γ \⊢ \t_{1} \: \T_{11} \→ \T_{12}$
      \quad $\Γ \⊢ \t_{2} \: \T_{12}$

      We use the IH on each premise and the result follows from \TApp.

      \Case\TTAbs:
      \quad $\t \= \λ \Y \<: \T_{1}.~ \t_{2}$
      \quad $\T \= \∀ \Y \<: \T_{1}.~ \T_{2}$
      \quad $\Γ, \Y \<: \T_{1} \⊢ \t_{2} \: \T_{2}$

      Using the IH with $\Γ_{IH} \= \Γ, \Y \<: \T_{1}$ we get $\Γ, \Y \<: \T_{1}, \X \<: \U \⊢ \t_{2} \: \T_{2}$.
      From \Cref{lem:permutation}, $\Γ, \X \<: \U, \Y \<: \T_{1}, \⊢ \t_{2} \: \T_{2}$.
      The result follows from \TTAbs.

      \Case\TTApp:
      \quad $\t \= \t_{1} \T_{2}$
      \quad $\T \= \[\Y \↦ \T_{2}] \T_{12}$
      \\\phantom{\textit{Case}~\TTApp:}
      \quad $\Γ \⊢ \t_{1} \: \(\∀ \Y \<: \T_{11}.~ \T_{12})$
      \quad $\Γ \⊢ \T_{2} \<: \T_{11}$

      Using the IH on the left premise we get $\Γ, \X \<: \U \⊢ \t_{1} \: \(\∀ \X \<: \T_{11}.~ \T_{12})$.
      Using the 2nd part of the lemma on the right premise we obtain $\Γ, \X \<: \U \⊢ \T_{2} \<: \T_{11}$.
      The result follows from \TTApp.

      \Case\TSub:
      \quad $\Γ \⊢ \t \: \S$
      \quad $\Γ \⊢ \S \<: \T$

      Using the IH on the left premise we get $\Γ, \S \<: \U \⊢ \t \: \S$.
      Using the 2nd part of the lemma on the right premise we obtain $\Γ, \X \<: \U \⊢ \S \<: \T$.
      The result follows from \TSub.

      \Case\TClass:
      \quad $\t \= \new \C$
      \quad $\T \= \{ \new \C \}$

      Using \TClass with context $\Γ, \X \<: \U$ directly leads to the desired result.

      \Case\TMatch:
      \quad $\t \= \t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$
      \quad $\T \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$
      \\\phantom{\textit{Case}~\TMatch:}
      \quad $\Γ \⊢ \t_s \: \T_s$
      \quad $\Γ, \x_i \: \C_i \⊢ \t_i \: \T_i$
      \quad $\Γ \⊢ \t_d \: \T_d$

      We use the IH on each premise and the result follows directly from \Cref{lem:permutation} and \TMatch.
    \end{itemize}
  \end{enumerate}
\end{proof}

### 2021-10-21 title-page.tex
Import FM proof

@@
 \begin{titlepage}
+\begin{otherlanguage}{french}
 \begin{center}
@@
 \vfill
-\normalsize THÈSE N° XXXX (\the\year)\\[10pt]
-PRÉSENTÉE LE \uppercase{\today}\\
+\normalsize THÈSE N. TBD (\the\year)\\[10pt]
+PRÉSENTÉE LE {\MakeUppercase\today}\\
 À LA FACULTÉ INFORMATIQUE ET COMMUNICATIONS\\
@@
 \vspace{2cm}
+\end{otherlanguage}
 \end{titlepage}

### 2021-10-21 article.tex
Wip import LTPwMT

Type-level programming is becoming more and more popular in the realm of functional programming.
However, the combination of type-level programming and subtyping remains largely unexplored in practical programming languages.
This paper presents \emph{match types}, a type-level equivalent of pattern matching.
Match types integrate seamlessly into programming languages with subtyping and, despite their simplicity, offer significant additional expressiveness.
We formalize the feature of match types in a calculus based on \SystemFsub and prove its soundness.
We practically evaluate our system by implementing match types in the Scala 3 reference compiler, thus making type-level programming readily available to a broad audience of programmers.

\section{Introduction}
\label{sec:introduction}

There is a growing interest in using \emph{type-level computation} to increase the expressivity of type systems, express additional constraints on the type level, and thereby improve the safety of general-purpose software.
What used to be an exclusive feature of dependently typed languages is slowly becoming accessible to everyday programmers.
GHC Haskell has been at the forefront of making this a reality and already provides several extensions to support type-level programming.
While Haskell is certainly not the only language moving towards dependent types, the trend seems to be limited to pure functional programming languages.

We believe that type-level programming is not necessarily incompatible with other programming paradigms and that the current division exists mainly due to a lack of attention from the research community.
Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\cite{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.

One notable exception is the TypeScript language, which recently introduced a new feature called \emph{conditional type}, a type-level ternary operator based on subtyping.
A conditional type, written \typescript{S extends T ? Tt : Tf}, reduces to \typescript{Tt} when \typescript{S} is a subtype of \typescript{T}, to \typescript{Tf} when \typescript{S} is not a subtype of \typescript{T}, and is left unreduced when types variables do not allow to draw a conclusion. Unfortunately, the algorithm used to reduce such conditional types is both \emph{unsound} and \emph{incomplete}.
Despite the unsoundness (discussed in \Cref{subsec:conditional-types-in-typescript}), the addition of conditional types to TypeScript illustrates the practical need and timeliness of this feature.

This paper presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
As the name suggests, match types allow programmers to express types that perform pattern matching on types:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
The example, which we explain in detail in \Cref{sec:overview}, defines the type \dotty{Elem} by matching on the type parameter \dotty{X}.
%
We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
Match types have received a great interest from the Scala community, and are already in active use.

In this paper, we explore the theoretical foundations of match types through the lens of a type system which extends \SystemFsub with pattern matching at the term and type level.
%
Our formalization serves two purposes: first, it gives a clear view on \emph{how} we integrated match types in Scala's type system and precisely describes the changes needed on the subtyping relation to make this integration possible. Second, thanks to a type safety proof based on the standard progress and preservation theorems, it gives confidence that the design of match types is sensible and our implementation is sound.

Conditional types provide \emph{concrete evidence} that our results are valuable beyond the context of Scala.
Our results are directly applicable to TypeScript's type system and provide a clear path to fixing the unsoundness introduced by conditional types.
Furthermore, we hope that match types can be useful as a reference for future designs of type-level programming features for languages with subtyping.

In summary, this paper makes the following contributions:
%
\begin{itemize}
  \item We introduce programming with match types in Scala by means of an example and highlight the interaction of type-level programming and subtyping (\Cref{sec:overview}).

  \item We formalize match types in the self-contained calculus \SystemFm and prove it sound, providing a theoretical basis of our implementation (\Cref{sec:formalization}). The paper is accompanied by a mechanization of \SystemFm, including proofs of progress and preservation.

  \item We describe our implementation of match types in the Scala compiler, discuss challenges, and relate the implementation to our formalization (\Cref{sec:implementation}).

  \item We evaluate match types in a case study, presenting a type-safe version of the NumPy library (\Cref{sec:evaluation}).

  \item We motivate the design of our formalization relative to prior work, we review the extensive related work on type families in Haskell, and discuss the unsoundness of conditional types in TypeScript (\Cref{sec:related-work}).
\end{itemize}

\section{Overview}
\label{sec:overview}

In this section, we offer a brief introduction of match types in Scala by inspecting the example from the previous section in more detail:
%
\begin{lstlisting}
type Elem[X] = X match {
  case String => Char
  case List[t] => Elem[t]
  case Any => X
}
\end{lstlisting}
%
This example defines a type \dotty{Elem} parametrized by one type parameter \dotty{X}. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
For example, the above type reduces as follows:
%
\begin{lstlisting}
Elem[String] =:= Char
Elem[Int] =:= Int
Elem[List[Int]] =:= Int
\end{lstlisting}
%
Here we use \dotty{S =:= T} to denote type equality between the two types \dotty{S} and \dotty{T}, witnessed by mutual subtyping.
To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
For example, although \dotty{String} is a subtype of both \dotty{String} and \dotty{Any} (the top of Scala's subtyping lattice), \dotty{Elem[String]} reduces to \dotty{Char} because the corresponding case appears first.

When the scrutinee type is a \dotty{List}, the match type \dotty{Elem} is defined recursively on the element type of the list.
Hence, in our example \dotty{Elem[List[Int]]} first reduces to the type \dotty{Elem[Int]}, and eventually to the type \dotty{Int}.

\subsection{A Lightweight Form of Dependent Typing}
Match types enable a lightweight form of dependent typing, since a term-level pattern matching expression can be typed accordingly at the type level as a match type.
Consider the following function definition:
%
\begin{lstlisting}
def elem[X](x: X): Elem[X] = x match {
  case x: String => x.charAt(0)
  case x: List[t] => elem(x.head)
  case x: Any => x
}
\end{lstlisting}
%
This definition is well-typed because the match expression in \dotty{elem}'s body has the exact same scrutinee and pattern types as \dotty{Elem[X]} (the function's return type).

Thanks to Scala's type inference, a call to the \dotty{elem} function can have a result type that \emph{depends} on a term-level parameter.
For instance, in the expression \dotty{elem(1)}, the Scala compiler infers the singleton type \dotty{X = 1} for \dotty{elem}'s type parameter.
This expression thus has type \dotty{Elem[1]}, which reduces to \dotty{Int} (via \dotty{Elem}'s third case).
Similarly, in \dotty{elem(x)}, the compiler infers the singleton type \dotty{X = x.type} and the expression has type \dotty{Elem[x.type]}, which might reduce further at the callsite depending on \dotty{x}'s type.

In both examples, singleton types create a dependency between a type parameter and a term, which, by transitivity, results in a lightweight form of dependent typing, that is, a dependency between a term parameter and a function's result type.

\subsection{Disjointness}
Our design of match types induces an additional constraint on match type reduction: the scrutinee type must be known to be \emph{disjoint} with all type patterns preceding the matching case. Informally, disjointness means that two types have no shared inhabitants.

The necessity for disjointness is best illustrated with an example. Consider \dotty{Seq[Int]}, the type of integer sequences. \dotty{Elem[Seq[Int]]} does not reduce:
\begin{enumerate}
  \item \dotty{Elem}'s first case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{String}.
  \item \dotty{Elem}'s second case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{List[Int]} (lists are sequences, but not the other way around).
  \item \dotty{Elem}'s third case is \emph{not considered} because \dotty{Seq[Int]} and \dotty{List[Int]} are not disjoint.
\end{enumerate}
Therefore, the reduction algorithm gets stuck on the second case and the overall type is irreducible.
Without disjointness, \dotty{Elem[Seq[Int]]} would reduce to \dotty{Seq[Int]} (via \dotty{Elem}'s third case), which would be unsound. For example, the expression \dotty{elem[Seq[Int]](List(1,2,3))} would have type \dotty{Seq[Int]}, but evaluates to the integer \dotty{1}.

\Cref{subsub:subtyping-and-disjointness} revisits this counterexample in a formal setting. \Cref{subsec:disjointness-in-scala} discusses our implementation of disjointness in the Scala compiler.

\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ \\and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to additions to \SystemFsub.}

\FMTypingRules{\SystemFm type system for a given set of classes $\C$ with class inheritance $\Ψ$ \\and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to additions to \SystemFsub.}

\clearpage
\section{Formalization}
\label{sec:formalization}

In this section, we formally present \SystemFm, an extension of \SystemFsub~\cite{cardelli1994extension} with pattern matching, opaque classes, and match types.
\Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
We discuss differences to \SystemFsub in the following subsections (\Cref{subsec:classes} and \ref{subsec:matches}).
In \Cref{subsec:type-safety}, we outline a proof of type safety for \SystemFm.
In \Cref{subsec:type-binding-extension}, we present an extension of \SystemFm with support for binding pattern variables in type patterns.

\subsection{Classes}
\label{subsec:classes}

\SystemFm is parametrized by a set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$.
The class inheritance forms a partial order on $\C$, that is, it is reflexive, antisymmetric and transitive.
The class disjointness is symmetric relation over $\C$.
Inheritance and disjointness must be consistent in the sense that $\(\A, \B) \∈ \Ξ$ implies that there is no class $\C$ such that $\(\C, \A) \∈ \Ψ$ and $\(\C, \B) \∈ \Ψ$.

Each class in $\C$ gives rise to a constructor (written $\new \C$), a type (written $\C$), and a constructor singleton type (written $\{ \new \C \})$.
The type $\C$ denotes all values that inherit $\C$, while the constructor singleton type $\{ \new \C \}$ denotes a single value: $\C$'s constructor call.
Subtyping between classes is dictated by $\Ψ$ via the \SPsi rule.

This parametric approach allows us to model class inheritance as found in object-oriented languages without the need for dedicated syntax for classes and data type definitions.
Although our approach might appear simplistic, it can easily model advanced object-oriented features such as multiple inheritance.
We discuss the encoding of Scala's types into \SystemFm in \Cref{subsec:disjointness-in-scala}.

Our type system refers to classes by names and therefore mixes structural and nominal types.
Names are useful to give a direct correspondence between runtime tags and compile-time types.
As we will see, runtime tags are essential to runtime type testing and play a central role in the evaluation of pattern matching.

\subsection{Matches}
\label{subsec:matches}

\SystemFm supports both pattern matching on the term level (\emph{match expressions}) as well as on the type-level (\emph{match types}).
Matches, both on terms and on types, are composed of a scrutinee, a list of cases and a default expression/type.
Each case consists of a \emph{type test} and a corresponding expression/type.
At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \cite{abadi1991dynamic}).
At the type level, a type test corresponds to a subtyping test with a particular type.
This disparity reflects the difference between runtime, where type tests are implemented using class tables, and compile time, where types are compared using the type system in its full extent.
We discuss the representation of Scala types at runtime in \Cref{subsec:types-at-runtime}.

Throughout this paper, we use the abbreviated syntax $\t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$ to denote an arbitrary number of cases, that is, $\∃ \n \∈ \ℕ.~ \t_s \match \{ \x_{1} \: \C_{1} \⇒ \t_{1}; \dots; \x_n \: \C_n \⇒ \t_n \} \otherwise \t_d$.

Match expressions and match types are related by the \TMatch typing rule.
This rule operates by typing each component of a match expression to then assemble the corresponding match type.

\begin{example}
\label{ex:simplereduction}
For example, given two disjoint classes $\A$ and $\B$, and an empty class inheritance ($\Ψ \= \Id, \Ξ \= \{ \(\A, \B), \(\B, \A) \}$); the following function:
\\\indent $\f \= \λ \X \<: \Top.~ \λ \x \: \X.~ \x \match \{ \a \: \A \⇒ \foo; \b \: \B \⇒ \bur \} \otherwise \buzz$
\\\noindent
can be typed precisely as
\\\indent $\f \hspace{4pt} \: \hspace{2.5pt} \∀ \X \<: \Top.~ \X \→ \X \match \{ \A \⇒ \Foo; \B \⇒ \Bur \} \otherwise \Buzz$
\\\noindent
where $\foo$, $\bur$ and $\buzz$ are expressions with types $\Foo$, $\Bur$ and $\Buzz$, respectively.
\end{example}

The cases of a match expression are evaluated \emph{sequentially}: the scrutinee is checked using the type test of each case, one after the other.
The overall expression reduces to the expression that corresponds to the first successful type test (\EMatch2).
When no type test succeeds, the match evaluates to its default expression (\EMatch3/4/5).
For instance, given the function $\f$ defined in \Cref{ex:simplereduction}, the expression $\(\f~\A~\(\new \A))$ evaluates to $\foo$ and $\(\f~\B~\(\new \B))$ to $\bur$.

% Even though \Fm is limited to type test patterns, more elaborate patterns can be expressed using type tests.
% For example, consider the following data type deconstruction pattern (left-hand side):

% list match {
%   case x :: xs => print(x)
%   case Nil => ()
% }

% list match {
%   case l : ::[Any] => print(l.head)
%   case Nil => ()
% }

% \noindent
% This pattern could alternatively be written as type test pattern (right-hand side), were \dotty{::[Any]} is the type of Scala's cons node.

\subsubsection*{Match Type Reduction}
The subtyping relation contains five rules for match type reduction, \SMatch1/2/3/4/5.
These rules are defined in pairs using the $\≡$ shorthand notation, where $\S \≡ \T$ means that $\S$ and $\T$ are in a mutual subtyping relation.
More precisely, \SMatch1/2 in \Cref{fig:FMTypingRules} corresponds to two typing rules with identical premises and symmetrical conclusion, and the same goes for \SMatch3/4.

The typing rules for match type reduction are best explained as generalizations of the evaluation relation.
Given a match type $\M \= \T_s \match \{ \C_i \⇒ \T_i \} \otherwise \T_d$, $\M$ reduces to $\T_i$ if and only if, for every value $\t_s$ in $\T_s$, the term level expression $\t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$ evaluates to $\t_i$.

The \SMatch1/2 rules correspond to the evaluation of a match expression to its $\n$th case (\EMatch2):

\SMatchAinline

The first premise ensures that the $\n$th case type test will succeed for every possible value in the scrutinee type $\T_s$.
Conversely, the second premise is a disjointness judgment, which ensures that no value in the scrutinee type would result in a successful type test for cases prior to the $\n$th case.
The \SMatch3/4 rules correspond to an evaluation to the default case (\EMatch2), and require disjointness between the scrutinee type and each type test type:

\SMatchBinline

Disjointness between two classes can be concluded directly using the \DXi rule which uses the class disjointness $\Ξ$.
Likewise, disjointness between a constructor singleton type and a class can be concluded directly by inspecting the class inheritance $\Ψ$ (\DPsi).
Function types and universal types are disjoint from classes as they are inhabited by different values (\DArrow, \DAll) and thus will never match.
The last disjointness rule, \DSub, states that if $\U$ and $\T$ are disjoint, then all subtypes of $\U$ are also disjoint with $\T$.

\begin{example}
We continue developing \Cref{ex:simplereduction} by showing how match type reduction rules can be used to conclude that $\(\f~\B~\(\new \B))$ has type $\Bur$.
Using \TTApp and \TApp, the expression can be typed as follows:
\\\indent
$\f~\B~\(\new \B) \hspace{2pt} \: \hspace{3pt} \B \match \{ \A \⇒ \Foo; \B \⇒ \Bur \} \otherwise \Buzz$
\\\noindent
Since our example assumes an empty class inheritance and $\(\A, \B) \∈ \Ξ$, the \SMatch1 rule gives:
\\\indent
$\Ø \⊢ \B \match \{ \A \⇒ \Foo; \B \⇒ \Bur \} \otherwise \Buzz \<: \Bur$
\\\noindent
Finally, using \TSub we get $\(\f~\B~\(\new \B)) \: \Bur$.
% This example also illustrates how \SystemFm enables simple forms of dependent typing.
% Indeed, the type of this function application \emph{depends} on its arguments; the function $\f$ is dependently typed.
\end{example}

\subsubsection*{Subtyping and Disjointness}\hspace{-1pt}
\label{subsub:subtyping-and-disjointness}
One might wonder what happens if we simplify the match type reduction rules by replacing premises of the form $\Γ \⊢ \disj(\T, \C)$ by seemingly equivalent premises of the form $\(\T, \C) \∉ \Ψ$.
Unfortunately, the resulting system would be unsound, which can be demonstrated with a counterexample.
Let us assume the function~$\f$ defined in \Cref{ex:simplereduction}, adding a new class $\E$ with $\Ψ \= \{ \(\E, \A), \(\E, \B) \}$ and $\Ξ \= \{ \}$.
Now, consider the term $\(\f~\B~\(\new \E))$. Since we have $\Ø \⊢ \E \<: \B$, this function application is well-typed and, given that $\(\E, \A) \∈ \Ψ$, evaluates to $\foo$. The term-level and type-level reductions are inconsistent!
The unsoundness arises when using $\(\B, \A) \∉ \Ψ$ with the modified \SMatch1 rule to wrongly conclude that $\(\f~\B~\(\new \E))$ has type $\Bur$. This would result in an inconsistency between types $\(\e \: \Bur)$ and evaluation $\(\e \⟶ \foo)$, and violate type soundness.
In \SystemFm, the match type obtained when typing $\(\f~\B~\(\new \E))$ does not reduce since the scrutinee type $\B$ is neither disjoint with, nor a subtype of the first pattern type test $\A$.
In this case, the match type is \emph{stuck}, and is assigned as is.
Stuck types can appear as the result of programming error, but can also be due to the local irreducibility of a match type.
For instance, the body of~$\f$ is typed with a stuck type, as shown in \Cref{ex:simplereduction}, but that type can later become reducible depending on type variable instantiations.

\subsection{Type Safety}
\label{subsec:type-safety}

We show the type safety of \SystemFm through the usual progress and preservation theorems.
This section provides an overview of the proof structure and states the involved lemmas and theorems.
Detailed proofs are available in the supplementary material of this paper in two different versions.
The first version is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
The second version is a mechanization of the proof in Coq,
using the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings.
Our mechanization uses a simplified representation of match types with exactly one case per match.
Matches with multiple cases can be expressed by nesting match types in default cases.

\structure{Structure of the type safety proof. Arrows represent implications between lemmas and theorems.}

% There is nothing special about this, omitted since it is standard.
% An important technical detail that we omitted in \Cref{fig:FMTypingRules} concerns the well-formedness of types on contexts.
% A type $\T$ is well-formed in $\Γ$ when all the free variables in $\T$ are bound in $\Γ$.
% A context $\Γ$ is well-formed if each name is bound at most once and if each binding is to a well-formed type in the context preceding its binding.
% To improve presentation, we have omitted all well-formedness premises from our typing rules.

\Cref{fig:structure} gives an overview of the proof structure by showing implications between the various lemmas and theorems.
The basic structure resembles that of \SystemFsub's standard safety proof from~\cite{pierce2002types}.
We continue our presentation by introducing the lemmas and theorems used in our type safety proof.

\subsubsection*{Preliminary Lemmas}
Our proof begins with preliminary technical lemmas: \prelemma{Permutation}, \prelemma{Weakening}, \prelemma{Strengthening}, \prelemma{Substitution}.
We omit stating these lemmas as they are entirely standard yet relatively lengthy given that they span the three relations of our system: typing, subtyping, and disjointness.
As usual, their proofs follow by mutual inductions on derivations.

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P$.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

\subsubsection*{Inversion of Subtyping}
\harpoon{Definition of the auxiliary relation $\⇌$, used to state inversion of subtyping (\Cref{lem:inversion-of-subtyping}).}

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
It represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

Similarly to inversion of subtyping, our canonical forms lemma is non-standard in that it uses a disjunction in its premise to account for match types.

\input{proofs/canonical-forms.tex}

\subsubsection*{Proof of Soundness}
The remaining proof of soundness is mostly standard.
\Cref{lem:inversion-of-typing} and \labelcref{lem:minimum-types} are simple inversions of typing rules.
\Cref{lem:minimum-types} is needed in the proof of preservation to recover subtyping bounds from typing judgments.
The proofs proceed by routine induction on derivations.
%
\input{proofs/inversion-of-typing.tex}
\input{proofs/minimum-types.tex}
%
With these lemmas in place, the proofs of progress and preservation are straightforward.
%
\input{proofs/progress.tex}
\input{proofs/preservation.tex}

\subsection{Type Binding Extension}
\label{subsec:type-binding-extension}

\FMExtension{\SystemFmB syntax, evaluation and typing rules for a given set of ground classes $\A$, set of parametric classes $\B$, class inheritance $\Ψ$, and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to changes made to \SystemFm.}

In this section, we present \SystemFmB, an extension of \SystemFm with support for binding pattern variables in type patterns.
\FmB is parametrized by two sets of classes, $\A$ and $\B$, representing non-parametric and parametric classes, respectively.
A parametric class, written $\B~\T$, takes exactly one type parameter\footnotemark.
This approach allows us to reuse most of \Fm's definitions.
Indeed, our formal development treats $\C$, $\Ψ$, and $\Ξ$ as mathematical objects, and is compatible with \FmB's new definition of classes.

\footnotetext{
The restriction to a single type parameter is for presentation purposes.
Both \SystemFmB's type system and type-safety proof can easily be adapted to support a variable number of binding variables.
}

In \Cref{fig:FMExtension}, we define \SystemFmB syntax and rules, where changes to \SystemFm are highlighted in gray.
\FmB's new syntax for match expressions and match types adds a \emph{pattern variable} to each construct.
In $\t_s \match \[\X] \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$, the pattern variable $\X$ is available to bind type parameters in $\C_i$ patterns.

The definitions of \SMatch3/4, \SMatch5, and \TMatch require minor adjustments to account for the pattern variable in typing contexts.
Note that pattern variables appear in contexts unconditionally, regardless of whether or not those variables are used in the corresponding patterns.

In the new subtyping rule for non-default match reduction, \BSMatch1/2, the first premise instantiates the pattern variable $\X$ to some type $\U$ such that the scrutinee type is a subtype of the $\n$th pattern:
\BSMatchInline
Here $\U$ is completely unconstrained: any instantiation of $\X$ such that $\T_s \<: \S_n$ would be admissible.
The disjointness judgments use a weaker upper bound for $\X$ than the subtyping judgment ($\X \<: \Top$ instead of $\X \<: \U$).
This is because the scrutinee type must be shown disjoint with non-matching pattern types for every possible instantiation of $\X$.
In an algorithmic system, $\U$ would be computed during type inference by constraint solving.

The new evaluation rule for non-default match reduction uses a similar mechanism: it looks for the \emph{first case} where the pattern variable can be instantiated such that the scrutinee inherits the corresponding pattern:

\BEMatchInline

The second premise rules out non-matching cases with a universal quantifier ranging over all types.
A concrete implementation would certainly opt for a more efficient approach, for instance by implementing $\Ψ$ as a lookup table.

\begin{example}
\label{ex:binding}
Consider the following class hierarchy with two ground classes: $\Char$ and $\String$, and a single parametric class $\List$, such that $\String$ extends $\List~\Char$:
%
\begin{flalign*}%
\indent
  \A &= \{ \Char, \String \}
& \B &= \{ \List \}
\hspace{200pt}\\\indent
  \Ψ &= \{ (\String, \List~\Char) \} ∪ \Id
& \Ξ &= \{ \}
\hspace{200pt}\\\indent
  \f &= \λ \x \: \Top.~\x \match [\X] \{ \xs \: \List~\X \⇒ \foo \} \otherwise \bur
\hspace{-100pt}
\end{flalign*}%
%
The function $\f$ matches its arguments against the $\List~\X$ pattern, where $\X$ is a pattern variable.
We examine the evaluation of two applications of $\f$:

\begin{enumerate}
  \item $\f \(\new \String)$ matches against $\List~\X$ with $\X \= \Char$ and evaluates to $\[\X \↦ \Char][\x \↦ \new \String] \foo$ via \BEMatch2 (since $\(\String, \List~\Char) \∈ \Ψ$).
  \item $\f \(\new \List~\Top)$ also matches $\List~\X$, this time with $\X \= \Top$, and evaluates to $\[\X \↦ \Top][\x \↦ \new \List~\Top] \foo$ via \BEMatch2. Here $\(\List~\Top, \List~\Top) \∈ \Ψ$ follows from $\Ψ$'s reflexivity.
\end{enumerate}
\end{example}

We established the type safety of \SystemFmB by adapting \SystemFm's pen-and-paper proof.
The required changes are lengthy, but relatively uninteresting; it boils down to additional bookkeeping of pattern variables in contexts.
The main takeaway from \FmB's type safety is that the proof does not require additional constraints on type $\U$ in \BSMatch1/2.
As a result, algorithmic implementations are free to use any mechanism to come up with instantiations of pattern variables.

\section{Implementation}
\label{sec:implementation}

Match types are implemented in Dotty, the reference compiler for Scala 3.
This section explains how our implementation relates to the formalization presented in \Cref{sec:formalization}.

In the compiler, match-type reduction happens during subtyping, just like in \SystemFm.
In order for subtyping to remain algorithmic, match type reduction rules are never used to introduce new match types, but only to simplify the ones present in the original program.
The reduction algorithm closely follows the typing rules of \Cref{sec:formalization}.
The scrutinee type is compared with each pattern sequentially.
If the scrutinee is a subtype of the first pattern type, the match type reduces.
Otherwise, if the scrutinee can be shown to be disjoint with the first pattern type, the algorithm proceeds to the next pattern.
If the algorithm reaches a pattern where neither subtyping nor disjointness can be concluded, the reduction is aborted and the match type remains unreduced.

\subsection{Disjointness in Scala}
\label{subsec:disjointness-in-scala}

Separate compilation is the biggest obstacle to concluding that two types are disjoint.
Indeed, in Scala, all traits and classes are extensible by default.
Because Scala programs are compiled with an open-world assumption, it is common for types to be effectively disjoint in the current compilation unit, but due to potential extensions in future compilations, the compiler must stay conservative.

Separate compilation is the reason why our formalization requires two different parameters to describe its class hierarchy.
One particular instantiation of \SystemFm can be thought of as a model of Scala's type system for a particular compilation unit, where $\C$ represents the set of classes declared \emph{so far}.
The class inheritance, $\Ψ$, remains valid for all subsequent compilation units: new class definitions do not alter the inheritance between previously defined classes.
However, the inheritance parameter ($\Ψ$) is, on its own, not sufficient to conclude that two classes are disjoint: new class definitions can introduce new overlaps between existing classes.
For this reason, our formalization uses a separate parameter to describe class disjointness ($\Ξ$).
To account for separate compilation, $\Ξ$ should only contain pairs of classes which would remain disjoint despite potential additions to the current set of classes.

Thankfully, Scala provides several ways to restrict extensibility.
The \emph{sealed} and \emph{final} annotations on traits and classes directly restrict the extensibility of annotated types: sealed types can only be extended in the same file as its declaration, thereby providing a way to enumerate all the children of a type.
Thus, disjointness of sealed traits and classes can be computed recursively by iterating over all the possible subtypes of that type.
The main distinction between traits and classes is that a class can extend at most one superclass.
This property allows the compiler to assert that classes are disjoint with a simple check: given two classes $\A$ and $\B$, if neither $\A \<: \B$ nor $\B \<: \A$, then no class could possibly extend both $\A$ and $\B$, and those two types are disjoint.

As an example, consider the following Scala definitions (left-hand side), and the corresponding instantiation of \SystemFm (right-hand side):

\vspace{-10pt}\noindent%
\begin{minipage}[t]{.475\linewidth}%
\begin{lstlisting}
sealed trait Part
final class Wheel extends Part
final class DiscBrake extends Part
\end{lstlisting}%
\vspace{-3pt}%
\begin{lstlisting}
trait Vehicle
class Bicycle extends Vehicle
class RoadBike extends Bicycle
\end{lstlisting}%
\vspace{-3pt}%
\begin{lstlisting}
class Helmet
\end{lstlisting}%
\end{minipage}%
%
\begin{minipage}[t]{.52\linewidth}%
\hfuzz=3pt
\begin{flalign*}%
       \C = \{ & \P, \W, \D, \V, \B, \R, \H \}\\[3pt]
\Ψ = \{ & (\W, \P), (\D, \P), (\B, \V), (\R, \B), (\R, \V) \}\\[3pt]
\Ξ = \{ & (\P, \V), (\P, \B), (\P, \R), (\P, \H),
\\      & (\W, \V), (\W, \B), (\W, \R), (\W, \H),
\\      & (\D, \V), (\D, \B), (\D, \R), (\D, \H),
\\      & (\W, \D), (\B, \H), (\R, \H) \}
\end{flalign*}%
\hfuzz=0pt
\end{minipage}

\noindent
The classes and inheritance relation are practically isomorphic between the two representation: Scala classes have a one-to-one correspondence to their \Fm counterparts (abbreviated with initials) and the inheritance only contains an additional entry for $\R$ and $\V$, obtained by transitivity ($\Ψ$'s reflexivity and $\Ξ$'s symmetry are omitted for brevity).

$\P$ is declared sealed, meaning that no additional parts can be defined outside of this compilation unit.
As a result, we can enumerate all parts to conclude that none are vehicles and $\(\P, \V) \∈ \Ξ$.
Note that this would not be the case if either $\W$ or $\D$ was declared non-final, since extending those classes would indirectly create new parts.
$\B$ and $\H$ are both classes that do not inherit each other, which implies that $\(\B, \H) \∈ \Ξ$ given that Scala classes can extend at most one class.
$\V$ and $\H$, however, cannot be concluded disjoint in those definitions.
If that turns out to be a desirable property, disjointness could easily be obtained by sealing $\V$ or finalizing $\H$.

\subsection{Empty Types}
\label{subsec:empty-types}

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called \dotty{Nothing}, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting any reduction.

To show why empty types are problematic, we can construct an example where, in the absence of an inhabitance check, the same match type could be reduced differently in two different contexts:
%
\begin{lstlisting}
type M[X] = X match {
  case Int => String
  case String => Int
}
class C {
  type X
  def f(bad: M[X & String]): Int = bad
}
class D extends C {
  type X = Int
}
\end{lstlisting}
%
In this example, the definition of \dotty{f} in \dotty{C} type-checks because \dotty{X & String} and \dotty{Int} are disjoint (since \dotty{String} and \dotty{Int} are disjoint) and \dotty{M[X & String]} reduces to \dotty{Int} (\dotty{M}'s second case applies).
Class \dotty{D} refines the definition of \dotty{C} by giving concrete definition of \dotty{X}.
The unsoundness manifests itself in the body of class \dotty{D}, where \dotty{X & String} is a subtype of \dotty{Int} and \dotty{M[X & String]} reduces to \dotty{String} (\dotty{M}'s first case applies).
There, it is possible to call the function~\dotty{f} with a string argument, which would result in a runtime error.
Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent \dotty{M[Int & String]} from reducing given that \dotty{Int & String} is not inhabited.

\subsection{Null Values}
\label{subsec:null-values}

In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \cite{nieto2020scala}.
Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.

% To allow easy migration from older versions, the strict treatment of nulls is still optional in Scala 3.0, enabled by the command line option \dotty{-Yexplicit-nulls}. The plan is to make strict null checking the default in the future.

\subsection{Variance}

Scala supports variance annotations on type parameters of higher-kinded types.
These annotations allow programmers to specify how the subtyping of annotated parameters influences the subtyping of the higher-kinded type.
For instance, \dotty{type F[+T]} defines a type \dotty{F} that is covariant in its first type parameter, meaning that \dotty{T1 <: T2} implies \dotty{F[T1] <: F[T2]}.
Contravariance, written \dotty{type G[-T]}, has the opposite meaning: \dotty{T1 <: T2} implies \dotty{G[T2] <: G[T1]}.

It would appear that co- and contravariant types are always overlapping, given that, for all types \dotty{X}, \dotty{F[Nothing] <:} \dotty{F[X]} and \dotty{G[Any] <: G[X]} (where \dotty{Nothing} and \dotty{Any} are Scala's bottom and top types).
However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the \dotty{Nothing} instantiation can be ruled out because no runtime program can produce a value of that type.

Scala tuples, for example, fall into this category.
\dotty{Tuple2}, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, \dotty{Tuple2[Int,} \dotty{X]} and \dotty{Tuple2[String,X]}, although \dotty{Tuple2[Nothing,X]} is a subtype of both, there is no runtime value of type \dotty{Tuple2[} \dotty{Nothing,X]} (since \dotty{Nothing} is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.

\subsection{Pattern Matching Exhaustivity}
\label{subsec:pattern-matching-exhaustivity}

The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \cite{fengyun2016generic}.
Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
\SystemFm uses default cases as a replacement for systematic exhaustivity checks or runtime exceptions.

\subsection{Types at Runtime}
\label{subsec:types-at-runtime}

% \olivier{
% Type erasure is the process by which type arguments of generic types (type constructors) are removed from programs during compilation.
% Scala's erasure affects pattern matching because it prevents certain patterns from being checked at runtime.
% For instance, both `List[String]` and `List[Int]` erase to `List[Object]`, which makes them indistinguishable at runtime.
% This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
% Match types reduce during type checking (before erasure) and are not affected by this limitation.
% Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.
% }

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \cite{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

Erasure directly affects the type tests that can be performed at runtime.
For example, while \dotty{case xs: List[Int] =>} is a syntactically valid pattern, it will lead to a compiler warning since the \dotty{Int} type parameter is eliminated by erasure and cannot be checked at runtime.

This restriction is reflected in our formalism by the difference between the evaluation rules for match expressions and the reduction rules for match types: evaluation is limited to inheritance checks on statically defined classes ($\(\C, \C_n) \∈ \Ψ$ in \EMatch2/3), as opposed to the match type reduction rules which are defined using the subtyping and type disjointness relations ($\Γ \⊢ \T_s \<: \S_n$ and $\Γ \⊢ \disj(\T_s, \S_m)$ in \SMatch1/2/3/4).

In \SystemFmB (\Cref{subsec:type-binding-extension}), match expressions support two sorts of parametric patterns: they can be either instantiated (\!\!$\match \{ \xs \: \List~\Int \}$, where $\Int \∈ \A$), or use a binding pattern variable (\!\!$\match \[\X] \{ \xs \: \List~\X \}$, where $\X$ is a pattern variable).
In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.

\subsection{Non-Termination}
\label{subsec:non-termination}

Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
Recursive match types can cause subtyping checks to loop indefinitely.
Our implementation does not check match types for termination, as any such check would necessarily limit expressiveness or convenience.
Instead, we detect divergence during match type reduction using a fuel mechanism.
The compiler is given an initial amount of fuel, which is consumed one unit at a time on every reduction step.
If the compiler runs out of fuel, the reduction is aborted with a "recursion limit exceeded" error.
The current implementation uses a fixed amount of initial fuel.
Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \cite{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.

\subsection{Inference}

\SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
Pattern matching in Scala supports many sorts of patterns~\cite{emir2007matching}, most of which do not have a match type counterpart.
Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
Instead, explicit type annotations must be provided.

\subsection{Caching}

Scala's type-checking algorithm makes heavy use of caching to improve its performance.
Special care must be taken when caching the result of match type reduction, given that the subtyping and disjointness checks are context-dependent.
Our implementation uses a context-aware cache for match types that automatically invalidates reduction results when match types are reduced in new contexts.
An example where naive caching would be incorrect can be found in \Cref{subsec:roles-in-haskell}.

\subsection{Size of the Implementation}

In terms of lines of code, our match type implementation is a relatively modest addition to the Scala compiler: the overall changes amount to around 1500 lines (excluding tests and documentation).

% Roughly speaking, it's probably less given the diff overlap...
% https://github.com/lampepfl/dotty/pull/4964 +822
% https://github.com/lampepfl/dotty/pull/5996 +364
% https://github.com/lampepfl/dotty/pull/6050 +144
% I don't count this one as I'm sure all these lines were touched afterwards:
% https://github.com/lampepfl/dotty/pull/6319
% https://github.com/lampepfl/dotty/pull/7364 +73
% https://github.com/lampepfl/dotty/pull/7364 +80
% https://github.com/lampepfl/dotty/pull/8024 +87

\section{Case Study: Shape-Safe NumPy}
\label{sec:evaluation}

In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\cite{harris2020numpy}.
The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\cite{chen2017typesafe, huang2017hasktorch}.
Our library uses match types to provide a shape-safe NumPy-like interface.
We believe that the approach presented in this case study is an improvement over the status quo because it does not use metaprogramming or any sort of convoluted encoding to express type-level operations.

% \subsection{Related Work}
% Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
% The Dex language~\cite{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
% Nexus~\cite{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
% However, it only guards against rank errors, and not ndarray size errors.

% Match types offer an abstraction that can be used to express what these solutions cannot. They allow us to program shape transformations and type constraints, in order to  achieve shape-safety for variable ranks.
% The implementation in this case study is inspired by the HaskTorch Haskell library~\cite{huang2017hasktorch}, which also represents shapes using HLists, and makes direct use of type-level computation to compute shape transformations.

\subsection{Shape Errors in Python}
In the example Python code below, the \python{img_batch} ndarray is a batch of 25 randomly generated RGB images of size 256$\×$256.
The code aims to compute a vector of length 25 containing the average grayscale color of each image in the batch, and then create a square 5$\×$5 image of the average grayscale colors.
However, this code contains a shape error and will throw an error at runtime.
%
\badreducecodesection
%
The error is in the call to \python{np.mean}, which takes as argument a list of axes to reduce along.
Unfortunately, the arguments to \python{np.mean} are off-by-one and result in a vector of length 3 (\python{img_batch}'s last dimension) instead of the intended length of 25 (\python{img_batch}'s first dimension); \python{avg_color} thus contains the average RGB color of the batch instead of the average grayscale color for each image.
The reshape operation will then fail at runtime, as it cannot reshape a 3-element vector into a 25-element matrix.
This error can be difficult to spot, given that NumPy's interface for reducing along multiple axes is index-based.
Runtime errors like this one can be particularly frustrating when they occur late in a long-running computation.

In the remainder of this section, we show how match types can be used to prevent this class of error.
After a preliminary introduction of singleton types (\Cref{subsec:singleton-types}), we introduce ndarray shapes at the type level using HLists (\Cref{subsec:array-shape}).
In \Cref{subsec:shape-match-types}, we show type-level implementations of the \python{np.mean} and \python{np.reshape} operations using match types.
Finally, we show how this newly defined API can detect and report the error from our original Python example.

\subsection{Singleton Types}
\label{subsec:singleton-types}

Scala supports singleton types, which are types inhabited by a single value~\cite{leontiev2014sip}.
For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
For instance, \dotty{type +[A <: Int,}\dotty{B <: Int]} represents the addition of integer singleton types.
Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
Representing numbers with singleton types is desirable for practical purposes, but not absolutely necessary for this case study (for instance, Peano numerals could be used instead).

\subsection{Type-Level Array Shape}
\label{subsec:array-shape}

The shape of an ndarray is a list of dimension lengths; we say that the shape $\(\a_{1}, \a_{2}, \dots, \a_n)$ has $\n$ dimensions.
For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\cite{kiselyov2004strongly}.
We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
\footnotetext{%
In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
%
\shapecodesection
%
This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.

To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
%
\ndarraycodesection

The goal of our presentation is to define type-safe operations on \dotty{NDArrays}.
Since our focus is on the type-level, we do not include value-level counterparts to \dotty{T} and \dotty{S} in the definition of \dotty{NDArray}, but this would be necessary in a complete implementation.

To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
\footnotetext{%
This snippet uses the triple question mark operator, Scala's standard notation for missing or omitted implementations.
}
%
\randomnormalcodesection

\subsection{Computation on Shapes with Match Types}
\label{subsec:shape-match-types}

Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
This constraint does not require any match types, but can simply be expressed as:
%
\multiplycodesection
%
However, we will need the additional expressiveness of match types to implement more complex constraints on array shapes, such as for reshapes (\ref{subsubsec:reshape}) and reductions (\ref{subsubsec:reduction}).

\subsubsection{Reshape}
\label{subsubsec:reshape}

An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape. The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
This can be naturally expressed with a match type:
%
\numelementscodesection

To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
%
\reshapecodesection
%
With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.

\subsubsection{Reduction}
\label{subsubsec:reduction}

The NumPy API provides a variety of functions to reduce along a set of axes of an ndarray, such as \python{np.mean(ndarray, axes)} or \python{np.var(ndarray, axes)}.
The \python{axes} parameter is a list of indices of dimensions, listing exactly those dimensions that will no longer be present in the output ndarray.
The dimension indices can be unordered and repeated, and out-of-bounds indices result in an error.
In the Python API, passing the \python{None} value instead of a list of axes reduces along all axes, meaning that the operation returns a scalar.
Note that this is the opposite of passing \dotty{Ø}, which means that we reduce over no axes (effectively a no-op).

This behavior is more complex than the previous examples, but can still be described accurately by a match type.
We use a match type called \dotty{Reduce} to compute the return shape of the operation.

\npmeancodesection

We implement reductions along a given list of indices with logic similar to two nested loops.
The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
%
\reducecodesection

\subsection{Shape safety}
\label{sec:shape-safety}

Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
%
\badreducescalacodesection
%
As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.

% \goodreducescalacodesection

\section{Related work}
\label{sec:related-work}

In this section, we provide a review of existing work and relate match types to dependently typed calculi with subtyping, intensional type analysis, type families and roles in Haskell, and conditional types in TypeScript.

\subsection{Dependently Typed Calculi with Subtyping}
\label{subsec:dependently-typed-calculi-with-subtyping}

There is a vast amount of literature on type systems combining subtyping with dependent types, justifying a full survey to relate it appropriately.
Instead, we offer a condensed summary of our reading journey and explain what led us to decide on using \SystemFsub as a foundation for our formalization.

Dependently typed calculi typically use the same language to describe terms and types.
This unification is also commonly used in the presence of subtyping~\cite{zwanenburg1999pure, hutchins2010pure, yang2017unifying}.
For systems with a complete term/type symmetry, this is a natural design, as it is concise and simplifies the meta-theory.
Unfortunately, the lack of distinction between term and type level renders these systems impractical for our purpose, given that our research takes place in the context of an existing language with a clear term/type separation.

Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\cite{aspinall1994singleton, stone2000deciding, courant2003strong}.
Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
This mechanism is appealing because it allows type system designers to cherry-pick the term constructs that should be allowed in types.
When multiple constructs are shared between terms and types, singleton types provide a clear economy of concepts.
In our study of match types, a minimal use of singleton types would result in sharing a single constructor between the term and the type languages: the constructor for matches. It is unclear if the benefits in doing so would outweigh the additional complexity.

Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\cite{essence2016amin}.
DOT does not directly support any form of type-level computation.
We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\cite{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.

After several attempts at formalizing match types within existing systems, we decided to pursue a simpler route of adding new constructs to a system without dependent types.
After all, the primary purpose of \SystemFm is to serve as a medium to concisely \emph{explain} our type-checking algorithm for match types.
For this reason, we built our work on top of \SystemFsub, which we believe should be the simplest, most familiar calculus among the systems cited in this section.

\subsection{Intensional Type Analysis}
\label{subsec:intensional-type-analysis}

In their work on intensional type analysis~\cite{harper1995compiling}, Harper and Morrisett introduce the $\λ^{ML}_i$ calculus that supports structural analysis of types.
In $\λ^{ML}_i$, types are represented as expressions that can be inspected by case analysis using a "typecase" construct, available both at the term and at the type level.
Match types can be seen as an extension of intensional type analysis to work with object-oriented class hierarchies and subtyping.
Whereas patterns in $\λ^{ML}_i$ are limited to a fixed set of disjoint types, match types need to deal with open class hierarchies, of which not all members are known at compile-time.
This means pattern types can overlap, and we need to perform an analysis of disjointness between the scrutinee type and each pattern type.
Disjointness allows for sound reduction in the presence of overlapping patterns and abstract scrutinee types, while retaining the natural sequential evaluation order of pattern matching.

\subsection{Type Families in Haskell}
\label{subsec:type-families-in-haskell}

Haskell's type families allow programmers to define type-level functions using pattern matching~\cite{chakravarty2005associated, schrijvers2008open}.
By default, type families are open, which means that a definition can spread across multiple files and compilations.
This flexibility induces a substantial restriction on type family definitions: patterns must not overlap.
One benefit of this restriction is that it prevents any ambiguity in the reduction of type families (patterns are pairwise disjoint), which is required given the distributed nature of definitions.
Open type families are well-suited to be used in conjunction with type classes, since both constructs have open-ended definitions with non-overlapping constraints.

Closed type families (CTFs), as introduced by~\citet{eisenberg2014closed}, allow for overlapping cases in type family definitions.
Unlike the open variant, CTF reduction is performed sequentially, based on unification and apartness checks.
In this regard, CTFs are closely related to match types.
In fact, if we replace unification checks with subtyping and apartness checks with disjointness, their reduction algorithm is practically identical to ours, from a high-level perspective.

By default, Haskell checks for termination of recursive type families, but this check can be disabled to increase type families' expressiveness.
Although the formalization presented in~\cite{eisenberg2014closed} does not cover non-terminating families, the paper discusses a soundness problem caused by non-termination.
The problem only occurs in the presence of repeated type bindings in patterns.
In Scala, match types (and pattern matching in general) do not allow repeated bindings and are therefore not affected by this problem.

% A significant difference between type-checking type families and match types is that type families are entirely handled during type inference, while match types reduce during subtyping.
% Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
% For instance, in a functional pearl on typed matrices~\cite{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.

\subsection{Roles in Haskell}
\label{subsec:roles-in-haskell}

Haskell's roles were introduced by~\citet{weirich2011generative} to fix a long-standing unsoundness caused by the interaction of open type families and the \haskell{newtype} construct.
We understand roles as type annotations which specify whether a given type can safely be nominally compared to other types, or if \emph{representational equality} (RE) should be used instead.
The word representation in RE refers to the runtime representation of a type.
In particular, RE dealiases \haskell{newtype} constructs.

In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
%
\begin{lstlisting}
trait AgeClass {
  type Age
  def addAge(a: Age, i: Int): Int
}
object AgeObject extends AgeClass {
  type Age = Int
  def addAge(a: Age, i: Int): Int = a + i
}
\end{lstlisting}
%
In this example, type \dotty{Age} is abstract in \dotty{AgeClass} and concrete in \dotty{AgeObject}.
In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates \dotty{Age} and \dotty{Int}.
Such type family would reduce differently in \dotty{AgeClass}, where the types are different, and in \dotty{AgeObject}, where those two types are synonyms, which can easily be exploited to obtain a runtime error.

Luckily, our design of match types is not affected by this issue.
The reason comes from the use of subtyping (and disjointness), which shields our implementation from incorrectly discriminating \dotty{Age} from \dotty{Int} when \dotty{Age} is abstract.
Consider the following match type definition (directly translated from the Haskell example):
%
\begin{lstlisting}
type M[X] = X match {
  case Age => Char
  case Int => Bool
}
\end{lstlisting}
%
Our algorithm would not reduce \dotty{M[Int]} to \dotty{Bool} in \dotty{AgeClass} as this reduction would require evidence that \dotty{Age} and \dotty{Int} are disjoint, which cannot be constructed when \dotty{Age} is an unbounded abstract type.

\subsection{Conditional Types in TypeScript}
\label{subsec:conditional-types-in-typescript}

TypeScript's conditional types, briefly mentioned in the introduction, are a type-level ternary operator based on subtyping.
Conditional types can be nested into a sequence of patterns that evaluate in order, making them similar to match types.

The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\cite{microsoft2020typescript}.
Given a type \typescript{S extends T ? Tt : Tf}, the TypeScript compiler first replaces all the type parameters in \typescript{S} and \typescript{T} by \typescript{any} (the top of TypeScript's subtyping lattice).
If the resulting types (after substitution) are not subtypes, the overall condition is reduced to \typescript{Tf}.
Unfortunately, this algorithm is both \emph{unsound} and \emph{incomplete}.

The unsoundness is caused by the incorrect widening of type parameters in contravariant position.
Although TypeScript does not have syntax for variance annotations, function types are covariant in their return type and contravariant in their arguments.
The conditional type unification algorithm wrongly approximates \typescript{X => string} to \typescript{any => string} and unifies the former with latter, which can lead to a runtime errors.

The incompleteness comes from the fact that type parameter approximation does not account for type parameter bounds.
Consider the following example:
%
\begin{lstlisting}[style=typescript]
type M<X> = X extends string ? A : B
function f<X extends string>: M<X> = new A
\end{lstlisting}
%
Here, TypeScript's reduction algorithm fails to recognize that \typescript{new A} can be typed as \typescript{M<X>}, even though \typescript{X} is clearly a subtype of \typescript{string} in \typescript{f}'s body.

Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\cite{bierman2014typescript}.
Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

\section{Conclusion}
In this paper, we introduced \emph{match types}, a lightweight mechanism for type-level programming that integrates seamlessly in subtyping-based programming languages.
We formalized match types in \SystemFm, a calculus based on \SystemFsub, and proved it sound.
Furthermore, we implemented match types in the Scala 3 compiler, making them readily available to a large audience of programmers.
A key insight for sound match types is the notion of disjointness, which complements subtyping in the match type reduction algorithm.
In the future, we plan to investigate inference of match types to avoid code duplication in programs that operate both at the term and the type level.

### 2021-10-22 article.tex
Import TLPwMT

@@
 
-\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ \\and class disjointness $\Ξ$.
+\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$.
 \hl{\text{Highlights}} correspond to additions to \SystemFsub.}
@@
 %
-\badreducecodesection
+\begin{lstlisting}[style=py]
+import numpy as np
+img_batch = np.random.normal(size=(25, 256, 256, 3))
+avg_colors = np.mean(img_batch, (0, 1, 2))
+avg_color_square = np.reshape(avg_colors, (5, 5))
+\end{lstlisting}
 %

### 2021-10-22 code-sections.tex
Import TLPwMT

\newcommand{\goodreducescalacodesection}{\lstinputlisting[style=scala, firstline=3, lastline=4, breaklines]{scala/src/main/scala/goodreduce.scala}}
\newcommand{\shapecodesection}{\lstinputlisting[style=scala, firstline=8, lastline=11, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\datatypecodesection}{\lstinputlisting[style=scala, firstline=19, lastline=19, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\ndarraycodesection}{\lstinputlisting[style=scala, firstline=23, lastline=23, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\randomnormalcodesection}{\lstinputlisting[style=scala, firstline=27, lastline=27, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\multiplycodesection}{\lstinputlisting[style=scala, firstline=31, lastline=31, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\numelementscodesection}{\lstinputlisting[style=scala, firstline=36, lastline=40, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\reshapecodesection}{\lstinputlisting[style=scala, firstline=44, lastline=45, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\reducecodesection}{\lstinputlisting[style=scala, firstline=50, lastline=66, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\npmeancodesection}{\lstinputlisting[style=scala, firstline=86, lastline=87, breaklines]{scala/src/main/scala/numpy.scala}}
\newcommand{\badreducescalacodesection}{\lstinputlisting[style=scala, firstline=2, lastline=4, breaklines]{scala/src/test/scala/badreduce.scala}}

### 2021-10-22 generate-code-sections.py

@@ -1,0 +1,53 @@
+#!/bin/python3
+import re
+import sys
+import os
+
+input_files = sys.argv[1:]
+if (len(input_files) == 0):
+    print("Expected command-line arguments")
+    print("Usages: ./generate-code-sections.py <...files>")
+    sys.exit(1)
+
+output_file = "code-sections.tex"
+listing_options = "breaklines"
+
+start_section = re.compile(r'\s*(?://|#)\s+start\s+section\s+(\w+)\s*')
+end_section = re.compile(r'\s*(?://|#)\s+end\s+section\s+(\w+)\s*')
+
+starts = {}
+finishes = {}
+
+for input_file in input_files:
+    with open(input_file, 'r') as f:
+        for i, line in enumerate(f.readlines()):
+            start_match = start_section.match(line)
+            if start_match:
+                starts[start_match.group(1)] = (input_file, i + 2) # minted start is inclusive and 1-based
+
+            end_match = end_section.match(line)
+            if end_match:
+                finishes[end_match.group(1)] = (input_file, i) # minted end is exclusive
+
+    if len(starts) == 0:
+        print(f"No sections defined in {input_file}!")
+        sys.exit(1)
+
+def dict_zip(*dicts):
+    return {k: [d[k] for d in dicts] for k in dicts[0].keys()}
+
+output = []
+for (section, positions) in dict_zip(starts, finishes).items():
+    if len(positions) < 2:
+        print(f"Ignoring section {section} as it does not have a start and a finish")
+    else:
+        [(start_file, start_line), (finish_file, finish_line)] = positions
+        if start_file != finish_file:
+            print(f"Start and finish for section '{section}' are in different files ({start_file}, {finish_file}")
+            sys.exit(1)
+        extension = os.path.splitext(start_file)[1][1:]
+        output.append(f"\\newcommand{{\\{section}codesection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
+
+with open(output_file, "w") as f:
+    f.write("\n".join(output))
+    print(f"Wrote {output_file}")

### 2021-10-22 scala/src/main/scala/goodreduce.scala

@@ -1,0 +1,9 @@
+val img_batch = random_normal(#:(25, #:(256, #:(256, #:(3, Ø)))))
+// start section goodreducescala
+val avg_colors = mean(
+  img_batch, #:(3, #:(1, #:(2, Ø))))
+// end section goodreducescala
+val avg_color_square: NDArray[Float, 5 #: 5 #: Ø] = reshape(avg_colors, #:(5, #:(5, Ø)))
+
+// SHOULD NOT COMPILE!! Uncomment
+// val wrong: NDArray[Float, 25 #: Ø.type] = mean(img_batch, #:(0, #:(1, #:(2, Ø))))

### 2021-10-22 scala/src/main/scala/numpy.scala

@@ -1,0 +1,88 @@
+sealed trait Shape
+final case class #:[H <: Int & Singleton, T <: Shape](head: H, tail: T) extends Shape
+case object Ø extends Shape
+
+// cheating a little bit here, but there's a footnote to explain it!
+object SimplifiedShape {
+// start section shape
+enum Shape {
+  case #:[H <: Int, T <: Shape](head: H, tail: T)
+  case Ø
+}
+// end section shape
+}
+
+type Ø = Ø.type
+type None = None.type
+
+// start section datatype
+trait DataType[T]
+// end section datatype
+
+// start section ndarray
+trait NDArray[T, S <: Shape]
+// end section ndarray
+
+// start section randomnormal
+def random_normal[S <: Shape](shape: S): NDArray[Float, S] = ???
+// end section randomnormal
+
+// start section multiply
+def multiply[T, S <: Shape](x: NDArray[T, S], y: NDArray[T, S]): NDArray[T, S] = ???
+// end section multiply
+
+import scala.compiletime.ops.int.*
+// start section numelements
+type NumElements[X <: Shape] <: Int =
+  X match {
+    case Ø => 1
+    case head #: tail => head * NumElements[tail]
+  }
+// end section numelements
+
+// start section reshape
+def reshape[T, From <: Shape, To <: Shape](arr: NDArray[T, From], newshape: To)
+    (implicit ev: NumElements[From] =:= NumElements[To]): NDArray[T, To] = ???
+// end section reshape
+
+import scala.compiletime.ops.int.+
+// start section reduce
+type Reduce[S <: Shape, Axes <: None | Shape] <: Shape =
+  Axes match {
+    case None => Ø
+    case Shape => Loop[S, Axes, 0]
+  }
+
+type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
+  S match {
+    case head #: tail => Contains[Axes, I] match {
+      case true => Loop[tail, Remove[Axes, I], I + 1]
+      case false => head #: Loop[tail, Axes, I + 1]
+    }
+    case Ø => Axes match {
+      case Ø => Ø
+      // otherwise, do not reduce further
+    }
+  }
+// end section reduce
+
+type Contains[Haystack <: Shape, Needle <: Int] <: Boolean = Haystack match {
+  case Ø => false
+  case head #: tail => head match {
+    case Needle => true
+    case _ => Contains[tail, Needle]
+  }
+}
+
[…]

### 2021-10-22 scala/src/test/scala/badreduce.scala

@@ -1,0 +1,5 @@
+// start section badreducescala
+val img_batch = random_normal(#:(25, #:(256, #:(256, #:(3, Ø)))))
+val avg_colors = mean(img_batch, #:(0, #:(1, #:(2, Ø))))
+val avg_color_square = reshape(avg_colors, #:(5, #:(5, Ø)))
+// end section badreducescala

### 2021-10-22 article.tex
s/paper/chapter/g

@@
 However, the combination of type-level programming and subtyping remains largely unexplored in practical programming languages.
-This paper presents \emph{match types}, a type-level equivalent of pattern matching.
+This chapter presents \emph{match types}, a type-level equivalent of pattern matching.
 Match types integrate seamlessly into programming languages with subtyping and, despite their simplicity, offer significant additional expressiveness.
@@
 
-This paper presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
+This chapter presents an alternative construct for type-level programming based on subtyping, which we call \emph{match types}.
 As the name suggests, match types allow programmers to express types that perform pattern matching on types:
@@
 
-In this paper, we explore the theoretical foundations of match types through the lens of a type system which extends \SystemFsub with pattern matching at the term and type level.
+In this chapter, we explore the theoretical foundations of match types through the lens of a type system which extends \SystemFsub with pattern matching at the term and type level.
 %
@@
 
-In summary, this paper makes the following contributions:
+In summary, this chapter makes the following contributions:
 %
@@
   \item We formalize match types in the self-contained calculus \SystemFm and prove it sound, providing a theoretical basis of our implementation (\Cref{sec:formalization}).
-The paper is accompanied by a mechanization of \SystemFm, including proofs of progress and preservation.
+The chapter is accompanied by a mechanization of \SystemFm, including proofs of progress and preservation.
 
@@
 
-Throughout this paper, we use the abbreviated syntax $\t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$ to denote an arbitrary number of cases, that is, $\∃ \n \∈ \ℕ.~ \t_s \match \{ \x_{1} \: \C_{1} \⇒ \t_{1}; \dots; \x_n \: \C_n \⇒ \t_n \} \otherwise \t_d$.
+Throughout this chapter, we use the abbreviated syntax $\t_s \match \{ \x_i \: \C_i \⇒ \t_i \} \otherwise \t_d$ to denote an arbitrary number of cases, that is, $\∃ \n \∈ \ℕ.~ \t_s \match \{ \x_{1} \: \C_{1} \⇒ \t_{1}; \dots; \x_n \: \C_n \⇒ \t_n \} \otherwise \t_d$.
 
@@
 This section provides an overview of the proof structure and states the involved lemmas and theorems.
-Detailed proofs are available in the supplementary material of this paper in two different versions.
-The first version is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
-The second version is a mechanization of the proof in Coq,
-using the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings.
-Our mechanization uses a simplified representation of match types with exactly one case per match.
+Detailed pen-and-paper proofs is available in the appendix of this thesis.
+A Coq mechanization of our proofs is available online\footnote{\url{https://zenodo.org/record/5568850}}.
+Our mechanization uses the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings, and a simplified representation of match types with exactly one case per match.
 Matches with multiple cases can be expressed by nesting match types in default cases.
@@
 \section{Conclusion}
-In this paper, we introduced \emph{match types}, a lightweight mechanism for type-level programming that integrates seamlessly in subtyping-based programming languages.
+In this chapter, we introduced \emph{match types}, a lightweight mechanism for type-level programming that integrates seamlessly in subtyping-based programming languages.
 We formalized match types in \SystemFm, a calculus based on \SystemFsub, and proved it sound.

### 2021-10-22 scala/generate-code-sections.py

@@ -9,7 +9,7 @@ if (len(input_files) == 0):
     print("Usages: ./generate-code-sections.py <...files>")
     sys.exit(1)
 
-output_file = "code-sections.tex"
+output_file = "scala/code-sections.tex"
 listing_options = "breaklines"
 
 start_section = re.compile(r'\s*(?://|#)\s+start\s+section\s+(\w+)\s*')

### 2021-10-26 scala/generate-code-sections.py
Shaping up the introduction

@@ -46,7 +46,7 @@ for (section, positions) in dict_zip(starts, finishes).items():
             print(f"Start and finish for section '{section}' are in different files ({start_file}, {finish_file}")
             sys.exit(1)
         extension = os.path.splitext(start_file)[1][1:]
-        output.append(f"\\newcommand{{\\{section}codesection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
+        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
 
 with open(output_file, "w") as f:
     f.write("\n".join(output))

### 2021-10-27 article.tex
Shaping up the introduction
s/cite/citep/g

@@
 Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
-Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\cite{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.
+Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.
 
@@
 
-In this section, we formally present \SystemFm, an extension of \SystemFsub~\cite{cardelli1994extension} with pattern matching, opaque classes, and match types.
+In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994extension} with pattern matching, opaque classes, and match types.
 \Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
@@
 Each case consists of a \emph{type test} and a corresponding expression/type.
-At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \cite{abadi1991dynamic}).
+At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \citep{abadi1991dynamic}).
 At the type level, a type test corresponds to a subtyping test with a particular type.
@@
 \Cref{fig:structure} gives an overview of the proof structure by showing implications between the various lemmas and theorems.
-The basic structure resembles that of \SystemFsub's standard safety proof from~\cite{pierce2002types}.
+The basic structure resembles that of \SystemFsub's standard safety proof from~\citep{pierce2002types}.
 We continue our presentation by introducing the lemmas and theorems used in our type safety proof.
@@
 
-In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \cite{nieto2020scala}.
+In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \citep{nieto2020scala}.
 Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.
@@
 The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
-Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \cite{fengyun2016generic}.
+Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \citep{fengyun2016generic}.
 Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
@@
 Scala's primary platform is the Java virtual machine (JVM).
-On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \cite{schinz2005compiling}.
+On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \citep{schinz2005compiling}.
 For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.
@@
 Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
-This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \cite{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.
+This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.
 
@@
 \SystemFm's type rules enable any match expression to be typed as a match type, but the situation is different in the full Scala language.
-Pattern matching in Scala supports many sorts of patterns~\cite{emir2007matching}, most of which do not have a match type counterpart.
+Pattern matching in Scala supports many sorts of patterns~\citep{emir2007matching}, most of which do not have a match type counterpart.
 Furthermore, typing match expressions as match types is not enabled by default in order to preserve backward compatibility.
@@
 In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
-To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\cite{harris2020numpy}.
+To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020numpy}.
 The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
-Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\cite{chen2017typesafe, huang2017hasktorch}.
+Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
 Our library uses match types to provide a shape-safe NumPy-like interface.
@@
 % \subsection{Related Work}
-% Shape and indexing errors in ndarrays is a widely acknowledged problem~\cite{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
-% The Dex language~\cite{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
-% Nexus~\cite{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
+% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
+% The Dex language~\citep{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
+% Nexus~\citep{chen2017typesafe} represents variable-rank ndarrays as HLists of dimension labels, and uses Scala's implicit resolution to express type-level transformations on shapes.
 % However, it only guards against rank errors, and not ndarray size errors.
@@
 They allow us to program shape transformations and type constraints, in order to  achieve shape-safety for variable ranks.
-% The implementation in this case study is inspired by the HaskTorch Haskell library~\cite{huang2017hasktorch}, which also represents shapes using HLists, and makes direct use of type-level computation to compute shape transformations.
+% The implementation in this case study is inspired by the HaskTorch Haskell library~\citep{huang2017hasktorch}, which also represents shapes using HLists, and makes direct use of type-level computation to compute shape transformations.
 
@@
 
-Scala supports singleton types, which are types inhabited by a single value~\cite{leontiev2014sip}.
+Scala supports singleton types, which are types inhabited by a single value~\citep{leontiev2014sip}.
 For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
@@
 For instance, a three-by-four matrix is a two-dimensional ndarray of shape $\(\3, \4)$.
-We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\cite{kiselyov2004strongly}.
+We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
 We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
@@
 %
-\shapecodesection
+\shapeCodeSection
 %
@@
 %
-\ndarraycodesection
+\ndarrayCodeSection
 
@@
 %
-\randomnormalcodesection
+\randomnormalCodeSection
 
@@
 %
-\multiplycodesection
+\multiplyCodeSection
 %
@@
 %
-\numelementscodesection
+\numelementsCodeSection
 
@@
 %
-\reshapecodesection
+\reshapeCodeSection
 %
@@
 
-\npmeancodesection
+\npmeanCodeSection
 
@@
 %
-\reducecodesection
+\reduceCodeSection
 
@@
 %
-\badreducescalacodesection
+\badreducescalaCodeSection
 %
@@
 
-% \goodreducescalacodesection
+% \goodreducescalaCodeSection
 
@@
 Dependently typed calculi typically use the same language to describe terms and types.
-This unification is also commonly used in the presence of subtyping~\cite{zwanenburg1999pure, hutchins2010pure, yang2017unifying}.
+This unification is also commonly used in the presence of subtyping~\citep{zwanenburg1999pure, hutchins2010pure, yang2017unifying}.
 For systems with a complete term/type symmetry, this is a natural design, as it is concise and simplifies the meta-theory.
@@
 
-Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\cite{aspinall1994singleton, stone2000deciding, courant2003strong}.
+Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994singleton, stone2000deciding, courant2003strong}.
 Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
@@
 
-Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\cite{essence2016amin}.
+Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{essence2016amin}.
 DOT does not directly support any form of type-level computation.
-We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\cite{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.
+We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.
 
@@
 
-In their work on intensional type analysis~\cite{harper1995compiling}, Harper and Morrisett introduce the $\λ^{ML}_i$ calculus that supports structural analysis of types.
+In their work on intensional type analysis~\citep{harper1995compiling}, Harper and Morrisett introduce the $\λ^{ML}_i$ calculus that supports structural analysis of types.
 In $\λ^{ML}_i$, types are represented as expressions that can be inspected by case analysis using a "typecase" construct, available both at the term and at the type level.
@@
 
-Haskell's type families allow programmers to define type-level functions using pattern matching~\cite{chakravarty2005associated, schrijvers2008open}.
+Haskell's type families allow programmers to define type-level functions using pattern matching~\citep{chakravarty2005associated, schrijvers2008open}.
 By default, type families are open, which means that a definition can spread across multiple files and compilations.
@@
 By default, Haskell checks for termination of recursive type families, but this check can be disabled to increase type families' expressiveness.
-Although the formalization presented in~\cite{eisenberg2014closed} does not cover non-terminating families, the paper discusses a soundness problem caused by non-termination.
+Although the formalization presented in~\citep{eisenberg2014closed} does not cover non-terminating families, the paper discusses a soundness problem caused by non-termination.
 The problem only occurs in the presence of repeated type bindings in patterns.
@@
 % Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
-% For instance, in a functional pearl on typed matrices~\cite{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.
+% For instance, in a functional pearl on typed matrices~\citep{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.
 
@@
 
-In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
+In their introductory example,~\citepauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
 Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:
@@
 
-The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\cite{microsoft2020typescript}.
+The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\citep{microsoft2020typescript}.
 Given a type \typescript{S extends T ?
@@
 
-Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\cite{bierman2014typescript}.
+Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014typescript}.
 Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

### 2021-10-27 scala/src/main/scala/safe-implicits.scala
Rename safe.scala

@@ -57,7 +57,7 @@ def f[Ps <: HList](ctx: Context["bool" :: Ps]): Context["bool" :: Ps] =
 // end section memImplicitFunction
 
 // start section memImplicitMain
-def main(ctx: Context[HNil]): Context[HNil] = {
+def main(ctx: Context[HNil]): Context[HNil] =
   ctx.malloc(32, "mem")
      .malloc(1, "bool")
      .call(f)
@@ -66,5 +66,4 @@ def main(ctx: Context[HNil]): Context[HNil] = {
      }
      .free("mem")
      .free("bool")
-}
 // end section memImplicitMain

### 2021-10-27 scala/src/main/scala/safe-matchtypes.scala

@@ -35,7 +35,7 @@ def f[Ps <: HList](ctx: Context["bool" :: Ps]): Context["bool" :: Ps] =
   }
 
 // start section memMatchtypeMain
-def main(ctx: Context[HNil]): Context[HNil] = {
+def main(ctx: Context[HNil]): Context[HNil] =
   ctx.malloc(32, "mem")
      .malloc(1, "bool")
      .call(f)
@@ -44,5 +44,4 @@ def main(ctx: Context[HNil]): Context[HNil] = {
      }
      .free("mem")
      .free("bool")
-}
 // end section memMatchtypeMain

### 2021-10-29 article.tex
Introduction

@@
 
-In their introductory example,~\citepauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
+In their introductory example,~\citeauthor{weirich2011generative} show how type family reduction can lead to unsoundness in the absence of role annotations.
 Their presentation also includes a hypothetical translation of their example to Standard ML, which translates directly to Scala as follows:

### 2021-10-29 scala/src/main/scala/safe-implicits.scala

@@ -6,55 +6,55 @@ sealed trait HNil extends HList
 case object HNil extends HNil
 case class ::[+H, +T <: HList](head: H, tail: T) extends HList
 
-// start section memImplicitHlistOps
+// start section memImplicitRemove
 trait Remove[V, Ps <: HList, O <: HList]
-object Remove {
-  implicit def casehead[V, Ps <: HList, O <: HList]: Remove[V, V :: Ps, Ps] = new Remove {}
+
+object Remove:
+  implicit def casehead[V, Ps <: HList, O <: HList]
+    : Remove[V, V :: Ps, Ps] = new Remove {}
+
   implicit def casetail[V, Ph, Pt <: HList, O <: HList]
-    (implicit ev: Remove[V, Pt, O]): Remove[V, Ph :: Pt, Ph :: O] = new Remove {}
-}
+    (implicit ev: Remove[V, Pt, O])
+    : Remove[V, Ph :: Pt, Ph :: O] = new Remove {}
+// end section memImplicitRemove
 
 trait NotIn[V, Ps <: HList]
-object NotIn {
+
+object NotIn:
   implicit def casenil[V]: NotIn[V, HNil] = new NotIn {}
   implicit def casecons[Ph, Pt <: HList, V]
     (implicit
       no: NotGiven[V =:= Ph],
       xs: NotIn[V, Pt]
     ): NotIn[V, Ph :: Pt] = new NotIn {}
-}
-// end section memImplicitHlistOps
 
 // start section memImplicitContext
-trait Context[Ps <: HList] {
-  def malloc[V <: Singleton]
-    (size: Int, v: V)
-    (implicit ev: NotIn[V, Ps])
-    : Context[V :: Ps] = ???
-
+trait Context[Ps <: HList]:
   def free[V <: Singleton, Out <: HList]
     (v: V)
     (implicit ev: Remove[V, Ps, Out])
-    : Context[Out] = ???
+    : Context[Out]
+// end section memImplicitContext
+
+  def malloc[V <: Singleton]
+    (size: Int, v: V)
+    (implicit ev: NotIn[V, Ps])
+    : Context[V :: Ps]
 
   def call[Out <: HList]
     (f: Context[Ps] => Context[Out])
-    : Context[Out] = ???
+    : Context[Out]
 
   def deref[V <: Singleton]
     (v: V)
     (body: Array[Byte] => Unit)
     (implicit ev: Remove[V, Ps, ?])
-    : Context[Ps] = ???
-}
[…]

### 2021-11-02 scala/src/main/scala/safe-implicits.scala
Introduction

@@ -7,15 +7,15 @@ case object HNil extends HNil
 case class ::[+H, +T <: HList](head: H, tail: T) extends HList
 
 // start section memImplicitRemove
-trait Remove[V, Ps <: HList, O <: HList]
+trait Remove[V, Ps <: HList, Out <: HList]
 
 object Remove:
-  implicit def casehead[V, Ps <: HList, O <: HList]
+  implicit def casehead[V, Ps <: HList, Out <: HList]
     : Remove[V, V :: Ps, Ps] = new Remove {}
 
-  implicit def casetail[V, Ph, Pt <: HList, O <: HList]
-    (implicit ev: Remove[V, Pt, O])
-    : Remove[V, Ph :: Pt, Ph :: O] = new Remove {}
+  implicit def casetail[V, Ph, Pt <: HList, Out <: HList]
+    (implicit ev: Remove[V, Pt, Out])
+    : Remove[V, Ph :: Pt, Ph :: Out] = new Remove {}
 // end section memImplicitRemove
 
 trait NotIn[V, Ps <: HList]

### 2021-11-02 scala/src/main/scala/safe-matchtypes.scala

@@ -6,23 +6,23 @@ sealed trait HNil extends HList
 case object HNil extends HNil
 case class ::[+H, +T <: HList](head: H, tail: T) extends HList
 
-// start section memMatchtypeHlistOps
 type NotIn[V, Ps] <: HList = Ps match {
   case V :: t => NotIn[V, Ps]
   case h :: t => h :: NotIn[V, t]
   case HNil => HNil
 }
 
+// start section memMatchtypeRemove
 type Remove[V, Ps] <: HList = Ps match {
   case V :: t => t
   case h :: t => h :: Remove[V, t]
 }
-// end section memMatchtypeHlistOps
+// end section memMatchtypeRemove
 
 // start section memMatchtypeContext
 trait Context[Ps <: HList] {
   def malloc[V <: Singleton](size: Int, v: V): Context[V :: NotIn[V, Ps]] = ???
-  def free[V <: Singleton, Out <: HList](v: V): Context[Remove[V, Ps]] = ???
+  def free[V <: Singleton](v: V): Context[Remove[V, Ps]] = ???
   def call[Out <: HList](f: Context[Ps] => Context[Out]): Context[Out] = ???
   def deref[V <: Singleton, EV <: Remove[V, Ps]](v: V)(body: Array[Byte] => Unit)
     : Context[Ps] = ???

### 2021-11-04 scala/src/main/scala/implicit-tutorial.scala
(Ab)Using Implicits

@@ -1,0 +1,29 @@
+object explicit {
+// start section explicitModulo
+val modulo: Int = 3
+def addm(x: Int, y: Int)(m: Int) = (x + y) % m
+addm(4, 5)(modulo)
+// end section explicitModulo
+}
+
+object impl {
+// start section implicitModulo
+implicit val modulo: Int = 3
+def addm(x: Int, y: Int)(implicit m: Int) = (x + y) % m
+addm(4, 5)
+// end section implicitModulo
+
+// start section ordExample
+trait Ord[T]:
+  def compare(a: T, b: T): Boolean
+
+def comp[T](x: T, y: T)(implicit ev: Ord[T]): Boolean =
+  ev.compare(x, y)
+
+implicit def intOrd: Ord[Int] = new Ord[Int]:
+  def compare(a: Int, b: Int): Boolean = a < b
+
+comp(1, 2)
+// end section ordExample
+
+}

### 2021-11-04 scala/src/main/scala/safe-implicits.scala

@@ -2,15 +2,15 @@ package foo
 import scala.util.NotGiven
 
 sealed trait HList
-sealed trait HNil extends HList
-case object HNil extends HNil
+case object HNil extends HList
 case class ::[+H, +T <: HList](head: H, tail: T) extends HList
+type HNil = HNil.type
 
 // start section memImplicitRemove
 trait Remove[V, Ps <: HList, Out <: HList]
 
 object Remove:
-  implicit def casehead[V, Ps <: HList, Out <: HList]
+  implicit def casehead[V, Ps <: HList]
     : Remove[V, V :: Ps, Ps] = new Remove {}
 
   implicit def casetail[V, Ph, Pt <: HList, Out <: HList]

### 2021-11-08 article.tex
Backport TLPwMT revisions

@@
 
-\clearpage
 \section{Formalization}
@@
 The class disjointness is symmetric relation over $\C$.
+
+The inheritance and disjointness parameters can be understood as a representation of a hierarchy of Scala traits and classes.
+For example, \dotty{trait C1; class C2 extends C1} is represented  in \Fm as $\Ψ \= \{ \(\C_1, \C_2) \};~ \Ξ \= \{ \}$.
+This representation also models the fact that certain types cannot possibly have common instances.
+For example, \dotty{class C3; class C4} is represented as $\Ψ \= \{ \};~ \Ξ \= \{ \(\C_3, \C_4), \(\C_4, \C_3) \}$, since Scala disallows multiple class inheritance.
 Inheritance and disjointness must be consistent in the sense that $\(\A, \B) \∈ \Ξ$ implies that there is no class $\C$ such that $\(\C, \A) \∈ \Ψ$ and $\(\C, \B) \∈ \Ψ$.
@@
 In \SystemFm, the match type obtained when typing $\(\f~\B~\(\new \E))$ does not reduce since the scrutinee type $\B$ is neither disjoint with, nor a subtype of the first pattern type test $\A$.
-In this case, the match type is \emph{stuck}, and is assigned as is.
-Stuck types can appear as the result of programming error, but can also be due to the local irreducibility of a match type.
-For instance, the body of~$\f$ is typed with a stuck type, as shown in \Cref{ex:simplereduction}, but that type can later become reducible depending on type variable instantiations.
+In this case, unreduced match type is assigned "as is".
+Unreduced types can appear as the result of programming error, but can also be due to the local irreducibility of a match type.
+For instance, the body of~$\f$ is typed with an unreduced type, as shown in \Cref{ex:simplereduction}, but that type can later become reducible depending on type variable instantiations.
 
@@
 This section provides an overview of the proof structure and states the involved lemmas and theorems.
-Detailed pen-and-paper proofs is available in the appendix of this thesis.
-A Coq mechanization of our proofs is available online\footnote{\url{https://zenodo.org/record/5568850}}.
-Our mechanization uses the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings, and a simplified representation of match types with exactly one case per match.
+Detailed proofs are available in the supplementary material of this paper, in two different versions.
+The first version, \cite{blanvillain2021type}, is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
+The second version, \cite{blanvillain2021artifact}, is a mechanization of the proof in Coq,
+using the locally nameless representation by \citet{aydemir2008engineering} to model variable bindings.
+Our mechanization uses a simplified representation of match types with exactly one case per match.
 Matches with multiple cases can be expressed by nesting match types in default cases.
@@
 \structure{Structure of the type safety proof. Arrows represent implications between lemmas and theorems.}
-
-% There is nothing special about this, omitted since it is standard.
-% An important technical detail that we omitted in \Cref{fig:FMTypingRules} concerns the well-formedness of types on contexts.
-% A type $\T$ is well-formed in $\Γ$ when all the free variables in $\T$ are bound in $\Γ$.
-% A context $\Γ$ is well-formed if each name is bound at most once and if each binding is to a well-formed type in the context preceding its binding.
-% To improve presentation, we have omitted all well-formedness premises from our typing rules.
 
@@
 We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
-Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P$.
+Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$.
+Elements of $\P$ can be understood as equivalence classes for \Fm's types.
 We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
@@
 %
-With these lemmas in place, the proofs of progress and preservation are straightforward.
+With these lemmas in hand, the proofs of progress and preservation are straightforward.
 %
@@
 A parametric class, written $\B~\T$, takes exactly one type parameter\footnotemark.
+We redefine $\C$ to be a syntactic object defined as $\C \::= \A \| \B~\T$.
+The class inheritance $\Ψ$ and class disjointness $\Ξ$ remain as binary relations on $\C$.
+A generic instantiation in the class hierarchy is represented as an element of $\Ψ$, for example, \dotty{A1 extends B2[A3]} is represented as $\(\A_1, \B_2~\A_3) \∈ \Ψ$.
+Generic inheritance is represented using multiple entries in $\Ψ$, for example, \dotty{B1[T] extends B2[T]} is represented as $\∀ \T. \(\B_1~\T, \B_2~\T) \∈ \Ψ$.
 This approach allows us to reuse most of \Fm's definitions.
@@
 
+\smallskip
 \noindent
@@
 Our library uses match types to provide a shape-safe NumPy-like interface.
+
+Scala programmers have a long history of using ad-hoc solutions for type-level programming~\citep{sabin2011shapeless, pilquist2013scodecs, blanvillain2016frameless}.
+These solutions have several downsides, such as being slow to compile and cumbersome to use.
+Match types aim at simplifying type-level programming by providing first-class language support.
 We believe that the approach presented in this case study is an improvement over the status quo because it does not use metaprogramming or any sort of convoluted encoding to express type-level operations.

### 2021-11-08 scala/generate-code-sections.py

@@ -46,7 +46,7 @@ for (section, positions) in dict_zip(starts, finishes).items():
             print(f"Start and finish for section '{section}' are in different files ({start_file}, {finish_file}")
             sys.exit(1)
         extension = os.path.splitext(start_file)[1][1:]
-        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[float, floatplacement=H, style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
+        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
 
 with open(output_file, "w") as f:
     f.write("\n".join(output))

### 2021-11-09 scala/src/main/scala/implicit-tutorial.scala
(Ab)Using Implicits

@@ -14,16 +14,28 @@ addm(4, 5)
 // end section implicitModulo
 
 // start section ordExample
-trait Ord[T]:
+trait Ordering[T]:
   def compare(a: T, b: T): Boolean
 
-def comp[T](x: T, y: T)(implicit ev: Ord[T]): Boolean =
+def comp[T](x: T, y: T)(implicit ev: Ordering[T]): Boolean =
   ev.compare(x, y)
 
-implicit def intOrd: Ord[Int] = new Ord[Int]:
-  def compare(a: Int, b: Int): Boolean = a < b
+implicit def intOrdering: Ordering[Int] =
+  new Ordering[Int]:
+    def compare(a: Int, b: Int): Boolean = a < b
 
 comp(1, 2)
 // end section ordExample
+
+
+// start section ordListExample
+implicit def listOrdering[T](implicit ev: Ordering[T]): Ordering[List[T]] =
+  new Ordering[List[T]]:
+    def compare(a: List[T], b: List[T]): Boolean =
+      (a, b) match
+        case (a :: as, b :: bs) => ev.compare(a, b) && compare(as, bs)
+        case (_, Nil) => false
+        case (Nil, _) => true
+// end section ordListExample
 
 }

### 2021-11-09 scala/src/main/scala/safe-implicits.scala

@@ -1,10 +1,17 @@
 package foo
 import scala.util.NotGiven
 
-sealed trait HList
-case object HNil extends HList
-case class ::[+H, +T <: HList](head: H, tail: T) extends HList
-type HNil = HNil.type
+// start section hlistEnumDefinition
+enum HList:
+  case HNil()
+  case ::[H, T <: HList](head: H, tail: T)
+// end section hlistEnumDefinition
+import HList._
+
+// sealed trait HList
+// case object HNil extends HList
+// case class ::[+H, +T <: HList](head: H, tail: T) extends HList
+// type HNil = HNil.type
 
 // start section memImplicitRemove
 trait Remove[V, Ps <: HList, Out <: HList]

### 2021-11-10 proofs/disjointness-subtyping-exclusivity.tex
(Ab)Using Implicits
xelatex

@@
 \begin{proof}
-  We prove mutual exclusivity of type disjointness and subtyping by first defining $\⟦\· \⟧_Γ$, a mapping from \SystemFm types (in a given context $\Γ$) into non-empty subsets of a newly defined set $\P$.
+  We prove mutual exclusivity of type disjointness and subtyping by first defining $\⟦\· \⟧_{\Γ}$, a mapping from \SystemFm types (in a given context $\Γ$) into non-empty subsets of a newly defined set $\P$.
 We then show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness relation in $\P$.
@@
   Indeed, suppose there exist two types $\S$ and $\T$ with $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \disj(\S, \T)$.
-  $\⟦ \S \⟧_Γ$ and $\⟦ \T \⟧_Γ$ are two non-empty sets which are both intersecting and disjoint, a contradiction.
+  $\⟦ \S \⟧_{\Γ}$ and $\⟦ \T \⟧_{\Γ}$ are two non-empty sets which are both intersecting and disjoint, a contradiction.
 
@@
 Elements of $\P$ can be understood as equivalence classes for \SystemFm values: $\Λ$ corresponds to all abstraction values, $\V$ corresponds to type abstraction value, and elements of $\C$ correspond to their respective constructors.
-  The definition of $\⟦\· \⟧_Γ$ makes this correspondence apparent.
+  The definition of $\⟦\· \⟧_{\Γ}$ makes this correspondence apparent.
 
-  We define $\⟦\· \⟧_Γ$, a mapping of \SystemFm types into subsets of $\P$ in a given context $\Γ$:
+  We define $\⟦\· \⟧_{\Γ}$, a mapping of \SystemFm types into subsets of $\P$ in a given context $\Γ$:
 
   \begin{align*}
-    \text{$\⟦ \Top \⟧_Γ$} &=
+    \text{$\⟦ \Top \⟧_{\Γ}$} &=
     \text{$\P$}\\
     %
-    \text{$\⟦ \X \⟧_Γ$} &=
-    \text{$\⟦ \T \⟧_Γ \quad$ where $\X \<: \T \∈ \Γ$}\\
+    \text{$\⟦ \X \⟧_{\Γ}$} &=
+    \text{$\⟦ \T \⟧_{\Γ} \quad$ where $\X \<: \T \∈ \Γ$}\\
     %
-    \text{$\⟦ \T_{1} \→ \T_{2} \⟧_Γ$} &=
+    \text{$\⟦ \T_{1} \→ \T_{2} \⟧_{\Γ}$} &=
     \text{$\{ \Λ \}$}\\
     %
-    \text{$\⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_Γ$} &=
+    \text{$\⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_{\Γ}$} &=
     \text{$\{ \V \}$}\\
     %
-    \text{$\⟦ \{ \new \C_{1} \} \⟧_Γ$} &=
+    \text{$\⟦ \{ \new \C_{1} \} \⟧_{\Γ}$} &=
     \text{$\{ \C_{1} \}$}\\
     %
-    \text{$\⟦ \C_{1} \⟧_Γ$} &=
-      \{ \c \∈ \C~|~(\c, \C_1) \∈ \Ψ \}\\
+    \text{$\⟦ \C_{1} \⟧_{\Γ}$} &=
+      \{ \c \∈ \C~\|~(\c, \C_1) \∈ \Ψ \}\\
     \⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_\Γ &=
       \begin{cases}
-        \text{$\⟦ \T_n \⟧_Γ$\phantom{\P} if $\⟦ \T_s \⟧_Γ \⊂ \⟦ \S_n \⟧_Γ$}\\
-        \phantom{\text{$\⟦ \T_n \⟧_Γ$}}\phantom{\P}\text{ and $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$}\\
-        \text{$\⟦ \T_d \⟧_Γ$\phantom{\P} if $\∀ \m.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$}\\
-        \P\phantom{\text{$\⟦ \T_n \⟧_Γ$}}\text{ otherwise}
+        \text{$\⟦ \T_n \⟧_{\Γ}$\phantom{\P} if $\⟦ \T_s \⟧_{\Γ} \⊂ \⟦ \S_n \⟧_{\Γ}$}\\
+        \phantom{\text{$\⟦ \T_n \⟧_{\Γ}$}}\phantom{\P}\text{ and $\∀ \m<n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \S_m \⟧_{\Γ} \= \{ \}$}\\
+        \text{$\⟦ \T_d \⟧_{\Γ}$\phantom{\P} if $\∀ \m.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \S_m \⟧_{\Γ} \= \{ \}$}\\
+        \P\phantom{\text{$\⟦ \T_n \⟧_{\Γ}$}}\text{ otherwise}
       \end{cases}
@@
   \begin{enumerate}
-    \item $\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$,
-    \item $\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$,
+    \item $\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \T \⟧_{\Γ}$,
+    \item $\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_{\Γ} \∩ \⟦ \T \⟧_{\Γ} \= \{ \}$,
   \end{enumerate}
@@
   \begin{enumerate}
-    \item ($\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$)
+    \item ($\Γ \⊢ \S \<: \T$ implies $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \T \⟧_{\Γ}$)
     \begin{itemize}
@@
 
-      $\⟦ \S \⟧_Γ \= \⟦ \T \⟧_Γ$ and the result is immediate.
+      $\⟦ \S \⟧_{\Γ} \= \⟦ \T \⟧_{\Γ}$ and the result is immediate.
 
@@
 
-      By the IH, $\⟦ \S \⟧_Γ \⊂ \⟦ \U \⟧_Γ$ and $\⟦ \U \⟧_Γ \⊂ \⟦ \T \⟧_Γ$. From subset transitivity, $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$, as required.
+      By the IH, $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \U \⟧_{\Γ}$ and $\⟦ \U \⟧_{\Γ} \⊂ \⟦ \T \⟧_{\Γ}$. From subset transitivity, $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \T \⟧_{\Γ}$, as required.
 
@@
 
-      $\⟦ \Top \⟧_Γ \= \P$ coincides with the codomain of $\⟦\· \⟧_Γ$.
-      Therefore, for any type $\T$, $\⟦ \T \⟧_Γ \⊂ \⟦ \Top \⟧_Γ$, as required.
+      $\⟦ \Top \⟧_{\Γ} \= \P$ coincides with the codomain of $\⟦\· \⟧_{\Γ}$.
+      Therefore, for any type $\T$, $\⟦ \T \⟧_{\Γ} \⊂ \⟦ \Top \⟧_{\Γ}$, as required.
 
@@
 
-      $\⟦ \{ \new \C_{1} \} \⟧_Γ \= \{ \C_{1} \}$ and
-      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$.
+      $\⟦ \{ \new \C_{1} \} \⟧_{\Γ} \= \{ \C_{1} \}$ and
+      $\⟦ \C_{1} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$.
       The result follows by reflexivity of $\Ψ$.
@@
 
-      By definition $\⟦ \X \⟧_Γ \= \⟦ \T \⟧_Γ$, and the result is immediate.
+      By definition $\⟦ \X \⟧_{\Γ} \= \⟦ \T \⟧_{\Γ}$, and the result is immediate.
 
@@
 
-      $\⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \⟦ \T_{1} \→ \T_{2} \⟧_Γ \= \{ \Λ \}$ and the result is immediate.
+      $\⟦ \S_{1} \→ \S_{2} \⟧_{\Γ} \= \⟦ \T_{1} \→ \T_{2} \⟧_{\Γ} \= \{ \Λ \}$ and the result is immediate.
 
@@
 
-      $\⟦ \∀ \X \<: \U_{1}.~ \S_{2} \⟧_Γ \= \⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_Γ \= \{ \V \}$ and the result is immediate.
+      $\⟦ \∀ \X \<: \U_{1}.~ \S_{2} \⟧_{\Γ} \= \⟦ \∀ \X \<: \U_{1}.~ \T_{2} \⟧_{\Γ} \= \{ \V \}$ and the result is immediate.
 
@@
 
-      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
-      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
-      By transitivity of $\Ψ$, $\⟦ \C_{1} \⟧_Γ \⊂ \⟦ \C_{2} \⟧_Γ$, as required.
+      $\⟦ \C_{1} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
+      $\⟦ \C_{2} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
+      By transitivity of $\Ψ$, $\⟦ \C_{1} \⟧_{\Γ} \⊂ \⟦ \C_{2} \⟧_{\Γ}$, as required.
 
@@
       \hfuzz=1.5pt
-      Using the IH we get $\⟦ \T_s \⟧_Γ \⊂ \⟦ \S_n \⟧_Γ$, $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_m \⟧_Γ \= \{ \}$ and $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_Γ \= \⟦ \T_n \⟧_Γ$.
-      This last equality implies both $\⟦ \T_{1} \⟧_Γ \⊂ \⟦ \T_{2} \⟧_Γ$ and $\⟦ \T_{2} \⟧_Γ \⊂ \⟦ \T_{1} \⟧_Γ$, as required.
+      Using the IH we get $\⟦ \T_s \⟧_{\Γ} \⊂ \⟦ \S_n \⟧_{\Γ}$, $\∀ \m<n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \S_m \⟧_{\Γ} \= \{ \}$ and $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_{\Γ} \= \⟦ \T_n \⟧_{\Γ}$.
+      This last equality implies both $\⟦ \T_{1} \⟧_{\Γ} \⊂ \⟦ \T_{2} \⟧_{\Γ}$ and $\⟦ \T_{2} \⟧_{\Γ} \⊂ \⟦ \T_{1} \⟧_{\Γ}$, as required.
 
@@
 
-      By the IH $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \S_n \⟧_Γ \= \{ \}$.
-      By definition, $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_Γ \= \⟦ \T_d \⟧_Γ$.
-      This equality implies both $\⟦ \T_{1} \⟧_Γ \⊂ \⟦ \T_{2} \⟧_Γ$ and $\⟦ \T_{2} \⟧_Γ \⊂ \⟦ \T_{1} \⟧_Γ$, the desired result for \SMatch3 and \SMatch4 respectively.
+      By the IH $\∀ \n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \S_n \⟧_{\Γ} \= \{ \}$.
+      By definition, $\⟦ \T_s \match \{ \S_i \⇒ \T_i \} \otherwise \T_d \⟧_{\Γ} \= \⟦ \T_d \⟧_{\Γ}$.
+      This equality implies both $\⟦ \T_{1} \⟧_{\Γ} \⊂ \⟦ \T_{2} \⟧_{\Γ}$ and $\⟦ \T_{2} \⟧_{\Γ} \⊂ \⟦ \T_{1} \⟧_{\Γ}$, the desired result for \SMatch3 and \SMatch4 respectively.
 
@@
 
-      By case analysis on $\⟦ \T \⟧_Γ$:
+      By case analysis on $\⟦ \T \⟧_{\Γ}$:
       \begin{enumerate}
-        \item $\⟦ \T \⟧_Γ \= \⟦ \T_n \⟧_Γ$
-        and $\⟦ \T_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$
-        and $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$:
+        \item $\⟦ \T \⟧_{\Γ} \= \⟦ \T_n \⟧_{\Γ}$
+        and $\⟦ \T_s \⟧_{\Γ} \⊂ \⟦ \U_n \⟧_{\Γ}$
+        and $\∀ \m<n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \U_m \⟧_{\Γ} \= \{ \}$:
 
-        The IH gives $\⟦ \S_s \⟧_Γ \⊂ \⟦ \T_s \⟧_Γ$.
-        Using $\∀ \m<n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$, we get $\∀ \m<n.~ \⟦ \S_s \⟧_Γ \∩ \⟦ \U_m \⟧_Γ \= \{ \}$.
-        Using $\⟦ \T_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$, we get $\⟦ \S_s \⟧_Γ \⊂ \⟦ \U_n \⟧_Γ$ (by subset transitivity).
-        Therefore, $\⟦ \S \⟧_Γ \= \⟦ \S_n \⟧_Γ$, and the result follows from the IH.
+        The IH gives $\⟦ \S_s \⟧_{\Γ} \⊂ \⟦ \T_s \⟧_{\Γ}$.
+        Using $\∀ \m<n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \U_m \⟧_{\Γ} \= \{ \}$, we get $\∀ \m<n.~ \⟦ \S_s \⟧_{\Γ} \∩ \⟦ \U_m \⟧_{\Γ} \= \{ \}$.
+        Using $\⟦ \T_s \⟧_{\Γ} \⊂ \⟦ \U_n \⟧_{\Γ}$, we get $\⟦ \S_s \⟧_{\Γ} \⊂ \⟦ \U_n \⟧_{\Γ}$ (by subset transitivity).
+        Therefore, $\⟦ \S \⟧_{\Γ} \= \⟦ \S_n \⟧_{\Γ}$, and the result follows from the IH.
 
-        \item $\⟦ \T \⟧_Γ \= \⟦ \T_d \⟧_Γ$ and
-        $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$:
+        \item $\⟦ \T \⟧_{\Γ} \= \⟦ \T_d \⟧_{\Γ}$ and
+        $\∀ \n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \U_n \⟧_{\Γ} \= \{ \}$:
 
-        The IH gives $\⟦ \S_s \⟧_Γ \⊂ \⟦ \T_s \⟧_Γ$.
-        Using $\∀ \n.~ \⟦ \T_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$ we get $\∀ \n.~ \⟦ \S_s \⟧_Γ \∩ \⟦ \U_n \⟧_Γ \= \{ \}$.
-        Therefore, $\⟦ \S \⟧_Γ \= \⟦ \S_d \⟧_Γ$, and the result follows from the IH.
+        The IH gives $\⟦ \S_s \⟧_{\Γ} \⊂ \⟦ \T_s \⟧_{\Γ}$.
+        Using $\∀ \n.~ \⟦ \T_s \⟧_{\Γ} \∩ \⟦ \U_n \⟧_{\Γ} \= \{ \}$ we get $\∀ \n.~ \⟦ \S_s \⟧_{\Γ} \∩ \⟦ \U_n \⟧_{\Γ} \= \{ \}$.
+        Therefore, $\⟦ \S \⟧_{\Γ} \= \⟦ \S_d \⟧_{\Γ}$, and the result follows from the IH.
 
-        \item $\⟦ \T \⟧_Γ \= \P$:
+        \item $\⟦ \T \⟧_{\Γ} \= \P$:
 
-        $\P$ is the codomain of $\⟦\· \⟧_Γ$, therefore $\⟦ \S \⟧_Γ \⊂ \⟦ \T \⟧_Γ$, as required.
+        $\P$ is the codomain of $\⟦\· \⟧_{\Γ}$, therefore $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \T \⟧_{\Γ}$, as required.
       \end{enumerate}
@@
 
-    \item ($\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$)
+    \item ($\Γ \⊢ \disj(\S, \T)$ implies $\⟦ \S \⟧_{\Γ} \∩ \⟦ \T \⟧_{\Γ} \= \{ \}$)
     \begin{itemize}
@@
 
-      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
-      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
-      By definition of Ξ, there is no class $\c$ such that both $\(\c, \C_{1}) \∈ \Ψ$ and $\(\c, \C_{2}) \∈ \Ψ$, and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \C_{2} \⟧_Γ \= \{ \}$.
+      $\⟦ \C_{1} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$ and
+      $\⟦ \C_{2} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
+      By definition of $\Ξ$, there is no class $\c$ such that both $\(\c, \C_{1}) \∈ \Ψ$ and $\(\c, \C_{2}) \∈ \Ψ$, and $\⟦ \C_{1} \⟧_{\Γ} \∩ \⟦ \C_{2} \⟧_{\Γ} \= \{ \}$.
 
@@
 
-      $\⟦ \{ \new \C_{1} \} \⟧_Γ \= \C_{1}$ and
-      $\⟦ \C_{2} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
-      Since $\(\C_{1}, \C_{2}) \∉ \Ψ$, $\C_{1} \∉ \⟦ \C_{2} \⟧_Γ$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \C_{2} \⟧_Γ \= \{ \}$.
+      $\⟦ \{ \new \C_{1} \} \⟧_{\Γ} \= \C_{1}$ and
+      $\⟦ \C_{2} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{2}) \∈ \Ψ \}$.
+      Since $\(\C_{1}, \C_{2}) \∉ \Ψ$, $\C_{1} \∉ \⟦ \C_{2} \⟧_{\Γ}$ and $\⟦ \C_{1} \⟧_{\Γ} \∩ \⟦ \C_{2} \⟧_{\Γ} \= \{ \}$.
 
@@
 
-      By the IH, $\⟦ \S \⟧_Γ \⊂ \⟦ \U \⟧_Γ$ and $\⟦ \U \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$.
-      $\⟦ \S \⟧_Γ \∩ \⟦ \T \⟧_Γ \= \{ \}$ follows directly.
+      By the IH, $\⟦ \S \⟧_{\Γ} \⊂ \⟦ \U \⟧_{\Γ}$ and $\⟦ \U \⟧_{\Γ} \∩ \⟦ \T \⟧_{\Γ} \= \{ \}$.
+      $\⟦ \S \⟧_{\Γ} \∩ \⟦ \T \⟧_{\Γ} \= \{ \}$ follows directly.
 
@@
 
-      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \{ \Λ \}$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \S_{1} \→ \S_{2} \⟧_Γ \= \{ \}$, as required.
+      $\⟦ \C_{1} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \S_{1} \→ \S_{2} \⟧_{\Γ} \= \{ \Λ \}$ and $\⟦ \C_{1} \⟧_{\Γ} \∩ \⟦ \S_{1} \→ \S_{2} \⟧_{\Γ} \= \{ \}$, as required.
 
@@
 
-      $\⟦ \C_{1} \⟧_Γ \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_Γ \= \{ \S \}$ and $\⟦ \C_{1} \⟧_Γ \∩ \⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_Γ \= \{ \}$, as required.
+      $\⟦ \C_{1} \⟧_{\Γ} \= \{ \c \∈ \C~\|~\(\c, \C_{1}) \∈ \Ψ \}$, $\⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_{\Γ} \= \{ \S \}$ and $\⟦ \C_{1} \⟧_{\Γ} \∩ \⟦ \∀ \X \<: \S_{1}.~ \S_{2} \⟧_{\Γ} \= \{ \}$, as required.
     \end{itemize}

### 2021-11-11 scala/generate-code-sections.py
Avoid page breaks in listings

@@ -46,7 +46,7 @@ for (section, positions) in dict_zip(starts, finishes).items():
             print(f"Start and finish for section '{section}' are in different files ({start_file}, {finish_file}")
             sys.exit(1)
         extension = os.path.splitext(start_file)[1][1:]
-        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
+        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[float, floatplacement=H, style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
 
 with open(output_file, "w") as f:
     f.write("\n".join(output))

### 2021-11-11 article.tex
xelatex
Fix typo

--- 2021-10-29 article.tex
@@
 This section provides an overview of the proof structure and states the involved lemmas and theorems.
-Detailed pen-and-paper proofs is available in the appendix of this thesis.
+Detailed pen-and-paper proofs are available in the appendix of this thesis.
 A Coq mechanization of our proofs is available online\footnote{\url{https://zenodo.org/record/5568850}}.

### 2021-11-13 article.tex
Fix bad boxes

--- 2021-11-08 article.tex
@@
 
-This parametric approach allows us to model class inheritance as found in object-oriented languages without the need for dedicated syntax for classes and data type definitions.
+This parametric approach allows us to model class inheritance as it is found in object-oriented languages, without the need for dedicated syntax for classes and data type definitions.
 Although our approach might appear simplistic, it can easily model advanced object-oriented features such as multiple inheritance.
@@
 \subsubsection*{Match Type Reduction}
-The subtyping relation contains five rules for match type reduction, \SMatch1/2/3/4/5.
+The subtyping relation contains 5 rules for match type reduction, \SMatch1/2/3/4/5.
 These rules are defined in pairs using the $\≡$ shorthand notation, where $\S \≡ \T$ means that $\S$ and $\T$ are in a mutual subtyping relation.
@@
 \subsubsection*{Inversion of Subtyping}
-\harpoon{Definition of the auxiliary relation $\⇌$, used to state inversion of subtyping (\Cref{lem:inversion-of-subtyping}).}
+\harpoon{Definition of the auxiliary relation $\⇌$, used to state inversion of subtyping.}
 
@@
 
-In the new subtyping rule for non-default match reduction, \BSMatch1/2, the first premise instantiates the pattern variable $\X$ to some type $\U$ such that the scrutinee type is a subtype of the $\n$th pattern:
+In the new subtyping rule for non-default match reduction, called \BSMatch1/2, the first premise instantiates the pattern variable $\X$ to some type $\U$ such that the scrutinee type is a subtype of the $\n$th pattern:
 \BSMatchInline
@@
 \hspace{200pt}\\\indent
-  \Ψ &= \{ (\String, \List~\Char) \} ∪ \Id
+  \Ψ &= \{ (\String, \List~\Char) \} \∪ \Id
 & \Ξ &= \{ \}

### 2021-11-15 scala/generate-code-sections.py
Remove CodeSection suffixes

@@ -46,7 +46,7 @@ for (section, positions) in dict_zip(starts, finishes).items():
             print(f"Start and finish for section '{section}' are in different files ({start_file}, {finish_file}")
             sys.exit(1)
         extension = os.path.splitext(start_file)[1][1:]
-        output.append(f"\\newcommand{{\\{section}CodeSection}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
+        output.append(f"\\newcommand{{\\{section}}}{{\lstinputlisting[style={extension}, firstline={start_line}, lastline={finish_line}, {listing_options}]{{{start_file}}}}}")
 
 with open(output_file, "w") as f:
     f.write("\n".join(output))

### 2021-11-15 scala/src/main/scala/numpy.scala

@@ -4,49 +4,49 @@ case object Ø extends Shape
 
 // cheating a little bit here, but there's a footnote to explain it!
 object SimplifiedShape {
-// start section shape
+// start section shapeEnum
 enum Shape {
   case #:[H <: Int, T <: Shape](head: H, tail: T)
   case Ø
 }
-// end section shape
+// end section shapeEnum
 }
 
 type Ø = Ø.type
 type None = None.type
 
-// start section datatype
+// start section datatypeTrait
 trait DataType[T]
-// end section datatype
+// end section datatypeTrait
 
-// start section ndarray
+// start section ndarrayTrait
 trait NDArray[T, S <: Shape]
-// end section ndarray
+// end section ndarrayTrait
 
-// start section randomnormal
+// start section randomnormalDef
 def random_normal[S <: Shape](shape: S): NDArray[Float, S] = ???
-// end section randomnormal
+// end section randomnormalDef
 
-// start section multiply
+// start section multiplyDef
 def multiply[T, S <: Shape](x: NDArray[T, S], y: NDArray[T, S]): NDArray[T, S] = ???
-// end section multiply
+// end section multiplyDef
 
 import scala.compiletime.ops.int.*
-// start section numelements
+// start section numelementsType
 type NumElements[X <: Shape] <: Int =
   X match {
     case Ø => 1
     case head #: tail => head * NumElements[tail]
   }
-// end section numelements
+// end section numelementsType
 
-// start section reshape
+// start section reshapeDef
 def reshape[T, From <: Shape, To <: Shape](arr: NDArray[T, From], newshape: To)
     (implicit ev: NumElements[From] =:= NumElements[To]): NDArray[T, To] = ???
-// end section reshape
+// end section reshapeDef
 
 import scala.compiletime.ops.int.+
-// start section reduce
+// start section reduceType
 type Reduce[S <: Shape, Axes <: None | Shape] <: Shape =
   Axes match {
     case None => Ø
@@ -64,7 +64,7 @@ type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
       // otherwise, do not reduce further
     }
   }
-// end section reduce
+// end section reduceType
 
 type Contains[Haystack <: Shape, Needle <: Int] <: Boolean = Haystack match {
   case Ø => false
@@ -82,7 +82,7 @@ type Remove[From <: Shape, Value <: Int & Singleton] <: Shape = From match {
   }
 }
[…]

### 2021-11-15 scala/src/main/scala/safe-implicits.scala
(Ab)Using Implicits

@@ -74,3 +74,36 @@ def main(ctx: Context[HNil]): Context[HNil] =
      .free("mem")
      .free("bool")
 // end section memImplicitMain
+
+trait js {
+// start section addEventListenerJS
+/** Attaches an event handler.
+ *
+ *  @param event     The event type, one of "mousedown", "mouseup",
+ *                   "mouseover", "mousewheel" and "contextmenu".
+ *  @param listener  The function to run when the event occurs.
+ */
+def addEventListener(event: String, listener: Event => Unit): Unit
+// end section addEventListenerJS
+
+}
+trait Event {
+
+// start section addEventListenerImplicitDef
+type EventTypes =
+  "mousedown" :: "mouseup" :: "mouseover" :: "mousewheel" :: "contextmenu" :: HNil
+
+def addEventListener[E <: Singleton]
+  (event: E, listener: Event => Unit)
+  (implicit ev: Remove[E, EventTypes, ?]): Unit
+// end section addEventListenerImplicitDef
+
+// start section addEventListenerImplicitCall
+addEventListener("mouseover", e => ())(
+  // Evidence that "mouseover" is in EventTypes,
+  // automatically infered by the compiler.
+  Remove.casetail(Remove.casetail(Remove.casehead))
+)
+// end section addEventListenerImplicitCall
+
+}

### 2021-11-17 scala/src/main/scala/safe-implicits.scala
(Ab)Using Implicits

@@ -8,11 +8,6 @@ enum HList:
 // end section hlistEnumDefinition
 import HList._
 
-// sealed trait HList
-// case object HNil extends HList
-// case class ::[+H, +T <: HList](head: H, tail: T) extends HList
-// type HNil = HNil.type
-
 // start section memImplicitRemove
 trait Remove[V, Ps <: HList, Out <: HList]
 
@@ -34,6 +29,27 @@ object NotIn:
       no: NotGiven[V =:= Ph],
       xs: NotIn[V, Pt]
     ): NotIn[V, Ph :: Pt] = new NotIn {}
+
+// start section removeAllWithPriority
+trait RemoveAll[V, Ps <: HList] { type Out <: HList }
+
+trait RemoveAllLowPrio:
+  type Aux[V, Ps <: HList, Out <: HList] = RemoveAll[V, Ps] { type Out }
+
+  implicit def casediff[V, Ph, Pt <: HList, Out <: HList]
+    (implicit ev: Aux[Ph, Pt, Out])
+    : Aux[V, Ph :: Pt, Ph :: Out] = new RemoveAll {}
+
+object RemoveAll extends RemoveAllLowPrio:
+  implicit def casenil[V]
+    : Aux[V, HNil, HNil] = new RemoveAll {}
+
+  implicit def casematch[V, Ps <: HList, Out <: HList]
+    (implicit ev: Aux[V, Ps, Out])
+    : Aux[V, V :: Ps, Out] = new RemoveAll {}
+// end section removeAllWithPriority
+
+val removeAllTest = implicitly[RemoveAll.Aux[1, 1 :: 2 :: 1 :: 3 :: HNil, 2 :: 3 :: HNil]]
 
 // start section memImplicitContext
 trait Context[Ps <: HList]:
@@ -78,7 +94,6 @@ def main(ctx: Context[HNil]): Context[HNil] =
 trait js {
 // start section addEventListenerJS
 /** Attaches an event handler.
- *
  *  @param event     The event type, one of "mousedown", "mouseup",
  *                   "mouseover", "mousewheel" and "contextmenu".
  *  @param listener  The function to run when the event occurs.
@@ -94,14 +109,15 @@ type EventTypes =
   "mousedown" :: "mouseup" :: "mouseover" :: "mousewheel" :: "contextmenu" :: HNil
 
 def addEventListener[E <: Singleton]
-  (event: E, listener: Event => Unit)
[…]

### 2021-11-22 scala/src/main/scala/safe-implicits.scala
(Ab)Using Implicits

@@ -108,12 +108,13 @@ trait Event {
 type EventTypes =
   "mousedown" :: "mouseup" :: "mouseover" :: "mousewheel" :: "contextmenu" :: HNil
 
-def addEventListener[E <: Singleton]
+def addEventListener[E <: String & Singleton]
   (event: E, handler: Event => Unit)
   (implicit ev: Remove[E, EventTypes, ?]): Unit
 // end section addEventListenerImplicitDef
 
 val myHandler: Event => Unit = e => ()
+addEventListener("mouseover", myHandler) // implicitly
 
 // start section addEventListenerImplicitCall
 addEventListener("mouseover", myHandler)(

### 2021-11-23 article.tex
Remove CodeSection suffixes
(Ab)Using Implicits

@@
 Each case consists of a \emph{type test} and a corresponding expression/type.
-At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase \citep{abadi1991dynamic}).
+At the term level, a type test consists of an inheritance check against a particular class (this is also known as a typecase~\citep{abadi1991dynamic}).
 At the type level, a type test corresponds to a subtyping test with a particular type.
@@
 
-In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null} \citep{nieto2020scala}.
+In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null}~\citep{nieto2020scala}.
 Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.
@@
 The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
-Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses \citep{fengyun2016generic}.
+Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{fengyun2016generic}.
 Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
@@
 Scala's primary platform is the Java virtual machine (JVM).
-On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased \citep{schinz2005compiling}.
+On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased~\citep{schinz2005compiling}.
 For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.
@@
 Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
-This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level \citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.
+This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.
 
@@
 %
-\shapeCodeSection
+\shapeEnum
 %
@@
 %
-\ndarrayCodeSection
+\ndarrayTrait
 
@@
 %
-\randomnormalCodeSection
+\randomnormalDef
 
@@
 %
-\multiplyCodeSection
+\multiplyDef
 %
@@
 %
-\numelementsCodeSection
+\numelementsType
 
@@
 %
-\reshapeCodeSection
+\reshapeDef
 %
@@
 
-\npmeanCodeSection
+\npmeanDef
 
@@
 %
-\reduceCodeSection
+\reduceType
 
@@
 %
-\badreducescalaCodeSection
+\badreducescala
 %
@@
 
-% \goodreducescalaCodeSection
+% \goodreducescala
 

### 2021-11-23 scala/src/main/scala/safe-implicits.scala

@@ -1,5 +1,4 @@
 package foo
-import scala.util.NotGiven
 
 // start section hlistEnumDefinition
 enum HList:
@@ -20,15 +19,24 @@ object Remove:
     : Remove[V, Ph :: Pt, Ph :: Out] = new Remove {}
 // end section memImplicitRemove
 
+// start section memImplicitNotIn
 trait NotIn[V, Ps <: HList]
 
 object NotIn:
   implicit def casenil[V]: NotIn[V, HNil] = new NotIn {}
-  implicit def casecons[Ph, Pt <: HList, V]
-    (implicit
-      no: NotGiven[V =:= Ph],
-      xs: NotIn[V, Pt]
-    ): NotIn[V, Ph :: Pt] = new NotIn {}
+
+  implicit def casecons[V, Ph, Pt <: HList]
+    (implicit xs: NotIn[V, Pt])
+    : NotIn[V, Ph :: Pt] = new NotIn {}
+
+  implicit def ambiguous1[V, Ph, Pt <: HList]
+    (implicit ev: V =:= Ph)
+    : NotIn[V, Ph :: Pt] = new NotIn {}
+
+  implicit def ambiguous2[V, Ph, Pt <: HList]
+    (implicit ev: V =:= Ph)
+    : NotIn[V, Ph :: Pt] = new NotIn {}
+// end section memImplicitNotIn
 
 // start section removeAllWithPriority
 trait RemoveAll[V, Ps <: HList] { type Out <: HList }
@@ -118,7 +126,7 @@ addEventListener("mouseover", myHandler) // implicitly
 
 // start section addEventListenerImplicitCall
 addEventListener("mouseover", myHandler)(
-  // Evidence that "mouseover" is in EventTypes, infered automatically.
+  // Evidence that "mouseover" is a valid event type, infered automatically.
   Remove.casetail(Remove.casetail(Remove.casehead))
 )
 // end section addEventListenerImplicitCall

### 2021-11-23 scala/src/test/scala/bad-implicits.scala

@@ -1,0 +1,9 @@
+object example {
+// start section implicitAmbiguity
+implicit val number: Int = 1
+implicit val modulo: Int = 3
+def addm(x: Int, y: Int)(implicit m: Int) = (x + y) % m
+addm(4, 5) // Error: ambiguous implicit arguments: both value number and
+           // value modulo match type Int of parameter m of method addm.
+// end section implicitAmbiguity
+}

### 2021-11-23 scala/src/main/scala/safe-implicits.scala
(Ab)Using Implicits

@@ -26,16 +26,13 @@ object NotIn:
   implicit def casenil[V]: NotIn[V, HNil] = new NotIn {}
 
   implicit def casecons[V, Ph, Pt <: HList]
-    (implicit xs: NotIn[V, Pt])
-    : NotIn[V, Ph :: Pt] = new NotIn {}
+    (implicit xs: NotIn[V, Pt]): NotIn[V, Ph :: Pt] = new NotIn {}
 
   implicit def ambiguous1[V, Ph, Pt <: HList]
-    (implicit ev: V =:= Ph)
-    : NotIn[V, Ph :: Pt] = new NotIn {}
+    (implicit ev: V =:= Ph): NotIn[V, Ph :: Pt] = new NotIn {}
 
   implicit def ambiguous2[V, Ph, Pt <: HList]
-    (implicit ev: V =:= Ph)
-    : NotIn[V, Ph :: Pt] = new NotIn {}
+    (implicit ev: V =:= Ph): NotIn[V, Ph :: Pt] = new NotIn {}
 // end section memImplicitNotIn
 
 // start section removeAllWithPriority

### 2021-11-24 article.tex
(Ab)Using Implicits

@@
 \subsection{A Lightweight Form of Dependent Typing}
-Match types enable a lightweight form of dependent typing, since a term-level pattern matching expression can be typed accordingly at the type level as a match type.
+Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.
 Consider the following function definition:
@@
 
-Haskell's type families allow programmers to define type-level functions using pattern matching~\citep{chakravarty2005associated, schrijvers2008open}.
+Haskell's type families~\citep{chakravarty2005associated, schrijvers2008open} allow programmers to define type-level functions using pattern matching.
 By default, type families are open, which means that a definition can spread across multiple files and compilations.

### 2021-11-24 scala/src/main/scala/safe-implicits.scala

@@ -35,6 +35,18 @@ object NotIn:
     (implicit ev: V =:= Ph): NotIn[V, Ph :: Pt] = new NotIn {}
 // end section memImplicitNotIn
 
+object NotGivenExample {
+import scala.util.NotGiven
+
+// start section memImplicitNotGiven
+implicit def casecons[V, Ph, Pt <: HList]
+  (implicit
+    no: NotGiven[V =:= Ph],
+    xs: NotIn[V, Pt]
+  ): NotIn[V, Ph :: Pt] = new NotIn {}
+// end section memImplicitNotGiven
+}
+
 // start section removeAllWithPriority
 trait RemoveAll[V, Ps <: HList] { type Out <: HList }
 

### 2021-11-24 article.tex
Down to 7 bad boxes...

@@
 One notable exception is the TypeScript language, which recently introduced a new feature called \emph{conditional type}, a type-level ternary operator based on subtyping.
-A conditional type, written \typescript{S extends T ?
-Tt : Tf}, reduces to \typescript{Tt} when \typescript{S} is a subtype of \typescript{T}, to \typescript{Tf} when \typescript{S} is not a subtype of \typescript{T}, and is left unreduced when types variables do not allow to draw a conclusion.
+A conditional type, written |S extends T ?
+Tt : Tf|, reduces to |Tt| when |S| is a subtype of |T|, to |Tf| when |S| is not a subtype of |T|, and is left unreduced when types variables do not allow to draw a conclusion.
 Unfortunately, the algorithm used to reduce such conditional types is both \emph{unsound} and \emph{incomplete}.
@@
 %
-The example, which we explain in detail in \Cref{sec:overview}, defines the type \dotty{Elem} by matching on the type parameter \dotty{X}.
+The example, which we explain in detail in \Cref{sec:overview}, defines the type |Elem| by matching on the type parameter |X|.
 %
@@
 %
-This example defines a type \dotty{Elem} parametrized by one type parameter \dotty{X}. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
+This example defines a type |Elem| parametrized by one type parameter |X|. The right-hand side is defined in terms of a match on the type parameter -- a \emph{match type}.
 A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
@@
 %
-Here we use \dotty{S =:= T} to denote type equality between the two types \dotty{S} and \dotty{T}, witnessed by mutual subtyping.
+Here we use |S =:= T| to denote type equality between the two types |S| and |T|, witnessed by mutual subtyping.
 To reduce a match type, the scrutinee is compared to each pattern, one after the other, using \emph{subtyping}.
-For example, although \dotty{String} is a subtype of both \dotty{String} and \dotty{Any} (the top of Scala's subtyping lattice), \dotty{Elem[String]} reduces to \dotty{Char} because the corresponding case appears first.
+For example, although |String| is a subtype of both |String| and |Any| (the top of Scala's subtyping lattice), |Elem[String]| reduces to |Char| because the corresponding case appears first.
 
-When the scrutinee type is a \dotty{List}, the match type \dotty{Elem} is defined recursively on the element type of the list.
-Hence, in our example \dotty{Elem[List[Int]]} first reduces to the type \dotty{Elem[Int]}, and eventually to the type \dotty{Int}.
+When the scrutinee type is a |List|, the match type |Elem| is defined recursively on the element type of the list.
+Hence, in our example |Elem[List[Int]]| first reduces to the type |Elem[Int]|, and eventually to the type |Int|.
 
@@
 %
-This definition is well-typed because the match expression in \dotty{elem}'s body has the exact same scrutinee and pattern types as \dotty{Elem[X]} (the function's return type).
+This definition is well-typed because the match expression in |elem|'s body has the exact same scrutinee and pattern types as |Elem[X]| (the function's return type).
 
-Thanks to Scala's type inference, a call to the \dotty{elem} function can have a result type that \emph{depends} on a term-level parameter.
-For instance, in the expression \dotty{elem(1)}, the Scala compiler infers the singleton type \dotty{X = 1} for \dotty{elem}'s type parameter.
-This expression thus has type \dotty{Elem[1]}, which reduces to \dotty{Int} (via \dotty{Elem}'s third case).
-Similarly, in \dotty{elem(x)}, the compiler infers the singleton type \dotty{X = x.type} and the expression has type \dotty{Elem[x.type]}, which might reduce further at the callsite depending on \dotty{x}'s type.
+Thanks to Scala's type inference, a call to the |elem| function can have a result type that \emph{depends} on a term-level parameter.
+For instance, in the expression |elem(1)|, the Scala compiler infers the singleton type |X = 1| for |elem|'s type parameter.
+This expression thus has type |Elem[1]|, which reduces to |Int| (via |Elem|'s third case).
+Similarly, in |elem(x)|, the compiler infers the singleton type |X = x.type| and the expression has type |Elem[x.type]|, which might reduce further at the callsite depending on |x|'s type.
 
@@
 
-The necessity for disjointness is best illustrated with an example. Consider \dotty{Seq[Int]}, the type of integer sequences. \dotty{Elem[Seq[Int]]} does not reduce:
+The necessity for disjointness is best illustrated with an example. Consider |Seq[Int]|, the type of integer sequences. |Elem[Seq[Int]]| does not reduce:
 \begin{enumerate}
-  \item \dotty{Elem}'s first case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{String}.
-  \item \dotty{Elem}'s second case does not apply because \dotty{Seq[Int]} is not a subtype of \dotty{List[Int]} (lists are sequences, but not the other way around).
-  \item \dotty{Elem}'s third case is \emph{not considered} because \dotty{Seq[Int]} and \dotty{List[Int]} are not disjoint.
+  \item |Elem|'s first case does not apply because |Seq[Int]| is not a subtype of |String|.
+  \item |Elem|'s second case does not apply because |Seq[Int]| is not a subtype of |List[Int]| (lists are sequences, but not the other way around).
+  \item |Elem|'s third case is \emph{not considered} because |Seq[Int]| and |List[Int]| are not disjoint.
 \end{enumerate}
 Therefore, the reduction algorithm gets stuck on the second case and the overall type is irreducible.
-Without disjointness, \dotty{Elem[Seq[Int]]} would reduce to \dotty{Seq[Int]} (via \dotty{Elem}'s third case), which would be unsound.
-For example, the expression \dotty{elem[Seq[Int]](List(1,2,3))} would have type \dotty{Seq[Int]}, but evaluates to the integer \dotty{1}.
+Without disjointness, |Elem[Seq[Int]]| would reduce to |Seq[Int]| (via |Elem|'s third case), which would be unsound.
+For example, the expression |elem[Seq[Int]](List(1,2,3))| would have type |Seq[Int]|, but evaluates to the integer |1|.
 
@@
 The inheritance and disjointness parameters can be understood as a representation of a hierarchy of Scala traits and classes.
-For example, \dotty{trait C1; class C2 extends C1} is represented  in \Fm as $\Ψ \= \{ \(\C_1, \C_2) \};~ \Ξ \= \{ \}$.
+For example, |trait C1; class C2 extends C1| is represented  in \Fm as $\Ψ \= \{ \(\C_1, \C_2) \};~ \Ξ \= \{ \}$.
 This representation also models the fact that certain types cannot possibly have common instances.
-For example, \dotty{class C3; class C4} is represented as $\Ψ \= \{ \};~ \Ξ \= \{ \(\C_3, \C_4), \(\C_4, \C_3) \}$, since Scala disallows multiple class inheritance.
+For example, |class C3; class C4| is represented as $\Ψ \= \{ \};~ \Ξ \= \{ \(\C_3, \C_4), \(\C_4, \C_3) \}$, since Scala disallows multiple class inheritance.
 Inheritance and disjointness must be consistent in the sense that $\(\A, \B) \∈ \Ξ$ implies that there is no class $\C$ such that $\(\C, \A) \∈ \Ψ$ and $\(\C, \B) \∈ \Ψ$.
@@
 For instance, given the function $\f$ defined in \Cref{ex:simplereduction}, the expression $\(\f~\A~\(\new \A))$ evaluates to $\foo$ and $\(\f~\B~\(\new \B))$ to $\bur$.
-
-% Even though \Fm is limited to type test patterns, more elaborate patterns can be expressed using type tests.
-% For example, consider the following data type deconstruction pattern (left-hand side):
-
-% list match {
-%   case x :: xs => print(x)
-%   case Nil => ()
-% }
-
-% list match {
-%   case l : ::[Any] => print(l.head)
-%   case Nil => ()
-% }
-
-% \noindent
-% This pattern could alternatively be written as type test pattern (right-hand side), were \dotty{::[Any]} is the type of Scala's cons node.
 
@@
 The class inheritance $\Ψ$ and class disjointness $\Ξ$ remain as binary relations on $\C$.
-A generic instantiation in the class hierarchy is represented as an element of $\Ψ$, for example, \dotty{A1 extends B2[A3]} is represented as $\(\A_1, \B_2~\A_3) \∈ \Ψ$.
-Generic inheritance is represented using multiple entries in $\Ψ$, for example, \dotty{B1[T] extends B2[T]} is represented as $\∀ \T. \(\B_1~\T, \B_2~\T) \∈ \Ψ$.
+A generic instantiation in the class hierarchy is represented as an element of $\Ψ$, for example, |A1 extends B2[A3]| is represented as $\(\A_1, \B_2~\A_3) \∈ \Ψ$.
+Generic inheritance is represented using multiple entries in $\Ψ$, for example, |B1[T] extends B2[T]| is represented as $\∀ \T. \(\B_1~\T, \B_2~\T) \∈ \Ψ$.
 This approach allows us to reuse most of \Fm's definitions.
@@
 An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
-The bottom of Scala's subtyping lattice, called \dotty{Nothing}, provides a direct way to refer to empty sets of values.
+The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
 Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
@@
 %
-In this example, the definition of \dotty{f} in \dotty{C} type-checks because \dotty{X & String} and \dotty{Int} are disjoint (since \dotty{String} and \dotty{Int} are disjoint) and \dotty{M[X & String]} reduces to \dotty{Int} (\dotty{M}'s second case applies).
-Class \dotty{D} refines the definition of \dotty{C} by giving concrete definition of \dotty{X}.
-The unsoundness manifests itself in the body of class \dotty{D}, where \dotty{X & String} is a subtype of \dotty{Int} and \dotty{M[X & String]} reduces to \dotty{String} (\dotty{M}'s first case applies).
-There, it is possible to call the function~\dotty{f} with a string argument, which would result in a runtime error.
-Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent \dotty{M[Int & String]} from reducing given that \dotty{Int & String} is not inhabited.
+In this example, the definition of |f| in |C| type-checks because |X & String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X & String]| reduces to |Int| (|M|'s second case applies).
+Class |D| refines the definition of |C| by giving concrete definition of |X|.
+The unsoundness manifests itself in the body of class |D|, where |X & String| is a subtype of |Int| and |M[X & String]| reduces to |String| (|M|'s first case applies).
+There, it is possible to call the function~|f| with a string argument, which would result in a runtime error.
+Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int & String]| from reducing given that |Int & String| is not inhabited.
 
@@
 
-In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \dotty{A | Null}~\citep{nieto2020scala}.
+In Scala 3, null values no longer inhabit every type: nullable types require explicit annotations of the form \lstinline!A | Null!~\citep{nieto2020scala}.
 Our implementation of subtyping and disjointness handles union types and therefore needs no particular treatment of null values.
-
-% To allow easy migration from older versions, the strict treatment of nulls is still optional in Scala 3.0, enabled by the command line option \dotty{-Yexplicit-nulls}.
+To allow easy migration from older versions, the strict treatment of nulls is still optional in Scala 3.0, enabled by the command line option |-Yexplicit-nulls|.
 The plan is to make strict null checking the default in the future.
@@
 These annotations allow programmers to specify how the subtyping of annotated parameters influences the subtyping of the higher-kinded type.
-For instance, \dotty{type F[+T]} defines a type \dotty{F} that is covariant in its first type parameter, meaning that \dotty{T1 <: T2} implies \dotty{F[T1] <: F[T2]}.
-Contravariance, written \dotty{type G[-T]}, has the opposite meaning: \dotty{T1 <: T2} implies \dotty{G[T2] <: G[T1]}.
+For instance, |type F[+T]| defines a type |F| that is covariant in its first type parameter, meaning that |T1 <: T2| implies |F[T1] <: F[T2]|.
+Contravariance, written |type G[-T]|, has the opposite meaning: |T1 <: T2| implies |G[T2] <: G[T1]|.
 
-It would appear that co- and contravariant types are always overlapping, given that, for all types \dotty{X}, \dotty{F[Nothing] <:} \dotty{F[X]} and \dotty{G[Any] <: G[X]} (where \dotty{Nothing} and \dotty{Any} are Scala's bottom and top types).
-However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the \dotty{Nothing} instantiation can be ruled out because no runtime program can produce a value of that type.
+It would appear that co- and contravariant types are always overlapping, given that, for all types |X|, |F[Nothing] <: F[X]| and |G[Any] <: G[X]| (where |Nothing| and |Any| are Scala's bottom and top types).
+However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the |Nothing| instantiation can be ruled out because no runtime program can produce a value of that type.
 
 Scala tuples, for example, fall into this category.
-\dotty{Tuple2}, the class for pairs, is defined as follows:
+|Tuple2|, the class for pairs, is defined as follows:
 %
@@
 %
-Given two instantiations of this this type, \dotty{Tuple2[Int,} \dotty{X]} and \dotty{Tuple2[String,X]}, although \dotty{Tuple2[Nothing,X]} is a subtype of both, there is no runtime value of type \dotty{Tuple2[} \dotty{Nothing,X]} (since \dotty{Nothing} is uninhabited), and, as a result, those two types are disjoint.
+Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
@@
 Erasure directly affects the type tests that can be performed at runtime.
-For example, while \dotty{case xs: List[Int] =>} is a syntactically valid pattern, it will lead to a compiler warning since the \dotty{Int} type parameter is eliminated by erasure and cannot be checked at runtime.
+For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it will lead to a compiler warning since the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
 
@@
 \subsection{Shape Errors in Python}
-In the example Python code below, the \python{img_batch} ndarray is a batch of 25 randomly generated RGB images of size 256$\×$256.
+In the example Python code below, the |img_batch| ndarray is a batch of 25 randomly generated RGB images of size 256$\×$256.
 The code aims to compute a vector of length 25 containing the average grayscale color of each image in the batch, and then create a square 5$\×$5 image of the average grayscale colors.
@@
 %
-The error is in the call to \python{np.mean}, which takes as argument a list of axes to reduce along.
-Unfortunately, the arguments to \python{np.mean} are off-by-one and result in a vector of length 3 (\python{img_batch}'s last dimension) instead of the intended length of 25 (\python{img_batch}'s first dimension); \python{avg_color} thus contains the average RGB color of the batch instead of the average grayscale color for each image.
+The error is in the call to |np.mean|, which takes as argument a list of axes to reduce along.
+Unfortunately, the arguments to |np.mean| are off-by-one and result in a vector of length 3 (|img_batch|'s last dimension) instead of the intended length of 25 (|img_batch|'s first dimension); |avg_color| thus contains the average RGB color of the batch instead of the average grayscale color for each image.
 The reshape operation will then fail at runtime, as it cannot reshape a 3-element vector into a 25-element matrix.
@@
 After a preliminary introduction of singleton types (\Cref{subsec:singleton-types}), we introduce ndarray shapes at the type level using HLists (\Cref{subsec:array-shape}).
-In \Cref{subsec:shape-match-types}, we show type-level implementations of the \python{np.mean} and \python{np.reshape} operations using match types.
+In \Cref{subsec:shape-match-types}, we show type-level implementations of the |np.mean| and |np.reshape| operations using match types.
 Finally, we show how this newly defined API can detect and report the error from our original Python example.
@@
 Scala supports singleton types, which are types inhabited by a single value~\citep{leontiev2014sip}.
-For instance, the singleton type \dotty{1} denotes the type containing the integer value \dotty{1}.
+For instance, the singleton type |1| denotes the type containing the integer value |1|.
 The Scala standard library contains several predefined match types to perform arithmetic operations at the type level.
-For instance, \dotty{type +[A <: Int,}\dotty{B <: Int]} represents the addition of integer singleton types.
+For instance, |type +[A <: Int, B <: Int]| represents the addition of integer singleton types.
 Internally, the compiler is special-cased to implement these arithmetic operations using constant folding.
@@
 We represent the shape of an ndarray at the type level using a heterogeneous type list, or HList for short~\citep{kiselyov2004strongly}.
-We define an HList called \dotty{Shape} as an ADT with two constructors\footnotemark, \dotty{#:} and \dotty{Ø}.
+We define an HList called |Shape| as an ADT with two constructors\footnotemark, |#:| and |Ø|.
 \footnotetext{%
-In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \dotty{Shape} data type.}
+In the interest of clarity, our presentation omits type bounds that are necessary to guide type inference, for example in the definition of the \lstinline!Shape! data type.}
 %
@@
 %
-This data type definition allows us to write lists of dimension sizes, both at the term level as \dotty{#:(3, #:(4, Ø))}, and at the type level as \dotty{#:[3, #:[4, Ø]]}.
-The HList type can equivalently be written with \dotty{#:} in infix notation, as \dotty{3 #: 4 #: Ø}.
+This data type definition allows us to write lists of dimension sizes, both at the term level as |#:(3, #:(4, Ø))|, and at the type level as |#:[3, #:[4, Ø]]|.
+The HList type can equivalently be written with |#:| in infix notation, as |3 #: 4 #: Ø|.
 
-To represent ndarrays, we define the \dotty{NDArray} type. This type is indexed by the ndarray element type (\dotty{T}), and by the ndarray shape (\dotty{S}), represented as an HList:
+To represent ndarrays, we define the |NDArray| type. This type is indexed by the ndarray element type (|T|), and by the ndarray shape (|S|), represented as an HList:
 %
@@
 
-The goal of our presentation is to define type-safe operations on \dotty{NDArrays}.
-Since our focus is on the type-level, we do not include value-level counterparts to \dotty{T} and \dotty{S} in the definition of \dotty{NDArray}, but this would be necessary in a complete implementation.
+The goal of our presentation is to define type-safe operations on |NDArrays|.
+Since our focus is on the type-level, we do not include value-level counterparts to |T| and |S| in the definition of |NDArray|, but this would be necessary in a complete implementation.
 
-To construct ndarrays, we define \dotty{random_normal}, which creates an ndarray of a given shape, where all elements are random \dotty{Float}s:\footnotemark
+To construct ndarrays, we define |random_normal|, which creates an ndarray of a given shape, where all elements are random |Float|s:\footnotemark
 \footnotetext{%
@@
 Encoding the types and shapes of ndarrays in the types allows us to readily provide type- and shape-safety for simple ndarray operations.
-For instance, the element-wise Hadamard product, written as \python{np.multiply(x, y)}, requires the \python{x} and \python{y} ndarrays to have the same shape and element types.
+For instance, the element-wise Hadamard product, written as |np.multiply(x, y)|, requires the |x| and |y| ndarrays to have the same shape and element types.
 This constraint does not require any match types, but can simply be expressed as:
@@
 
-An operation commonly used in NumPy is \python{np.reshape}, which changes the shape of an ndarray, but does not change its values.
+An operation commonly used in NumPy is |np.reshape|, which changes the shape of an ndarray, but does not change its values.
 A restriction imposed by the NumPy API is that the output shape must have the same number of elements as the input shape.
-The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}\dotty{2 #: 3 #: 4 #: Ø}\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
-Note that an ndarray of shape \dotty{Ø} is a scalar, and thus has a single element.
+The number of elements of a shape is the product of the sizes of its dimensions; an ndarray of shape\hspace{4pt}|2 #: 3 #: 4 #: Ø|\hspace{4pt}has \begin{math}2\times3\times4 = 24\end{math} elements.
+Note that an ndarray of shape |Ø| is a scalar, and thus has a single element.
 This can be naturally expressed with a match type:
@@
 To restrict reshaping to be applicable only on valid shapes, the type system must support encoding type equality constraints.
-For this, we make use of Scala's implicit parameters, and of the \dotty{=:=} type equality constraint that is a part of Scala's standard library.
+For this, we make use of Scala's implicit parameters, and of the |=:=| type equality constraint that is a part of Scala's standard library.
 %
@@
 %
-With this definition of \dotty{reshape}, the compiler will only accept a usage of \dotty{reshape} if it is able to prove that the number of elements in the old shape is the same as in the new shape.
+With this definition of |reshape|, the compiler will only accept a usage of |reshape| if it is able to prove that the number of elements in the old shape is the same as in the new shape.
 This example illustrates how match types can be used in concert with existing features like singleton types and implicit resolution to express powerful constraints.
@@
 
-The NumPy API provides a variety of functions to reduce along a set of axes of an ndarray, such as \python{np.mean(ndarray, axes)} or \python{np.var(ndarray, axes)}.
-The \python{axes} parameter is a list of indices of dimensions, listing exactly those dimensions that will no longer be present in the output ndarray.
+The NumPy API provides a variety of functions to reduce along a set of axes of an ndarray, such as |np.mean(ndarray, axes)| or |np.var(ndarray, axes)|.
+The |axes| parameter is a list of indices of dimensions, listing exactly those dimensions that will no longer be present in the output ndarray.
 The dimension indices can be unordered and repeated, and out-of-bounds indices result in an error.
-In the Python API, passing the \python{None} value instead of a list of axes reduces along all axes, meaning that the operation returns a scalar.
-Note that this is the opposite of passing \dotty{Ø}, which means that we reduce over no axes (effectively a no-op).
+In the Python API, passing the |None| value instead of a list of axes reduces along all axes, meaning that the operation returns a scalar.
+Note that this is the opposite of passing |Ø|, which means that we reduce over no axes (effectively a no-op).
 
 This behavior is more complex than the previous examples, but can still be described accurately by a match type.
-We use a match type called \dotty{Reduce} to compute the return shape of the operation.
+We use a match type called |Reduce| to compute the return shape of the operation.
 
@@
 We implement reductions along a given list of indices with logic similar to two nested loops.
-The outer loop, \dotty{Loop}, traverses the shape and counts the current index.
-The inner loops are implemented using \dotty{Contains} and \dotty{Remove}, standard operations on HLists (omitted).
-When \dotty{Loop} reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
+The outer loop, |Loop|, traverses the shape and counts the current index.
+The inner loops are implemented using |Contains| and |Remove|, standard operations on HLists (omitted).
+When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
 %
 \reduceType
+\reduceLoop
 
@@
 
-Having defined \dotty{random_normal}, \dotty{reshape} and \dotty{mean}, we can rewrite our original Python example in Scala:
+Having defined |random_normal|, |reshape| and |mean|, we can rewrite our original Python example in Scala:
 %
@@
 %
-As expected, the call to \dotty{mean} returns a tensor of shape \dotty{3 #: Ø}.
-Therefore, the call to \dotty{reshape} does not type-check since the input shape has 3 elements instead of 25.
-If we fix the off by one error to reduce along the correct indices, \dotty{1 #: 2 #:} \dotty{3 #: Ø}, the call to \dotty{reshape} type-checks and \dotty{avg_color} has shape \dotty{25 #: Ø}, as expected.
+As expected, the call to |mean| returns a tensor of shape |3 #: Ø|.
+Therefore, the call to |reshape| does not type-check since the input shape has 3 elements instead of 25.
+If we fix the off by one error to reduce along the correct indices, |1 #: 2 #: 3 #: Ø|, the call to |reshape| type-checks and |avg_color| has shape |25 #: Ø|, as expected.
 
@@
 
-% A significant difference between type-checking type families and match types is that type families are entirely handled during type inference, while match types reduce during subtyping.
-% Influencing Haskell's global type inference is notoriously difficult, which can, unfortunately, hinder the usability of type families.
-% For instance, in a functional pearl on typed matrices~\citep{santos2020match}, the authors report that their first (simpler) attempt was deemed inconclusive due to limitations of the Haskell type inference when combining type families with pattern matching.
-
 \subsection{Roles in Haskell}
@@
 
-Haskell's roles were introduced by~\citet{weirich2011generative} to fix a long-standing unsoundness caused by the interaction of open type families and the \haskell{newtype} construct.
+Haskell's roles were introduced by~\citet{weirich2011generative} to fix a long-standing unsoundness caused by the interaction of open type families and the |newtype| construct.
 We understand roles as type annotations which specify whether a given type can safely be nominally compared to other types, or if \emph{representational equality} (RE) should be used instead.
 The word representation in RE refers to the runtime representation of a type.
-In particular, RE dealiases \haskell{newtype} constructs.
+In particular, RE dealiases |newtype| constructs.
 
@@
 %
-In this example, type \dotty{Age} is abstract in \dotty{AgeClass} and concrete in \dotty{AgeObject}.
-In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates \dotty{Age} and \dotty{Int}.
-Such type family would reduce differently in \dotty{AgeClass}, where the types are different, and in \dotty{AgeObject}, where those two types are synonyms, which can easily be exploited to obtain a runtime error.
+In this example, type |Age| is abstract in |AgeClass| and concrete in |AgeObject|.
+In the pre-role Haskell equivalent of this example, the unsoundness comes when the above definitions are combined with a type family that discriminates |Age| and |Int|.
+Such type family would reduce differently in |AgeClass|, where the types are different, and in |AgeObject|, where those two types are synonyms, which can easily be exploited to obtain a runtime error.
 
 Luckily, our design of match types is not affected by this issue.
-The reason comes from the use of subtyping (and disjointness), which shields our implementation from incorrectly discriminating \dotty{Age} from \dotty{Int} when \dotty{Age} is abstract.
+The reason comes from the use of subtyping (and disjointness), which shields our implementation from incorrectly discriminating |Age| from |Int| when |Age| is abstract.
 Consider the following match type definition (directly translated from the Haskell example):
@@
 %
-Our algorithm would not reduce \dotty{M[Int]} to \dotty{Bool} in \dotty{AgeClass} as this reduction would require evidence that \dotty{Age} and \dotty{Int} are disjoint, which cannot be constructed when \dotty{Age} is an unbounded abstract type.
+Our algorithm would not reduce |M[Int]| to |Bool| in |AgeClass| as this reduction would require evidence that |Age| and |Int| are disjoint, which cannot be constructed when |Age| is an unbounded abstract type.
 
@@
 The TypeScript language specification briefly describes the algorithm used to reduce conditional types in the presence of type variables~\citep{microsoft2020typescript}.
-Given a type \typescript{S extends T ?
-Tt : Tf}, the TypeScript compiler first replaces all the type parameters in \typescript{S} and \typescript{T} by \typescript{any} (the top of TypeScript's subtyping lattice).
-If the resulting types (after substitution) are not subtypes, the overall condition is reduced to \typescript{Tf}.
+Given a type |S extends T ? Tt : Tf|, the TypeScript compiler first replaces all the type parameters in |S| and |T| by |any| (the top of TypeScript's subtyping lattice).
+If the resulting types (after substitution) are not subtypes, the overall condition is reduced to |Tf|.
 Unfortunately, this algorithm is both \emph{unsound} and \emph{incomplete}.
@@
 Although TypeScript does not have syntax for variance annotations, function types are covariant in their return type and contravariant in their arguments.
-The conditional type unification algorithm wrongly approximates \typescript{X => string} to \typescript{any => string} and unifies the former with latter, which can lead to a runtime errors.
+The conditional type unification algorithm wrongly approximates |X => string| to |any => string| and unifies the former with latter, which can lead to a runtime errors.
 
@@
 %
-Here, TypeScript's reduction algorithm fails to recognize that \typescript{new A} can be typed as \typescript{M<X>}, even though \typescript{X} is clearly a subtype of \typescript{string} in \typescript{f}'s body.
+Here, TypeScript's reduction algorithm fails to recognize that |new A| can be typed as |M<X>|, even though |X| is clearly a subtype of |string| in |f|'s body.
 

### 2021-11-24 scala/src/main/scala/numpy.scala

@@ -52,7 +52,9 @@ type Reduce[S <: Shape, Axes <: None | Shape] <: Shape =
     case None => Ø
     case Shape => Loop[S, Axes, 0]
   }
+// end section reduceType
 
+// start section reduceLoop
 type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
   S match {
     case head #: tail => Contains[Axes, I] match {
@@ -64,7 +66,7 @@ type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
       // otherwise, do not reduce further
     }
   }
-// end section reduceType
+// end section reduceLoop
 
 type Contains[Haystack <: Shape, Needle <: Int] <: Boolean = Haystack match {
   case Ø => false
@@ -83,6 +85,5 @@ type Remove[From <: Shape, Value <: Int & Singleton] <: Shape = From match {
 }
 
 // start section npmeanDef
-def mean[T, S <: Shape, A <: Shape](arr: NDArray[T, S], axes: A): NDArray[T,
-    Reduce[S, A]] = ???
+def mean[T, S <: Shape, A <: Shape](arr: NDArray[T, S], axes: A): NDArray[T, Reduce[S, A]] = ???
 // end section npmeanDef

### 2021-11-25 scala/src/main/scala/safe-implicits.scala
(Ab)Using Implicits

@@ -68,13 +68,13 @@ object RemoveAll extends RemoveAllLowPrio:
 
 val removeAllTest = implicitly[RemoveAll.Aux[1, 1 :: 2 :: 1 :: 3 :: HNil, 2 :: 3 :: HNil]]
 
-// start section memImplicitContext
+// start section memImplicitContextFree
 trait Context[Ps <: HList]:
   def free[V <: Singleton, Out <: HList]
     (v: V)
     (implicit ev: Remove[V, Ps, Out])
     : Context[Out]
-// end section memImplicitContext
+// end section memImplicitContextFree
 
   def malloc[V <: Singleton]
     (size: Int, v: V)
@@ -90,6 +90,16 @@ trait Context[Ps <: HList]:
     (body: Array[Byte] => Unit)
     (implicit ev: Remove[V, Ps, ?])
     : Context[Ps]
+
+object Ctx0 {
+// start section memImplicitContextMalloc
+trait Context[Ps <: HList]:
+  def malloc[V <: Singleton]
+    (size: Int, v: V)
+    (implicit ev: NotIn[V, Ps])
+    : Context[V :: Ps]
+// end section memImplicitContextMalloc
+}
 
 def f[Ps <: HList](ctx: Context["bool" :: Ps]): Context["bool" :: Ps] =
   ctx.deref("bool") { b =>

### 2021-11-30 article.tex
Add Dep. Scala preprint

@@
 Unfortunately, most of the existing research conducted in this domain is not directly applicable to languages with \emph{subtyping}.
-Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994singleton, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.
+Although the combination of subtyping and type-level programming has been studied extensively on the theoretical side, through the means of dependently typed systems~\citep{aspinall1994subtyping, zwanenburg1999pure, stone2000deciding, courant2003strong, hutchins2010pure, yang2017unifying}, the practical side remains largely unexplored.
 
@@
 
-In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994extension} with pattern matching, opaque classes, and match types.
+In this section, we formally present \SystemFm, an extension of \SystemFsub~\citep{cardelli1994an} with pattern matching, opaque classes, and match types.
 \Cref{fig:FMSyntaxEvaluation} defines \Fm's syntax and evaluation relation. \Cref{fig:FMTypingRules} defines \Fm's type system, composed of three relations: typing, subtyping, and type disjointness.
@@
 The Scala compiler checks for pattern matching exhaustivity to prevent runtime exceptions caused by missing cases.
-Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{fengyun2016generic}.
+Exhaustivity checking uses static knowledge about the class hierarchy (such as the sealed and final annotations) to check that every value in the scrutinee type is covered by the pattern clauses~\citep{liu2016a}.
 Non-exhaustive patterns are compiled with an additional "catch-all" case which throws a runtime exception.
@@
 Although this seems to be sufficient for most practical purposes, we plan on making it configurable.
-This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, vilhelm2015dependently, eisenberg2016dependent}.
+This mechanism is completely standard and already used in other programing languages with unbounded recursion at the type level~\citep{eisenberg2014closed, sjoberg2015a, eisenberg2016dependent}.
 
@@
 In this section, we present a case study to show how match types can be used to express complex type constraints, which in turn can prevent certain programming errors at compile time.
-To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020numpy}.
+To this end, we outline the type-level implementation of a library for multidimensional arrays which mimics the NumPy API~\citep{harris2020array}.
 The goal of our library is to provide a \emph{shape}-safe interface for manipulating $\n$-dimensional arrays (abbreviated ndarrays), where array shapes and indices are checked for errors at compile-time rather than at runtime.
-Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
+Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019machine, rush2019tensor}, and several solutions have already been proposed, notably in the form of libraries that rely on type-level programming~\citep{chen2017typesafe, huang2017hasktorch}.
 Our library uses match types to provide a shape-safe NumPy-like interface.
@@
 % \subsection{Related Work}
-% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019stuck, rush2019namedtensor}, and several solutions have been suggested.
+% Shape and indexing errors in ndarrays is a widely acknowledged problem~\citep{barham2019machine, rush2019tensor}, and several solutions have been suggested.
 % The Dex language~\citep{maclaurin2019dex} introduces an experimental language with a Hindley--Milner-based type system with types for index sets, but cannot express variable-rank ndarrays and lacks the type-level arithmetic needed for typing reshapes.
@@
 
-Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994singleton, stone2000deciding, courant2003strong}.
+Singleton types provide an interesting middle ground between unified and separate syntax and have also been studied in conjunction with dependent types and subtyping~\citep{aspinall1994subtyping, stone2000deciding, courant2003strong}.
 Singleton types give a mechanism to refer to terms \emph{in} types, usually by means of a set-like syntax.
@@
 
-Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{essence2016amin}.
+Dependent Object Types (DOT) are, to this day, the most significant effort in formalizing Scala's type system~\citep{amin2017type}.
 DOT does not directly support any form of type-level computation.
-We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017simple, giarrusso2020soundness}, extending DOT remains too big of a challenge to concisely describe language extensions.
+We considered using DOT as a starting point for our work, however, despite the recent effort to simplify DOT's soundness proof~\citep{rapoport2017a, giarrusso2020scala}, extending DOT remains too big of a challenge to concisely describe language extensions.
 
@@
 
-Haskell's type families~\citep{chakravarty2005associated, schrijvers2008open} allow programmers to define type-level functions using pattern matching.
+Haskell's type families~\citep{chakravarty2005associated, schrijvers2008type} allow programmers to define type-level functions using pattern matching.
 By default, type families are open, which means that a definition can spread across multiple files and compilations.
@@
 
-Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014typescript}.
+Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014understanding}.
 Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.

### 2021-11-30 preprint.tex
Add Dep. Scala preprint

\newcommand{\oursystem}{$\lambda${\tiny $\mathop{\protect\vphantom{X}}^\text{nd}_{<:\{\}}$}\xspace}
\newcommand\FR{System~FR\xspace}
\newcommand\singleton[1]{\lbrace #1 \rbrace}

Type-level programming is an increasingly popular way to obtain additional type safety.
Unfortunately, it remains a second-class citizen in the majority of industrially-used programming languages.
We propose a new dependently-typed system with subtyping and singleton types whose goal is to enable type-level programming in an accessible style.
At the heart of our system lies a non-deterministic choice operator.
We argue that embracing non-determinism is crucial for bringing dependent types to a broader audience of programmers, since real-world programs will inevitably interact with imprecisely-typed, or even impure code.
Furthermore, we show that singleton types combined with the choice operator can serve as a replacement for many type functions of interest in practice.
We establish the soundness of our approach using the Coq proof assistant.
Our soundness approach models non-determinism using additional function arguments to represent choices.
We represent type-level computation using singleton types and existential types that quantify over choice arguments.
To demonstrate the practicality of our type system, we present an implementation as a modification of the Scala compiler.
We provide a case study in which we develop a strongly-typed wrapper for Spark datasets.

\section{Introduction}

Dependent types have been met with considerable interest from the research community in recent years.
Their primary application so far has been in proof assistants such as Agda~\citep{norell2007towards}, Coq~\citep{bertot2004interactive} and Idris~\citep{brady2013idris}, where they provide a sound and expressive foundation for theorem proving.
However, dependent types are still largely absent from general-purpose programming languages, despite a long history of lightweight approaches~\citep{xi1998eliminating}.
In the context of Haskell, much research has gone into extending the language to support computations on types, for instance in the form of functional dependencies~\citep{jones2000type}, type families~\citep{kiselyov2010fun} and promoted datatypes~\citep{yorgey2012giving}.
These techniques have seen adoption by Haskell programmers, showing that there is a real demand for such mechanisms.
Furthermore, recent research has explored how dependent types could be added to the language for the same purpose~\citep{eisenberg2016dependent, weirich2017a}.
In a largely orthogonal direction, inference for dependent refinement types is reaching significant maturity~\citep{vazou2018gradual, vazou2017refinement,vazou2015bounded}.

Dependently-typed languages often rely on a unified syntax to describe both terms and types.
The simplicity of this approach is unfortunately at odds with the design of most programming languages, where types and terms are expressed using separate syntactic categories.
Singleton types provide a simple solution to this problem by allowing every term to be represented as a type.
The singleton type of a term therefore gives us the most precise specification for that term.

In this paper, we report on our attempt at combining an industrial mixed-paradigm language, Scala, with dependent types.
We offer both a formalization of our type system and a discussion of the challenges faced in a practical implementation.
It is our hope that the present paper will serve as a guide for other language implementors interested in pushing the limit of their type systems by adding dependent types.

Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
Instead, our focus is on improving type safety by increasing the expressive power of the type system.

We present \oursystem, a dependently-typed calculus with subtyping and singleton types.
The main novelty of our calculus is a new approach to expressing type-level computation that, at first, seems diametrically opposed to the purity other systems favor.
A new term is added for non-deterministic choice from a base type, similar to Floyd's choice operator~\citep{floyd1967nondeterministic}.
Designing a sound system in the presence of non-determinism is challenging.
Our solution provides systematic translation of non-determinism using additional parameters that are existentially quantified at a syntactically well-defined point.
Consequently, a term in \oursystem may reduce to different values.
Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
As a result, our type system is capable of type computations by manipulating types which are based on terms, but can nonetheless contain more than a single value.
In combination with subtyping, this allows us to seamlessly integrate with impure, or imprecisely-typed programs.

Our contributions are as follows:

\begin{itemize}
\item
  We present our calculus \oursystem, which illustrates the novel elements of our extension to Scala.
  The type system of \oursystem combines dependent types, subtyping and a generalization of singleton types to non-deterministic terms.
  We demonstrate how the interplay of these features allows us to leverage term-level programs for type-level computation.
\item
  We provide a soundness proof of \oursystem by reusing reducibility semantics of \FR~\citep{hamza2019system}.
  Using its semantics we prove the soundness of our rules.
  These proofs are mechanized using the Coq proof assistant~\citep{bertot2004interactive}.
  The formalization is available in the additional materials.
\item
  We show a concrete use-case of our system by implementing it as an extension of Scala (\Cref{sec:extending-scala}), and using it to develop a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
  Thanks to dependent types, we can statically ensure the type safety of database operations such as join and filter.
  We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
\end{itemize}

\section{Motivating Examples}
\label{sec:motivating-example}

We begin by motivating why dependent types are desirable in general purpose programming, and how one might use them to improve type safety.
In our first example, we design an API that keeps track of database tables' schemas in the type.
We demonstrate how dependently-typed list operations can be used to compute schemas resulting from join operations at the type level.
Our second example shows how to build a safer version of the zip operation on lists that only accepts equally-sized arguments.
The examples in this section are written in our dependently-typed extension of Scala described in \Cref{sec:extending-scala}.

\subsection{Safe Join}

As a first step, we show how our system supports type-level programming in the style of term-level programs.
Consider the following definition of the list datatype, which is standard Scala up the |dependent| keyword:

\begin{lstlisting}
sealed trait Lst { (*\text{\dots}*) }
dependent case class Cons(head: Any, tail: Lst) extends Lst
dependent case class Nil() extends Lst
\end{lstlisting}

\noindent
We can define list concatenation in the usual functional style of Scala\footnote{Our examples use the indentation-based syntax introduced in Scala 3.0.}, that is, using pattern matching and recursion:

\begin{lstlisting}
sealed trait Lst:
  dependent def concat(that: Lst) <: Lst =
    this match
      case Cons(x, xs) => Cons(x, concat(xs, that))
      case Nil() => that
\end{lstlisting}

\noindent
By annotating a method as |dependent|, the user instructs our system that the result type of |concat| should be as precise as its implementation.
Effectively, this means that the body of |concat| is lifted to the type level, and will be partially evaluated at every call site to compute a precise result type which \emph{depends} on the given inputs.
For recursive |dependent| methods such as |concat|, we infer types that include calls to |concat| itself.
The |<:| annotation lets us provide an upper bound on |concat|'s result type, which will be used while type checking the method's definition.
Finally, by qualifying the definition of |Cons| and |Nil| as |dependent| we also allow their constructors and extractors to be lifted to the type level.
Using these definitions, we can now request the precise type whenever we manipulate lists by annotating the new |val| binding with |dependent|:

\begin{lstlisting}
dependent val l1 = Cons("A", Nil())
dependent val l2 = Cons("B", Nil())
dependent val l3 = l1.concat(l2)
l3.size: { 2 }
l3: { Cons("A", Cons("B", Nil())) }
\end{lstlisting}

Enclosing a pure term in braces (|{|~\ldots{}~|}|) denotes the singleton type of that term.
In the last two lines of this example we are therefore asking our system to prove that |l3| has size 2 and is equivalent to |Cons("A", Cons("B", Nil()))|.

In Scala we often deal with impure or imprecisely-typed code, however.
To integrate with such terms, we provide the |choose[T]| construct.
Operationally, we interpret |choose[T]| as a non-deterministic choice from |T|, which can be modeled faithfully on the type level as an existentially quantified inhabitant of |T| in a singleton type.
Thus, we equate |{ choose[T] }| to |T|, and when typing an impure term such as |Cons(readString(), Nil())| we can assign the precise type |{ Cons(choose[String], Nil()) }|.
Returning to the previous example, this means that even in the presence of impurity, we can perform useful type-level computation and checking:

\begin{lstlisting}
dependent val l2 = Cons(readString(), Nil())
dependent val l3 = l1.concat(l2)
l3: { Cons("A", Cons(choose[String], Nil())) }
\end{lstlisting}

In a style similar to |concat|, we can define remove on |Lst|:

\begin{lstlisting}
sealed trait Lst:
  dependent def remove(e: String) <: Lst =
    this match
      case Cons(head, tail) =>
        if (e == head) tail
        else Cons(head, tail.remove(e))
      case _ => throw new Error("element not found")
\end{lstlisting}

\noindent
Removing |"B"| yields the expected result, while trying to remove |"C"| from |l3| leads to a \emph{compilation error}, since the given program will provably fail at runtime.

\begin{lstlisting}
l3.remove("B"): { Cons("A", Nil()) }
l3.remove("C") // Error: element not found
\end{lstlisting}

The lists we defined so far can be used to implement a type-safe interface for database tables.

\begin{lstlisting}
dependent case class Table(schema: Lst, data: spark.DataFrame):
  dependent def join(right: Table, col: String) <: Table =
    val s1 = this.schema.remove(col)
    val s2 = right.schema.remove(col)
    val newSchema = Cons(col, s1.concat(s2))
    val newData = this.data.join(right.data, col)
    new Table(newSchema, newData)
\end{lstlisting}

\noindent
In this example, we wrap a weakly-typed implementation of Spark's |DataFrame| in the |dependent| class |Table|.
The first argument of this class represents the schema of the table as a precisely-typed list.
The second argument is the underlying |DataFrame|.
In the implementation of |join|, we execute the join operation on the underlying tables (|newData|) and compute the resulting schema corresponding to that join (|newSchema|).
By annotating the |join| method as |dependent|, the resulting schema is reflected in the type:

\begin{lstlisting}
dependent val schema1 = Cons("age", Cons("name", Nil()))
dependent val schema2 = Cons("name", Cons("unit", Nil()))
dependent val table1  = Table(schema1, (*\text{\dots}*))
dependent val table2  = Table(schema2, (*\text{\dots}*))
dependent val joined  = table1.join(table2, "name")
joined: { Table(Cons("name", Cons("age", Cons("unit", Nil()))), choose[DataFrame]) }
\end{lstlisting}

\noindent
Reflecting table schemas in types increases type safety over the existing weakly-typed interface.
For instance, it becomes possible to raise compile-time errors when a user tries to use non-existent columns.
This is an improvement over the underlying Spark implementation that would instead fail at runtime.

\subsection{Safe Zip}
\label{safe-zip}

Our first example demonstrated how dependent methods allow inference of precise types.
Conversely, we can also use singleton types to constrain method parameters further.
In this example, our goal is to write a safer wrapper for functions like |zip| that should only be applicable to lists of the same length.
To accomplish this, we can constrain the second parameter of |zip| as follows:

\begin{lstlisting}
def safeZip(xs: Lst, ys: { sizedLike(xs) }) = unsafeZip(xs, ys)
\end{lstlisting}

\noindent
Here we would like |{ sizedLike(xs) }| to be inhabited by all lists of equal length as |xs|, regardless of their elements'values.
How can this be achieved, given that |sizedLike(xs)| is a term? By exploiting the non-deterministic interpretation of |choose[T]|, we can provide a succinct definition for |sizedLike|:

\begin{lstlisting}
dependent def sizedLike(xs: Lst) <: Lst =
  xs match
    case Nil() => Nil()
    case Cons(x, ys) => Cons(choose[Any], sizedLike(ys))
\end{lstlisting}

\noindent
Consider, for instance, the meaning of |{ sizedLike(xs) }| for |xs = Cons(1, Cons(2, Nil()))|.
After reduction, we obtain |{ Cons(choose[Any], Cons(choose[Any], Nil())) }|, which is a type that represents all lists of size 2.
Thus |safeZip| requires that every caller prove that |xs| and |ys| are of the same length, which ensures that the underlying implementation in |unsafeZip| will never fail or truncate elements from one of the lists.

Note that, unlike |concat| and |remove| that can be used both on the term and the type level, |sizedLike| is here intended to be used as a type function, but not at runtime.

\section{Implementation}
\label{sec:extending-scala}

In this section we give an overview of how we extended Scala with dependent types.
This development was an experiment to explore the feasibility of adding dependent types in Scala.
We implemented our prototype as an extension of Dotty, the reference compiler for future versions of the Scala language.
Our presentation focuses on several facets of the implementation that are not reflected in our formalism.

On a syntactic level, our Scala extension consists of three additions:

\begin{itemize}
\item
  the singleton types syntax |{ t }|,
\item
  the |dependent| modifier for methods, values and classes,
\item
  the |choose[T]| construct.
\end{itemize}

\noindent
The newly-introduced singleton type syntax enables a subset of Scala expressions to be used in types.
This subset approximately corresponds to the core functional subset of Scala, plus the |choose[T]| construct, as illustrated in \oursystem.
Within this subset, the main differences between our formalism and implementation lie in the handling of pattern matching.

\subsection{Pattern Matching}
\label{pattern-matching}

Pattern matching in Scala supports a wide range of matching techniques~\citep{emir2007matching}.
For example, \emph{extractor patterns} rely on user-defined methods to extract values from objects.
As a result, these custom extractors can contain arbitrary side effects.
Our implementation limits the kind of patterns available in types to the two simplest forms: decomposition of case classes and the type-tests/type-casts patterns.

During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics, that is, patterns are checked top-to-bottom, and type-tests are evaluated using runtime type information available after type erasure.

For example, consider the following pattern matching expression:

\begin{lstlisting}
s match { case _: T1 => v1 case _: T2 => v2 }
\end{lstlisting}

\noindent
When used in a type, this expression reduces to |v1| if the scrutinee's type is a subtype of |T1|.
In order to reduce to |v2|, type normalization must make sure |T1| and the scrutinee's type are disjoint, namely that the dynamic type of |s| cannot possibly be smaller than |T1|.
Disjointness proofs are built using static knowledge about the class hierarchy and make use of the guarantees implied by the |sealed| and |final| qualifiers, which are Scala's way of declaring closed-type hierarchies.

\subsection{Two Modes of Type Inference}

In order to retain backwards-compatibility, our system supports two modes of type inference: the precise inference mode which infers singleton types, and the default inference mode that corresponds to Scala's current type-inference algorithm.
Concretely, users opt into our new inference mode using the |dependent| qualifier on methods, values, and classes.

When inferring the result type of a |dependent| method, our system lifts the method's body into a type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |choose[T]|.

For example, given the following definition:

\begin{lstlisting}
dependent def getName(personalized: Boolean) =
  if (personalized) readString() else "Joe"
\end{lstlisting}

\noindent
our system infers the following result type:

\begin{lstlisting}
{ if (personalized) choose[String] else "Joe" }
\end{lstlisting}

Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
However, in the case of a |dependent| method, an explicit result type is only used as an upper bound for the actual precise result type and will only be used to type-check the method's body.
At other call sites, the (precise) inferred result type is used.
Bounds of dependent methods are written using a special syntax (|<: T|), which emphasizes the difference from normal result types (|: T|).

\subsection{Approximating Side Effects}

\paragraph{State}

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Thanks to |choose[T]| we can still formulate precise result types when terms depend on the result of side-effectful operations.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, but we can still assign a better type than |Lst| to an expression like |Cons(z, Nil())|, that is, |{ Cons(choose[Int], Nil()) }|.

\paragraph{Exceptions}

Similarly to how we model other side effects, exceptions are approximated in types.
Our type-inference algorithm uses a new error type, |Error(e)|, which we infer when raising an exception with |throw e|.
Exception handlers are typed imprecisely using the default mode of type-inference.
Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{e1; e2}| is simply |{ e2 }|.
However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
In these cases, our type system can prove that the program execution will encounter exceptional behavior, and reports a compilation error.
This approach is conservative in that it might reject programs that recover from exceptions.
Also note that this is a sanity check, rather than a guarantee of no exceptions occurring at runtime.
That is, depending on which rules are used during subtyping, it is possible to succeed without entering type normalization, resulting in such errors going undetected.
Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
It would be interesting to explore the addition of an effect system to our Scala extension and formalization.

\subsection{Virtual Dispatch}

Our extension does not model virtual dispatch explicitly in singleton types.
Instead, the result type of a method call |t.m(|\ldots{}|)| is always the result type of |m| in |t|'s static type.
Consequently, |dependent| methods effectively become |final|, given that only a provably-equivalent implementation could be used to override it.

Special care must be taken when an imprecisely-typed method is overridden with a |dependent| one.
In this situation, the result type of a method invocation can lose precision depending on type of the receiver.
Calls to the |equals| methods are a common example of this: |equals| is defined at the top of Scala's type hierarchy as referential equality and can be overridden arbitrarily.
Given a class Foo with a |dependent| overrides of |equals|, calls to |Foo.equals(Any)| and |Any.equals(Foo)| are not equivalent; the former precisely reflects the equality defined in |Foo| whereas the latter merely returns a |Boolean|.

\subsection{Termination}

We distinguish two important aspects of termination.

The first question is whether type-checked programs are guaranteed to terminate.
For simplicity, our work side-steps this question, requiring bounds for recursion.
A more general solution would be to compute or infer such bounds using measure functions, as done in \FR~\citep{hamza2019system}.
Another approach would be to extend our translation of non-determinism to permit non-termination.
We consider this aspect orthogonal to the objectives of this paper.
Our work targets general-purpose programming language whose type safety is defined with regards to its runtime semantics and that may include non-terminating interactive computations.

The second question is termination of our type checker.
Non-termination of type checking implies that the type checker can give three possible answers, "type correct", "type incorrect" or "do not know" (or timeout).
Treating "do not know" as "type incorrect" makes the non-termination unproblematic from a soundness perspective.
A similar argument is made for other dependently-typed languages with unbounded recursion, such as Dependent Haskell~\citep{eisenberg2016dependent} or Cayenne~\citep{augustsson1998cayenne}.
In practice, our system deals with infinite loops using a fuel mechanism.
Every evaluation step consumes a unit of fuel, and an error is reported when the compiler runs out of fuel.
The default fuel limit can be increased via a compiler flag to enable arbitrarily long compilation times.

\section{Use Case}
\label{sec:use-case}

In this section, we extend the motivating example presented in \Cref{sec:motivating-example} by building a type-safe interface for Spark datasets.
We use dependent types to implement a simple domain-specific type checker for the SQL-like expressions used in Spark.
We then compare the compilation time of our dependently-typed interface against an equivalent encoding based on implicits.

\subsection{A Type-Safe Database Interface}

The type-safe interface presented in this section illustrates the expressive power of our system and is implemented purely as a library.
For brevity, our presentation only covers a small part of Spark's dataset interface, but the approach can be scaled to cover that interface in its entirety.
The type safety of database queries is a canonical example and has been studied in many different settings~\citep{leijen1999domain, kazerounian2019type, meijer2006linq, chlipala2010ur}.

The example built in \Cref{sec:motivating-example} uses lists of column names to represent schemas.
A straightforward improvement is to also track the type of columns as part of the schema.
Instead of using column names directly, we introduce the following |Column| class with a phantom type parameter |T| for the column type, and a field |name| for the column name:

\begin{lstlisting}
dependent case class Column[T](name: String) { (*\text{\dots}*) }
\end{lstlisting}

Table schemas become lists of |Column|-s and thereby gain precision.
The definition of |join| given in \Cref{sec:motivating-example} can be adapted to this new schema encoding to prevent joining two tables that have columns with matching names but different types.

A large proportion of the weakly-typed Spark interface is dedicated to building expressions on table columns.
Such expressions can currently be built from strings, in a subset of SQL, or using a Scala DSL which is essentially untyped.

The lack of type safety for column expressions can be particularly dangerous when mixing columns of different types.
The pitfall is caused by Spark's inconsistency: depending on types of columns and operations involved, programs will either crash at runtime, or, more dangerously, data will be silently converted from one type to another.

By keeping track of column types it becomes possible to enforce the well-typedness of column expressions.
As an example, consider the following Spark program:

\begin{lstlisting}
table.filter(table.col("a") + table.col("b") === table.col("c"))
\end{lstlisting}

We would like our interface to enforce the following safety properties:

\begin{itemize}
\item
  Columns $a$, $b$ and $c$ are part of the schema of
  |table|.
\item
  Addition is well-defined on columns $a$ and $b$.
\item
  The result of adding columns $a$ and $b$ can be compared with
  column $c$.
\item
  The overall column expression yields a |Boolean|, which
  conforms to filter's argument type.
\end{itemize}

Automatic conversions during equality checks can be prevented by restricting column equality to expressions of the same type |T|:

\begin{lstlisting}
dependent case class Column[T](k: String):
  def ===(that: Column[T]): Column[Boolean] = Column(s"(${this.k} === ${that.k})")
\end{lstlisting}

Addition in Spark is defined between numeric types and characters.
The result type of an addition depends on the operand types.
For numeric types, Spark will pick the larger of the operand types according to the following ordering: |Double > Long > Int > Byte|.
The situation is quite surprising with characters as any addition involving a |Char| will result in a |Double|.

Dependent types can be used to precisely model these conversions.
We define a type function to compute the result type of additions:

\begin{lstlisting}
def addRes(a: Any, b: Any) =
  (a, b) match
    case (_: Char, _: Char | Byte | Int  | Long | Double) => choose[Double]
    case (_: Byte, _: Byte | Int  | Long | Double)        => b
    case (_: Int,  _: Int  | Long | Double)               => b
    case (_: Long, _: Long | Double)                      => b
    case (_: Double, _: Double)                           => choose[Double]
    case (_: Byte | Int | Long | Double, _)               => addRes(b, a)
    case _ => throw new Error("incompatible types in addition")
type AddRes[A, B] = { addRes(choose[A], choose[B]) }
\end{lstlisting}

Also note the use of recursion in the second-to-last case, to avoid duplicating symmetric cases.
The |AddRes| type can be used to define a |Column| addition that accurately models Spark's runtime:

\begin{lstlisting}
dependent case class Column[T] private (k: String):
  dependent def +[U](that: Column[U]) <: Column[_] =
    Column[AddRes[T, U]](s"(${this.k} + ${that.k})")
\end{lstlisting}

Allowing programmers to construct |Column|-s from string literals would defeat the purpose of a type-safe interface.
Instead, programmers should extract columns from a |Table|'s schema.
For that purpose, we implement the |col| method on |Table| and annotate the |Column| constructor as private.

\begin{lstlisting}
dependent case class Table(schema: Lst, data: spark.DataFrame):
  dependent def col(name: String) <: Column[_] =
    dependent def find(key: String, list: Lst) <: Any =
      list match
        case Cons(head: Column[_], tail) =>
          if (head.k == key) head else find(key, tail)
        case _ => throw new Error("column not found in schema")
    find(name, schema)
  dependent def filter(predicate: Column[Boolean]) <: Table =
    new Table(this.schema, this.data.filter(predicate.k))
\end{lstlisting}

The |col| method is implemented using a nested dependent method to find the column corresponding to the given name.
Thanks to the dependent annotation, the type-checker is able to statically evaluate calls to |col|.
Assuming the table's schema contains a column |a| of type |Int| and columns |b| and |c| of type |Long|, the compiler will be able to infer types as follows:

\begin{lstlisting}
val pred = (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("a") \phantom{.} }*) + (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("b") \phantom{.}\phantom{.} }*) === (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("c") \phantom{.}\phantom{.} }*)
// Infers: (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Int]("a") \}}}*)   (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("b") \}}}*)     (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("c") \}}}*)
\end{lstlisting}

\noindent
Given our definitions of column addition and equality, the overall |pred| expression is typed as |Column[Boolean]|.
All the safety properties stated above are therefore enforced by the dependently-typed interface presented in this section.

\subsection{Comparison to an Existing Technique}

% \measurements{Comparing the compilation times of two implementations of list concatenation and join, log.\ scale.}
% TODO OLIVIER

Programmers have managed to find clever encodings that circumvent the lack of first-class support for type-level programming in many languages.
These encodings can be very cumbersome, as they often entail poor error reporting and a negative impact on compilation times~\citep{mcbride2002faking},~\citep{kiselyov2004strongly}.
In Scala, implicits are the primary mechanism by which programmers implement type-level programming~\citep{odersky2018simplicitly}.

Frameless~\citep{blanvillain2016frameless} is a Scala library that implements a type-safe interface for Spark by making heavy use of implicits.
Most type-level computations in this library are performed on the heterogeneous lists provided by Shapeless~\citep{sabin2011shapeless}.

We compared the dependently-typed Spark interface presented in this section against the implicit-based implementation of Frameless.
To do so, we isolated the implicit-based implementation of the |join| operation on table schemas, and compared its compilation time against the dependently-typed version presented in this section.
To evaluate the scalability of both approaches we generated test cases with varying schema sizes and compiled each test case in isolation.
A similar comparison is done for list concatenation, which constitutes a building block of |join|.

% \Cref{measurements} shows that, in both benchmarks, the dependently-typed implementation compiles faster than the version with implicits, and compilation time scales better with the size of the input.
% TODO OLIVIER
In the join benchmark, we see that the implicit-based implementation exceeds 30 seconds of compilation time around the 200 columns mark, and continues to grow quadratically.
This can be explained by the nature of implicit resolution, which might backtrack during its search.
The compilation time of the dependently-typed implementation grows linearly and stays below one second until the 350 columns mark.
We were able to observe similar trends in the concatenation benchmark.
These measurements were obtained by averaging 120 compilations on a warm compiler, and have been performed on an i7-7700K Processor running Oracle JVM 1.8.0 on Linux.

\section{Related Work}

As of today, Haskell is perhaps closest to becoming dependently-typed among the general-purpose programming languages used in industry.
Haskell's type families~\citep{kiselyov2010fun} provide a direct way to express type-level computations.
Other language extensions such as functional dependencies~\citep{jones2000type} and promoted datatypes~\citep{yorgey2012giving} are also moving Haskell towards dependent types.
Nevertheless, programming in Haskell remains significantly different from using full-spectrum dependently-typed languages.
A significant difference is that Haskell imposes a strict separation between terms and types.
As a result, writing dependently-typed programs in Haskell often involves code duplication between types and terms.
These redundancies can be somewhat avoided using the singletons package~\citep{eisenberg2012dependently}, which uses meta-programming to automatically generate types from datatypes and function definitions.

In the context of Haskell, Eisenberg's work on Dependent Haskell~\citep{eisenberg2016dependent} is closest to ours, in that it adds first-class support for dependent types to an established language, in a backwards-compatible way.
Dependent Haskell supports general recursion without termination checks, which makes it less suitable for theorem proving.
While we share similar goals, our work is differentiated by the contrasting paradigms of Scala and Haskell.
Like many object-oriented languages, Scala is primarily built around subtyping and does not restrict the use of side effects.
Furthermore, Eisenberg's system provides control over the relevance of values and type parameters.
In contrast, our system does not support any erasure annotations and simply follows Scala's canonical erasure strategy: types are systematically erased to JVM types, and terms are left untouched.
Weirich established a fully mechanized type safety proof for the core of Dependent Haskell using the Coq proof assistant~\citep{weirich2017a}.

Cayenne is a Haskell-like language with dependent types introduced in 1998 by Augustsson~\citep{augustsson1998cayenne}.
Like Dependent Haskell, it resembles our system in its treatment of termination, and differs by being a purely functional programming language.
Cayenne's treatment of erasure is similar to Scala's: types are systematically erased.
Augustsson proves that Cayenne's erasure is semantics-preserving, but does not provide any other metatheoretical results.

Adding dependent types to object-oriented languages is a remarkably under-explored area of research.
A notable exception is the recent work of Kazerounian et al.~on adding dependent types to Ruby~\citep{kazerounian2019type}.
Their goals are very much aligned with ours: using type-level programming to increase program safety.
Given the extremely dynamic nature of Ruby, it is unsurprising that their solution greatly differs from ours.
In their work, type checking happens entirely at runtime and has to be performed at every function invocation to account for possible changes in function definitions.
Safety is obtained by inserting dynamic checks, similarly to gradual typing.

The work of Campos and Vasconcelos on DOL (Dependent Object-oriented Language)~\citep{campos2018dependent} shares similar goals but is limited to inequality constraints on integer parameters (in the style of~\citep{xi1998eliminating}).

Dependently-typed lambda calculi with subtyping were described at least as far back as 1988 by Cardelli~\citep{cardelli1988structural}.
His type system is much more expressive than ours and allows bounded quantification over both types and terms using the notion of a |Type| type and power types.
Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide semantics for his system and leaves the equivalence relation among types unspecified.

In~\citep{aspinall1994singleton} Aspinall introduces ${\lambda}_{\leq}{}$, a dependently-typed system with subtyping \emph{and} singleton types that resembles ours in its type language.
His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
Aspinall's system also has primitive types and allows for atomic subtyping among them, but no congruence rules, hence partially-widened forms cannot be represented.

System ${\lambda}P_{\leq}$~\citep{aspinall1996subtyping} combines subtyping and dependent types in the Edinburgh Logical Framework.
In this work, Aspinall et al.~propose a type-checking algorithm for ${\lambda}P_{\leq}$ which they show to be complete and terminating.
Their system uses a kinding relation to ensure well-formedness of type applications.
A kind system is not required in \oursystem as we emulate type applications inside singleton types.

In~\citep{stone2000deciding}, Stone and Harper describe a dependently-typed calculus with singleton kinds and subkinding.
Their type-and-kind system is similar to Aspinall's ${\lambda}_{\leq}{}$ term-and-type system, but operates one level up the hierarchy.

More recently, Courant~\citep{courant2003strong} developed a variant of Aspinall's ${\lambda}_{\leq}{}$ with a type-inference algorithm that he proves to be sound and complete.
The main takeaway from Courant's work is the inclusion of a coercion rule in delta reduction.
These coercions are used to "tag" variables with their declared type, which prevents these types from being lost during substitution.
Our formalism resembled Courant's system, it shares the \textsc{SubSing} subtyping rule (SUB/SINGL in Courant's work), and $\beta\delta$-reduction.

Pure Type Systems~\citep{barendregt1991introduction} provide a unified presentation of systems of dependently-typed $\lambda$-calculus by using a single syntactic category for both terms and types.

In~\citep{zwanenburg1999pure}, Zwanenburg defines an extension of pure type systems that include both subtyping and bounded quantification.
A central design decision of his system is that subtyping rules do not depend on typing rules.
The absence of circularity simplifies both the theory and the metatheory, at the cost of having to define subtyping on pseudoterms rather than only well-typed terms.
Another limitation of Zwanenburg's theory is that it cannot be extended with a Top-type.

Pure Subtype Systems~\citep{hutchins2010pure} is another framework with unified syntax; it differs from traditional approaches in that it uses a single relation, subtyping, that subsumes typing, subtyping, and type evaluation as found in our system.
Their system allows for partially-widened types similar to ours and also enables the computation with different levels of precision.
For instance, it can conclude that Int + 5 can be approximated as Int.
The paper presents a partial investigation of the metatheory, but the proof of soundness remains incomplete.
Nevertheless, Hutchins reports that he has not been able to construct a counter-example, even with the addition of fixpoints.

In~\citep{yang2017unifying}, Yang and Oliveira propose a dependently-typed generalization of System $F_{\leq}$ with unified syntax and a single relation that subsumes typing and subtyping.
In their system, type computations are driven by cast operators: each reduction or expansion step requires an annotation to explicitly instruct the type checker to take a step.
Explicit casts make it possible to allow general recursion without compromising decidability of type checking.
It would be interesting to study variants of \oursystem based on explicit casts instead of our finitized fix.

Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
Even though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their original goals largely orthogonal to ours.

### 2021-11-30 preprint.tex
Fix a few bad boxes

@@
 Dependent types have been met with considerable interest from the research community in recent years.
-Their primary application so far has been in proof assistants such as Agda~\citep{norell2007towards}, Coq~\citep{bertot2004interactive} and Idris~\citep{brady2013idris}, where they provide a sound and expressive foundation for theorem proving.
+Their primary application so far has been in proof assistants such as Coq~\citep{bertot2004interactive} and Agda~\citep{norell2007towards}, where they provide a sound and expressive foundation for theorem proving.
 However, dependent types are still largely absent from general-purpose programming languages, despite a long history of lightweight approaches~\citep{xi1998eliminating}.
@@
 \item
-  We provide a soundness proof of \oursystem by reusing reducibility semantics of \FR~\citep{hamza2019system}.
+  We provide a soundness proof of \oursystem by reusing the reducibility semantics of \FR \citep{hamza2019system}.
   Using its semantics we prove the soundness of our rules.
@@
 \noindent
-In this example, we wrap a weakly-typed implementation of Spark's |DataFrame| in the |dependent| class |Table|.
+In this example, we wrap a Spark's |DataFrame| in the |dependent| class |Table|.
 The first argument of this class represents the schema of the table as a precisely-typed list.
@@
 
-Special care must be taken when an imprecisely-typed method is overridden with a |dependent| one.
+Special care must be taken when an imprecisely-typed method is overridden with a dependent one.
 In this situation, the result type of a method invocation can lose precision depending on type of the receiver.
@@
 Given our definitions of column addition and equality, the overall |pred| expression is typed as |Column[Boolean]|.
-All the safety properties stated above are therefore enforced by the dependently-typed interface presented in this section.
+Thus, the dependently-typed interface presented in this section successfully enforces all the safety properties stated above.
 
@@
 Adding dependent types to object-oriented languages is a remarkably under-explored area of research.
-A notable exception is the recent work of Kazerounian et al.~on adding dependent types to Ruby~\citep{kazerounian2019type}.
+A notable exception is the recent work of~\cite{kazerounian2019type} on adding dependent types to Ruby.
 Their goals are very much aligned with ours: using type-level programming to increase program safety.
@@
 Safety is obtained by inserting dynamic checks, similarly to gradual typing.
-
-The work of Campos and Vasconcelos on DOL (Dependent Object-oriented Language)~\citep{campos2018dependent} shares similar goals but is limited to inequality constraints on integer parameters (in the style of~\citep{xi1998eliminating}).
-
-Dependently-typed lambda calculi with subtyping were described at least as far back as 1988 by Cardelli~\citep{cardelli1988structural}.
-His type system is much more expressive than ours and allows bounded quantification over both types and terms using the notion of a |Type| type and power types.
-Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide semantics for his system and leaves the equivalence relation among types unspecified.
-
-In~\citep{aspinall1994singleton} Aspinall introduces ${\lambda}_{\leq}{}$, a dependently-typed system with subtyping \emph{and} singleton types that resembles ours in its type language.
-His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
-Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
-Aspinall's system also has primitive types and allows for atomic subtyping among them, but no congruence rules, hence partially-widened forms cannot be represented.
-
-System ${\lambda}P_{\leq}$~\citep{aspinall1996subtyping} combines subtyping and dependent types in the Edinburgh Logical Framework.
-In this work, Aspinall et al.~propose a type-checking algorithm for ${\lambda}P_{\leq}$ which they show to be complete and terminating.
-Their system uses a kinding relation to ensure well-formedness of type applications.
-A kind system is not required in \oursystem as we emulate type applications inside singleton types.
-
-In~\citep{stone2000deciding}, Stone and Harper describe a dependently-typed calculus with singleton kinds and subkinding.
-Their type-and-kind system is similar to Aspinall's ${\lambda}_{\leq}{}$ term-and-type system, but operates one level up the hierarchy.
-
-More recently, Courant~\citep{courant2003strong} developed a variant of Aspinall's ${\lambda}_{\leq}{}$ with a type-inference algorithm that he proves to be sound and complete.
-The main takeaway from Courant's work is the inclusion of a coercion rule in delta reduction.
-These coercions are used to "tag" variables with their declared type, which prevents these types from being lost during substitution.
-Our formalism resembled Courant's system, it shares the \textsc{SubSing} subtyping rule (SUB/SINGL in Courant's work), and $\beta\delta$-reduction.
-
-Pure Type Systems~\citep{barendregt1991introduction} provide a unified presentation of systems of dependently-typed $\lambda$-calculus by using a single syntactic category for both terms and types.
-
-In~\citep{zwanenburg1999pure}, Zwanenburg defines an extension of pure type systems that include both subtyping and bounded quantification.
-A central design decision of his system is that subtyping rules do not depend on typing rules.
-The absence of circularity simplifies both the theory and the metatheory, at the cost of having to define subtyping on pseudoterms rather than only well-typed terms.
-Another limitation of Zwanenburg's theory is that it cannot be extended with a Top-type.
-
-Pure Subtype Systems~\citep{hutchins2010pure} is another framework with unified syntax; it differs from traditional approaches in that it uses a single relation, subtyping, that subsumes typing, subtyping, and type evaluation as found in our system.
-Their system allows for partially-widened types similar to ours and also enables the computation with different levels of precision.
-For instance, it can conclude that Int + 5 can be approximated as Int.
-The paper presents a partial investigation of the metatheory, but the proof of soundness remains incomplete.
-Nevertheless, Hutchins reports that he has not been able to construct a counter-example, even with the addition of fixpoints.
-
-In~\citep{yang2017unifying}, Yang and Oliveira propose a dependently-typed generalization of System $F_{\leq}$ with unified syntax and a single relation that subsumes typing and subtyping.
-In their system, type computations are driven by cast operators: each reduction or expansion step requires an annotation to explicitly instruct the type checker to take a step.
-Explicit casts make it possible to allow general recursion without compromising decidability of type checking.
-It would be interesting to study variants of \oursystem based on explicit casts instead of our finitized fix.
-
-Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
-Even though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their original goals largely orthogonal to ours.

### 2021-12-01 preprint.tex
Add preprint figure

@@
-\newcommand{\oursystem}{$\lambda${\tiny $\mathop{\protect\vphantom{X}}^\text{nd}_{<:\{\}}$}\xspace}
-\newcommand\FR{System~FR\xspace}
-\newcommand\singleton[1]{\lbrace #1 \rbrace}
-
 Type-level programming is an increasingly popular way to obtain additional type safety.
@@
 \begin{lstlisting}
-val pred = (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("a") \phantom{.} }*) + (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("b") \phantom{.}\phantom{.} }*) === (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("c") \phantom{.}\phantom{.} }*)
-// Infers: (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Int]("a") \}}}*)   (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("b") \}}}*)     (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("c") \}}}*)
+val pred =    table.col("a")    +    table.col("b")    ===   table.col("c")
+// Infers: { Column[Int]("a") }   { Column[Long]("b") }   { Column[Long]("c") }
 \end{lstlisting}
+% val pred = (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("a") \phantom{.} }*) + (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("b") \phantom{.}\phantom{.} }*) === (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("c") \phantom{.}\phantom{.} }*)
+% // Infers: (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Int]("a") \}}}*)   (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("b") \}}}*)     (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("c") \}}}*)
 
@@
 
-% \measurements{Comparing the compilation times of two implementations of list concatenation and join, log.\ scale.}
-% TODO OLIVIER
+\dependentVsImplicitBenchmarks{Comparing the compilation times of two implementations of list concatenation and join, logarithmic scale.}
 
@@
 
-% \Cref{measurements} shows that, in both benchmarks, the dependently-typed implementation compiles faster than the version with implicits, and compilation time scales better with the size of the input.
-% TODO OLIVIER
+\Cref{fig:dependentVsImplicitBenchmarks} shows that, in both benchmarks, the dependently-typed implementation compiles faster than the version with implicits, and compilation time scales better with the size of the input.
+
 In the join benchmark, we see that the implicit-based implementation exceeds 30 seconds of compilation time around the 200 columns mark, and continues to grow quadratically.

### 2021-12-08 scala/src/main/scala/numpy-match-types.scala
Rename Reduce to ReduceAxes

@@ -47,7 +47,7 @@ def reshape[T, From <: Shape, To <: Shape](arr: NDArray[T, From], newshape: To)
 
 import scala.compiletime.ops.int.+
 // start section reduceType
-type Reduce[S <: Shape, Axes <: None | Shape] <: Shape =
+type ReduceAxes[S <: Shape, Axes <: None | Shape] <: Shape =
   Axes match {
     case None => Ø
     case Shape => Loop[S, Axes, 0]
@@ -85,5 +85,5 @@ type Remove[From <: Shape, Value <: Int & Singleton] <: Shape = From match {
 }
 
 // start section npmeanDef
-def mean[T, S <: Shape, A <: Shape](arr: NDArray[T, S], axes: A): NDArray[T, Reduce[S, A]] = ???
+def mean[T, S <: Shape, A <: Shape](arr: NDArray[T, S], axes: A): NDArray[T, ReduceAxes[S, A]] = ???
 // end section npmeanDef

### 2021-12-17 bench.sh
Setup gnuplot

@@ -1,16 +1,15 @@
 #!/bin/sh
 
 set -eu
-
-BASEDIR=$(realpath "$(dirname "$0")")
+test "$(pwd)" = "$(realpath "$(dirname "$0")")" || (echo "wrong dir"; exit 1)
 
 # Generate sources:
-# rm -rf "$BASEDIR/scala/target/generated/"
-# mkdir -p "$BASEDIR/scala/target/generated/"
+# rm -rf "scala/target/generated/"
+# mkdir -p "scala/target/generated/"
 # git grep --name-only //X" | grep scala | while read s; do
-#   cp "$s" "$BASEDIR/scala/target/generated/"
+#   cp "$s" "scala/target/generated/"
 # done
-# find "$BASEDIR/scala/target/generated/" -name "*.scala" | while read s; do
+# find "scala/target/generated/" -name "*.scala" | while read s; do
 #   seq -w 1 8 256 | while read i; do
 #     replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
 #     echo "$s.$i.scala"
@@ -22,7 +21,7 @@ BASEDIR=$(realpath "$(dirname "$0")")
 # Mesure bytecode size:
 # rm -f bytecode-size.log
 # rm -f *.class *.tasty
-# find "$BASEDIR/scala/target/generated/" -name "*.scala" | sort | while read s; do
+# find "scala/target/generated/" -name "*.scala" | sort | while read s; do
 #   if echo "$s" | grep -q "dependent"; then
 #     /usr/bin/time -f "%e" -o "/tmp/time" $HOME/workspace/thesis/bin/dotc -J-Xss4m "$s"
 #   else
@@ -33,10 +32,21 @@ BASEDIR=$(realpath "$(dirname "$0")")
 # done
 
 # Mesure compilation times:
-find "$BASEDIR/scala/target/generated/" -name "*.scala" | sort | while read s; do
-  if echo "$s" | grep -q "dependent"; then
-    (cd "$BASEDIR/ddotty/" && sbt "dotty-bench/jmh:run 60 60 1 $s" | tee "$s.log")
-  else
-    (cd "$HOME/workspace/dotty-master/" && sbt "scala3-bench/jmh:run 60 60 1 $s" | tee "$s.log")
-  fi
[…]

### 2021-12-17 process.sh

@@ -1,0 +1,25 @@
+#!/bin/sh
+
+set -eux
+
+BASEDIR=$(dirname "$0")
+
+rm "$BASEDIR/"*.csv || true
+
+cat "$BASEDIR"/*.log         |\
+grep -E -A4 "Running|Result" |\
+grep -E "±|scala"            |\
+paste -s -d' \n'             |\
+while read res; do
+  # [info] Running (fork) dotty.tools.benchmarks.Bench 40 10 1 /path/to/src-gen/dependent-remove-140.scala   295.236 ±(99.9%) 17.036 ms/op [Average]
+  file=$(echo "$res" | grep -oP '(?<=src-gen/).+(?=.scala)')
+  avg=$(echo "$res" | grep -oP '(?<=   )\d+.\d+(?=.*99.9)')
+  err=$(echo "$res" | grep -oP '(?<= )\d+.\d+(?= ms/op)')
+  bench=$(echo "$file" | grep -oP '.+(?=-\d+)')
+  size=$(echo "$file" | grep -oP '\d+')
+  echo "$size,$avg,$err" >> "$BASEDIR/$bench.csv"
+done
+
+ls "$BASEDIR/"*.csv | while read f; do
+  sort -n "$f" -o "$f"
+done

### 2022-01-12 bench.sh
Update graphs

@@ -4,19 +4,19 @@ set -eu
 test "$(pwd)" = "$(realpath "$(dirname "$0")")" || (echo "wrong dir"; exit 1)
 
 # Generate sources:
-# rm -rf "scala/target/generated/"
-# mkdir -p "scala/target/generated/"
-# git grep --name-only //X" | grep scala | while read s; do
-#   cp "$s" "scala/target/generated/"
-# done
-# find "scala/target/generated/" -name "*.scala" | while read s; do
-#   seq -w 1 8 256 | while read i; do
-#     replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
-#     echo "$s.$i.scala"
-#     cat $s | sed "s#0 :: //X#$replacement#g" > "$s.$i.scala"
-#   done
-#   rm $s
-# done
+rm -rf "scala/target/generated/"
+mkdir -p "scala/target/generated/"
+git grep --name-only "//X" | grep scala | while read s; do
+  cp "$s" "scala/target/generated/"
+done
+find "scala/target/generated/" -name "*.scala" | while read s; do
+  seq -w 1 8 256 | while read i; do
+    replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
+    echo "$s.$i.scala"
+    cat $s | sed "s#0 :: //X#$replacement#g" > "$s.$i.scala"
+  done
+  rm $s
+done
 
 # Mesure bytecode size:
 # rm -f bytecode-size.log
@@ -41,12 +41,12 @@ test "$(pwd)" = "$(realpath "$(dirname "$0")")" || (echo "wrong dir"; exit 1)
 # done
 
 # Generate CSV:
-rm -rf figures/*.csv
-find "scala/target/generated/" -name "*.log" | sort | while read l; do
-  r=$(grep "±(99.9%)" "$l")
-  avg=$(echo "$r" | grep -oP '\d+.\d+(?=.*99.9)')
-  err=$(echo "$r" | grep -oP '(?<= )\d+.\d+(?= ms/op)')
-  bench=$(echo "$l" | grep -oP '(?<=generated/)[^.]+(?=.scala)')
-  size=$(echo "$l" | grep -oP '\d+')
-  echo "$size,$avg,$err" >> "figures/$bench.csv"
-done
+# rm -rf figures/*.csv
[…]

### 2022-01-12 scala/src/main/scala/numpy-match-types.scala

@@ -95,8 +95,8 @@ object Bench {
     type ::[A <: Int, B <: Shape] = #:[A, B]
 
     type A = (
-      0 :: //X
-      Ø
+      0 #: 1 #: 2 #: 3 #: 4 #: 5 #: 6 #: 7 #: 8 #: 9 #: 10 #: 11 #: 12 #: 13 #: 14 #: 15 #: 16 #: 17 #: 18 #: 19 #: 20 #: 21 #: 22 #: 23 #: 24 #: 25 #: 26 #: 27 #: 28 #: 29 #: 30 #: 31 #: 32 #: 33 #: 34 #: 35 #: 36 #: 37 #: 38 #: 39 #: 40 #: 41 #: 42 #: 43 #: 44 #: 45 #: 46 #: 47 #: 48 #: 49 #: 50 #: 51 #: 52 #: 53 #: 54 #: 55 #: 56 #: 57 #: 58 #: 59 #: 60 #: 61 #: 62 #: 63 #: 64 #: 65 #: 66 #: 67 #: 68 #: 69 #: 70 #: 71 #: 72 #: 73 #: 74 #: 75 #: 76 #: 77 #: 78 #: 79 #: 80 #: 81 #: 82 #: 83 #: 84 #: 85 #: 86 #: 87 #: 88 #: 89 #: 90 #: 91 #: 92 #: 93 #: 94 #: 95 #: 96 #: 97 #: 98 #: 99 #: 100 #: 101 #: 102 #: 103 #: 104 #: 105 #: 106 #: 107 #: 108 #: 109 #: 110 #: 111 #: 112 #: 113 #: 114 #: 115 #: 116 #: 117 #: 118 #: 119 #: 120 #: 121 #: 122 #: 123 #: 124 #: 125 #: 126 #: 127 #: 128 #: 129 #: 130 #: 131 #: 132 #: 133 #: 134 #: 135 #: 136 #: 137 #: 138 #: 139 #: 140 #: 141 #: 142 #: 143 #: 144 #: 145 #: 146 #: 147 #: 148 #: 149 #: 150 #: Ø
+      // 0 #: 1 #: 2 #: 3 #: 4 #: 5 #: 6 #: 7 #: 8 #: 9 #: 10 #: Ø
     )
 
     type Reduced = ReduceAxes[A, A]

### 2022-01-14 scala/src/main/scala/numpy-implicits.scala
Fix scala example compilation

@@ -43,26 +43,24 @@ object NumpyImplicits {
 
   trait Loop[S <: Shape, Axes <: Shape, I <: Int] { type Out <: Shape }
 
-  trait LoopLowPrio {
-    val instance: Nothing = (new Loop {}).asInstanceOf
-
+  object Loop {
     type Aux[S <: Shape, Axes <: Shape, I <: Int, O <: Shape] =
       Loop[S, Axes, I] { type Out = O }
 
+    val instance: Nothing = (new Loop {}).asInstanceOf
+
     implicit def caseNil[I <: Int]: Aux[Ø, Ø, I, Ø] = instance
 
-    implicit def caseConsFalse[Sh <: Int, St <: Shape, Axes <: Shape, I <: Int, RemoveOut <: Shape, Out <: Shape]
-      (implicit
-        lo: Aux[St, Axes, Succ[I], Out]
-      ): Aux[Sh :: St, Axes, I, Sh :: Out] = instance
-  }
-
-  object Loop extends LoopLowPrio {
     implicit def caseConsTrue[Sh <: Int, St <: Shape, Axes <: Shape, I <: Int, RemoveOut <: Shape, Out <: Shape]
       (implicit
         rm: Remove.Aux[Axes, I, RemoveOut],
         lo: Aux[St, RemoveOut, Succ[I], Out]
       ): Aux[Sh :: St, Axes, I, Out] = instance
+
+    implicit def caseConsFalse[Sh <: Int, St <: Shape, Axes <: Shape, I <: Int, Out <: Shape]
+      (implicit
+        lo: Aux[St, Axes, Succ[I], Out]
+      ): Aux[Sh :: St, Axes, I, Sh :: Out] = instance
   }
 
   object Bench {

### 2022-01-14 scala/src/main/scala/numpy-match-types.scala

@@ -1,7 +1,7 @@
 object NumpyMatchTypes {
 
 sealed trait Shape
-final case class #:[+H <: Int, +T <: Shape](head: H, tail: T) extends Shape
+final case class #:[+H <: Int & Singleton, +T <: Shape](head: H, tail: T) extends Shape
 case object Ø extends Shape
 
 // cheating a little bit here, but there's a footnote to explain it!
@@ -92,11 +92,11 @@ def mean[T, S <: Shape, A <: Shape](arr: NDArray[T, S], axes: A): NDArray[T, Red
 
 object Bench {
   def main(args: Array[String]): Unit = {
-    type ::[A <: Int, B <: Shape] = #:[A, B]
+    type ::[A <: Int & Singleton, B <: Shape] = #:[A, B]
 
     type A = (
-      0 #: 1 #: 2 #: 3 #: 4 #: 5 #: 6 #: 7 #: 8 #: 9 #: 10 #: 11 #: 12 #: 13 #: 14 #: 15 #: 16 #: 17 #: 18 #: 19 #: 20 #: 21 #: 22 #: 23 #: 24 #: 25 #: 26 #: 27 #: 28 #: 29 #: 30 #: 31 #: 32 #: 33 #: 34 #: 35 #: 36 #: 37 #: 38 #: 39 #: 40 #: 41 #: 42 #: 43 #: 44 #: 45 #: 46 #: 47 #: 48 #: 49 #: 50 #: 51 #: 52 #: 53 #: 54 #: 55 #: 56 #: 57 #: 58 #: 59 #: 60 #: 61 #: 62 #: 63 #: 64 #: 65 #: 66 #: 67 #: 68 #: 69 #: 70 #: 71 #: 72 #: 73 #: 74 #: 75 #: 76 #: 77 #: 78 #: 79 #: 80 #: 81 #: 82 #: 83 #: 84 #: 85 #: 86 #: 87 #: 88 #: 89 #: 90 #: 91 #: 92 #: 93 #: 94 #: 95 #: 96 #: 97 #: 98 #: 99 #: 100 #: 101 #: 102 #: 103 #: 104 #: 105 #: 106 #: 107 #: 108 #: 109 #: 110 #: 111 #: 112 #: 113 #: 114 #: 115 #: 116 #: 117 #: 118 #: 119 #: 120 #: 121 #: 122 #: 123 #: 124 #: 125 #: 126 #: 127 #
[…]

### 2022-01-17 bench.sh
Talk about RemoveAll

@@ -9,7 +9,7 @@ mkdir -p "scala/target/generated/"
 git grep --name-only "//X" | grep scala | while read s; do
   cp "$s" "scala/target/generated/"
 done
-find "scala/target/generated/" -name "*.scala" | while read s; do
+find "$(pwd)/scala/target/generated/" -name "*.scala" | while read s; do
   seq -w 1 8 256 | while read i; do
     replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
     echo "$s.$i.scala"
@@ -21,7 +21,7 @@ done
 # Mesure bytecode size:
 # rm -f bytecode-size.log
 # rm -f *.class *.tasty
-# find "scala/target/generated/" -name "*.scala" | sort | while read s; do
+# find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
 #   if echo "$s" | grep -q "dependent"; then
 #     /usr/bin/time -f "%e" -o "/tmp/time" $HOME/workspace/thesis/bin/dotc -J-Xss4m "$s"
 #   else
@@ -32,7 +32,7 @@ done
 # done
 
 # Mesure compilation times:
-# find "scala/target/generated/" -name "*.scala" | sort | while read s; do
+# find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
 #   if echo "$s" | grep -q "dependent"; then
 #     (cd "ddotty/" && sbt "dotty-bench/jmh:run 60 60 1 $s" | tee "$s.log")
 #   else
@@ -42,7 +42,7 @@ done
 
 # Generate CSV:
 # rm -rf figures/*.csv
-# find "scala/target/generated/" -name "*.log" | sort | while read l; do
+# find "$(pwd)/scala/target/generated/" -name "*.log" | sort | while read l; do
 #   r=$(grep "±(99.9%)" "$l")
 #   avg=$(echo "$r" | grep -oP '\d+.\d+(?=.*99.9)')
 #   err=$(echo "$r" | grep -oP '(?<= )\d+.\d+(?= ms/op)')

### 2022-01-17 scala/src/main/scala/safe-implicits.scala

@@ -3,7 +3,7 @@ object SafeImplicits {
 // start section hlistEnumDefinition
 enum HList:
   case HNil()
-  case ::[H, T <: HList](head: H, tail: T)
+  case ::[+H, +T <: HList](head: H, tail: T)
 // end section hlistEnumDefinition
 import HList._
 
@@ -47,26 +47,67 @@ implicit def casecons[V, Ph, Pt <: HList]
 // end section memImplicitNotGiven
 }
 
-// start section removeAllWithPriority
-trait RemoveAll[V, Ps <: HList] { type Out <: HList }
+// start section removeAllDefinition
+trait RemoveAll[V, Ps <: HList, Out <: HList]
+// end section removeAllDefinition
 
-trait RemoveAllLowPrio:
-  type Aux[V, Ps <: HList, Out <: HList] = RemoveAll[V, Ps] { type Out }
-
-  implicit def casediff[V, Ph, Pt <: HList, Out <: HList]
-    (implicit ev: Aux[Ph, Pt, Out])
-    : Aux[V, Ph :: Pt, Ph :: Out] = new RemoveAll {}
-
-object RemoveAll extends RemoveAllLowPrio:
+object RemoveBroken {
+// start section removeAllBroken
+object RemoveAll:
   implicit def casenil[V]
-    : Aux[V, HNil, HNil] = new RemoveAll {}
+    : RemoveAll[V, HNil, HNil] = new RemoveAll {}
 
   implicit def casematch[V, Ps <: HList, Out <: HList]
-    (implicit ev: Aux[V, Ps, Out])
-    : Aux[V, V :: Ps, Out] = new RemoveAll {}
-// end section removeAllWithPriority
+    (implicit ev: RemoveAll[V, Ps, Out])
+    : RemoveAll[V, V :: Ps, Out] = new RemoveAll {}
 
-val removeAllTest = implicitly[RemoveAll.Aux[1, 1 :: 2 :: 1 :: 3 :: HNil, 2 :: 3 :: HNil]]
+  implicit def casedoesnt[V, Ph, Pt <: HList, Out <: HList]
+    (implicit ev: RemoveAll[V, Pt, Out])
+    : RemoveAll[V, Ph :: Pt, Ph :: Out] = new RemoveAll {}
+// end section removeAllBroken
+}
+
+// start section removeAllPrioritized
[…]

### 2022-01-21 article.tex
Rename Reduce to ReduceAxes
(Ab)Using Implicits conclusion

@@
-Type-level programming is becoming more and more popular in the realm of functional programming.
-However, the combination of type-level programming and subtyping remains largely unexplored in practical programming languages.
-This chapter presents \emph{match types}, a type-level equivalent of pattern matching.
-Match types integrate seamlessly into programming languages with subtyping and, despite their simplicity, offer significant additional expressiveness.
-We formalize the feature of match types in a calculus based on \SystemFsub and prove its soundness.
-We practically evaluate our system by implementing match types in the Scala 3 reference compiler, thus making type-level programming readily available to a broad audience of programmers.
-
 \section{Introduction}
@@
 This behavior is more complex than the previous examples, but can still be described accurately by a match type.
-We use a match type called |Reduce| to compute the return shape of the operation.
+We use a match type called |ReduceAxes| to compute the return shape of the operation.
 

### 2022-01-26 bench.sh
Publish benchmarks.zip and sources.zip

@@ -1,52 +1,0 @@
-#!/bin/sh
-
-set -eu
-test "$(pwd)" = "$(realpath "$(dirname "$0")")" || (echo "wrong dir"; exit 1)
-
-# Generate sources:
-rm -rf "scala/target/generated/"
-mkdir -p "scala/target/generated/"
-git grep --name-only "//X" | grep scala | while read s; do
-  cp "$s" "scala/target/generated/"
-done
-find "$(pwd)/scala/target/generated/" -name "*.scala" | while read s; do
-  seq -w 1 8 256 | while read i; do
-    replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
-    echo "$s.$i.scala"
-    cat $s | sed "s#0 :: //X#$replacement#g" > "$s.$i.scala"
-  done
-  rm $s
-done
-
-# Mesure bytecode size:
-# rm -f bytecode-size.log
-# rm -f *.class *.tasty
-# find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
-#   if echo "$s" | grep -q "dependent"; then
-#     /usr/bin/time -f "%e" -o "/tmp/time" $HOME/workspace/thesis/bin/dotc -J-Xss4m "$s"
-#   else
-#     /usr/bin/time -f "%e" -o /tmp/time $HOME/workspace/dotty-master/bin/scalac -J-Xss4m "$s"
-#   fi
-#   echo "$s: $(cat *.class | wc -c) ($(cat /tmp/time)s)" | tee -a bytecode-size.log
-#   rm -f *.class *.tasty
-# done
-
-# Mesure compilation times:
-# find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
-#   if echo "$s" | grep -q "dependent"; then
-#     (cd "ddotty/" && sbt "dotty-bench/jmh:run 60 60 1 $s" | tee "$s.log")
-#   else
-#     (cd "$HOME/workspace/dotty-master/" && sbt "scala3-bench/jmh:run 60 60 1 $s" | tee "$s.log")
-#   fi
-# done
-
-# Generate CSV:
-# rm -rf figures/*.csv
-# find "$(pwd)/scala/target/generated/" -name "*.log" | sort | while read l; do
-#   r=$(grep "±(99.9%)" "$l")
[…]

### 2022-01-26 process.sh

@@ -1,25 +1,0 @@
-#!/bin/sh
-
-set -eux
-
-BASEDIR=$(dirname "$0")
-
-rm "$BASEDIR/"*.csv || true
-
-cat "$BASEDIR"/*.log         |\
-grep -E -A4 "Running|Result" |\
-grep -E "±|scala"            |\
-paste -s -d' \n'             |\
-while read res; do
-  # [info] Running (fork) dotty.tools.benchmarks.Bench 40 10 1 /path/to/src-gen/dependent-remove-140.scala   295.236 ±(99.9%) 17.036 ms/op [Average]
-  file=$(echo "$res" | grep -oP '(?<=src-gen/).+(?=.scala)')
-  avg=$(echo "$res" | grep -oP '(?<=   )\d+.\d+(?=.*99.9)')
-  err=$(echo "$res" | grep -oP '(?<= )\d+.\d+(?= ms/op)')
-  bench=$(echo "$file" | grep -oP '.+(?=-\d+)')
-  size=$(echo "$file" | grep -oP '\d+')
-  echo "$size,$avg,$err" >> "$BASEDIR/$bench.csv"
-done
-
-ls "$BASEDIR/"*.csv | while read f; do
-  sort -n "$f" -o "$f"
-done

### 2022-01-26 run-benchmarks.sh

@@ -1,0 +1,55 @@
+#!/bin/sh
+
+set -eu
+test "$(pwd)" = "$(realpath "$(dirname "$0")")" || (echo "wrong dir"; exit 1)
+
+test -d ddotty || git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7 ddotty
+test -d dotty || git clone git@github.com:lampepfl/dotty.git --branch 3.1.2-RC1 dotty
+
+# Generate sources:
+rm -rf "scala/target/generated/"
+mkdir -p "scala/target/generated/"
+git grep --name-only "//X" | grep scala | while read s; do
+  cp "$s" "scala/target/generated/"
+done
+find "$(pwd)/scala/target/generated/" -name "*.scala" | while read s; do
+  seq -w 1 8 256 | while read i; do
+    replacement=$(seq 0 $(expr $i - 1) | xargs -i printf "{} :: ")
+    echo "$s.$i.scala"
+    cat $s | sed "s#0 :: //X#$replacement#g" > "$s.$i.scala"
+  done
+  rm $s
+done
+
+# Mesure bytecode size:
+rm -f bytecode-size.log
+rm -f *.class *.tasty
+find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
+  if echo "$s" | grep -q "dependent"; then
+    /usr/bin/time -f "%e" -o "/tmp/time" $HOME/workspace/thesis/bin/dotc -J-Xss4m "$s"
+  else
+    /usr/bin/time -f "%e" -o /tmp/time $HOME/workspace/dotty-master/bin/scalac -J-Xss4m "$s"
+  fi
+  echo "$s: $(cat *.class | wc -c) ($(cat /tmp/time)s)" | tee -a bytecode-size.log
+  rm -f *.class *.tasty
+done
+
+# Mesure compilation times:
+find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
+  if echo "$s" | grep -q "dependent"; then
+    (cd "ddotty/" && sbt "dotty-bench/jmh:run 60 60 1 $s" | tee "$s.log")
+  else
+    (cd "dotty/" && sbt "scala3-bench/jmh:run 60 60 1 $s" | tee "$s.log")
+  fi
+done
+
+# Generate CSV:
+rm -rf figures/*.csv
[…]

### 2022-01-27 article.tex
More attribution!

@@
 
-  \item We evaluate match types in a case study, presenting a type-safe version of the NumPy library (\Cref{sec:evaluation}).
+  \item We evaluate match types in a case study, presenting a type-safe version of the NumPy library (\Cref{sec:case-study-shape-safe-num-py}).
 
@@
 \section{Case Study: Shape-Safe NumPy}
-\label{sec:evaluation}
+\label{sec:case-study-shape-safe-num-py}
 

### 2022-01-28 run-benchmarks.sh
Add dotty submodule

@@ -21,16 +21,16 @@ find "$(pwd)/scala/target/generated/" -name "*.scala" | while read s; do
   rm $s
 done
 
-# Mesure bytecode size:
+# Mesure bytecode sizes:
 rm -f bytecode-size.log
 rm -f *.class *.tasty
 find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
   if echo "$s" | grep -q "dependent"; then
-    /usr/bin/time -f "%e" -o "/tmp/time" $HOME/workspace/thesis/bin/dotc -J-Xss4m "$s"
+    ddotty/bin/dotc -J-Xss4m "$s"
   else
-    /usr/bin/time -f "%e" -o /tmp/time $HOME/workspace/dotty-master/bin/scalac -J-Xss4m "$s"
+    dotty/bin/scalac -J-Xss4m "$s"
   fi
-  echo "$s: $(cat *.class | wc -c) ($(cat /tmp/time)s)" | tee -a bytecode-size.log
+  echo "$s: $(cat *.class | wc -c)" | tee -a bytecode-size.log
   rm -f *.class *.tasty
 done
 
@@ -43,7 +43,7 @@ find "$(pwd)/scala/target/generated/" -name "*.scala" | sort | while read s; do
   fi
 done
 
-# Generate CSV:
+# Generate CSV files:
 rm -rf figures/*.csv
 find "$(pwd)/scala/target/generated/" -name "*.log" | sort | while read l; do
   r=$(grep "±(99.9%)" "$l")

### 2022-01-30 scala/src/main/scala/tutorial.scala
Performance-Based Evaluation

@@ -1,4 +1,4 @@
-object ImplicitTutorial {
+object Tutorial {
 
 object explicit {
 // start section explicitModulo
@@ -39,6 +39,18 @@ implicit def listOrdering[T](implicit ev: Ordering[T]): Ordering[List[T]] =
         case (_, Nil) => false
         case (Nil, _) => true
 // end section ordListExample
+
+import scala.compiletime.ops.int.+
+
+// start section geqDefinition
+// GEQ[X, Y] compiles if X is greater than Y
+type GEQ[A <: Int, B <: Int] = A match {
+  case B => true
+  case _ => GEQ[A, B + 1]
+}
+// end section geqDefinition
+
+type A = GEQ[5, 1]
 
 }
 

### 2022-01-31 preprint.tex
Performance-Based Evaluation

@@
 \subsection{Safe Join}
+\label{sec:safe-join}
 
@@
 dependent val joined  = table1.join(table2, "name")
-joined: { Table(Cons("name", Cons("age", Cons("unit", Nil()))), choose[DataFrame]) }
+joined: { Table(
+            Cons("name", Cons("age", Cons("unit", Nil()))),
+            choose[DataFrame]) }
 \end{lstlisting}

### 2022-01-31 scala/src/main/scala/tutorial.scala

@@ -43,7 +43,7 @@ implicit def listOrdering[T](implicit ev: Ordering[T]): Ordering[List[T]] =
 import scala.compiletime.ops.int.+
 
 // start section geqDefinition
-// GEQ[X, Y] compiles if X is greater than Y
+// GEQ[A, B] compiles if A is greater than B
 type GEQ[A <: Int, B <: Int] = A match {
   case B => true
   case _ => GEQ[A, B + 1]
@@ -51,6 +51,16 @@ type GEQ[A <: Int, B <: Int] = A match {
 // end section geqDefinition
 
 type A = GEQ[5, 1]
+
+/*
+// start section geqExample
+5 <: 1? no
+5 <: 1+1? no
+5 <: 1+1+1? no
+5 <: 1+1+1+1? no
+5 <: 1+1+1+1+1? yes
+// end section geqExample
+*/
 
 }
 

### 2022-02-01 preprint.tex
Generalizing Scala's Singleton Types

@@
 Unfortunately, it remains a second-class citizen in the majority of industrially-used programming languages.
-We propose a new dependently-typed system with subtyping and singleton types whose goal is to enable type-level programming in an accessible style.
-At the heart of our system lies a non-deterministic choice operator.
-We argue that embracing non-determinism is crucial for bringing dependent types to a broader audience of programmers, since real-world programs will inevitably interact with imprecisely-typed, or even impure code.
-Furthermore, we show that singleton types combined with the choice operator can serve as a replacement for many type functions of interest in practice.
-We establish the soundness of our approach using the Coq proof assistant.
-Our soundness approach models non-determinism using additional function arguments to represent choices.
-We represent type-level computation using singleton types and existential types that quantify over choice arguments.
-To demonstrate the practicality of our type system, we present an implementation as a modification of the Scala compiler.
-We provide a case study in which we develop a strongly-typed wrapper for Spark datasets.
+We propose a new dependently-typed system with subtyping and singleton types whose goal is enabling type-level programming in an accessible style.
+To this end, we have prototyped our system as an extension of the Scala programming language.
+We demonstrate the practicality of our system with a case study in which we develop a strongly-typed wrapper for Spark datasets.
+Through our formalization and implementation in the context of an industrial-strength compiler, we hope to provide a reference point for other language designers interested in adding dependent types.
 
@@
 Furthermore, recent research has explored how dependent types could be added to the language for the same purpose~\citep{eisenberg2016dependent, weirich2017a}.
-In a largely orthogonal direction, inference for dependent refinement types is reaching significant maturity~\citep{vazou2018gradual, vazou2017refinement,vazou2015bounded}.
 
@@
 The simplicity of this approach is unfortunately at odds with the design of most programming languages, where types and terms are expressed using separate syntactic categories.
-Singleton types provide a simple solution to this problem by allowing every term to be represented as a type.
-The singleton type of a term therefore gives us the most precise specification for that term.
+Singleton types provide a simple solution to this problem by allowing terms to be represented as types.
 
+XXXX
 In this paper, we report on our attempt at combining an industrial mixed-paradigm language, Scala, with dependent types.
@@
 
-We present \oursystem, a dependently-typed calculus with subtyping and singleton types.
-The main novelty of our calculus is a new approach to expressing type-level computation that, at first, seems diametrically opposed to the purity other systems favor.
-A new term is added for non-deterministic choice from a base type, similar to Floyd's choice operator~\citep{floyd1967nondeterministic}.
-Designing a sound system in the presence of non-determinism is challenging.
-Our solution provides systematic translation of non-determinism using additional parameters that are existentially quantified at a syntactically well-defined point.
-Consequently, a term in \oursystem may reduce to different values.
 Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
@@
 
-Our contributions are as follows:
+With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
+Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automat
+ically using an additional keyword.
+This effectively allows users to execute programs involving functions and pattern matching to the type level.
+
+We implemented our type system on top of Dotty, the next-generation compiler of Scala.
+Working in the context of a production compiler we designed our implementation with backward-compatibility and interoperability in mind
+.
+Our system is backward-compatible in the sense that all existing programs continue to compile, but also interoperable, in that existing
+code can be used from dependently-typed programs and vice versa.
+As we will see, subtyping plays a key role in achieving interoperability.
+
+The remaining of the chapter is organized as follows:
 
@@
 \item
-  We present our calculus \oursystem, which illustrates the novel elements of our extension to Scala.
-  The type system of \oursystem combines dependent types, subtyping and a generalization of singleton types to non-deterministic terms.
-  We demonstrate how the interplay of these features allows us to leverage term-level programs for type-level computation.
+  We begin by motivating why dependent types are desirable and how one might use them to improve type safety (\Cref{sec:motivating-example}).
+Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
 \item
-  We provide a soundness proof of \oursystem by reusing the reducibility semantics of \FR \citep{hamza2019system}.
-  Using its semantics we prove the soundness of our rules.
-  These proofs are mechanized using the Coq proof assistant~\citep{bertot2004interactive}.
-  The formalization is available in the additional materials.
+  We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}).
+This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in adding dependent types to their language with subtyping.
 \item
-  We show a concrete use-case of our system by implementing it as an extension of Scala (\Cref{sec:extending-scala}), and using it to develop a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
+  We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
   Thanks to dependent types, we can statically ensure the type safety of database operations such as join and filter.
@@
 \end{itemize}
+
+Our formalization is out of the scope of this thesis...
 
@@
 
-\subsection{Safe Zip}
-\label{safe-zip}
-
-Our first example demonstrated how dependent methods allow inference of precise types.
-Conversely, we can also use singleton types to constrain method parameters further.
-In this example, our goal is to write a safer wrapper for functions like |zip| that should only be applicable to lists of the same length.
-To accomplish this, we can constrain the second parameter of |zip| as follows:
-
-\begin{lstlisting}
-def safeZip(xs: Lst, ys: { sizedLike(xs) }) = unsafeZip(xs, ys)
-\end{lstlisting}
-
-\noindent
-Here we would like |{ sizedLike(xs) }| to be inhabited by all lists of equal length as |xs|, regardless of their elements'values.
-How can this be achieved, given that |sizedLike(xs)| is a term? By exploiting the non-deterministic interpretation of |choose[T]|, we can provide a succinct definition for |sizedLike|:
-
-\begin{lstlisting}
-dependent def sizedLike(xs: Lst) <: Lst =
-  xs match
-    case Nil() => Nil()
-    case Cons(x, ys) => Cons(choose[Any], sizedLike(ys))
-\end{lstlisting}
-
-\noindent
-Consider, for instance, the meaning of |{ sizedLike(xs) }| for |xs = Cons(1, Cons(2, Nil()))|.
-After reduction, we obtain |{ Cons(choose[Any], Cons(choose[Any], Nil())) }|, which is a type that represents all lists of size 2.
-Thus |safeZip| requires that every caller prove that |xs| and |ys| are of the same length, which ensures that the underlying implementation in |unsafeZip| will never fail or truncate elements from one of the lists.
-
-Note that, unlike |concat| and |remove| that can be used both on the term and the type level, |sizedLike| is here intended to be used as a type function, but not at runtime.
-
 \section{Implementation}
@@
 This development was an experiment to explore the feasibility of adding dependent types in Scala.
-We implemented our prototype as an extension of Dotty, the reference compiler for future versions of the Scala language.
+We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
 Our presentation focuses on several facets of the implementation that are not reflected in our formalism.

### 2022-02-01 preprint.tex
More attribution!

@@
-Type-level programming is an increasingly popular way to obtain additional type safety.
-Unfortunately, it remains a second-class citizen in the majority of industrially-used programming languages.
-We propose a new dependently-typed system with subtyping and singleton types whose goal is enabling type-level programming in an accessible style.
-To this end, we have prototyped our system as an extension of the Scala programming language.
-We demonstrate the practicality of our system with a case study in which we develop a strongly-typed wrapper for Spark datasets.
-Through our formalization and implementation in the context of an industrial-strength compiler, we hope to provide a reference point for other language designers interested in adding dependent types.
-
 \section{Introduction}

### 2022-02-03 preprint.tex
Generalizing Scala's Singleton Types

@@
 
-XXXX
-In this paper, we report on our attempt at combining an industrial mixed-paradigm language, Scala, with dependent types.
-We offer both a formalization of our type system and a discussion of the challenges faced in a practical implementation.
-It is our hope that the present paper will serve as a guide for other language implementors interested in pushing the limit of their type systems by adding dependent types.
-
+In this paper, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
 Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
@@
 
+TODO: reformulate
 Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
@@
 With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
-Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automat
-ically using an additional keyword.
+Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using an new language keyword.
 This effectively allows users to execute programs involving functions and pattern matching to the type level.
-
-We implemented our type system on top of Dotty, the next-generation compiler of Scala.
-Working in the context of a production compiler we designed our implementation with backward-compatibility and interoperability in mind
-.
-Our system is backward-compatible in the sense that all existing programs continue to compile, but also interoperable, in that existing
-code can be used from dependently-typed programs and vice versa.
-As we will see, subtyping plays a key role in achieving interoperability.
 
@@
 \item
-  We begin by motivating why dependent types are desirable and how one might use them to improve type safety (\Cref{sec:motivating-example}).
+  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}).
 Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
 \item
+  TODO
+  We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
+  We describe how we extended Scala with dependent types.
+  This development was an experiment to explore the feasibility of adding dependent types in Scala.
+  We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
+  Our presentation focuses on several facets of the implementation that are not reflected in our formalism.
+
   We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}).
@@
   We show a concrete use-case of our system by implementing a strongly-typed wrapper for Apache Spark~\citep{zaharia2016apache} (\Cref{sec:use-case}).
-  Thanks to dependent types, we can statically ensure the type safety of database operations such as join and filter.
+  Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
+  TODO: keep this?
   We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
@@
 
-Our formalization is out of the scope of this thesis...
+TODO: pretend that formalization doesn't exist?
+The said report covers more material than what is included in this chapter, in particular, our formalization and the associated metatheory are out of the scope of this dissertation \citep[Section 3 and 4]{schmid2020coming}.
 
@@
 
-We begin by motivating why dependent types are desirable in general purpose programming, and how one might use them to improve type safety.
-In our first example, we design an API that keeps track of database tables' schemas in the type.
-We demonstrate how dependently-typed list operations can be used to compute schemas resulting from join operations at the type level.
-Our second example shows how to build a safer version of the zip operation on lists that only accepts equally-sized arguments.
-The examples in this section are written in our dependently-typed extension of Scala described in \Cref{sec:extending-scala}.
+We begin by motivating why type-level programming is desirable in general purpose programming.
+In our first example, we design an API that keeps track of database tables' schemas in the type, and uses that information to improve type safety.
+The examples in this section are written in our Scala extension described in \Cref{sec:extending-scala}.
 
@@
 \noindent
-We can define list concatenation in the usual functional style of Scala\footnote{Our examples use the indentation-based syntax introduced in Scala 3.0.}, that is, using pattern matching and recursion:
+We can define list concatenation in the usual functional style of Scala, that is, using pattern matching and recursion:
 
@@
 By annotating a method as |dependent|, the user instructs our system that the result type of |concat| should be as precise as its implementation.
-Effectively, this means that the body of |concat| is lifted to the type level, and will be partially evaluated at every call site to compute a precise result type which \emph{depends} on the given inputs.
+Effectively, this means that the body of |concat| is lifted to the type level in a singleton type, and will be partially evaluated at every call site to compute a precise result type which \emph{depends} on the given inputs.
 For recursive |dependent| methods such as |concat|, we infer types that include calls to |concat| itself.
@@
 Finally, by qualifying the definition of |Cons| and |Nil| as |dependent| we also allow their constructors and extractors to be lifted to the type level.
-Using these definitions, we can now request the precise type whenever we manipulate lists by annotating the new |val| binding with |dependent|:
+Using these definitions, we can now request the precise type whenever we manipulate lists by annotating |val| bindings as |dependent|:
 
@@
 Enclosing a pure term in braces (|{|~\ldots{}~|}|) denotes the singleton type of that term.
-In the last two lines of this example we are therefore asking our system to prove that |l3| has size 2 and is equivalent to |Cons("A", Cons("B", Nil()))|.
-
-In Scala we often deal with impure or imprecisely-typed code, however.
-To integrate with such terms, we provide the |choose[T]| construct.
-Operationally, we interpret |choose[T]| as a non-deterministic choice from |T|, which can be modeled faithfully on the type level as an existentially quantified inhabitant of |T| in a singleton type.
-Thus, we equate |{ choose[T] }| to |T|, and when typing an impure term such as |Cons(readString(), Nil())| we can assign the precise type |{ Cons(choose[String], Nil()) }|.
-Returning to the previous example, this means that even in the presence of impurity, we can perform useful type-level computation and checking:
-
-\begin{lstlisting}
-dependent val l2 = Cons(readString(), Nil())
-dependent val l3 = l1.concat(l2)
-l3: { Cons("A", Cons(choose[String], Nil())) }
-\end{lstlisting}
-
-In a style similar to |concat|, we can define remove on |Lst|:
+In the last two lines of this example, we are therefore asking our system to prove that |l3| has size 2 and is equivalent to |Cons("A", Cons("B", Nil()))|.
+Similarly, we can define remove on |Lst|:
 
@@
             Cons("name", Cons("age", Cons("unit", Nil()))),
-            choose[DataFrame]) }
+            _: DataFrame) }
 \end{lstlisting}
@@
 
-In this section we give an overview of how we extended Scala with dependent types.
-This development was an experiment to explore the feasibility of adding dependent types in Scala.
+In this section we give an overview of how we extended Scala with a generalized notion of singleton types, and how we manipulate these types during type-checking.
+Our presentation is divided into three parts.
+First, we introduce new types that reflect a subset of Scala's term language at the type level.
+We then introduce the |dependent| qualifier, which influences type inference to assign generalized singleton types to the annotated definition.
+Finally, we give an informal description of our type evaluation algorithm.
+
+This development was an experiment to explore the feasibility of adding a lightweight form of dependent types to Scala.
 We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
-Our presentation focuses on several facets of the implementation that are not reflected in our formalism.
 
-On a syntactic level, our Scala extension consists of three additions:
+\subsection{Reflecting terms in types}
+
+The fundamental difference between our and Scala's existing type system is that we take Scala's notion of singleton types, that is, precise types for variable bindings and literals, and extend them to cover Scala's core functional expressions.
+Concretely we add new forms of types for the following constructs:
 
 \begin{itemize}
-\item
-  the singleton types syntax |{ t }|,
-\item
-  the |dependent| modifier for methods, values and classes,
-\item
-  the |choose[T]| construct.
+  \item Variable bindings and member selections
+  \item Primitive literals
+  \item Method calls
+  \item If-then-else expressions
+  \item Pattern matching expressions
+  \item Constructor calls
 \end{itemize}
 
-\noindent
-The newly-introduced singleton type syntax enables a subset of Scala expressions to be used in types.
-This subset approximately corresponds to the core functional subset of Scala, plus the |choose[T]| construct, as illustrated in \oursystem.
-Within this subset, the main differences between our formalism and implementation lie in the handling of pattern matching.
+Such new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
+For instance, suppose $\text{foo}$ stands for the expression
+
+\begin{lstlisting}
+if (x % 2 == 0) "even" else "odd"
+\end{lstlisting}
+
+\noindent then $\text{foo}$ can be typed as |{| $\text{foo}$ |}|.
+More precisely, |{| $\text{foo}$ |}| is the singleton type of $\text{foo}$ and corresponds to the unique value of $\text{foo}$ in a given context.
+
+The first two constructs in this list have antecedents in Scala.
+Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{scalaSpec}, which is equivalent to |{ x }| in our system.
+Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{sip23}.
+
+In addition to base types and singleton types for every pure term, our system also supports types that lie in-between.
+We add a new form, |_: T|, which supplements expressions in types of form |{| $e$ |}| by allowing the user to intersperse terms and base types.
+
+For instance, consider the following list and its singleton type: |Cons(Nil, Nil): { Cons(Nil, Nil) }|.
+In our system, this term may also be typed less precisely, such as:
+
+\begin{itemize}
+  \item |Cons(Nil, Nil): { Cons(_: Any, Nil   ) }|
+  \item |Cons(Nil, Nil): { Cons(Nil   , _: Lst) }|
+  \item |Cons(Nil, Nil): { Cons(_: Any, _: Lst) }|
+\end{itemize}
+
+all of which are subsumed by |Cons(Nil, Nil): Lst|.
 
@@
 
-When inferring the result type of a |dependent| method, our system lifts the method's body into a type.
+When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
 This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
-When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |choose[T]|.
+When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|.
 
@@
 \begin{lstlisting}
-{ if (personalized) choose[String] else "Joe" }
+{ if (personalized) (_: String) else "Joe" }
 \end{lstlisting}
@@
 In this section, we extend the motivating example presented in \Cref{sec:motivating-example} by building a type-safe interface for Spark datasets.
-We use dependent types to implement a simple domain-specific type checker for the SQL-like expressions used in Spark.
-We then compare the compilation time of our dependently-typed interface against an equivalent encoding based on implicits.
+We use our Scala extension to implement a simple domain-specific type checker for the SQL-like expressions used in Spark.
 
@@
   (a, b) match
-    case (_: Char, _: Char | Byte | Int  | Long | Double) => choose[Double]
+    case (_: Char, _: Char | Byte | Int  | Long | Double) => _: Double
     case (_: Byte, _: Byte | Int  | Long | Double)        => b
@@
     case (_: Long, _: Long | Double)                      => b
-    case (_: Double, _: Double)                           => choose[Double]
+    case (_: Double, _: Double)                           => _: Double
     case (_: Byte | Int | Long | Double, _)               => addRes(b, a)
     case _ => throw new Error("incompatible types in addition")
-type AddRes[A, B] = { addRes(choose[A], choose[B]) }
+type AddRes[A, B] = { addRes(_: A, _: B) }
 \end{lstlisting}
@@
 \end{lstlisting}
-% val pred = (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("a") \phantom{.} }*) + (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("b") \phantom{.}\phantom{.} }*) === (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{ \phantom{.} table.col("c") \phantom{.}\phantom{.} }*)
-% // Infers: (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Int]("a") \}}}*)   (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("b") \}}}*)     (*\setlength{\fboxrule}{-1pt}\fcolorbox{light-gray}{light-gray}{\textit{\{ Column[Long]("c") \}}}*)
 
@@
 In contrast, our system does not support any erasure annotations and simply follows Scala's canonical erasure strategy: types are systematically erased to JVM types, and terms are left untouched.
-Weirich established a fully mechanized type safety proof for the core of Dependent Haskell using the Coq proof assistant~\citep{weirich2017a}.
+\citeauthor{weirich2017a} established a fully mechanized type safety proof for the core of Dependent Haskell~\citep{weirich2017a}.
 

### 2022-02-04 article.tex
Performance-Based Evaluation
Generalizing Scala's Singleton Types

@@
 This section provides an overview of the proof structure and states the involved lemmas and theorems.
-Detailed proofs are available in the supplementary material of this paper, in two different versions.
+Detailed proofs are available in the supplementary material of this thesis, in two different versions.
 The first version, \cite{blanvillain2021type}, is a pen-and-paper proof where \SystemFm is exactly as presented in \Cref{fig:FMTypingRules}.
@@
 \subsection{Caching}
+\label{subsec:caching}
 
@@
 Although the situation is concerning, it might not be as bad as it seems given that soundness is a non-goal of TypeScript's type system~\citep{bierman2014understanding}.
-Nevertheless, we believe that the results of this paper are directly applicable to conditional types and could be used to improve TypeScript's type checker.
+Nevertheless, we believe that the results of this chapter are directly applicable to conditional types and could be used to improve TypeScript's type checker.
 

### 2022-02-04 preprint.tex
Generalizing Scala's Singleton Types

@@
 
-In this paper, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
+In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
 Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
@@
 \item
-  TODO
+  We describe how we extended Scala with a generalization of singleton types.
   We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
-  We describe how we extended Scala with dependent types.
-  This development was an experiment to explore the feasibility of adding dependent types in Scala.
-  We implemented our prototype as an extension of Dotty, the reference Scala~3 compiler.
-  Our presentation focuses on several facets of the implementation that are not reflected in our formalism.
-
-  We describe how we extended Scala with dependent types (\Cref{sec:extending-scala}).
-This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in adding dependent types to their language with subtyping.
+  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system,
+  as well as programming language designers interested dependent types.
 \item
@@
   Thanks to our generalized singleton types, we can statically ensure the type safety of database operations such as join and filter.
-  TODO: keep this?
   We compare our implementation with an equivalent implicit-based one and show remarkable compilation time savings.
 \end{itemize}
-
-TODO: pretend that formalization doesn't exist?
-The said report covers more material than what is included in this chapter, in particular, our formalization and the associated metatheory are out of the scope of this dissertation \citep[Section 3 and 4]{schmid2020coming}.
 
@@
 dependent val joined  = table1.join(table2, "name")
-joined: { Table(
-            Cons("name", Cons("age", Cons("unit", Nil()))),
-            _: DataFrame) }
+joined: { Table(Cons("name", Cons("age", Cons("unit", Nil()))), _: DataFrame) }
 \end{lstlisting}
@@
 Our presentation is divided into three parts.
-First, we introduce new types that reflect a subset of Scala's term language at the type level.
+
+First, we introduce new types that reflect a subset of Scala's term language at the type level, and give an informal description of our type evaluation algorithm.
 We then introduce the |dependent| qualifier, which influences type inference to assign generalized singleton types to the annotated definition.
-Finally, we give an informal description of our type evaluation algorithm.
+Finally, we discuss how our extension interacts with other aspects of Scala, such as side effects, virtual dispatch, and recursion.
 
@@
 
-\subsection{Reflecting terms in types}
+\subsection{Reflecting Terms in Types}
 
 The fundamental difference between our and Scala's existing type system is that we take Scala's notion of singleton types, that is, precise types for variable bindings and literals, and extend them to cover Scala's core functional expressions.
-Concretely we add new forms of types for the following constructs:
+Concretely we add new types for the following constructs:
 
@@
 
-Such new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
+These new types can be expressed using the |{| $e$ |}| type syntax, where $e$ is an expression in the core functional subset of Scala listed above.
 For instance, suppose $\text{foo}$ stands for the expression
@@
 The first two constructs in this list have antecedents in Scala.
-Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{scalaSpec}, which is equivalent to |{ x }| in our system.
-Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{sip23}.
+Types for bindings are available since the early days of Scala with the |x.type| syntax \citep[Section 3.2.1]{odersky2006scala}, which is equivalent to |{ x }| in our system.
+Types for primitive literals have recently been added to the language: literals for booleans, strings and the various numeric types are made available in the type language \citep{leontiev2014sip}.
 
@@
 \begin{itemize}
-  \item |Cons(Nil, Nil): { Cons(_: Any, Nil   ) }|
-  \item |Cons(Nil, Nil): { Cons(Nil   , _: Lst) }|
+  \item |Cons(Nil, Nil): { Cons(_: Any, Nil) }|
+  \item |Cons(Nil, Nil): { Cons(Nil, _: Lst) }|
   \item |Cons(Nil, Nil): { Cons(_: Any, _: Lst) }|
@@
 all of which are subsumed by |Cons(Nil, Nil): Lst|.
+
+\subsection{Type Evaluation}
+
+Our system evaluates types using a call-by-value partial evaluator, which we embedded in the type-checker.
+The evaluation rules for the term in generalized singleton types are completely standard up to dynamic type tests.
+
+We perform evaluation of types during subtyping.
+For instance, given the following definition of a |paritiy| function:
+
+\begin{lstlisting}
+dependent def parity(x: Int) =
+  if (x % 2 == 0) "even" else "odd"
+\end{lstlisting}
+
+in order to prove the well-typedness of
+
+\begin{lstlisting}
+val p: { "odd" } = parity(5)
+\end{lstlisting}
+
+\noindent
+the type-checker will ensure that |{ parity(5) }| is a subtype of |{ "odd" }|.
+In the process, the left-hand side is evaluated as follows:
+
+\begin{lstlisting}
+  { parity(5) }                          (*$(1)$*)
+= { if (5 % 2 == 0) "even" else "odd" }  (*$(2)$*)
+= { if (false) "even" else "odd" }       (*$(3)$*)
+= { "odd" }
+\end{lstlisting}
+
+\noindent
+In (1), the method call to |parity| is replaced by its result type, where the method parameter has been substituted by the singleton type of the argument, |{ 5 }|.
+(2) evaluates the boolean expression to |false|.
+(3) reduces the |if| expression to its else branch.
+
+In general, our evaluator will execute operations on concrete primitive values of types such as |Boolean|, |Int| and |String|, i.e., perform constant-folding.
 
@@
 \begin{lstlisting}
-s match { case _: T1 => v1 case _: T2 => v2 }
+s match {
+  case _: T1 => v1
+  case _: T2 => v2
+}
 \end{lstlisting}
@@
 In order to reduce to |v2|, type normalization must make sure |T1| and the scrutinee's type are disjoint, namely that the dynamic type of |s| cannot possibly be smaller than |T1|.
-Disjointness proofs are built using static knowledge about the class hierarchy and make use of the guarantees implied by the |sealed| and |final| qualifiers, which are Scala's way of declaring closed-type hierarchies.
+Disjointness proofs are built using static knowledge about the class hierarchy, such as the |sealed| and |final| qualifiers, which are Scala's way of declaring closed-type hierarchies.
 
@@
 
+\noindent
+Note that we could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type  explicitly.
+
 Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
@@
 The situation is analogous for |dependent| methods.
-Thanks to |choose[T]| we can still formulate precise result types when terms depend on the result of side-effectful operations.
 Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
-For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, but we can still assign a better type than |Lst| to an expression like |Cons(z, Nil())|, that is, |{ Cons(choose[Int], Nil()) }|.
+For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, since reading a mutable variable is a side effect.
+However, we can still assign precise types to expressions containing |z|, for instance, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
 
@@
 Despite these shortcomings, our treatment of exceptions results in a practical way to raise compile-time errors.
-It would be interesting to explore the addition of an effect system to our Scala extension and formalization.
+It would be interesting to explore the addition of an effect system to our Scala extension.
 
@@
 The first question is whether type-checked programs are guaranteed to terminate.
-For simplicity, our work side-steps this question, requiring bounds for recursion.
-A more general solution would be to compute or infer such bounds using measure functions, as done in \FR~\citep{hamza2019system}.
-Another approach would be to extend our translation of non-determinism to permit non-termination.
-We consider this aspect orthogonal to the objectives of this paper.
-Our work targets general-purpose programming language whose type safety is defined with regards to its runtime semantics and that may include non-terminating interactive computations.
+Proving termination or totality of dependent programs is a non-goal of our system.
+Unlike proof assistants, Scala programs do not manipulate proof terms.
+Consequently, the lack of totality checks does not affect Scala's present notion of safety.
+Exceptions or infinite loops in the evaluation of a `dependent` method would prevent the completion of type-checking.
 
 The second question is termination of our type checker.
-Non-termination of type checking implies that the type checker can give three possible answers, "type correct", "type incorrect" or "do not know" (or timeout).
+Non-termination of type checking implies that the type checker can give three possible answers, "type correct", "type incorrect" or "do not know" (when type checking times out).
 Treating "do not know" as "type incorrect" makes the non-termination unproblematic from a soundness perspective.

### 2022-02-06 preprint.tex
Generalizing Scala's Singleton Types: typos

@@
 In the context of Haskell, much research has gone into extending the language to support computations on types, for instance in the form of functional dependencies~\citep{jones2000type}, type families~\citep{kiselyov2010fun} and promoted datatypes~\citep{yorgey2012giving}.
-These techniques have seen adoption by Haskell programmers, showing that there is a real demand for such mechanisms.
+These techniques have seen vivid adoption by Haskell programmers, showing that there is a real demand for such mechanisms.
 Furthermore, recent research has explored how dependent types could be added to the language for the same purpose~\citep{eisenberg2016dependent, weirich2017a}.
@@
 With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
-Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using an new language keyword.
-This effectively allows users to execute programs involving functions and pattern matching to the type level.
+Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using a new language keyword.
+This effectively allows users to execute programs involving functions and pattern matching at the type level.
 
@@
 The |<:| annotation lets us provide an upper bound on |concat|'s result type, which will be used while type checking the method's definition.
-Finally, by qualifying the definition of |Cons| and |Nil| as |dependent| we also allow their constructors and extractors to be lifted to the type level.
+Finally, by qualifying the definition of |Cons| and |Nil| as |dependent|, we also allow their constructors and extractors to be lifted to the type level.
 Using these definitions, we can now request the precise type whenever we manipulate lists by annotating |val| bindings as |dependent|:
@@
 
-The lists we defined so far can be used to implement a type-safe interface for database tables.
+The lists we defined so far can be used to implement a type-safe interface for database tables:
 
@@
 
-all of which are subsumed by |Cons(Nil, Nil): Lst|.
+\noindent
+all of which are subsumed by
+
+\begin{itemize}
+  \item |Cons(Nil, Nil): Lst|.
+\end{itemize}
 
@@
 We perform evaluation of types during subtyping.
-For instance, given the following definition of a |paritiy| function:
+For instance, consider the following definition of a |parity| function:
 
@@
 
-in order to prove the well-typedness of
+\noindent
+In order to prove the well-typedness of
 
@@
 
-During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics, that is, patterns are checked top-to-bottom, and type-tests are evaluated using runtime type information available after type erasure.
+During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics.
+Patterns are checked top-to-bottom, and type-tests are evaluated using runtime type information available after type erasure.
 
@@
 
-In order to retain backwards-compatibility, our system supports two modes of type inference: the precise inference mode which infers singleton types, and the default inference mode that corresponds to Scala's current type-inference algorithm.
+In order to retain backwards-compatibility, our system supports two modes of type inference: the precise inference mode which infers singleton types, and the default inference mode that corresponds to Scala's type-inference algorithm.
 Concretely, users opt into our new inference mode using the |dependent| qualifier on methods, values, and classes.
@@
 \noindent
-Note that we could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type  explicitly.
+We could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type  explicitly.
 
@@
 At other call sites, the (precise) inferred result type is used.
-Bounds of dependent methods are written using a special syntax (|<: T|), which emphasizes the difference from normal result types (|: T|).
+Bounds of dependent methods are written using a special syntax (|<: T|).
 
 \subsection{Approximating Side Effects}
-
-\paragraph{State}
 
@@
 Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
-For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type, since reading a mutable variable is a side effect.
-However, we can still assign precise types to expressions containing |z|, for instance, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
-
-\paragraph{Exceptions}
+For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type.
+However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
 
@@
 However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.
-In these cases, our type system can prove that the program execution will encounter exceptional behavior, and reports a compilation error.
+In these cases, our type system can prove that the program execution will encounter exceptional behavior, and report a compilation error.
 This approach is conservative in that it might reject programs that recover from exceptions.
@@
 
-We distinguish two important aspects of termination.
+We distinguish two important aspects of termination: the termination of type-checked programs and the termination of our type checker.
 
-The first question is whether type-checked programs are guaranteed to terminate.
-Proving termination or totality of dependent programs is a non-goal of our system.
+Proving termination or totality of programs is a non-goal of our system.
 Unlike proof assistants, Scala programs do not manipulate proof terms.

### 2022-02-06 preprint.tex
Generalizing Scala's Singleton Types

@@
 
-In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form a dependently-typed programming.
-Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring totality of functions through termination checks.
+In this chapter, we report on our attempt to generalize Scala's singleton types to support type-level programming, as well as a lightweight form of dependently-typed programming.
+Unlike proof assistants, we do not aim to use types as a general-purpose logic, which would favor designs ensuring the totality of functions through termination checks.
 Instead, our focus is on improving type safety by increasing the expressive power of the type system.
 
-TODO: reformulate
-Our system generalizes the traditional notion of singleton type: when the lifted term $t$ contains a non-deterministic choice, the resulting type $\singleton{t}$ denotes the set of values that $t$ could possibly reduce to.
-As a result, our type system is capable of type computations by manipulating types which are based on terms, but can nonetheless contain more than a single value.
-In combination with subtyping, this allows us to seamlessly integrate with impure, or imprecisely-typed programs.
-
-With these goals in mind, we extended Scala's type system to lift programs and partially evaluate them as part of type-checking.
-Users can manipulate dependent types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using a new language keyword.
+With these goals in mind, we extended Scala's type system to lift programs to the type level and partially evaluate them as part of type-checking.
+Users can manipulate those types either directly, using new syntactic forms in Scala's type language, or have them inferred automatically using a new language keyword.
 This effectively allows users to execute programs involving functions and pattern matching at the type level.
@@
 \item
-  We begin by motivating why type-level programming in desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}).
+  We begin by motivating why type-level programming is desirable and how one might use our Scala extension to improve type safety (\Cref{sec:motivating-example}).
 Our example demonstrates how to design a strongly-typed API in a functional style accessible to programmers.
 \item
-  We describe how we extended Scala with a generalization of singleton types.
+  We describe how we extended Scala with a generalization of singleton types (\Cref{sec:extending-scala}).
   We prototyped our type system on top of Dotty, the reference Scala~3 compiler.
-  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system,
-  as well as programming language designers interested dependent types.
+  This practical introduction would be of interest to any Scala programmer willing to learn how to use our system, as well as programming language designers interested in dependent types.
 \item
@@
 \end{itemize}
+
+Our original presentation of generalized singleton types contains a formalization \citep[Section 3 and 4]{schmid2020coming}, which is out of the scope of this chapter.
 
@@
 
-For instance, consider the following list and its singleton type: |Cons(Nil, Nil): { Cons(Nil, Nil) }|.
+For instance, consider the following list and its singleton type:
+
+\begin{itemize}
+  \item |Cons(Nil, Nil): { Cons(Nil, Nil) }|
+\end{itemize}
+
 In our system, this term may also be typed less precisely, such as:

### 2022-02-07 symposium.tex
Type and Null Safety for Regexes

\section{Introduction}

The Scala standard library includes a package to manipulate regular expressions.
This package is based on Java's implementation of regular expression, and thus benefits from the high performance of the JVM's regex engine.
The Scala implementation innovates with its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
The package's documentation starts with the following example:

% val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
% "2004-01-20" match {
%   case date(year, month, day) =>
%     s"$year was a good year for PLs."
% }

\noindent
The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capture groups, and shields its users from off-by-one error.

While the syntax of this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
Furthermore, the values coming out of capture groups can be null (when using optional captures) and thus require an additional layer of validation.
Those shortcomings might appear benign when looking at small examples, but can easily turn into bugs when dealing with a large-scale code base.

In this chapter, we propose a new design for Scala's regular expression library which provides both type-safe and null-safe capture group extraction.
Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement while still enjoying additional safety.

Our contribution are as follows:

\begin{itemize}
  \item ...
  \item ...
  \item ...
  \item ...
\end{itemize}

\section{Architecture}

In this section...

\section{Type-Level}

In this section...

\section{Term-Level}

In this section...

\section{Evaluation}

In this section...

### 2022-02-07 scala/src/main/scala/regex.scala
Add regex implementation

@@ -1,0 +1,230 @@
+import java.util.regex.Pattern
+import annotation.experimental
+import compiletime.ops.string.{+ => ++, _}
+import compiletime.ops.int.{+, -, Max, <}
+import Lib._
+
+@experimental
+object Regex {
+  def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
+    new Regex(regex)
+
+  type Compile[R <: String] =
+    FlattenReverse[Loop[R, 0, Length[R], HNil, 0]]
+
+  type Charat[R <: String, At <: Int] =
+    Substring[R, At, At + 1]
+
+  type Nullmark[R <: String, At <: Int, Hi <: Int] =
+    At match
+      case Hi => false
+      case _ =>
+        Charat[R, At] match
+          case "?" | "*" | "|" => true
+          case "+" => Nullmark[R, At + 1, Hi]
+          case _ => false
+
+  type Nullable[R <: String, At <: Int, Hi <: Int] =
+    At match
+      case 1 => Nullable0[R, At, Hi, 0]
+      case _ => Charat[R, At - 2] match
+        case "|" => true
+        case _ => Nullable0[R, At, Hi, 0]
+
+  type Nullable0[R <: String, At <: Int, Hi <: Int, Ops <: Int] =
+    At match
+      case Hi => false
+      case _ =>
+        Charat[R, At] match
+          case ")" => Ops match
+            case 0 => Nullmark[R, At + 1, Hi]
+            case _ => Nullable0[R, At + 1, Hi, Ops - 1]
+          case "(" => Nullable0[R, At + 1, Hi, Ops + 1]
+          case _   => Nullable0[R, At + 1, Hi, Ops]
+
+  type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: HList, Nw <: Int] <: HList =
+    Lo match
+      case Hi => Acc
+      case _   => Charat[R, Lo] match
+        case ")"  =>
+          Loop[R, Lo + 1, Hi, Acc, Max[0, Nw - 1]]
+        case "("  =>
+          Nw match
+            case 0 =>
+              Nullable[R, Lo + 1, Hi] match
+                case true =>
[…]

### 2022-02-07 scala/src/test/scala/bad-implicits.scala

@@ -1,5 +1,6 @@
 object BadImplicits {
 
+/*
 // start section implicitAmbiguity
 implicit val number: Int = 1
 implicit val modulo: Int = 3
@@ -7,5 +8,6 @@ def addm(x: Int, y: Int)(implicit m: Int) = (x + y) % m
 addm(4, 5) // Error: ambiguous implicit arguments: both value number and
            // value modulo match type Int of parameter m of method addm.
 // end section implicitAmbiguity
+*/
 
 }

### 2022-02-07 scala/src/test/scala/badreduce.scala

@@ -2,10 +2,12 @@ object BadReduce {
 
 import NumpyMatchTypes._
 
+/*
 // start section badreducescala
 val img_batch = random_normal(#:(25, #:(256, #:(256, #:(3, Ø)))))
 val avg_colors = mean(img_batch, #:(0, #:(1, #:(2, Ø))))
 val avg_color_square = reshape(avg_colors, #:(5, #:(5, Ø)))
 // end section badreducescala
+*/
 
 }

### 2022-02-07 scala/src/test/scala/regex-tests.scala

@@ -1,0 +1,39 @@
+import annotation.experimental
+
+@experimental
+object Tests {
+  object TypeLevel {
+    def check[A <: String, B](implicit ev: Regex.Compile[A] =:= B): Unit = ()
+    type S = String
+    type O = Option[String]
+
+    check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
+    check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
+    check["(\\d{4})?-(\\d{2})*-(\\d{2})++", (O, O, S)]
+    check["((A)(B(C)))", (S, S, S, S)]
+    check["((A)(B(C)))?", (O, O, O, O)]
+    check["((A)(B(C))?)", (S, S, O, O)]
+    check["((A)|(B(C)))", (S, O, O, O)]
+    check["(B)|C", O]
+    check["C|(A)", O]
+    check[".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*", (S, S, O)]
+  }
+
+  def assertEquals(expected: Any, actual: Any): Unit =
+    assert(
+      expected == actual,
+      s"expected: $expected, actual: $actual")
+
+  def main(args: Array[String]): Unit = {
+    val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
+    assertEquals("2004", "2004-01-20" match
+      case date(year, month, day) => year)
+
+    val everyday = Regex(".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*")
+    assertEquals(("Scala.js", "Scala"),
+      "Write Scala.js everyday!" match
+        case everyday(a, b, None) => (a, b))
+
+    println("OK!")
+  }
+}

### 2022-02-07 symposium.tex
Type and Null Safety for Regexes

@@
 
+\regexDocumentation
 % val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
-% "2004-01-20" match {
+% "2004-01-20" match
 %   case date(year, month, day) =>
 %     s"$year was a good year for PLs."
-% }
 
@@
 
-Our contribution are as follows:
-
-\begin{itemize}
-  \item ...
-  \item ...
-  \item ...
-  \item ...
-\end{itemize}
-
 \section{Architecture}
 
-In this section...
+We present our library in terms of 3 components:
+
+\begin{enumerate}
+  \item The type-level capture group analysis, which uses match types to inspect the user-provided regular expression and compute the corresponding tuple-based type representation (\Cref{sec:type-level}).
+
+  \item The term-level capture group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).
+
+  \item The user-level interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capture group extraction.
+\end{enumerate}
+
+We begin our presentation with the user-level interface, which takes the form of relatively simple Scala class and companion object:
+
+\regexUserLevel
+% object Regex:
+%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
+%     new Regex(regex)
+%
+% class Regex[P] private (val regex: String):
+%   val pattern = Pattern.compile(regex)
+%   def unapply(s: String): Option[P] =
+%     val m = pattern.matcher(s)
+%     if (m.matches())
+%       val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
+%       Some(toTuple[P](regex, a))
+%     else
+%       None
+
+The |apply| method is the entry point of our library.
+It takes a regular expression and returns an instance of |Regex|.
+Instead of directly accepting |String| argument, that method requires a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression (the |Singleton| bound is necessary to guide Scala's type inference).
+We use that type-level string immediately to instantiate |Compile|, which is our type-level capture group analysis component.
+We store the output of that analysis as |Regex|'s type argument.
+
+The implementation of the |Regex| class is straightforward.
+Similarly its counterpart in the original Scala implementation, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
+The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.
+
+The main difference between Scala's original |Regex| implementation and ours lies in the signature the |unapply| method, which determines the nature of extraction performed by that method.
+By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capture groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
+We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
+In our implementation, this is done via the |toTuple| method, which is our
+term-level capture group processing component.
+As a result, our implementation never returns null values, which removes the null-checking burden of from the user.
+
+For instance, in the following example, we use a regular expression with an optional capture group to extract the integral and fractional parts of a rational number:
+
+\regexRational
+% val rational = Regex("(\\d+)\\.?(\\d+)?")
+% "3.1415" match
+%   case rational(i, Some(f)) =>
+%     val n = i.size + f.size
+%     s"This number is $n digits long"
+
+\noindent
+Here, our implementation instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which it then uses in |unapply|'s return type to obtain a precise representation of the regex's capture groups.
 
 \section{Type-Level}
+\label{sec:type-level}
 
-In this section...
+The main challenge...
 
 \section{Term-Level}
+\label{sec:term-level}
 

### 2022-02-07 scala/src/main/scala/regex.scala

@@ -1,51 +1,39 @@
 import java.util.regex.Pattern
 import annotation.experimental
-import compiletime.ops.string.{+ => ++, _}
+import compiletime.ops.string._
 import compiletime.ops.int.{+, -, Max, <}
-import Lib._
+import Package.Lib._
 
 @experimental
-object Regex {
+object Package {
+
+// start section regexUserLevel
+object Regex:
   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
     new Regex(regex)
 
-  type Compile[R <: String] =
-    FlattenReverse[Loop[R, 0, Length[R], HNil, 0]]
+class Regex[P] private (val regex: String):
+  val pattern = Pattern.compile(regex)
+  def unapply(s: String): Option[P] =
+    val m = pattern.matcher(s)
+    if (m.matches())
+      val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
+      Some(toTuple[P](regex, a))
+    else
+      None
+// end section regexUserLevel
 
-  type Charat[R <: String, At <: Int] =
+object Lib {
+  type CharAt[R <: String, At <: Int] =
     Substring[R, At, At + 1]
 
-  type Nullmark[R <: String, At <: Int, Hi <: Int] =
-    At match
-      case Hi => false
-      case _ =>
-        Charat[R, At] match
-          case "?" | "*" | "|" => true
-          case "+" => Nullmark[R, At + 1, Hi]
-          case _ => false
+  type Compile[R <: String] =
+    Reverse[Loop[R, 0, Length[R], EmptyTuple, 0]]
 
-  type Nullable[R <: String, At <: Int, Hi <: Int] =
-    At match
-      case 1 => Nullable0[R, At, Hi, 0]
-      case _ => Charat[R, At - 2] match
-        case "|" => true
-        case _ => Nullable0[R, At, Hi, 0]
-
-  type Nullable0[R <: String, At <: Int, Hi <: Int, Ops <: Int] =
-    At match
-      case Hi => false
-      case _ =>
-        Charat[R, At] match
-          case ")" => Ops match
[…]

### 2022-02-07 scala/src/test/scala/regex-tests.scala

@@ -1,9 +1,10 @@
 import annotation.experimental
+import Package._
 
 @experimental
 object Tests {
   object TypeLevel {
-    def check[A <: String, B](implicit ev: Regex.Compile[A] =:= B): Unit = ()
+    def check[A <: String, B](implicit ev: Lib.Compile[A] =:= B): Unit = ()
     type S = String
     type O = Option[String]
 
@@ -23,6 +24,23 @@ object Tests {
     assert(
       expected == actual,
       s"expected: $expected, actual: $actual")
+
+  def examples: Unit = {
+// start section regexDocumentation
+val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
+"2004-01-20" match
+  case date(y, m, d) =>
+    s"$y was a good year for PLs."
+// end section regexDocumentation
+
+// start section regexRational
+val rational = Regex("(\\d+)\\.?(\\d+)?")
+"3.1415" match
+  case rational(i, Some(f)) =>
+    val n = i.size + f.size
+    s"This number is $n digits long"
+// end section regexRational
+  }
 
   def main(args: Array[String]): Unit = {
     val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

@@
 \noindent
-Here, our implementation instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which it then uses in |unapply|'s return type to obtain a precise representation of the regex's capture groups.
+Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capture groups.
 
@@
 
-The main challenge...
+The purpose of our type-level component is two-fold:
+
+\begin{enumerate}
+  \item Identify the capture groups of the input regular expression.
+  \item Determine which of those capture group are optional, that is, whether or not the regex engine can assign null values for each of those groups.
+\end{enumerate}
+
+\noindent
+The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| element, depending of the optionality analysis.
+
+We present our implementation incrementally, starting from simple but incomplete solution, and progressively building up towards our final solution.
+The first version of our type-level program is limited to capture group identification.
+The implementation is straightforward:
+
+\begin{lstlisting}
+import compiletime.ops.string.{Length, CharAt}
+import compiletime.ops.int.+
+
+type Compile[R <: String] =
+  Reverse[Loop[R, 0, Length[R], EmptyTuple]]
+
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
+  Lo match
+    case Hi => Acc
+    case _   => CharAt[R, Lo] match
+      case '('  => Loop[R, Lo + 1, Hi, String *: Acc]
+      case '\\' => Loop[R, Lo + 2, Hi, Acc]
+      case _ => Loop[R, Lo + 1, Hi, Acc]
+\end{lstlisting}
+
+\noindent
+The |Length| and |CharAt| types are special cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
+Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integer just like we would at the term level.
+The |EmptyTuple| and |*:| are Scala~3's new generic representation of tuples.
+Those types remove the need for heterogeneous lists and allow us to manipulate tuples directly.
+
+Attentive readers might have noticed that our handling of backslash, regex's escape character, might lead to off-by-one errors.
+Be reassured, this is not an oversight!
+Regular expressions with trailing backslashes result in runtime exceptions (reported as "unexpected internal error").
+Even though detecting there type of error at compile time is out of the scope of our implementation, doing so is still desirable.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into an call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error.
+
+Our first implementation of |Compile| treats every opening parethesis as the start of a capture group, which does no do honor syntax of Java's regular expression.
+Indeed, Java also support several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+
+\regexIsCapturing
+% type IsCapturing[R <: String, At <: Int] =
+%   CharAt[R, At] match
+%     case "?" => CharAt[R, At + 1] match
+%       case "<" => CharAt[R, At + 2] match
+%         case "!" => false // zero-width negative lookbehind
+%         case _ => true    // named-capturing group
+%       case _ => false     // other non-capturing special constructs
+%     case _ => true        // unnamed-capturing group
 

### 2022-02-08 scala/src/main/scala/regex.scala

@@ -24,150 +24,168 @@ class Regex[P] private (val regex: String):
 // end section regexUserLevel
 
 object Lib {
-  type CharAt[R <: String, At <: Int] =
-    Substring[R, At, At + 1]
 
-  type Compile[R <: String] =
-    Reverse[Loop[R, 0, Length[R], EmptyTuple, 0]]
+type CharAt[R <: String, At <: Int] =
+  Substring[R, At, At + 1]
 
-  type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
-    Lo match
-      case Hi => Acc
-      case _   => CharAt[R, Lo] match
-        case ")"  =>
-          Loop[R, Lo + 1, Hi, Acc, Max[0, Nw - 1]]
-        case "("  =>
-          Nw match
-            case 0 =>
-              Nullable[R, Lo + 1, Hi] match
-                case true =>
-                  Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
-                case false =>
-                  Loop[R, Lo + 1, Hi, String *: Acc, 0]
-            case _ =>
-              Loop[R, Lo + 1, Hi, Option[String] *: Acc, Nw + 1]
-        case "\\" =>
-          Lo + 1 match
-            case Hi => Acc
-            case _ => Loop[R, Lo + 2, Hi, Acc, Nw]
-        case _ => Loop[R, Lo + 1, Hi, Acc, Nw]
+type Compile[R <: String] =
+  Reverse[Loop[R, 0, Length[R], EmptyTuple, 0]]
 
-  type Nullmark[R <: String, At <: Int, Hi <: Int] =
-    At match
-      case Hi => false
-      case _ =>
-        CharAt[R, At] match
-          case "?" | "*" | "|" => true
-          case "+" => Nullmark[R, At + 1, Hi]
-          case _ => false
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
+  Lo match
+    case Hi => Acc
+    case _   => CharAt[R, Lo] match
+      case ")"  =>
+        Loop[R, Lo + 1, Hi, Acc, Max[0, Nw - 1]]
+      case "("  =>
+        Nw match
+          case 0 =>
+            Nullable[R, Lo + 1, Hi] match
[…]

### 2022-02-08 scala/src/test/scala/regex-tests.scala

@@ -18,6 +18,11 @@ object Tests {
     check["(B)|C", O]
     check["C|(A)", O]
     check[".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*", (S, S, O)]
+
+    check["((?:A)(B(C)))", (S, S, S)]
+    check["((A)(?:B(C)))?", (O, O, O)]
+    check["((A)(B(?:C))?)", (S, S, O)]
+    check["(?:(A)|(B(C)))", (O, O, O)]
   }
 
   def assertEquals(expected: Any, actual: Any): Unit =

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes typos

@@
 
-We begin our presentation with the user-level interface, which takes the form of relatively simple Scala class and companion object:
+We begin our presentation with the user-level interface, which takes the form of a relatively simple Scala class and companion object:
 
@@
 The implementation of the |Regex| class is straightforward.
-Similarly its counterpart in the original Scala implementation, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
+Similarly to its counterpart in the original Scala implementation, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
 The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.
 
-The main difference between Scala's original |Regex| implementation and ours lies in the signature the |unapply| method, which determines the nature of extraction performed by that method.
+The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
 By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capture groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
@@
 term-level capture group processing component.
-As a result, our implementation never returns null values, which removes the null-checking burden of from the user.
+As a result, our implementation never returns null values, which removes the null-checking burden from the user.
 
@@
   \item Identify the capture groups of the input regular expression.
-  \item Determine which of those capture group are optional, that is, whether or not the regex engine can assign null values for each of those groups.
+  \item Determine which of those capture groups are optional, that is, whether or not the regex engine can assign null values for each of those groups.
 \end{enumerate}
@@
 
-We present our implementation incrementally, starting from simple but incomplete solution, and progressively building up towards our final solution.
+We present our implementation incrementally, starting from a simple but incomplete solution, and progressively building up towards our final solution.
 The first version of our type-level program is limited to capture group identification.
@@
 \noindent
-The |Length| and |CharAt| types are special cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
+The |Length| and |CharAt| types are special-cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
 Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integer just like we would at the term level.
@@
 Regular expressions with trailing backslashes result in runtime exceptions (reported as "unexpected internal error").
-Even though detecting there type of error at compile time is out of the scope of our implementation, doing so is still desirable.
-If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into an call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error.
+Even though detecting these types of errors at compile time is out of the scope of our implementation, doing so is still desirable.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error.
 
-Our first implementation of |Compile| treats every opening parethesis as the start of a capture group, which does no do honor syntax of Java's regular expression.
+Our first implementation of |Compile| treats every opening parenthesis as the start of a capture group, which does not do honor to the syntax of Java's regular expression.
 Indeed, Java also support several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes typos

@@
 \noindent
-The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| element, depending of the optionality analysis.
+The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending of the optionality analysis.
 
@@
 The |Length| and |CharAt| types are special-cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
-Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integer just like we would at the term level.
+Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integers just like we would at the term level.
 The |EmptyTuple| and |*:| are Scala~3's new generic representation of tuples.
@@
 Our first implementation of |Compile| treats every opening parenthesis as the start of a capture group, which does not do honor to the syntax of Java's regular expression.
-Indeed, Java also support several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+Indeed, Java also supports several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
 

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

--- 2022-02-08 symposium.tex
@@
 The implementation of the |Regex| class is straightforward.
-Similarly to its counterpart in the original Scala implementation, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
+Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
 The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.
@@
 \noindent
-The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| element, depending of the optionality analysis.
+The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending of the nullability analysis.
 
 We present our implementation incrementally, starting from a simple but incomplete solution, and progressively building up towards our final solution.
+
+\subsection{Capture Group Identification}
+
 The first version of our type-level program is limited to capture group identification.
@@
 
-\begin{lstlisting}
-import compiletime.ops.string.{Length, CharAt}
-import compiletime.ops.int.+
-
-type Compile[R <: String] =
-  Reverse[Loop[R, 0, Length[R], EmptyTuple]]
-
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
-  Lo match
-    case Hi => Acc
-    case _   => CharAt[R, Lo] match
-      case '('  => Loop[R, Lo + 1, Hi, String *: Acc]
-      case '\\' => Loop[R, Lo + 2, Hi, Acc]
-      case _ => Loop[R, Lo + 1, Hi, Acc]
-\end{lstlisting}
+\regexFirstIteration
+% import compiletime.ops.string._
+% import compiletime.ops.int.+
+%
+% type Compile[R <: String] =
+%   Reverse[Loop[R, 0, Length[R], EmptyTuple]]
+%
+% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
+%   Lo match
+%     case Hi => Acc
+%     case _   => CharAt[R, Lo] match
+%       case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
+%       case "\\" => Loop[R, Lo + 2, Hi, Acc]
+%       case _ => Loop[R, Lo + 1, Hi, Acc]
 
@@
 The |Length| and |CharAt| types are special-cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
-Similarly, the |+| type from |compiletime.ops.int| allows us to manipulate integer just like we would at the term level.
+Similarly, the |+| type (from |compiletime.ops.int|) allows us to manipulate integers just like we would at the term level.
 The |EmptyTuple| and |*:| are Scala~3's new generic representation of tuples.
-Those types remove the need for heterogeneous lists and allow us to manipulate tuples directly.
+Those types remove the need for heterogeneous lists and lets us manipulate tuples directly.
+
+\subsection{Out-Of-Bound Errors}
 
@@
 
+\subsection{Ruling Out Non-Capturing}
+
 Our first implementation of |Compile| treats every opening parenthesis as the start of a capture group, which does not do honor to the syntax of Java's regular expression.
-Indeed, Java also support several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, and must all be discriminated from named and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+Indeed, Java also supports several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, which must all be discriminated from named- and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+Our second implementation (omitted) uses the following |IsCapturing| predicate type to rule out non-capturing groups:
 
 \regexIsCapturing
-% type IsCapturing[R <: String, At <: Int] =
+% type IsCapturing[R <: String, At <: Int] <: Boolean =
 %   CharAt[R, At] match
@@
 %     case _ => true        // unnamed-capturing group
+
+\noindent
+Similar to our handling of backslash characters, this implementation intentionally does not guard against out-of-bounds errors, since these errors correspond to ill-formed regular expressions.
+
+\subsection{Nullability Analysis}
+
+\regexIsNullable
+% type IsNullable[R <: String, At <: Int, Hi <: Int, Ops <: Int] <: Boolean =
+%   CharAt[R, At] match
+%     case ")" => Ops match
+%       case 0 => IsMarked[R, At + 1, Hi]
+%       case _ => IsNullable[R, At + 1, Hi, Ops - 1]
+%     case "(" => IsNullable[R, At + 1, Hi, Ops + 1]
+%     case "\\" => IsNullable[R, At + 2, Hi, Ops]
+%     case _    => IsNullable[R, At + 1, Hi, Ops]
+%
+% type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+%   At match
+%     case Hi => false
+%     case _ =>
+%       CharAt[R, At] match
+%         case "?" | "*" | "|" => true
+%         case "+" => IsMarked[R, At + 1, Hi]
+%         case _ => false
 

### 2022-02-08 scala/src/main/scala/regex.scala

@@ -35,12 +35,21 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
   Lo match
     case Hi => Acc
     case _   => CharAt[R, Lo] match
-      case ")"  =>
+      case "|" =>
+        CharAt[R, Lo + 1] match {
+          case "(" =>
+            IsCapturing[R, Lo + 2] match {
+              case false => Loop[R, Lo + 2, Hi, Acc, Nw + 1]
+              case true  => Loop[R, Lo + 2, Hi, Option[String] *: Acc, Nw + 1]
+            }
+          case _ => Loop[R, Lo + 2, Hi, Acc, Nw]
+        }
+      case ")" =>
         Loop[R, Lo + 1, Hi, Acc, Max[0, Nw - 1]]
-      case "("  =>
+      case "(" =>
         Nw match
           case 0 =>
-            Nullable[R, Lo + 1, Hi] match
+            IsNullable[R, Lo + 1, Hi, 0] match
               case true =>
                 IsCapturing[R, Lo + 1] match {
                   case false => Loop[R, Lo + 1, Hi, Acc, 1]
@@ -60,7 +69,7 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
       case _ => Loop[R, Lo + 1, Hi, Acc, Nw]
 
 // start section regexIsCapturing
-type IsCapturing[R <: String, At <: Int] =
+type IsCapturing[R <: String, At <: Int] <: Boolean =
   CharAt[R, At] match
     case "?" => CharAt[R, At + 1] match
       case "<" => CharAt[R, At + 2] match
@@ -70,33 +79,25 @@ type IsCapturing[R <: String, At <: Int] =
     case _ => true        // unnamed-capturing group
 // end section regexIsCapturing
 
-type Nullmark[R <: String, At <: Int, Hi <: Int] =
+// start section regexIsNullable
+type IsNullable[R <: String, At <: Int, Hi <: Int, Ops <: Int] <: Boolean =
+  CharAt[R, At] match
+    case ")" => Ops match
+      case 0 => IsMarked[R, At + 1, Hi]
+      case _ => IsNullable[R, At + 1, Hi, Ops - 1]
+    case "(" => IsNullable[R, At + 1, Hi, Ops + 1]
[…]

### 2022-02-08 scala/src/test/scala/regex-tests.scala

@@ -45,6 +45,31 @@ val rational = Regex("(\\d+)\\.?(\\d+)?")
     val n = i.size + f.size
     s"This number is $n digits long"
 // end section regexRational
+
+import Lib.{CharAt, Reverse}
+
+// start section regexFirstIteration
+import compiletime.ops.string._
+import compiletime.ops.int.+
+
+type Compile[R <: String] =
+  Reverse[Loop[R, 0, Length[R], EmptyTuple]]
+
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
+  Lo match
+    case Hi => Acc
+    case _   => CharAt[R, Lo] match
+      case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
+      case "\\" => Loop[R, Lo + 2, Hi, Acc]
+      case _ => Loop[R, Lo + 1, Hi, Acc]
+// end section regexFirstIteration
+
+def check[A <: String, B](implicit ev: Compile[A] =:= B): Unit = ()
+type S = String
+check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
+check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
+check["((A)(B(C)))", (S, S, S, S)]
+check["((B(C)))", (S, S, S)]
   }
 
   def main(args: Array[String]): Unit = {

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

@@
 \noindent
-The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capture groups, and shields its users from off-by-one error.
+The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields its users from off-by-one error.
 
@@
 First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
-Furthermore, the values coming out of capture groups can be null (when using optional captures) and thus require an additional layer of validation.
+Furthermore, the values coming out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
 Those shortcomings might appear benign when looking at small examples, but can easily turn into bugs when dealing with a large-scale code base.
 
-In this chapter, we propose a new design for Scala's regular expression library which provides both type-safe and null-safe capture group extraction.
+In this chapter, we propose a new design for Scala's regular expression library which provides both type-safe and null-safe capturing group extraction.
 Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
@@
 \begin{enumerate}
-  \item The type-level capture group analysis, which uses match types to inspect the user-provided regular expression and compute the corresponding tuple-based type representation (\Cref{sec:type-level}).
+  \item The type-level capturing group analysis, which uses match types to inspect the user-provided regular expression and compute the corresponding tuple-based type representation (\Cref{sec:type-level}).
 
-  \item The term-level capture group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).
+  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).
 
-  \item The user-level interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capture group extraction.
+  \item The user-level interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capturing group extraction.
 \end{enumerate}
@@
 Instead of directly accepting |String| argument, that method requires a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression (the |Singleton| bound is necessary to guide Scala's type inference).
-We use that type-level string immediately to instantiate |Compile|, which is our type-level capture group analysis component.
+We use that type-level string immediately to instantiate |Compile|, which is our type-level capturing group analysis component.
 We store the output of that analysis as |Regex|'s type argument.
@@
 The implementation of the |Regex| class is straightforward.
-Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capture groups whenever the matching succeeds.
+Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capturing groups whenever the matching succeeds.
 The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.
@@
 The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
-By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capture groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
+By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
 We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
 In our implementation, this is done via the |toTuple| method, which is our
-term-level capture group processing component.
+term-level capturing group processing component.
 As a result, our implementation never returns null values, which removes the null-checking burden from the user.
 
-For instance, in the following example, we use a regular expression with an optional capture group to extract the integral and fractional parts of a rational number:
+For instance, in the following example, we use a regular expression with an optional capturing group to extract the integral and fractional parts of a rational number:
 
@@
 \noindent
-Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capture groups.
+Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
 
@@
 \begin{enumerate}
-  \item Identify the capture groups of the input regular expression.
-  \item Determine which of those capture groups are optional, that is, whether or not the regex engine can assign null values for each of those groups.
+  \item Identify the capturing groups of the input regular expression.
+  \item Determine which of those capturing groups are optional, that is, whether or not the regex engine can assign null values for each of those groups.
 \end{enumerate}
@@
 
-The first version of our type-level program is limited to capture group identification.
+The first version of our type-level program is limited to capturing group identification.
 The implementation is straightforward:
@@
 
-\subsection{Ruling Out Non-Capturing}
+\subsection{Non-Capturing Groups}
 
-Our first implementation of |Compile| treats every opening parenthesis as the start of a capture group, which does not do honor to the syntax of Java's regular expression.
-Indeed, Java also supports several other special constructs which start with an opening parethesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, which must all be discriminated from named- and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not do honor to the syntax of Java's regular expression.
+Indeed, Java also supports several other special constructs which start with an opening parenthesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, which must all be discriminated from named- and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
 Our second implementation (omitted) uses the following |IsCapturing| predicate type to rule out non-capturing groups:
@@
 
-\regexIsNullable
-% type IsNullable[R <: String, At <: Int, Hi <: Int, Ops <: Int] <: Boolean =
+Establishing the nullability of capturing groups is more complicated than it sounds.
+At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{For brevity, our presentation does not account for the regex alternative operator (\textbar) which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
+In our implementation, we do account for that operator, which we handle analogously.}:
+
+\regexNaiveIsNullable
+% type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
 %   CharAt[R, At] match
-%     case ")" => Ops match
-%       case 0 => IsMarked[R, At + 1, Hi]
-%       case _ => IsNullable[R, At + 1, Hi, Ops - 1]
-%     case "(" => IsNullable[R, At + 1, Hi, Ops + 1]
-%     case "\\" => IsNullable[R, At + 2, Hi, Ops]
-%     case _    => IsNullable[R, At + 1, Hi, Ops]
+%     case ")" => IsMarked[R, At + 1, Hi]
+%     case _   => IsNullable[R, At + 1, Hi]
 %
@@
 %       CharAt[R, At] match
-%         case "?" | "*" | "|" => true
+%         case "?" | "*" => true
 %         case "+" => IsMarked[R, At + 1, Hi]
 %         case _ => false
+
+\noindent
+Unfortunately, this naive solution is incorrect for regular expressions with nested capturing groups.
+For example, in |(A(B)?)|, our naive solution incorrectly labels the first group as optional.
+To overcome this problem, we add a parameter to the definition of |IsNullable|, which we use keep to track of the nesting level:
+
+\regexIsNullable
+% type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
+%   CharAt[R, At] match
+%     case ")" => Lvl match
+%       case 0 => IsMarked[R, At + 1, Hi]
+%       case _ => IsNullable[R, At + 1, Hi, Lvl - 1]
+%     case "(" => IsNullable[R, At + 1, Hi, Lvl + 1]
+%     case "\\" => IsNullable[R, At + 2, Hi, Lvl]
+%     case _    => IsNullable[R, At + 1, Hi, Lvl]
+
+Nested capturing groups bring another complication to the nullability analysis, caused by the interaction between the nullability of inner and outer groups.
+When an outer group is deemed optional, it overrides the nullability of all its inner groups, which also become optional.
+For instance, in |(A(B))?|, both the first (|A(B)|) and the second (|B|) capturing group are optional.
 

### 2022-02-08 scala/src/main/scala/regex.scala

@@ -31,7 +31,7 @@ type CharAt[R <: String, At <: Int] =
 type Compile[R <: String] =
   Reverse[Loop[R, 0, Length[R], EmptyTuple, 0]]
 
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] <: Tuple =
   Lo match
     case Hi => Acc
     case _   => CharAt[R, Lo] match
@@ -39,15 +39,15 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
         CharAt[R, Lo + 1] match {
           case "(" =>
             IsCapturing[R, Lo + 2] match {
-              case false => Loop[R, Lo + 2, Hi, Acc, Nw + 1]
-              case true  => Loop[R, Lo + 2, Hi, Option[String] *: Acc, Nw + 1]
+              case false => Loop[R, Lo + 2, Hi, Acc, Lvl + 1]
+              case true  => Loop[R, Lo + 2, Hi, Option[String] *: Acc, Lvl + 1]
             }
-          case _ => Loop[R, Lo + 2, Hi, Acc, Nw]
+          case _ => Loop[R, Lo + 2, Hi, Acc, Lvl]
         }
       case ")" =>
-        Loop[R, Lo + 1, Hi, Acc, Max[0, Nw - 1]]
+        Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
       case "(" =>
-        Nw match
+        Lvl match
           case 0 =>
             IsNullable[R, Lo + 1, Hi, 0] match
               case true =>
@@ -62,11 +62,11 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Nw <: Int] <: Tuple =
                 }
           case _ =>
             IsCapturing[R, Lo + 1] match {
-              case false => Loop[R, Lo + 1, Hi, Acc, Nw + 1]
-              case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Nw + 1]
+              case false => Loop[R, Lo + 1, Hi, Acc, Lvl + 1]
+              case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
             }
[…]

### 2022-02-08 scala/src/test/scala/regex-tests.scala

@@ -70,6 +70,46 @@ check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
 check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
 check["((A)(B(C)))", (S, S, S, S)]
 check["((B(C)))", (S, S, S)]
+
+object NaiveIsNullable {
+// start section regexNaiveIsNullable
+type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
+  CharAt[R, At] match
+    case ")" => IsMarked[R, At + 1, Hi]
+    case _   => IsNullable[R, At + 1, Hi]
+
+type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+  At match
+    case Hi => false
+    case _ =>
+      CharAt[R, At] match
+        case "?" | "*" => true
+        case "+" => IsMarked[R, At + 1, Hi]
+        case _ => false
+// end section regexNaiveIsNullable
+}
+
+object Last {
+import compiletime.ops.int.Max
+import compiletime.ops.int.-
+import Lib.IsNullable
+// start section regexLastIteration
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] =
+  Lo match
+    case Hi => Acc
+    case _   => CharAt[R, Lo] match
+      case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
+      case "(" =>
+        Lvl match
+          case 0 => IsNullable[R, Lo + 1, Hi, 0] match
+            case true =>  Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
+            case false => Loop[R, Lo + 1, Hi, Acc, 0]
+          case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
+      case "\\" => Loop[R, Lo + 2, Hi, Acc, Lvl]
+      case _ => Loop[R, Lo + 1, Hi, Acc, Lvl]
+// end section regexLastIteration
+}
+
   }
 
   def main(args: Array[String]): Unit = {

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

@@
 
-  \item The user-level interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capturing group extraction.
+  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capturing group extraction.
 \end{enumerate}
 
-We begin our presentation with the user-level interface, which takes the form of a relatively simple Scala class and companion object:
+We begin our presentation with the user interface, which takes the form of a relatively simple Scala class and companion object:
 
@@
 The first version of our type-level program is limited to capturing group identification.
-The implementation is straightforward:
+The implementation is straightforward, it iterates through the regex's characters and accumulates a |String| type for every opening parenthesis:
 
@@
 Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not do honor to the syntax of Java's regular expression.
-Indeed, Java also supports several other special constructs which start with an opening parenthesis, such as non-capturing groups, |(?:X)|, lookaheads, |(?=X)| \& |(?!X)|, and lookbehinds, |(?<=X)| \& |(?<!X)|, which must all be discriminated from named- and unnamed-capturing groups, |(?<name>X)| \& |(X)|.
+Indeed, Java also supports several other special constructs which start with an opening parenthesis, for instance:
+
+\begin{itemize}
+  \item non-capturing groups, |(?:X)|,
+  \item lookaheads, |(?=X)| and |(?!X)|,
+  \item lookbehinds, |(?<=X)| and |(?<!X)|.
+\end{itemize}
+
+\noindent
+To correctly identify capturing groups (which can be either named, |(?<name>X)|, or unnamed, |(X)|), we must differentiate them from other special constructs.
 Our second implementation (omitted) uses the following |IsCapturing| predicate type to rule out non-capturing groups:
@@
 Establishing the nullability of capturing groups is more complicated than it sounds.
-At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{For brevity, our presentation does not account for the regex alternative operator (\textbar) which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
-In our implementation, we do account for that operator, which we handle analogously.}:
+At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{
+For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
+In our implementation, we do account for that operator, which we handle analogously.
+}.
+For instance, in the following naive implementation of |IsNullable|, we look for the first closing parenthesis and inspect the following characters to determine if a capturing group is nullable:
 
@@
 Unfortunately, this naive solution is incorrect for regular expressions with nested capturing groups.
-For example, in |(A(B)?)|, our naive solution incorrectly labels the first group as optional.
-To overcome this problem, we add a parameter to the definition of |IsNullable|, which we use keep to track of the nesting level:
+For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
+To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner groups from the closing parenthesis of the outer group:
 
@@
 
-Nested capturing groups bring another complication to the nullability analysis, caused by the interaction between the nullability of inner and outer groups.
-When an outer group is deemed optional, it overrides the nullability of all its inner groups, which also become optional.
+Nested capturing groups bring another complication to the nullability analysis, which is caused by the interaction between inner and outer groups.
+When an outer group is deemed optional, this overrides the nullability of all its inner groups, which also become optional.
 For instance, in |(A(B))?|, both the first (|A(B)|) and the second (|B|) capturing group are optional.
+
+The final iteration of our type-level program handles nested groups by operating in two different modes.
+While iterating through the regex's characters, our algorithm can either be outside of an optional group (|Lvl=0|), or inside an option group (|Lvl>0|), in which case it overrides the nullability of newly encountered groups.
+Our implementation is as follows (|IsCapturing| omitted for brevity):
+
+\regexLastIteration
+% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] =
+%   Lo match
+%     case Hi => Acc
+%     case _  => CharAt[R, Lo] match
+%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
+%       case "(" =>
+%         Lvl match
+%           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
+%             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
+%             case false => Loop[R, Lo + 1, Hi, Acc, 0]
+%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
+%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Lvl]
+%       case _ => Loop[R, Lo + 1, Hi, Acc, Lvl]
 

### 2022-02-08 scala/src/main/scala/regex.scala

@@ -34,7 +34,7 @@ type Compile[R <: String] =
 type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] <: Tuple =
   Lo match
     case Hi => Acc
-    case _   => CharAt[R, Lo] match
+    case _  => CharAt[R, Lo] match
       case "|" =>
         CharAt[R, Lo + 1] match {
           case "(" =>
@@ -73,10 +73,10 @@ type IsCapturing[R <: String, At <: Int] <: Boolean =
   CharAt[R, At] match
     case "?" => CharAt[R, At + 1] match
       case "<" => CharAt[R, At + 2] match
-        case "!" => false // zero-width negative lookbehind
-        case _ => true    // named-capturing group
-      case _ => false     // other non-capturing special constructs
-    case _ => true        // unnamed-capturing group
+        case "=" | "!" => false // lookbehinds
+        case _ => true          // named-capturing group
+      case _ => false           // other special constructs
+    case _ => true              // unnamed-capturing group
 // end section regexIsCapturing
 
 // start section regexIsNullable

### 2022-02-08 scala/src/test/scala/regex-examples.scala

@@ -1,0 +1,86 @@
+import annotation.experimental
+import Package._
+import Lib.{CharAt, Reverse}
+import compiletime.ops.int.Max
+import compiletime.ops.int.-
+
+@experimental
+object Examples {
+
+// start section regexDocumentation
+val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
+"2004-01-20" match
+  case date(y, m, d) =>
+    s"$y was a good year for PLs."
+// end section regexDocumentation
+
+// start section regexRational
+val rational = Regex("(\\d+)\\.?(\\d+)?")
+"3.1415" match
+  case rational(i, Some(f)) =>
+    val n = i.size + f.size
+    s"This number is $n digits long"
+// end section regexRational
+
+// start section regexFirstIteration
+import compiletime.ops.string._
+import compiletime.ops.int.+
+
+type Compile[R <: String] =
+  Reverse[Loop[R, 0, Length[R], EmptyTuple]]
+
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
+  Lo match
+    case Hi => Acc
+    case _  => CharAt[R, Lo] match
+      case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
+      case "\\" => Loop[R, Lo + 2, Hi, Acc]
+      case _ => Loop[R, Lo + 1, Hi, Acc]
+// end section regexFirstIteration
+
+def check[A <: String, B](implicit ev: Compile[A] =:= B): Unit = ()
+type S = String
+check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
+check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
+check["((A)(B(C)))", (S, S, S, S)]
+check["((B(C)))", (S, S, S)]
+
+object NaiveIsNullable {
+// start section regexNaiveIsNullable
+type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
+  CharAt[R, At] match
+    case ")" => IsMarked[R, At + 1, Hi]
+    case _   => IsNullable[R, At + 1, Hi]
+
+type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+  At match
+    case Hi => false
+    case _ =>
+      CharAt[R, At] match
+        case "?" | "*" => true
[…]

### 2022-02-08 scala/src/test/scala/regex-tests.scala

@@ -30,88 +30,6 @@ object Tests {
       expected == actual,
       s"expected: $expected, actual: $actual")
 
-  def examples: Unit = {
-// start section regexDocumentation
-val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
-"2004-01-20" match
-  case date(y, m, d) =>
-    s"$y was a good year for PLs."
-// end section regexDocumentation
-
-// start section regexRational
-val rational = Regex("(\\d+)\\.?(\\d+)?")
-"3.1415" match
-  case rational(i, Some(f)) =>
-    val n = i.size + f.size
-    s"This number is $n digits long"
-// end section regexRational
-
-import Lib.{CharAt, Reverse}
-
-// start section regexFirstIteration
-import compiletime.ops.string._
-import compiletime.ops.int.+
-
-type Compile[R <: String] =
-  Reverse[Loop[R, 0, Length[R], EmptyTuple]]
-
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
-  Lo match
-    case Hi => Acc
-    case _   => CharAt[R, Lo] match
-      case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
-      case "\\" => Loop[R, Lo + 2, Hi, Acc]
-      case _ => Loop[R, Lo + 1, Hi, Acc]
-// end section regexFirstIteration
-
-def check[A <: String, B](implicit ev: Compile[A] =:= B): Unit = ()
-type S = String
-check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
-check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
-check["((A)(B(C)))", (S, S, S, S)]
-check["((B(C)))", (S, S, S)]
-
-object NaiveIsNullable {
-// start section regexNaiveIsNullable
-type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
-  CharAt[R, At] match
-    case ")" => IsMarked[R, At + 1, Hi]
-    case _   => IsNullable[R, At + 1, Hi]
-
-type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
-  At match
-    case Hi => false
-    case _ =>
-      CharAt[R, At] match
-        case "?" | "*" => true
-        case "+" => IsMarked[R, At + 1, Hi]
[…]

### 2022-02-08 symposium.tex
Type and Null Safety for Regexes

@@
 
-The Scala standard library includes a package to manipulate regular expressions.
-This package is based on Java's implementation of regular expression, and thus benefits from the high performance of the JVM's regex engine.
-The Scala implementation innovates with its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
+The Scala standard library contains a package to manipulate regular expressions.
+This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
+The Scala implementation innovates in its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
 The package's documentation starts with the following example:
@@
 \noindent
-The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields its users from off-by-one error.
+The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.
 
@@
 First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
-Furthermore, the values coming out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
-Those shortcomings might appear benign when looking at small examples, but can easily turn into bugs when dealing with a large-scale code base.
+Furthermore, the values that come out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
+Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale code base.
 
-In this chapter, we propose a new design for Scala's regular expression library which provides both type-safe and null-safe capturing group extraction.
+In this chapter, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
 Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
-We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement while still enjoying additional safety.
+We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
 
@@
 
-  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed by the type-level component (\Cref{sec:term-level}).
+  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed at the type level (\Cref{sec:term-level}).
 
-  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe capturing group extraction.
+  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe experience.
 \end{enumerate}
@@
 It takes a regular expression and returns an instance of |Regex|.
-Instead of directly accepting |String| argument, that method requires a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression (the |Singleton| bound is necessary to guide Scala's type inference).
-We use that type-level string immediately to instantiate |Compile|, which is our type-level capturing group analysis component.
-We store the output of that analysis as |Regex|'s type argument.
+Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string (the |Singleton| bound is necessary to guide Scala's type inference).
+We use that type-level string to instantiate |Compile|, our type-level analysis component.
 
 The implementation of the |Regex| class is straightforward.
-Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the regular expression, and to extract the capturing groups whenever the matching succeeds.
-The |Regex| class defines an |unapply| method, which the Scala compiler will invoke when using an instance of that class as a pattern matching extractor \citep{emir2007matching}.
+Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against its regular expression, and to extract the capturing groups when the matching succeeds.
+The |Regex| class defines an |unapply| method, which the Scala compiler invokes when using instances of that class as pattern matching extractors \citep{emir2007matching}.
 
@@
 
-\subsection{Capture Group Identification}
+\subsection{Capturing Group Identification}
 
@@
 For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
-In our implementation, we do account for that operator, which we handle analogously.
+Similarly, we omit handling the "at least n times" quantifiers which can lead to nullable capturing groups.
+Our implementation accounts for both operators.
 }.
@@
 For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
-To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner groups from the closing parenthesis of the outer group:
+To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups:
 
@@
 
-The final iteration of our type-level program handles nested groups by operating in two different modes.
-While iterating through the regex's characters, our algorithm can either be outside of an optional group (|Lvl=0|), or inside an option group (|Lvl>0|), in which case it overrides the nullability of newly encountered groups.
-Our implementation is as follows (|IsCapturing| omitted for brevity):
+We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
+The algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
+This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.
+
+With this latest improvement, we develop our implementation into its final iteration (|IsCapturing| omitted for brevity):
 
 \regexLastIteration
-% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] =
+% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
 %   Lo match
@@
 %     case _  => CharAt[R, Lo] match
-%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
+%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
 %       case "(" =>
-%         Lvl match
+%         Opt match
 %           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
@@
 %             case false => Loop[R, Lo + 1, Hi, Acc, 0]
-%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
-%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Lvl]
-%       case _ => Loop[R, Lo + 1, Hi, Acc, Lvl]
+%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
+%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
+%       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]
+
+This concludes the development of our nullability analysis and completes the presentation of the type-level component of our library.
+Despite the scarcity of Scala's standard library at the type level, we managed to achieve our ends while keeping our implementation relatively concise and, hopefully, understandable.
+In its final iteration, our capturing group analysis is, to the best of our knowledge, on par with Java's implementation of regular expression.
 

### 2022-02-08 scala/src/main/scala/regex.scala

@@ -31,7 +31,7 @@ type CharAt[R <: String, At <: Int] =
 type Compile[R <: String] =
   Reverse[Loop[R, 0, Length[R], EmptyTuple, 0]]
 
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] <: Tuple =
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] <: Tuple =
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
@@ -39,15 +39,15 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] <: Tuple 
         CharAt[R, Lo + 1] match {
           case "(" =>
             IsCapturing[R, Lo + 2] match {
-              case false => Loop[R, Lo + 2, Hi, Acc, Lvl + 1]
-              case true  => Loop[R, Lo + 2, Hi, Option[String] *: Acc, Lvl + 1]
+              case false => Loop[R, Lo + 2, Hi, Acc, Opt + 1]
+              case true  => Loop[R, Lo + 2, Hi, Option[String] *: Acc, Opt + 1]
             }
-          case _ => Loop[R, Lo + 2, Hi, Acc, Lvl]
+          case _ => Loop[R, Lo + 2, Hi, Acc, Opt]
         }
       case ")" =>
-        Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
+        Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
       case "(" =>
-        Lvl match
+        Opt match
           case 0 =>
             IsNullable[R, Lo + 1, Hi, 0] match
               case true =>
@@ -62,11 +62,11 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] <: Tuple 
                 }
           case _ =>
             IsCapturing[R, Lo + 1] match {
-              case false => Loop[R, Lo + 1, Hi, Acc, Lvl + 1]
-              case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
+              case false => Loop[R, Lo + 1, Hi, Acc, Opt + 1]
+              case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
             }
[…]

### 2022-02-08 scala/src/test/scala/regex-examples.scala

@@ -67,19 +67,19 @@ object Last {
 import Lib.IsNullable
 
 // start section regexLastIteration
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Lvl <: Int] =
+type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
-      case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Lvl - 1]]
+      case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
       case "(" =>
-        Lvl match
+        Opt match
           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
             case false => Loop[R, Lo + 1, Hi, Acc, 0]
-          case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Lvl + 1]
-      case "\\" => Loop[R, Lo + 2, Hi, Acc, Lvl]
-      case _ => Loop[R, Lo + 1, Hi, Acc, Lvl]
+          case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
+      case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
+      case _ => Loop[R, Lo + 1, Hi, Acc, Opt]
 // end section regexLastIteration
 }
 

### 2022-02-09 article.tex
We Ain't Need No Dependent Types!

@@
 \subsection{A Lightweight Form of Dependent Typing}
+\label{subsec:a-lightweight-form-of-dependent-typing}
+
 Match types enable a lightweight form of dependent typing, since term-level pattern matching expressions can be typed accordingly at the type level as a match types.

### 2022-02-09 symposium.tex
We Ain't Need No Dependent Types!

@@
 
-  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the representation computed at the type level (\Cref{sec:term-level}).
+  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed (\Cref{sec:term-level}).
 
@@
 %     if (m.matches())
-%       val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
-%       Some(toTuple[P](regex, a))
+%       val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
+%       Some(transform[P](regex, a))
 %     else
@@
 We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
-In our implementation, this is done via the |toTuple| method, which is our
-term-level capturing group processing component.
+In our implementation, this is done via the |transform| method, which is our
+term-level processing component.
 As a result, our implementation never returns null values, which removes the null-checking burden from the user.
@@
 \subsection{Non-Capturing Groups}
+\label{subsec:non-capturing-groups}
 
@@
 We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
-The algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
+Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
 This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.
 
-With this latest improvement, we develop our implementation into its final iteration (|IsCapturing| omitted for brevity):
+With this latest improvement, we advance our implementation to its final iteration (we omit usages of |IsCapturing|, from \Cref{subsec:non-capturing-groups}, for brevity):
 
@@
 
-In this section...
+The term-level component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
+Concretely, this component's job consists in transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
+That transformation must conform to the representation computed at the type level.
+
+In Scala, all type parameters are erased, which can lead so some friction when trying to write programs whose execution \emph{depends} on a type, such as in the problem at hand.
+In this section, we propose two different solutions to that problem.
+
+\begin{enumerate}
+  \item In \Cref{subsec:we-ain-t-need-no-dependent-types}, we proceed by sheer force of code duplication: we translate the entirety of our type-level algorithm into term-level functions and use a cast to correlate the two.
+
+  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show...
+\end{enumerate}
+
+\subsection{We Ain't Need No Dependent Types!}
+\label{subsec:we-ain-t-need-no-dependent-types}
+
+The first implementation of our term-level component duplicates the type-level definitions presented in \Cref{sec:type-level} to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
+In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of option's constructor:
+
+\regexTermLvlLoop
+% val identity: String => Any = { x => assert(x != null); x }
+% type Accumulator = List[String => Any]
+%
+% def loop(r: String, lo: Int, hi: Int, acc: Accumulator, opt: Int): Accumulator =
+%   lo match
+%     case `hi` => acc
+%     case _  => r.charAt(lo) match
+%       case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
+%       case '(' =>
+%         opt match
+%           case 0 => isNullable(r, lo + 1, hi, 0) match
+%             case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
+%             case false => loop(r, lo + 1, hi, identity :: acc, 0)
+%           case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
+%       case '\\' => loop(r, lo + 2, hi, acc, opt)
+%       case _ => loop(r, lo + 1, hi, acc, opt)
+
+\noindent
+After having computed those functions, we simply do a pointwise application with the raw output of the regex engine, package the result into a tuple, \emph{et voilà!}
+
+\regexTransform
+% def transform[P](pattern: String, arr: Array[String]): P =
+%   val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
+%   val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
+%   assert(arr.size == fs.size, "Unexpected number of capturing groups")
+%   ts.asInstanceOf[P]
+
+\emph{In principle}, Scala~3's type system is capable of doing a lightweight form of dependent typing (\Cref{subsec:a-lightweight-form-of-dependent-typing}), where it...
+
+\begin{lstlisting}
+def isMarked
+  [R <: String, At <: Int, Hi <: Int]
+  (r: R, at: At, hi: Hi)
+  : IsMarked[R, At, Hi]
+\end{lstlisting}
+
+\subsection{Implicit-Based Extractor Synthesis}
+\label{subsec:implicit-based-extractor-synthesis}
 

### 2022-02-09 scala/src/main/scala/regex.scala

@@ -17,8 +17,8 @@ class Regex[P] private (val regex: String):
   def unapply(s: String): Option[P] =
     val m = pattern.matcher(s)
     if (m.matches())
-      val a = Array.tabulate(m.groupCount) { i => m.group(i + 1) }
-      Some(toTuple[P](regex, a))
+      val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
+      Some(transform[P](regex, a))
     else
       None
 // end section regexUserLevel
@@ -96,69 +96,91 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
     case _ =>
       CharAt[R, At] match
         case "?" | "*" | "|" => true
-        case "{" => CharAt[R, At + 1, Hi] match
+        case "{" => CharAt[R, At + 1] match
           case "0" => true
           case _ => false
         case "+" => IsMarked[R, At + 1, Hi]
         case _ => false
 
-def toTuple[P](pattern: String, arr: Array[String]): P = {
+def transform[P](pattern: String, arr: Array[String]): P =
   if (arr.size == 1)
     arr(0).asInstanceOf[P]
   else
     val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
     val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
-    assert(arr.size == fs.size)
+    assert(arr.size == fs.size, "Unexpected number of capturing groups")
     ts.asInstanceOf[P]
-}
 
-def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], nw: Int): List[String => Any] =
+def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[String => Any] =
   lo match
     case `hi` => acc
-    case _   => r.charAt(lo) match
-      case ')'  =>
-        loop(r, lo + 1, hi, acc, 0.max(nw - 1))
-      case '('  =>
-        nw match
+    case _  => r.charAt(lo) match
+      case '|' =>
+        r.charAt(lo + 1) match {
+          case '(' =>
+            isCapturing(r, lo + 2) match {
+              case false => loop(r, lo + 2, hi, acc, opt + 1)
+              case true  => loop(r, lo + 2, hi, Option.apply :: acc, opt + 1)
+            }
+          case _ => loop(r, lo + 2, hi, acc, opt)
+        }
+      case ')' =>
+        loop(r, lo + 1, hi, acc, 0.max(opt - 1))
+      case '(' =>
+        opt match
[…]

### 2022-02-09 scala/src/test/scala/regex-examples.scala

@@ -81,6 +81,36 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
       case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]
 // end section regexLastIteration
+
+def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean = false
+
+// start section regexTermLvlLoop
+val identity: String => Any = { x => assert(x != null); x }
+
+def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int)
+    : List[String => Any] =
+  lo match
+    case `hi` => acc
+    case _  => r.charAt(lo) match
+      case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
+      case '(' =>
+        opt match
+          case 0 => isNullable(r, lo + 1, hi, 0) match
+            case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
+            case false => loop(r, lo + 1, hi, identity :: acc, 0)
+          case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
+      case '\\' => loop(r, lo + 2, hi, acc, opt)
+      case _ => loop(r, lo + 1, hi, acc, opt)
+// end section regexTermLvlLoop
+
+// start section regexTransform
+def transform[P](pattern: String, arr: Array[String]): P =
+  val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
+  val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
+  assert(arr.size == fs.size, "Unexpected number of capturing groups")
+  ts.asInstanceOf[P]
+// end section regexTransform
+
 }
 
 }

### 2022-02-09 symposium.tex
Type and Null Safety for Regexes

@@
 The implementation of the |Regex| class is straightforward.
-Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against its regular expression, and to extract the capturing groups when the matching succeeds.
-The |Regex| class defines an |unapply| method, which the Scala compiler invokes when using instances of that class as pattern matching extractors \citep{emir2007matching}.
+Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the given regular expression, and to extract the content of the capturing groups when the matching succeeds.
+These operations operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.
 
 The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
-By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ corresponds to the tuple arity, and that the type of those groups will correspond to the tuple types.
+By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity, and that the type of those groups corresponds to the type of the tuple elements.
 We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
@@
 
-For instance, in the following example, we use a regular expression with an optional capturing group to extract the integral and fractional parts of a rational number:
+For instance, in the following example usage of our library, we use a regular expression with an optional capturing group to extract the integral and fractional parts of a rational number:
 
@@
 Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
+The resulting code is type-safe: if we change the pattern to omit the option unpacking (replacing |Some(f)| by |f|), the compiler would raise a type error on the call to |.size| (size is defined on strings but not options).
+
+In the following section, we explain our technique to analysis regular expressions and statically compute a type-level representation of the regex's capturing groups (|P| in the example above).
 
@@
 \begin{enumerate}
-  \item Identify the capturing groups of the input regular expression.
-  \item Determine which of those capturing groups are optional, that is, whether or not the regex engine can assign null values for each of those groups.
+  \item Identify the capturing groups of a regular expression.
+  \item Determine which of those capturing groups are optional, that is, whether or not the regex engine could possibly assign null values for each of those groups.
 \end{enumerate}
@@
 \noindent
-The |Length| and |CharAt| types are special-cased by the compiler: when their first argument is a literal singleton string, the compiler evaluates those types via their corresponding term-level implementation.
-Similarly, the |+| type (from |compiletime.ops.int|) allows us to manipulate integers just like we would at the term level.
-The |EmptyTuple| and |*:| are Scala~3's new generic representation of tuples.
-Those types remove the need for heterogeneous lists and lets us manipulate tuples directly.
+The |Length| and |CharAt| types (from |ops.string|) are special-cased by the compiler: when their first argument is a known string literal, the compiler evaluates those types via their corresponding term-level implementation.
+Similarly, the |+| type (from |ops.int|) allows us to manipulate integers just like we would at the term level.
+The |EmptyTuple| and |*:| types are Scala~3's new generic representation of tuples.
+Those types remove the need for heterogeneous lists and let us manipulate tuples directly.
 
@@
 
-Attentive readers might have noticed that our handling of backslash, regex's escape character, might lead to off-by-one errors.
+Attentive readers might have noticed that our handling of backslash, regex's escape character, might result in off-by-one errors.
+Indeed, the |Loop| type terminates when |Lo = Hi|, but increases |Lo| by 2 to skip over escaped characters.
+
 Be reassured, this is not an oversight!
-Regular expressions with trailing backslashes result in runtime exceptions (reported as "unexpected internal error").
-Even though detecting these types of errors at compile time is out of the scope of our implementation, doing so is still desirable.
-If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error.
+Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as "unexpected internal error").
+Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that error at runtime.
 
@@
 For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
-Similarly, we omit handling the "at least n times" quantifiers which can lead to nullable capturing groups.
+Similarly, we omit handling the "at least $n$ times" quantifiers which can lead to nullable capturing groups (when $n=0$).
 Our implementation accounts for both operators.
@@
 Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
-This last improvement leads to the final iteration of our type-level program handles nested groups by operating in two different modes.
 
@@
 
-This concludes the development of our nullability analysis and completes the presentation of the type-level component of our library.
+This concludes the development of our nullability analysis and completes our presentation of the type-level component of our library.
 Despite the scarcity of Scala's standard library at the type level, we managed to achieve our ends while keeping our implementation relatively concise and, hopefully, understandable.
@@
 
-The first implementation of our term-level component duplicates the type-level definitions presented in \Cref{sec:type-level} to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
-In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of option's constructor:
+The first implementation of our term-level component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
+In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of the option constructor:
 

### 2022-02-09 symposium.tex
Type and Null Safety for Regexes

@@
 
-\emph{In principle}, Scala~3's type system is capable of doing a lightweight form of dependent typing (\Cref{subsec:a-lightweight-form-of-dependent-typing}), where it...
+This solution has the benefit of being conceptually simple.
+We needed to write an implementation that conforms to our type-level program, and we did!
+By duplicating that program, and using an unsafe case to convince the compiler of our good intentions.
 
-\begin{lstlisting}
-def isMarked
-  [R <: String, At <: Int, Hi <: Int]
-  (r: R, at: At, hi: Hi)
-  : IsMarked[R, At, Hi]
-\end{lstlisting}
+The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language that supports type- and term-level polymorphism.
+In a language with shared term and type syntax, we could write our analysis generically and derive two programs from it, one for each level.
+
+In principle, Scala~3's type inference should be able to solve half of the problem with its ability to correlate match types and match expression.
+In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
+In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
+In practice, at the time of writing, this solution does not play well with predefined types from the standard library, because the compiler lacks the knowledge necessary to correlate those types with their term-level counterpart.
+For instance, the compiler does not recognize |CharAt["foo",0]| as a valid type for |"foo".chatAt(0)|.
+
+In the long run, it would be interesting to investigate extending Scala with a "precise" mode of type inference, that would automatically infer match types and singleton types, whenever possible.
+In \Cref{chap:generalizing-singleton-types}, we demonstrated the feasibility of this approach by generalizing Scala's singleton types.
+A promising avenue would be to revisit that line of work through the lens of type inference.
 

### 2022-02-09 scala/src/main/scala/regex.scala

@@ -23,6 +23,45 @@ class Regex[P] private (val regex: String):
       None
 // end section regexUserLevel
 
+abstract class Sanetizer[T](val i: Int):
+  def mutate(arr: Array[Any]): Unit
+
+object Sanetizer:
+  implicit val basecase1: Sanetizer[EmptyTuple] =
+    new Sanetizer[EmptyTuple](0):
+      def mutate(arr: Array[Any]): Unit = ()
+
+  implicit def stringcase[T <: Tuple]
+    (implicit ev: Sanetizer[T]): Sanetizer[String *: T] =
+      new Sanetizer[String *: T](ev.i + 1):
+        def mutate(arr: Array[Any]): Unit =
+          assert(arr(arr.size - i) != null)
+          ev.mutate(arr)
+
+  implicit def optioncase[T <: Tuple]
+    (implicit ev: Sanetizer[T]): Sanetizer[Option[String] *: T] =
+      new Sanetizer[Option[String] *: T](ev.i + 1):
+        def mutate(arr: Array[Any]): Unit =
+          arr(arr.size - i) = Option(arr(arr.size - i))
+          ev.mutate(arr)
+
+object Regex2:
+  def apply[R <: String & Singleton](regex: R)(implicit san: Sanetizer[Compile[R]]): Regex2[Compile[R]] =
+    new Regex2(regex, san)
+
+class Regex2[P] private (val regex: String, val san: Sanetizer[P]):
+  val pattern = Pattern.compile(regex)
+  def unapply(s: String): Option[P] =
+    val m = pattern.matcher(s)
+    if (m.matches())
+      val arr = Array.tabulate[Any](m.groupCount)(i => m.group(i + 1))
+      assert(arr.size == san.i, "Unexpected number of capturing groups")
+      san.mutate(arr)
+      val tuple = Tuple.fromArray(arr)
+      Some(tuple.asInstanceOf[P])
+    else
+      None
+
 object Lib {
 
 type CharAt[R <: String, At <: Int] =
@@ -211,61 +250,6 @@ type Reverse[L <: Tuple] = L match {
[…]

### 2022-02-09 scala/src/test/scala/regex-tests.scala

@@ -31,11 +31,11 @@ object Tests {
       s"expected: $expected, actual: $actual")
 
   def main(args: Array[String]): Unit = {
-    val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
+    val date = Regex2("(\\d{4})-(\\d{2})-(\\d{2})")
     assertEquals("2004", "2004-01-20" match
       case date(year, month, day) => year)
 
-    val everyday = Regex(".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*")
+    val everyday = Regex2(".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*")
     assertEquals(("Scala.js", "Scala"),
       "Write Scala.js everyday!" match
         case everyday(a, b, None) => (a, b))

### 2022-02-09 symposium.tex
Type and Null Safety for Regexes

@@
 The term-level component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
-Concretely, this component's job consists in transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
+Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
 That transformation must conform to the representation computed at the type level.
@@
 This solution has the benefit of being conceptually simple.
-We needed to write an implementation that conforms to our type-level program, and we did!
-By duplicating that program, and using an unsafe case to convince the compiler of our good intentions.
+We needed to write an implementation that conforms to our type-level program, and we did!XXX
+By duplicating that program, and using an unsafe cast to convince the compiler of our good intentions.
 
@@
 
-In principle, Scala~3's type inference should be able to solve half of the problem with its ability to correlate match types and match expression.
+In principle, Scala~3's type inference should be able to solve half of the problem with its ability to correlate match types and match expressions.
 In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.

### 2022-02-10 symposium.tex
Implicit-Based Extractor Synthesis

@@
 This solution has the benefit of being conceptually simple.
-We needed to write an implementation that conforms to our type-level program, and we did!XXX
-By duplicating that program, and using an unsafe cast to convince the compiler of our good intentions.
+We needed to write an implementation that conforms to our type-level program, and this is literally what we did, by duplicating that program and using an unsafe cast to convince the compiler of our good intentions.
 
@@
 
+The second implementation of our term-level component takes a completely different approach.
+Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized program.
+That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine for a particular regular expression.
+
+Our implementation consists of a type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements.
+That design is analogous to the recursive implicit
+\Cref{sec:recursive-implicit-resolution}
+
+The |Sanitizer| type class defines a single method that mutates an array in place to warp its nullable elements into options:
+
+\regexSanitizerTypeClass
+% abstract class Sanitizer[T](val i: Int):
+%   def mutate(arr: Array[Any]): Unit
+%
+% object Sanitizer:
+%   implicit val basecase: Sanitizer[EmptyTuple] =
+%     new Sanitizer[EmptyTuple](0):
+%       def mutate(arr: Array[Any]): Unit = ()
+%
+%   implicit def stringcase[T <: Tuple]
+%     (implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
+%       new Sanitizer[String *: T](ev.i + 1):
+%         def mutate(arr: Array[Any]): Unit =
+%           assert(arr(arr.size - i) != null)
+%           ev.mutate(arr)
+%
+%   implicit def optioncase[T <: Tuple]
+%     (implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
+%       new Sanitizer[Option[String] *: T](ev.i + 1):
+%         def mutate(arr: Array[Any]): Unit =
+%           arr(arr.size - i) = Option(arr(arr.size - i))
+%           ev.mutate(arr)
+
+From a design standpoint, this solution is appealing as it leads to no duplicated code and no duplicated computations: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
+However, this solution comes with an additional cost in terms of compilation time.
+In the following section, we evaluate the performance of our library and discuss the trade-offs between the first and second implementation of our term-level component.
+
 \section{Evaluation}

### 2022-02-10 scala/src/main/scala/regex.scala

@@ -23,33 +23,35 @@ class Regex[P] private (val regex: String):
       None
 // end section regexUserLevel
 
-abstract class Sanetizer[T](val i: Int):
+// start section regexSanitizerTypeClass
+abstract class Sanitizer[T](val i: Int):
   def mutate(arr: Array[Any]): Unit
 
-object Sanetizer:
-  implicit val basecase1: Sanetizer[EmptyTuple] =
-    new Sanetizer[EmptyTuple](0):
+object Sanitizer:
+  implicit val basecase: Sanitizer[EmptyTuple] =
+    new Sanitizer[EmptyTuple](0):
       def mutate(arr: Array[Any]): Unit = ()
 
   implicit def stringcase[T <: Tuple]
-    (implicit ev: Sanetizer[T]): Sanetizer[String *: T] =
-      new Sanetizer[String *: T](ev.i + 1):
+    (implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
+      new Sanitizer[String *: T](ev.i + 1):
         def mutate(arr: Array[Any]): Unit =
           assert(arr(arr.size - i) != null)
           ev.mutate(arr)
 
   implicit def optioncase[T <: Tuple]
-    (implicit ev: Sanetizer[T]): Sanetizer[Option[String] *: T] =
-      new Sanetizer[Option[String] *: T](ev.i + 1):
+    (implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
+      new Sanitizer[Option[String] *: T](ev.i + 1):
         def mutate(arr: Array[Any]): Unit =
           arr(arr.size - i) = Option(arr(arr.size - i))
           ev.mutate(arr)
+// end section regexSanitizerTypeClass
 
 object Regex2:
-  def apply[R <: String & Singleton](regex: R)(implicit san: Sanetizer[Compile[R]]): Regex2[Compile[R]] =
+  def apply[R <: String & Singleton](regex: R)(implicit san: Sanitizer[Compile[R]]): Regex2[Compile[R]] =
     new Regex2(regex, san)
 
-class Regex2[P] private (val regex: String, val san: Sanetizer[P]):
+class Regex2[P] private (val regex: String, val san: Sanitizer[P]):
   val pattern = Pattern.compile(regex)
   def unapply(s: String): Option[P] =
     val m = pattern.matcher(s)

### 2022-02-10 scala/src/test/scala/regex-bench.scala
Add regex benchmarks

@@ -1,0 +1,421 @@
+import scala.util.matching.Regex
+// import Package._
+
+object RegexBenchmarks {
+  def quote (s: String): String = if (s == null) "null" else "\"" + escape(s) + "\""
+  def escape(s: String): String = s.flatMap(escapedChar)
+
+  def escapedChar(ch: Char): String = ch match {
+    case '\b' => "\\b"
+    case '\t' => "\\t"
+    case '\n' => "\\n"
+    case '\f' => "\\f"
+    case '\r' => "\\r"
+    case '"'  => "\\\""
+    case '\'' => "\\\'"
+    case '\\' => "\\\\"
+    case _    => if (ch.isControl) "\\0" + Integer.toOctalString(ch.toInt)
+                 else              String.valueOf(ch)
+  }
+  var i = 0
+
+  def main(args: Array[String]): Unit = {
+    def test(r: Regex, s: String, o: String): Unit =
+      val m = r.pattern.matcher(s)
+      if (m.matches())
+        if (m.groupCount == 0)
+          println((new Exception).getStackTrace().map(_.getLineNumber()).toList.tail.head)
+
+        import Package.Lib._
+        val fns0 = loop(r.toString, 0, r.toString.length, Nil, 0).reverse
+        val fns1 = fns0.map(_.apply("")).map {
+          case _: String    => { (x: String) => x }
+          case _: Option[_] => { (x: String) => s"Some($x)" }
+        }
+        val extracts = Array.tabulate(m.groupCount)(i => quote(m.group(i + 1))).toList.mkString("(", ", ", ")")
+        val patterns = Array.tabulate(m.groupCount)(i => s"g$i").toList.zip(fns1).map((x, f) => f(x)).mkString("(", ", ", ")")
[…]

### 2022-02-10 scala/src/test/scala/regex-tests.scala

@@ -30,7 +30,7 @@ object Tests {
       expected == actual,
       s"expected: $expected, actual: $actual")
 
-  def main(args: Array[String]): Unit = {
+  def mains(args: Array[String]): Unit = {
     val date = Regex2("(\\d{4})-(\\d{2})-(\\d{2})")
     assertEquals("2004", "2004-01-20" match
       case date(year, month, day) => year)

### 2022-02-10 scala/regex/src/main/scala/bench/RegexBench1.scala
Setup jmh

@@ -1,34 +1,39 @@
-import scala.util.matching.Regex
+package bench
+
+import RegexPackage._
 import annotation.experimental
+import org.openjdk.jmh.annotations._
 
 @experimental
-object RegexBench3 {
-def main(args: Array[String]): Unit = {
+class RegexBench1 {
+
+@Benchmark
+def main(): Unit = {
 
 val r1 = Regex("((((((((((a))))))))))"); "a" match { case r1(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
 val r2 = Regex("((((((((((a))))))))))\\10"); "aa" match { case r2(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
 val r3 = Regex("(((((((((a)))))))))"); "a" match { case r3(g0, g1, g2, g3, g4, g5, g6, g7, g8) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8) == ("a", "a", "a", "a", "a", "a", "a", "a", "a")) }
-val r4 = Regex("((?:aaaa|bbbb)cccc)?"); "aaaacccc" match { case r4(g0) => assert((g0) == ("aaaacccc")) }
-val r5 = Regex("((?:aaaa|bbbb)cccc)?"); "bbbbcccc" match { case r5(g0) => assert((g0) == ("bbbbcccc")) }
+val r4 = Regex("((?:aaaa|bbbb)cccc)?"); "aaaacccc" match { case r4(Some(g0)) => assert((g0) == ("aaaacccc")) }
+val r5 = Regex("((?:aaaa|bbbb)cccc)?"); "bbbbcccc" match { case r5(Some(g0)) => assert((g0) == ("bbbbcccc")) }
 val r6 = Regex("((?i)a)b"); "ab" match { case r6(g0) => assert((g0) == ("a")) }
 val r7 = Regex("((?i)a)b"); "Ab" match { case r7(g0) => assert((g0) == ("A")) }
 val r8 = Regex("((?i:a))b"); "ab" match { case r8(g0) => assert((g0) == ("a")) }
[…]

### 2022-02-10 scala/regex/src/main/scala/bench/RegexBench2.scala

@@ -1,9 +1,14 @@
-import Package._
+package bench
+
+import RegexPackage._
 import annotation.experimental
+import org.openjdk.jmh.annotations._
 
 @experimental
-object RegexBench2 {
-def mains(args: Array[String]): Unit = {
+class RegexBench2 {
+
+@Benchmark
+def run(): Unit = {
 
 val r1 = Regex2("((((((((((a))))))))))"); "a" match { case r1(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
 val r2 = Regex2("((((((((((a))))))))))\\10"); "aa" match { case r2(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }

### 2022-02-10 scala/regex/src/main/scala/bench/RegexBench3.scala

@@ -1,0 +1,197 @@
+package bench
+
+import RegexPackage._
+import annotation.experimental
+import org.openjdk.jmh.annotations._
+
+@experimental
+class RegexBench3 {
+
+@Benchmark
+def run(): Unit = {
+
+val r1 = Regex3("((((((((((a))))))))))"); "a" match { case r1(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
+val r2 = Regex3("((((((((((a))))))))))\\10"); "aa" match { case r2(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
+val r3 = Regex3("(((((((((a)))))))))"); "a" match { case r3(g0, g1, g2, g3, g4, g5, g6, g7, g8) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8) == ("a", "a", "a", "a", "a", "a", "a", "a", "a")) }
+val r4 = Regex3("((?:aaaa|bbbb)cccc)?"); "aaaacccc" match { case r4(g0) => assert((g0) == ("aaaacccc")) }
+val r5 = Regex3("((?:aaaa|bbbb)cccc)?"); "bbbbcccc" match { case r5(g0) => assert((g0) == ("bbbbcccc")) }
+val r6 = Regex3("((?i)a)b"); "ab" match { case r6(g0) => assert((g0) == ("a")) }
+val r7 = Regex3("((?i)a)b"); "Ab" match { case r7(g0) => assert((g0) == ("A")) }
+val r8 = Regex3("((?i:a))b"); "ab" match { case r8(g0) => assert((g0) == ("a")) }
+val r9 = Regex3("((?i:a))b"); "Ab" match { case r9(g0) => assert((g0) == ("A")) }
+val r10 = Regex3("(([a-c])b*?\\2)*"); "ababbbcbc" match { case r10(g0, g1) => assert((g0, g1) == ("cbc", "c")) }
+val r11 = Regex3("(([a-c])b*?\\2){3}"); "ababbbcbc" match { case r11(g0, g1) => assert((g0, g1) == ("cbc", "c")) }
[…]

### 2022-02-10 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -1,11 +1,13 @@
+package bench
+
 import java.util.regex.Pattern
 import annotation.experimental
 import compiletime.ops.string._
 import compiletime.ops.int.{+, -, Max, <}
-import Package.Lib._
 
 @experimental
-object Package {
+object RegexPackage {
+import Lib._
 
 // start section regexUserLevel
 object Regex:
@@ -73,6 +75,20 @@ class Regex2[P] private (val regex: String, val san: Sanitizer[P]):
         Some(arr(0).asInstanceOf[P])
       else
         Some(Tuple.fromArray(arr).asInstanceOf[P])
+    else
+      None
+
+object Regex3:
+  def apply(regex: String): Regex3 =
+    new Regex3(regex)
+
+class Regex3 private (val regex: String):
+  val pattern = Pattern.compile(regex)
+  def unapplySeq(s: String): Option[Seq[String]] =
+    val m = pattern.matcher(s)
+    if (m.matches())
+      val arr = Array.tabulate(m.groupCount)(i => m.group(i + 1))
+      Some(arr)
     else
       None
 
@@ -252,7 +268,7 @@ def isPiped(r: String, at: Int, hi: Int, lvl: Int): Int =
 // get ride of those types. Furthermore, the code responsible for
 // generating _n extractors for tuples does not know about generic tuples,
 // so we also manually flattern those types here...
-type Reverse[L <: Tuple] = L match {
+type Reverse[L <: Tuple] = L match
   case EmptyTuple => Unit
   case x1 *: EmptyTuple => x1
   case x1 *: x2 *: EmptyTuple => (x2, x1)
@@ -276,7 +292,6 @@ type Reverse[L <: Tuple] = L match {
   case x1 *: x2 *: x3 *: x4 *: x5 *: x6 *: x7 *: x8 *: x9 *: x10 *: x11 *: x12 *: x13 *: x14 *: x15 *: x16 *: x17 *: x18 *: x19 *: x20 *: EmptyTuple => (x20, x19, x18, x17, x16, x15, x14, x13, x12, x11, x10, x9, x8, x7, x6, x5, x4, x3, x2, x1)
[…]

### 2022-02-10 scala/src/main/scala/regex-examples.scala

@@ -1,11 +1,14 @@
+package bench
+
 import annotation.experimental
-import Package._
+import bench._
+import RegexPackage._
 import Lib.{CharAt, Reverse}
 import compiletime.ops.int.Max
 import compiletime.ops.int.-
 
 @experimental
-object Examples {
+object RegexExamples {
 
 // start section regexDocumentation
 val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
@@ -136,5 +139,27 @@ object Sanitizer:
 // end section regexSanitizerTypeClass
 
 }
+
+def checkLib[A <: String, B](implicit ev: Lib.Compile[A] =:= B): Unit = ()
+type O = Option[String]
+
+checkLib["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
+checkLib["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
+checkLib["(\\d{4})?-(\\d{2})*-(\\d{2})++", (O, O, S)]
+checkLib["((A)(B(C)))", (S, S, S, S)]
+checkLib["((A)(B(C)))?", (O, O, O, O)]
+checkLib["((A)(B(C))?)", (S, S, O, O)]
+checkLib["((A)|(B(C)))", (S, O, O, O)]
+checkLib["(B)|C", O]
+checkLib["C|(A)", O]
+checkLib[".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*", (S, S, O)]
+
+checkLib["((?:A)(B(C)))", (S, S, S)]
+checkLib["((A)(?:B(C)))?", (O, O, O)]
+checkLib["((A)(B(?:C))?)", (S, S, O)]
+checkLib["(?:(A)|(B(C)))", (O, O, O)]
+checkLib["(A)(B)|(C)(D)", (O, O, O, O)]
+checkLib["(A)(B)(C)(D)|", (O, O, O, O)]
+checkLib["((A)|(B)(C))(D)", (S, O, O, O, S)]
 
 }

### 2022-02-10 scala/src/test/scala/regex-bench-gen.scala

@@ -1,232 +1,0 @@
-import scala.util.matching.Regex
-
-object RegexBenchGen {
-  def quote (s: String): String = if (s == null) "null" else "\"" + escape(s) + "\""
-  def escape(s: String): String = s.flatMap(escapedChar)
-
-  def escapedChar(ch: Char): String = ch match {
-    case '\b' => "\\b"
-    case '\t' => "\\t"
-    case '\n' => "\\n"
-    case '\f' => "\\f"
-    case '\r' => "\\r"
-    case '"'  => "\\\""
-    case '\'' => "\\\'"
-    case '\\' => "\\\\"
-    case _    => if (ch.isControl) "\\0" + Integer.toOctalString(ch.toInt)
-                 else              String.valueOf(ch)
-  }
-  var i = 0
-
-  def mains(args: Array[String]): Unit = {
-    def test(r: Regex, s: String, o: String): Unit =
-      val m = r.pattern.matcher(s)
-      if (m.matches())
-        if (m.groupCount == 0)
-          println((new Exception).getStackTrace().map(_.getLineNumber()).toList.tail.head)
-
-        import Package.Lib._
-        val fns0 = compile(r.toString)
-        val fns1 = fns0.map(_.apply("")).map {
-          case _: String    => { (x: String) => x }
-          case _: Option[_] => { (x: String) => s"Some($x)" }
-        }
-        val extracts = Array.tabulate(m.groupCount)(i => quote(m.group(i + 1))).toList
-        val patterns = Array.tabulate(m.groupCount)(i => s"g$i").toList
-          .zip(fns1).map((x, f) => f(x))
-          .zip(extracts).map { case (p, e) => if (e == "null") then "None" else p }
[…]

### 2022-02-10 scala/src/test/scala/regex-bench1.scala

@@ -1,192 +1,0 @@
-import Package._
-import annotation.experimental
-
-@experimental
-object RegexBench {
-def main(args: Array[String]): Unit = {
-
-val r1 = Regex("((((((((((a))))))))))"); "a" match { case r1(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
-val r2 = Regex("((((((((((a))))))))))\\10"); "aa" match { case r2(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
-val r3 = Regex("(((((((((a)))))))))"); "a" match { case r3(g0, g1, g2, g3, g4, g5, g6, g7, g8) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8) == ("a", "a", "a", "a", "a", "a", "a", "a", "a")) }
-val r4 = Regex("((?:aaaa|bbbb)cccc)?"); "aaaacccc" match { case r4(Some(g0)) => assert((g0) == ("aaaacccc")) }
-val r5 = Regex("((?:aaaa|bbbb)cccc)?"); "bbbbcccc" match { case r5(Some(g0)) => assert((g0) == ("bbbbcccc")) }
-val r6 = Regex("((?i)a)b"); "ab" match { case r6(g0) => assert((g0) == ("a")) }
-val r7 = Regex("((?i)a)b"); "Ab" match { case r7(g0) => assert((g0) == ("A")) }
-val r8 = Regex("((?i:a))b"); "ab" match { case r8(g0) => assert((g0) == ("a")) }
-val r9 = Regex("((?i:a))b"); "Ab" match { case r9(g0) => assert((g0) == ("A")) }
-val r10 = Regex("(([a-c])b*?\\2)*"); "ababbbcbc" match { case r10(Some(g0), Some(g1)) => assert((g0, g1) == ("cbc", "c")) }
-val r11 = Regex("(([a-c])b*?\\2){3}"); "ababbbcbc" match { case r11(g0, g1) => assert((g0, g1) == ("cbc", "c")) }
[…]

### 2022-02-10 scala/src/test/scala/regex-tests.scala

@@ -1,48 +1,0 @@
-import annotation.experimental
-import Package._
-
-@experimental
-object Tests {
-  object TypeLevel {
-    def check[A <: String, B](implicit ev: Lib.Compile[A] =:= B): Unit = ()
-    type S = String
-    type O = Option[String]
-
-    check["(\\d{4})-(\\d{2})-(\\d{2})", (S, S, S)]
-    check["(\\()-(\\d{2})-(\\d{2})", (S, S, S)]
-    check["(\\d{4})?-(\\d{2})*-(\\d{2})++", (O, O, S)]
-    check["((A)(B(C)))", (S, S, S, S)]
-    check["((A)(B(C)))?", (O, O, O, O)]
-    check["((A)(B(C))?)", (S, S, O, O)]
-    check["((A)|(B(C)))", (S, O, O, O)]
-    check["(B)|C", O]
-    check["C|(A)", O]
-    check[".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*", (S, S, O)]
-
-    check["((?:A)(B(C)))", (S, S, S)]
-    check["((A)(?:B(C)))?", (O, O, O)]
-    check["((A)(B(?:C))?)", (S, S, O)]
-    check["(?:(A)|(B(C)))", (O, O, O)]
-    check["(A)(B)|(C)(D)", (O, O, O, O)]
-    check["(A)(B)(C)(D)|", (O, O, O, O)]
-    check["((A)|(B)(C))(D)", (S, O, O, O, S)]
-  }
-
-  def assertEquals(expected: Any, actual: Any): Unit =
-    assert(
-      expected == actual,
-      s"expected: $expected, actual: $actual")
-
-  def mains(args: Array[String]): Unit = {
-    val date = Regex2("(\\d{4})-(\\d{2})-(\\d{2})")
-    assertEquals("2004", "2004-01-20" match
-      case date(year, month, day) => year)
-
-    val everyday = Regex2(".*\\s(([A-Za-z]{5}(hum)?).js)\\s.*")
-    assertEquals(("Scala.js", "Scala"),
-      "Write Scala.js everyday!" match
-        case everyday(a, b, None) => (a, b))
-
-    println("OK!")
-  }
-}

### 2022-02-11 symposium.tex
Add regex benchmarks
Evaluation

@@
 
-  \item The term-level capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed (\Cref{sec:term-level}).
+  \item The rnutime capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed (\Cref{sec:term-level}).
 
@@
 In our implementation, this is done via the |transform| method, which is our
-term-level processing component.
+runtime processing component.
 As a result, our implementation never returns null values, which removes the null-checking burden from the user.
@@
 
-The term-level component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
+The runtime component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
 Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
@@
 
-The first implementation of our term-level component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
+The first implementation of our runtime component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
 In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of the option constructor:
@@
 
-The second implementation of our term-level component takes a completely different approach.
+The second implementation of our runtime component takes a completely different approach.
 Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized program.
@@
 However, this solution comes with an additional cost in terms of compilation time.
-In the following section, we evaluate the performance of our library and discuss the trade-offs between the first and second implementation of our term-level component.
+In the following section, we evaluate the performance of our library and discuss the trade-offs between the first and second implementation of our runtime component.
 
@@
 
-In this section...
+To assess the correctness of our library, we use the QT3TS test suite \citep{w3c1994xquery}, which contains about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
+We select the subset of those tests which exercise successful matching with capturing groups, which amounts to a total of 183 tests.
+Each test consists of one capturing group extraction, which we check againts the reference extraction.
+
+\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code duplicated runtime (Dup) and with its implicit-based runtime (Impl) on our test suite (lower is better).}
+
+In this section... \Cref{fig:regexBars}.
+
+\section{Related Work}
+
+They did it all: \citep{spishak2012type}
+Also this talk, from Stephanie: \citep{weirich2014examples}

### 2022-02-11 preprint.tex
Remove useless subsection
Evaluation

@@
 
-\section{Motivating Examples}
+\section{Motivating Example}
 \label{sec:motivating-example}
@@
 The examples in this section are written in our Scala extension described in \Cref{sec:extending-scala}.
-
-\subsection{Safe Join}
-\label{sec:safe-join}
 
@@
 We were able to observe similar trends in the concatenation benchmark.
-These measurements were obtained by averaging 120 compilations on a warm compiler, and have been performed on an i7-7700K Processor running Oracle JVM 1.8.0 on Linux.
+We obtained those measurements by averaging 120 independent compilations on a warm compiler, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0 on Linux.
 

### 2022-02-11 symposium.tex
Evaluation

@@
 
-To assess the correctness of our library, we use the QT3TS test suite \citep{w3c1994xquery}, which contains about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
-We select the subset of those tests which exercise successful matching with capturing groups, which amounts to a total of 183 tests.
-Each test consists of one capturing group extraction, which we check againts the reference extraction.
+To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
+We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form a single benchmark.
+We use this benchmark to evaluate the performance of our library in terms of compilation times and execution times.
 
-\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code duplicated runtime (Dup) and with its implicit-based runtime (Impl) on our test suite (lower is better).}
+\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code duplicated runtime (Dup) and with its implicit-based runtime (Impl).
+Lower is better.}
 
-In this section... \Cref{fig:regexBars}.
+In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.
+
+We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
+With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted from \Cref{fig:regexBars}).
 

### 2022-02-12 symposium.tex
Type-Safe Regular Expressions: related work

@@
 The second implementation of our runtime component takes a completely different approach.
-Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized program.
+Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
 That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine for a particular regular expression.
@@
 Our implementation consists of a type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements.
-That design is analogous to the recursive implicit
-\Cref{sec:recursive-implicit-resolution}
-
-The |Sanitizer| type class defines a single method that mutates an array in place to warp its nullable elements into options:
+The |Sanitizer| type class defines a single method that mutates an array in place to warp its nullable elements in options:
 
@@
 From a design standpoint, this solution is appealing as it leads to no duplicated code and no duplicated computations: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
-However, this solution comes with an additional cost in terms of compilation time.
+This combination of type-level programming and implicit-based code synthesis is reminiscent of the compile-time variant of multi-stage metaprogramming \citep{stucki2018practical}.
+In those terms, our approach corresponds to a multi-staged program whose stage 0 happens entirely within Scala's type checker.
+
 In the following section, we evaluate the performance of our library and discuss the trade-offs between the first and second implementation of our runtime component.
@@
 
+XXX
+
 We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
@@
 
-They did it all: \citep{spishak2012type}
+\citeauthor{spishak2012type} developed a type system to validate regular expression syntax and capturing group extraction in Java programs \citep{spishak2012type}.
+Their tool takes the form of a compiler plugin that enhances Java's type system to track the number of capturing groups using type annotations.
+Unlike our system, their are not concerned with null safety and does not detect optional capturing groups.
+In practice, their tool operates similarly to a linter: it reports false positives and sometimes requires manual annotations.
+
+\citep{mainland2007s}'s quasiquote
+
 Also this talk, from Stephanie: \citep{weirich2014examples}
+dependently-typed flavor of GHC Haskell which requires no less than 20 language extensions, whereas our implementation compiles with vanilla Scala~3.

### 2022-02-12 scala/regex/src/main/scala/bench/RegexPackage.scala
Fix implicit benchmark

@@ -60,12 +60,12 @@ object Sanitizer:
         arr(0) = Option(arr(0))
 
 object Regex2:
-  def apply[R <: String & Singleton](regex: R)(implicit san: Sanitizer[Compile[R]]): Regex2[Compile[R]] =
-    new Regex2(regex, san)
+  def apply[R <: String & Singleton](regex: R): Regex2[Compile[R]] =
+    new Regex2(regex)
 
-class Regex2[P] private (val regex: String, val san: Sanitizer[P]):
+class Regex2[P] private (val regex: String):
   val pattern = Pattern.compile(regex)
-  def unapply(s: String): Option[P] =
+  def unapply(s: String)(implicit san: Sanitizer[P]): Option[P] =
     val m = pattern.matcher(s)
     if (m.matches())
       val arr = Array.tabulate[Any](m.groupCount)(i => m.group(i + 1))

### 2022-02-12 symposium.tex
Type-Safe Regular Expressions: evaluation

@@
 \regexDocumentation
-% val date = new Regex("(\\d{4})-(\\d{2})-(\\d{2})")
-% "2004-01-20" match
-%   case date(year, month, day) =>
-%     s"$year was a good year for PLs."
 
@@
 \regexUserLevel
-% object Regex:
-%   def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
-%     new Regex(regex)
-%
-% class Regex[P] private (val regex: String):
-%   val pattern = Pattern.compile(regex)
-%   def unapply(s: String): Option[P] =
-%     val m = pattern.matcher(s)
-%     if (m.matches())
-%       val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
-%       Some(transform[P](regex, a))
-%     else
-%       None
 
@@
 \regexRational
-% val rational = Regex("(\\d+)\\.?(\\d+)?")
-% "3.1415" match
-%   case rational(i, Some(f)) =>
-%     val n = i.size + f.size
-%     s"This number is $n digits long"
 
@@
 \regexFirstIteration
-% import compiletime.ops.string._
-% import compiletime.ops.int.+
-%
-% type Compile[R <: String] =
-%   Reverse[Loop[R, 0, Length[R], EmptyTuple]]
-%
-% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
-%   Lo match
-%     case Hi => Acc
-%     case _  => CharAt[R, Lo] match
-%       case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
-%       case "\\" => Loop[R, Lo + 2, Hi, Acc]
-%       case _ => Loop[R, Lo + 1, Hi, Acc]
 
@@
 \regexIsCapturing
-% type IsCapturing[R <: String, At <: Int] <: Boolean =
-%   CharAt[R, At] match
-%     case "?" => CharAt[R, At + 1] match
-%       case "<" => CharAt[R, At + 2] match
-%         case "!" => false // zero-width negative lookbehind
-%         case _ => true    // named-capturing group
-%       case _ => false     // other non-capturing special constructs
-%     case _ => true        // unnamed-capturing group
 
@@
 \regexNaiveIsNullable
-% type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
-%   CharAt[R, At] match
-%     case ")" => IsMarked[R, At + 1, Hi]
-%     case _   => IsNullable[R, At + 1, Hi]
-%
-% type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
-%   At match
-%     case Hi => false
-%     case _ =>
-%       CharAt[R, At] match
-%         case "?" | "*" => true
-%         case "+" => IsMarked[R, At + 1, Hi]
-%         case _ => false
 
@@
 \regexIsNullable
-% type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
-%   CharAt[R, At] match
-%     case ")" => Lvl match
-%       case 0 => IsMarked[R, At + 1, Hi]
-%       case _ => IsNullable[R, At + 1, Hi, Lvl - 1]
-%     case "(" => IsNullable[R, At + 1, Hi, Lvl + 1]
-%     case "\\" => IsNullable[R, At + 2, Hi, Lvl]
-%     case _    => IsNullable[R, At + 1, Hi, Lvl]
 
@@
 \regexLastIteration
-% type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
-%   Lo match
-%     case Hi => Acc
-%     case _  => CharAt[R, Lo] match
-%       case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
-%       case "(" =>
-%         Opt match
-%           case 0 => IsNullable[R, Lo + 1, Hi, 0] match
-%             case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
-%             case false => Loop[R, Lo + 1, Hi, Acc, 0]
-%           case _ => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
-%       case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
-%       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]
 
@@
 
-In Scala, all type parameters are erased, which can lead so some friction when trying to write programs whose execution \emph{depends} on a type, such as in the problem at hand.
+In Scala, all type parameters are erased, which leads to some friction when wanting to write programs whose execution \emph{depends} types, such as in the problem at hand.
 In this section, we propose two different solutions to that problem.
@@
 
-  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show...
+  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show how to use implicits to perform type-directed code synthesis: we use the output of our type-level analysis as an input of implicit search to generate a type-specialized capture group sanitizer.
 \end{enumerate}
@@
 The first implementation of our runtime component duplicates the type-level definitions presented in the previous section to compute a list of sanitizing functions that correspond to the given regex's capturing groups.
-In a nutshell, instead of accumulating |String| and |Option[String]| types, we rewrite the outer loop of our algorithm to accumulate identity functions and eta-expansions of the option constructor:
+In a nutshell, instead of accumulating |String| and |Option[String]| in a type, we accumulate identity functions and eta-expansions of the option constructor in a list:
 
 \regexTermLvlLoop
-% val identity: String => Any = { x => assert(x != null); x }
-% type Accumulator = List[String => Any]
-%
-% def loop(r: String, lo: Int, hi: Int, acc: Accumulator, opt: Int): Accumulator =
-%   lo match
-%     case `hi` => acc
-%     case _  => r.charAt(lo) match
-%       case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
-%       case '(' =>
-%         opt match
-%           case 0 => isNullable(r, lo + 1, hi, 0) match
-%             case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
-%             case false => loop(r, lo + 1, hi, identity :: acc, 0)
-%           case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
-%       case '\\' => loop(r, lo + 2, hi, acc, opt)
-%       case _ => loop(r, lo + 1, hi, acc, opt)
 
@@
 \regexTransform
-% def transform[P](pattern: String, arr: Array[String]): P =
-%   val fs = loop(pattern, 0, pattern.length, Nil, 0).reverse
-%   val ts = Tuple.fromArray(arr.zip(fs).map { (x, f) => f(x) })
-%   assert(arr.size == fs.size, "Unexpected number of capturing groups")
-%   ts.asInstanceOf[P]
 
@@
 
-The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language that supports type- and term-level polymorphism.
-In a language with shared term and type syntax, we could write our analysis generically and derive two programs from it, one for each level.
+The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language with support for type- and term-level polymorphism.
+In a language with shared term and type syntax, we could write our analysis generically, and derive two programs from it, one for each level.
 
-In principle, Scala~3's type inference should be able to solve half of the problem with its ability to correlate match types and match expressions.
+In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions.
 In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
 In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
-In practice, at the time of writing, this solution does not play well with predefined types from the standard library, because the compiler lacks the knowledge necessary to correlate those types with their term-level counterpart.
-For instance, the compiler does not recognize |CharAt["foo",0]| as a valid type for |"foo".chatAt(0)|.
+In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
+For instance, the compiler does not recognize |CharAt["foo", 0]| as a valid type for |"foo".chatAt(0)|.
 
@@
 In \Cref{chap:generalizing-singleton-types}, we demonstrated the feasibility of this approach by generalizing Scala's singleton types.
-A promising avenue would be to revisit that line of work through the lens of type inference.
+A promising avenue would be to revisit that line of work with the less ambitious goal of improving Scala's type inference.
 
@@
 
-The second implementation of our runtime component takes a completely different approach.
+The second version of our runtime component takes a completely different approach.
 Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
-That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine for a particular regular expression.
+That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine by following the shape of the tuple type computed at the type level.
 
-Our implementation consists of a type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements.
-The |Sanitizer| type class defines a single method that mutates an array in place to warp its nullable elements in options:
+Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the results of our type-level analysis).
+The |Sanitizer| type class defines a single method that mutates an array in place to warp nullable elements in options:
 
 \regexSanitizerTypeClass
-% abstract class Sanitizer[T](val i: Int):
-%   def mutate(arr: Array[Any]): Unit
-%
-% object Sanitizer:
-%   implicit val basecase: Sanitizer[EmptyTuple] =
-%     new Sanitizer[EmptyTuple](0):
-%       def mutate(arr: Array[Any]): Unit = ()
-%
-%   implicit def stringcase[T <: Tuple]
-%     (implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
-%       new Sanitizer[String *: T](ev.i + 1):
-%         def mutate(arr: Array[Any]): Unit =
-%           assert(arr(arr.size - i) != null)
-%           ev.mutate(arr)
-%
-%   implicit def optioncase[T <: Tuple]
-%     (implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
-%       new Sanitizer[Option[String] *: T](ev.i + 1):
-%         def mutate(arr: Array[Any]): Unit =
-%           arr(arr.size - i) = Option(arr(arr.size - i))
-%           ev.mutate(arr)
 
-From a design standpoint, this solution is appealing as it leads to no duplicated code and no duplicated computations: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
-This combination of type-level programming and implicit-based code synthesis is reminiscent of the compile-time variant of multi-stage metaprogramming \citep{stucki2018practical}.
+From a design standpoint, this solution is appealing as it leads to no duplicated code or computation: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
+This combination of type-level programming and implicit-based code synthesis is reminiscent of multi-stage metaprogramming, in its compile-time variant \citep{stucki2018practical}.
 In those terms, our approach corresponds to a multi-staged program whose stage 0 happens entirely within Scala's type checker.
@@
 To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
-We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form a single benchmark.
-We use this benchmark to evaluate the performance of our library in terms of compilation times and execution times.
+We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
+We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.
 
-\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code duplicated runtime (Dup) and with its implicit-based runtime (Impl).
+In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code-duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.
+
+\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code-duplicated runtime (Dup) and with its implicit-based runtime (Impl).
 Lower is better.}
 
-In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.
+We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
+With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
+We run our benchmarks on the latest version of Scala~3 (3.1.2 at the time of writing).
 
-XXX
+\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline, which corresponds to an average of +0.3~ms per regex.
+This corresponds to the cost of match type reduction.
+The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost implicit resolution.
+Both variants are thus relatively cheap which validates their utility for practical uses.
+When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of the type checker \citep[2.11.3]{petrashko2017design}.
 
-We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
-With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted from \Cref{fig:regexBars}).
+\paragraph{Execution Times} At runtime, the code-duplicated variant adds 51~\textmu s over the baseline.
+Although this difference is three orders of magnitudes smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
+The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation timed).
+To our surprise, our implicit-based runtime outperforms Scala's standard library implementation.
+We suspect that this difference is due to the poor interaction between the JVM's just-in-time compiler and the code generated for variadic extractors (|unapplySeq|, used by the Std implementation).
+Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight grain in performance.
 
+\bigskip
+
+In view of these results, we decided to use our implicit-based runtime in the published version of our library.
 \section{Related Work}

### 2022-02-12 symposium.tex
Abstract and stuff

@@
 This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
-The Scala implementation innovates in its presentation: it uses the power and flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
+The Scala implementation innovates in its presentation: it uses the flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
 The package's documentation starts with the following example:
@@
 \noindent
-The extractor pattern, |case date(y,m,d) =>|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.
+The extractor pattern, |case date(y,m,d)|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.
 
@@
 In this chapter, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
-Our design makes extensive use of match types, Scala~3 new feature for type-level programming, to statically analyze regular expressions during type checking and generate type-specialized extractors.
+Our design makes extensive use of match types, Scala's new feature for type-level programming, to statically analyze regular expressions during type checking.
 We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
@@
 
-  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show how to use implicits to perform type-directed code synthesis: we use the output of our type-level analysis as an input of implicit search to generate a type-specialized capture group sanitizer.
+  \item In \Cref{subsec:implicit-based-extractor-synthesis}, we show how to use implicits to perform type-directed code synthesis: we use the output of our type-level analysis as an input of implicit search to generate a type-specialized capturing group sanitizer.
 \end{enumerate}
@@
 
-From a design standpoint, this solution is appealing as it leads to no duplicated code or computation: we perform our capture group analysis at the type-level, once and for all, and synthesize code accordingly.
+From a design standpoint, this solution is appealing as it leads to no duplicated code or computation: we perform our capturing group analysis at the type-level, once and for all, and synthesize code accordingly.
 This combination of type-level programming and implicit-based code synthesis is reminiscent of multi-stage metaprogramming, in its compile-time variant \citep{stucki2018practical}.
@@
 
-\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code-duplicated runtime (Dup) and with its implicit-based runtime (Impl).
-Lower is better.}
+\regexBars{A comparison of the compilation time (left) and execution time (right) of Scala's standard regex library (Std) against our library with its code-duplicated runtime (Dup) and with its implicit-based runtime (Impl).}
 
@@
 With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
-We run our benchmarks on the latest version of Scala~3 (3.1.2 at the time of writing).
+We run our benchmarks on the latest version of Scala~3 at the time of writing (3.1.2).
 
-\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline, which corresponds to an average of +0.3~ms per regex.
+\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline (an average of +0.3~ms per regex).
 This corresponds to the cost of match type reduction.
 The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost implicit resolution.
-Both variants are thus relatively cheap which validates their utility for practical uses.
-When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of the type checker \citep[2.11.3]{petrashko2017design}.
+Both variants are thus relatively cheap, which validates their utility for practical uses.
+When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[2.11.3]{petrashko2017design}.
 
-\paragraph{Execution Times} At runtime, the code-duplicated variant adds 51~\textmu s over the baseline.
+\paragraph{Execution Times} At runtime, our code-duplicated variant adds 51~\textmu s over the baseline.
 Although this difference is three orders of magnitudes smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
-The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation timed).
+The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation time).
 To our surprise, our implicit-based runtime outperforms Scala's standard library implementation.
 We suspect that this difference is due to the poor interaction between the JVM's just-in-time compiler and the code generated for variadic extractors (|unapplySeq|, used by the Std implementation).
-Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight grain in performance.
+Our implicit-based implementation appears to be more prone to inlining and partial evaluation since it does not involve generic sequences, which could explain the slight gain in performance.
 
@@
 Their tool takes the form of a compiler plugin that enhances Java's type system to track the number of capturing groups using type annotations.
-Unlike our system, their are not concerned with null safety and does not detect optional capturing groups.
+Unlike our system, their solution is not concerned with null safety and does not detect optional capturing groups.
 In practice, their tool operates similarly to a linter: it reports false positives and sometimes requires manual annotations.

### 2022-02-12 symposium.tex
Type-Safe Regular Expressions: conclusion

@@
 In view of these results, we decided to use our implicit-based runtime in the published version of our library.
+
 \section{Related Work}
@@
 dependently-typed flavor of GHC Haskell which requires no less than 20 language extensions, whereas our implementation compiles with vanilla Scala~3.
+
+\section{Conclusion}
+
+In this chapter, we introduced a new design for type-safe regular expressions in Scala.
+We presented our type-level analysis, which identifies capturing groups and computes their nullability.
+We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
+We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis has a marginal impact on compilation times.
+
+\paragraph{Future Work}
+We propose to use a dedicated data type to represent the regex alternative operator.
+In our current design, we map every optional regex operator to an |Option[String]|, which does not always accurately reflects the structure of regular expressions.
+For instance, in \lstinline!"(A)(B)|(C)"!
+(which matches either |AB| or |C|), our system represents capturing as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
+This representation fails to account for the fact that the first and second capturing groups have identical nullability status, and that this status is opposite to the nullability of the third group.
+Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the regular expression.

### 2022-02-13 symposium.tex
Rephrase

@@
 \paragraph{Future Work}
-We propose to use a dedicated data type to represent the regex alternative operator.
-In our current design, we map every optional regex operator to an |Option[String]|, which does not always accurately reflects the structure of regular expressions.
+We propose to extend our approach with a dedicated data type for the regex alternative operator.
+In our current design, we map every optional regex operator to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
 For instance, in \lstinline!"(A)(B)|(C)"!
-(which matches either |AB| or |C|), our system represents capturing as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
-This representation fails to account for the fact that the first and second capturing groups have identical nullability status, and that this status is opposite to the nullability of the third group.
+(which matches either |AB| or |C|), our system represents capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
+This representation fails to account for the fact that the first and second groups have identical nullability status, and that this status is opposite to the nullability of the third group.
 Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the regular expression.

### 2022-02-13 symposium.tex
Related work

@@
 
-\citeauthor{spishak2012type} developed a type system to validate regular expression syntax and capturing group extraction in Java programs \citep{spishak2012type}.
+There is surprisingly little literature concerned with the safety of regular expression capturing groups.
+In a recent survey paper on the correctness of regular expressions, \citet{li2021ensuring} classified the publications in this area into several categories, ranging from empirical studies to regular expression synthesis and repair.
+However, they only listed a single paper about static checking: the type system developed by \citet{spishak2012type} to validate regular expression syntax and capturing group usage in Java programs.
 Their tool takes the form of a compiler plugin that enhances Java's type system to track the number of capturing groups using type annotations.
@@
 
-\citep{mainland2007s}'s quasiquote
+\citep{cheplyaka2011regex} regex-applicative
 
@@
 We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
-We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis has a marginal impact on compilation times.
+We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.
 

### 2022-02-13 scala/src/main/scala/regex-examples.scala

@@ -96,12 +96,11 @@ def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int)
     case `hi` => acc
     case _  => r.charAt(lo) match
       case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
-      case '(' =>
-        opt match
-          case 0 => isNullable(r, lo + 1, hi, 0) match
-            case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
-            case false => loop(r, lo + 1, hi, identity :: acc, 0)
-          case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
+      case '(' => opt match
+        case 0 => isNullable(r, lo + 1, hi, 0) match
+          case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
+          case false => loop(r, lo + 1, hi, identity :: acc, 0)
+        case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
       case '\\' => loop(r, lo + 2, hi, acc, opt)
       case _ => loop(r, lo + 1, hi, acc, opt)
 // end section regexTermLvlLoop

### 2022-02-13 symposium.tex
Related work

@@
 We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
+The source code of our implementation is available online\footnote{\texttt{https://gist.github.com/OlivierBlanvillain/418dcb74cbd2e12eb4a90b7336f40b17}}.
 
@@
 
-\citep{cheplyaka2011regex} regex-applicative
+The Haskell ecosystem contains several packages for regular expressions, with various levels of static safety.
+The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
+The main drawback of this solution lies in the library's learning curves: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are constrained to learn a new syntax, which is arguably more complicated.
 
-Also this talk, from Stephanie: \citep{weirich2014examples}
-dependently-typed flavor of GHC Haskell which requires no less than 20 language extensions, whereas our implementation compiles with vanilla Scala~3.
+Stephanie Weirich presented, on several occasions \citep{weirich2014examples}, a regular expression library as an example of a dependently-typed program written in Haskell.
+The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
+To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
+In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala~3.
+An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at runtime), whereas her library also contains a complete regular expression engine.
 

### 2022-02-13 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -78,20 +78,6 @@ class Regex2[P] private (val regex: String):
     else
       None
 
-object Regex3:
-  def apply(regex: String): Regex3 =
-    new Regex3(regex)
-
-class Regex3 private (val regex: String):
-  val pattern = Pattern.compile(regex)
-  def unapplySeq(s: String): Option[Seq[String]] =
-    val m = pattern.matcher(s)
-    if (m.matches())
-      val arr = Array.tabulate(m.groupCount)(i => m.group(i + 1))
-      Some(arr)
-    else
-      None
-
 object Lib {
 
 type CharAt[R <: String, At <: Int] =
@@ -101,7 +87,6 @@ type Compile[R <: String] =
   Reverse[Loop[R, 0, Length[R], EmptyTuple, IsPiped[R, 0, Length[R], 0]]]
 
 type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] <: Tuple =
-
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
@@ -216,7 +201,6 @@ def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[
       case '\\' => loop(r, lo + 2, hi, acc, opt)
       case _ => loop(r, lo + 1, hi, acc, opt)
 
-// start section regexisCapturing
 def isCapturing(r: String, at: Int): Boolean =
   r.charAt(at) match
     case '?' => r.charAt(at + 1) match
@@ -225,9 +209,7 @@ def isCapturing(r: String, at: Int): Boolean =
         case _ => true          // named-capturing group
       case _ => false           // other special constructs
     case _ => true              // unnamed-capturing group
-// end section regexisCapturing
 
-// start section regexisNullable
 def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean =
   r.charAt(at) match
     case ')' => lvl match
@@ -236,7 +218,6 @@ def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean =
     case '(' => isNullable(r, at + 1, hi, lvl + 1)
     case '\\' => isNullable(r, at + 2, hi, lvl)
     case _    => isNullable(r, at + 1, hi, lvl)
-// end section regexisNullable
 
 def isMarked(r: String, at: Int, hi: Int): Boolean =
   at match
[…]

### 2022-02-14 ack.tex
Acknowledgements

Advisor:

Martin Odersky

Thesis Committee:

Rachid Guerraoui
Paolo Giarrusso
Richard Eisenberg
Viktor Kuncak

Scala center colleagues:

Guillaume Masse
Jorge Vicente
Ólafur Páll Geirsson
Travis Lee
Darja Jovanovic
Jamie Thompson
Julien Richard-Foy

LAMP colleagues:

Aggelos Biboudis
Aleksander Boruch-Gruszecki
Allan Renucci
Anatolii Kmetiuk
Denys Shabalin
Dmitry Petrashko
Felix Mulder
Fengyun Liu
Jonathan Brachthäuser
Manohar Jonnalagedda
Matt Bovel
Nicolas Alexander Stucki
Paolo Giarrusso
Sébastien Doeraene
Sandro Stucki

LAMP colleagues:

Georg Schmid
Jad Hamza

Lightbend:

Adriaan Moors
Dale Wijnand

Students:

Maxime Kjaer
Andrea Veneziano

Staff:

Fabien Salvi
Natascha Fontana
Faille France

Friends:

Mia Primorac
Hermina Petric Maretić

Family:

Alicia Blanvillain
Soledad Andany
Christian Blanvillain
Leila

Girlfriend:

Dragana Milovancevic

### 2022-02-14 symposium.tex
Fix typos

@@
 \begin{enumerate}
-  \item The type-level capturing group analysis, which uses match types to inspect the user-provided regular expression and compute the corresponding tuple-based type representation (\Cref{sec:type-level}).
+  \item The type-level capturing group analysis, which uses match types to inspect the user-provided regular expression and compute tuple representation of the expression's capturing groups.
+  This component takes the form of a parametric type called |Compile|, which we present in \Cref{sec:type-level}.
 
-  \item The rnutime capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed (\Cref{sec:term-level}).
+  \item The runtime capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed.
+  This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.
 
@@
 Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string (the |Singleton| bound is necessary to guide Scala's type inference).
-We use that type-level string to instantiate |Compile|, our type-level analysis component.
+We use that type-level string to instantiate the |Compile| type, which is our type-level analysis component.
 
@@
 The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
-By returning an optional tuple, we tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity, and that the type of those groups corresponds to the type of the tuple elements.
-We take advantage of that flexibility to enrich the results of Java's regex engine by wrapping nullable values into options.
-In our implementation, this is done via the |transform| method, which is our
-runtime processing component.
-As a result, our implementation never returns null values, which removes the null-checking burden from the user.
+In our implementation, we instantiate the type |P| to a tuple and use that type in the result type of the |unapply| method to tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity.
+Furthermore, the type of those groups corresponds to the type of the tuple elements.
+This allows us to enrich the results of Java's regex engine by wrapping nullable values into options.
+Our implementation performs this wrapping via the |transform| method, which is our runtime processing component.
+As a result, our implementation never returns null values and removes the null-checking burden from the user.
 
@@
 
-Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not do honor to the syntax of Java's regular expression.
+Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not honor the syntax of Java's regular expression.
 Indeed, Java also supports several other special constructs which start with an opening parenthesis, for instance:
@@
 
-In Scala, all type parameters are erased, which leads to some friction when wanting to write programs whose execution \emph{depends} types, such as in the problem at hand.
+In Scala, all type parameters are erased, which leads to some friction when wanting to write programs whose execution \emph{depends} on types, such as in the problem at hand.
 In this section, we propose two different solutions to that problem.
@@
 
-\subsection{We Ain't Need No Dependent Types!}
+\subsection{We Don't Need No Dependent Types!}
 \label{subsec:we-ain-t-need-no-dependent-types}
@@
 In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
-For instance, the compiler does not recognize |CharAt["foo", 0]| as a valid type for |"foo".chatAt(0)|.
+For instance, the compiler does not recognize |CharAt["foo", 0]| as a valid type for |"foo".charAt(0)|.
 

### 2022-02-14 ack.tex
Acknowledgements

@@
-Advisor:
+First and foremost, I would like to thank my advisor, Martin Odersky, for his help and support throughout my PhD study, and for providing me with this unique opportunity to work on my favorite language, Scala.
 
-Martin Odersky
+I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and taking part in the committee.
 
-Thesis Committee:
+I would like to thank all my colleagues from LAMP, LARA and Scala Center.
+You were the source of so many interesting discussions, cheerful moments, and great memories, it was really a pleasure to work along your side.
+Aggelos Biboudis,
+Aleksander Boruch-Gruszecki,
+Allan Renucci,
+Anatolii Kmetiuk,
+Darja Jovanovic,
+Denys Shabalin,
+Dmitry Petrashko,
+Fabien Salvi,
+Felix Mulder,
+Fengyun Liu,
+Guillaume Martres,
+Guillaume Masse,
+Jad Hamza,
+Jamie Thompson,
+Jonathan Brachthäuser,
+Jorge Vicente,
+Julien Richard-Foy,
+Manohar Jonnalagedda,
+Matt Bovel,
+Natascha Fontana,
+Nicolas Alexander Stucki,
+Ólafur Páll Geirsson,
+Paolo Giarrusso,
+Romain Edelmann,
+Sandro Stucki,
+Sébastien Doeraene,
+Simon Guilloud, and
+Travis Lee---thank you!
+If it wasn't for you, I would have not have gotten here.
 
-Rachid Guerraoui
-Paolo Giarrusso
-Richard Eisenberg
-Viktor Kuncak
+A special thanks to Georg Schmid for being such an amazing friend, colleague and co-author over the past five years.
+I will never forget what we went through together, the paper deadlines, the hacking sessions, the bike rides, all games...
+It's crazy how much we did and how much we learned together.
+And of course, thank you Mia Primorac for being there for all the fun parts!
 
-Scala center colleagues:
+I would like to thank my parents, Soledad and Christian, and my little sister, Alicia, for being there for me all these years.
+It was truly a pleasure to have you close by throughout my PhD.
+I must also thank the latest arrival in family, Leila, for being such an endless source of joy and cuteness.
 
-Guillaume Masse
-Jorge Vicente
-Ólafur Páll Geirsson
-Travis Lee
-Darja Jovanovic
-Jamie Thompson
-Julien Richard-Foy
-
-LAMP colleagues:
-
-Aggelos Biboudis
-Aleksander Boruch-Gruszecki
-Allan Renucci
-Anatolii Kmetiuk
-Denys Shabalin
-Dmitry Petrashko
-Felix Mulder
-Fengyun Liu
-Jonathan Brachthäuser
-Manohar Jonnalagedda
-Matt Bovel
-Nicolas Alexander Stucki
-Paolo Giarrusso
-Sébastien Doeraene
-Sandro Stucki
-
-LAMP colleagues:
-
-Georg Schmid
-Jad Hamza
-
-Lightbend:
-
-Adriaan Moors
-Dale Wijnand
-
-Students:
-
-Maxime Kjaer
-Andrea Veneziano
-
-Staff:
-
-Fabien Salvi
-Natascha Fontana
-Faille France
-
-Friends:
-
-Mia Primorac
-Hermina Petric Maretić
-
-Family:
-
-Alicia Blanvillain
-Soledad Andany
-Christian Blanvillain
-Leila
-
-Girlfriend:
-
-Dragana Milovancevic
+Last but not the least, I would like to thank Dragana Milovancevic for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
+You are the best.

### 2022-02-14 thesis.tex
Acknowledgements

\input{headers.tex}

\begin{document}

\frontmatter
\input{title-page}
\setcounter{page}{0}

\chapter{Acknowledgements}

\input{ack.tex}

\bigskip
\noindent\textit{Lausanne, \today}
\hfill Olivier Blanvillain

\cleardoublepage
\chapter*{Abstract} % max 3499 characters
\markboth{Abstract}{Abstract}
\addcontentsline{toc}{chapter}{Abstract (English/Français)}

Over the past decade, the Scala community has shown great interest in using type-level programming to obtain additional type safety.
Unfortunately, the lack of support from the Scala compiler has been a barrier to the adoption of that technique, notably due to its negative impact on compilation times.
In this thesis, we present three techniques for type-level programming in Scala.
First, we explain the status quo, implicits, and show how we can divert them from their intended use to write ad hoc type-level programs.
Second, we propose a generalization of Scala's singleton types, which adds the ability to manipulate term-level programs at the type level.
Third, we introduce match types, a type-level equivalent of pattern matching, which we implemented in the Scala~3 compiler.
Throughout this dissertation, we demonstrate the practicality of our newly introduced techniques, by the means of case studies and examples.
Our performance evaluation shows that our new techniques outperform the status quo in terms of binary sizes and compilation times.

\paragraph{Keywords} Programming Languages, Type Systems, Compilers, Scala.

\begin{otherlanguage}{french}
\cleardoublepage
\chapter*{Résumé}
\markboth{Résumé}{Résumé}

Au cours de la dernière décennie, la communauté Scala a montré un grand intérêt pour l'utilisation de la programmation au niveau du système de type afin d'améliorer la sûreté du typage.
Malheureusement, le manque de support dans le compilateur Scala a été un frein pour l'adoption de cette technique, notamment en raison de son impact négatif sur les temps de compilation.
Dans cette thèse, nous présentons trois techniques de programmation au niveau du système de type en Scala.
Tout d'abord, nous expliquons le statu quo, les implicites, et montrons comment nous pouvons les détourner de leur usage prévu pour écrire des programmes au niveau des types.
Deuxièmement, nous proposons une généralisation des types singleton de Scala, qui ajoute la possibilité de manipuler et d'exécuter des programmes au niveau des types.
Troisièmement, nous introduisons les match types, un équivalent du pattern matching au niveau des types, que nous avons implémenté dans le compilateur Scala~3.
Tout au long de cette dissertation, nous démontrons le caractère pratique de nos nouvelles techniques, au moyen d'études de cas et d'exemples.
Notre évaluation des performances montre que nos nouvelles techniques surpassent le statu quo en termes de tailles binaires et de temps de compilation.

\paragraph{Mots clés} Langages de Programmation, Systèmes de Types, Compilateurs, Scala.

\end{otherlanguage}

\hypersetup{linkcolor=black}
\tableofcontents
\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{List of Figures}
\listoffigures
\hypersetup{linkcolor=purplish}
\mainmatter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%  Thesis Content  %%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\chapter{Introduction}
\label{chap:introduction}

In March 2017, our research group went on a ski retreat in the Swiss Alps.
After a full day of skiing on the Diablerets massif, we gathered for a lab dinner.
Denys Shabalin, who was working on Scala Native at the time~\citep{shabalin2020just}, started a conversation about manual memory management.
The discussion revolved around the following question: could a Rust-like ownership system be viable for Scala?
Denys' answer was clear: ownership is fundamentally at odds with the way Scala handles references, and without making deep changes to Scala's type system, the task was simply impossible.

The next day, I came up with a toy domain-specific language (DSL) that implements the basis of a linear type system using type-level programming.
Here is an example of a short program written in this DSL:

\memImplicitMain

\noindent
The type argument of |Context| is a type-level list of strings that corresponds to the memory regions allocated at each program point.
Methods of |Context| use type-level programming techniques to enforce the following properties:

\begin{enumerate}
  \item memory regions must be allocated (|malloc|) \emph{before} they are deallocated (|free|),
  \item all memory regions must be deallocated by the end of the program,
  \item dereferencing (|deref|) is only allowed on previously allocated regions.
\end{enumerate}

\noindent
The implementation makes use of Scala's implicits to enforce these properties.
While this small DSL is obviously too simplistic to be of any practical use, it demonstrates the power of type-level programming.

I was delighted with my solution!
Denys, however, was not impressed.
I attribute this apathy to his dislike of implicits.
Despite their widespread usage, implicits are notorious for their complexity~\citep{kvrikava2019scala}.
In particular, using implicits for type-level computations requires carefully crafted definitions following a specific pattern.
To give the reader a glimpse of this pattern, we show the definition of the |free| method on |Context|:

\memImplicitContextFree

\noindent
This definition uses three parameter lists, one for type parameters (|V| and |Out|), one for a value parameter (|v|), and one of an implicit value parameter (|ev|).
Only the second parameter list is intended to be specified at use site; the type and implicit parameters are meant to be inferred.
When a users write |.free("mem")| they only set the |v| parameter (|v="mem"|); the compiler takes care of finding valid assignments for |V|, |Out| and |ev|.
The implicit parameter of type |Remove[V,Ps,Out]| is the entry point to the world of type-level programming.
It specifies how |V| (constrained by to be |v|'s type), |Ps| (defined in the class), and |Out| (unconstrained) are interrelated:

\memImplicitRemove

\noindent
Implicit-prefixed definitions can be understood as \emph{facts} and \emph{rules} of a logic program.
The first definition, |casehead|, specifies a fact: the result of removing |V| from the list |V::Ps| is |Ps|, which is expressed as an instance of |Remove[V,V::Ps,Ps]|.
The second definition, |casetail|, specifies a rule: if the result of removing |V| from the list |Pt| is |Out|, the results of removing |V| from the list |Ph::Pt| is |Ph::Out| (\Cref{sec:hlists-remove} develops this example in more details).
When a user writes |ctx.free("mem")| on a |ctx| of type |Context["bool"::"mem"::HNil]|, the compiler uses |casehead| and |casetail| to compute a type |Out| such that |Out| is the results of removing |"mem"| from |"bool"::"mem"::HNil|.

In retrospect, I have to agree with Denys' judgment at the time: this style of programming is convoluted, to say the least.
Aesthetics and pragmatism aside, programming with implicits requires a complete paradigm shift.
Instead of using pattern matching and functions, implicits require algorithms to be expressed using relations and constraints, which makes the task harder than it should be.

\emph{Can we do better?} This is the question that motivates the work presented in this dissertation.

Our contributions are as follows:

\begin{itemize}
  \item In \Cref{chap:ab-using-implicits}, we present techniques for programming at the type-level with implicits.
  In particular, we develop the example presented in the introduction, and show how to use Scala's implicit resolution mechanism to compute types, in a style that resembles logic programming.

  \item In \Cref{chap:generalizing-singleton-types}, we propose a generalization of Scala's singleton types, whose goal is to enable type-level programming in an accessible style.
  Concretely, we extend Scala's type system with the ability to lift term-level programs to the type level and evaluate those programs during type-checking.
  We implement our system as an extension of the Scala~3 compiler, and show its practicality with a case study in which we develop a strongly-typed interface for Spark datasets.

  \item In \Cref{chap:match-types}, we introduce match types, a type-level equivalent of pattern matching.
  Match types integrate seamlessly into programming languages with subtyping and, despite their simplicity, offer significant additional expressiveness.
  We formalize match types in a self-contained calculus based on \SystemFsub and prove its soundness.
  We demonstrate the practical value of our system by implementing match types in the Scala~3 compiler, thus making type-level programming readily available to a broad audience of programmers.

  \item In \Cref{chap:type-safe-regular-expressions}, we propose a new design for type-safe regular expressions in Scala.
  Our approach makes extensive use of match types to identify capturing groups during type checking.
  We walk the reader through our design, step by step, providing detailed explanations along the way.
  Our implementation is on par with Java's regular expressions and only has a marginal impact on compilation times.

  \item In \Cref{chap:performance-evaluation}, we evaluate the performance of the various type-level programming techniques introduced in prior chapters, when confronted with large type-level programs.
  We show that the techniques introduced in \Cref{chap:generalizing-singleton-types} and~\ref{chap:match-types} systematically outperform implicits in terms of compilation times and binary sizes.
\end{itemize}

\chapter{(Ab)Using Implicits}
\label{chap:ab-using-implicits}

Scala's implicit parameters have outgrown their roots as a simple syntactic construct to the extent that they provide basic support for type-level programming.
In this chapter, we present techniques for implicit-based type-level programming in Scala.
In particular, through extended examples, we show how to use Scala's implicit resolution mechanism to compute types, in a style that resembles logic programming.

\subsection*{Attribution}

The first section of this chapter is based on the introduction to implicits from~\citep{odersky2018simplicitly}, which was written in collaboration with Martin Odersky, Fengyun Liu, Aggelos Biboudis, Heather Miller and Sandro Stucki, and published in POPL'18.

Code samples for the remaining of this chapter are based on the implementation of heterogeneous lists from the Shapeless library~\citep{sabin2011shapeless}.

\section{Implicit Parameters: Overview}

\emph{Implicit parameters} offer a convenient way to write code without the need to pass all arguments explicitly.
The ability to omit function arguments gives rise to many interesting coding styles and patterns.
On every call to functions with implicit parameters, the compiler looks for an implicit definition in scope to satisfy the call.
So, instead of passing a parameter explicitly:

\explicitModulo

\noindent
we can mark a set of parameters as implicit (a single parameter in this example) and let the compiler retrieve the missing argument for us.
In the following example, |addm| is a method with one implicit parameter and |modulo| is an implicit definition:

\implicitModulo

The process of implicit parameter discovery performed by the compiler is called \emph{implicit resolution}.
The resolution algorithm looks for implicits in the current scope and in the companion objects of all classes associated with the query type.
In the previous example, the implicit definition is declared in the current scope.
Since that definition has type |Int|, the compiler resolves the method call by passing |modulo| automatically.

\subsubsection{The type class pattern}

Implicits can be used to implement type classes~\citep{wadler1989how} as a design pattern~\citep{oliveira2010type}.
We give an example of an implementation of the |Ordering| type class.
This example consists of three parts:

\begin{enumerate}
  \item |Ordering[T]|, which is a regular trait with a single method, |compare|,
  \item the generic function |comp|, which compares two arguments and accepts an implicit argument, providing an \emph{implicit evidence} that these two values can be compared,
  \item the implicit definition |intOrdering|, which provides an \emph{instance} of the |Ordering| type class for integers.
\end{enumerate}

\ordExample

We have briefly introduced implicit parameters and showed how they can be used to avoid clutter in function applications.
In the next section, we present recursive implicit resolution.

\section{Recursive Implicit Resolution}
\label{sec:recursive-implicit-resolution}

Implicit methods can themselves take implicit parameters.
For example, we can define the lexicographic list ordering as follows:

\ordListExample

\noindent
This definition is parametrized by the list's element type and by an ordering of that type, passed as an implicit parameter.
Since |listOrdering| is itself implicit, it defines a \emph{rule}: it allows the compiler to materialize an implicit |Ordering[List[T]]|, given an implicit |Ordering[T]| (for any type |T|).

Parametrized implicit definitions can lead to recursive implicit resolution. For example, the compiler will use |listOrdering| twice to synthesize an implicit |Ordering[List[List[T]]]|.
This is where the (type-level) fun begins!

\subsubsection{Heterogeneous lists}

A heterogeneous list, or |HList| for short~\citep{kiselyov2004strongly}, is a datatype capable of storing data of different types.
In Scala~3, we can define the |HList| datatype as follows:

\hlistEnumDefinition

\noindent
The |::| constructor offers an interesting symmetry between the term and type level, which allows |HList| types to capture the same structure that their term-level counterparts.
For example, the term |::(1,::(2,HNil()))| can be typed as |::[1,::[2,HNil]]|, which is a perfect reification of that term (we use literal singleton types to represent constant literals at the type level~\citep{leontiev2014sip}).

\subsubsection{HList's remove}
\label{sec:hlists-remove}

Scala's implicits allow us to define type-level operations for heterogeneous lists.
We develop the example presented in \Cref{chap:introduction} by looking into the remove operation on |HList|s.
Remove operation takes as argument an element and a list, and returns that list with the first occurrence of the element removed.
This operation should yield an error if the element is not part of the list.

An implicit-based type-level operation typically takes the form of a trait, the operation's entry point, and several implicit definitions, one for each "case" of the operation's algorithm.
Let us consider the implementation of the remove operation in more detail:

\memImplicitRemove

\noindent
The |Remove| trait takes 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the element removed).
The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.

When the given element is not part of the list, implicit resolution fails with an "implicit not found" error, which indicates the incorrect use of |Remove|.
More precisely, the implicit search first iterates through the list by repeatedly using |casetail| until it reaches the end of the list, at which point it fails to find an implicit value of type |Remove[V,HNil,Out]| (neither |casehead| nor |casetail| produce an instantiation of |Remove| with |Ps=HNil|).

As an example usage of |Remove|, consider the following stringly-typed, JavaScript-inspired method:

\addEventListenerJS

\noindent
This method's documentation specifies 5 valid alternatives for the |event| argument, but that constraint is not reflected in the method's type signature.
In JavaScript, calling this method with an erroneous element event type is a \emph{no-op}, which can make this kind of error particularly hard to spot.

Instead of specifying that constraint in the documentation, we can represent the valid event types in a |HList|, and use an implicit parameter of type |Remove| to statically enforce that property:

\addEventListenerImplicitDef

\noindent
The |Singleton| type bound is a marker that instructs type inference to preserve literal singleton types (by default, the compiler widens those types).
The "|?|" type is Scala~3's new syntax for wildcard types~\citep[Wildcard Arguments in Types]{odersky2013scala}.

With this updated signature, the compiler is able to detect invalid event types at compile time.
When given a valid event type |E|, the compiler synthesizes an implicit evidence of type |Remove[E,EventTypes,?]| which witnesses |E|'s validity.
After implicit resolution, calls to the |addEventListener| method are expanded to automatically insert an implicit parameter for the second parameter list, such as in the following example:

\addEventListenerImplicitCall

This concludes our presentation of the implicit-based encoding of type-level computation.
Despite its verbosity, this pattern generalizes to arbitrary recursive computations and enables Scala programmers to write elaborate type-level programs solely based on implicits.
In the next section, we discuss ambiguities and priorities between implicit definitions.

\section{Ambiguities and Priorities}

Scala's implicit resolution relies on an intricate priority system to establish the precedence of implicit definitions.
Implicit-based programs sometimes rely on ambiguities and priorities of implicit definition, as we will see in this section through a series of examples.

\subsection{Implicit Ambiguities}

The Scala compiler rejects programs with \emph{ambiguous} implicit definitions.
For instance, if we write two identically-typed implicit definitions in the same scope, the compiler will consider them ambiguous and report an error:

\implicitAmbiguity

While ambiguities typically indicate programming errors, we can also use them purposefully to implement error cases of a type-level program.
As an example, consider the |NotIn| operation on |HList| that is only defined if the given element is \emph{not} part of the list.
We define |NotIn| using two ambiguous implicits; the implementation is lengthy, but straightforward:

\memImplicitNotIn

\noindent
The |casenil| and |casecons| implicits simply iterate through the list.
The |ambiguous| implicits are two identical definitions that will cause the compiler to raise an error if |V| and |Ph| are equal, for any element |Ph| of the list ("|=:=|" is a type from Scala's standard library that witnesses mutual subtyping).

As an example usage of |NotIn|, we revisit the DSL presented in \Cref{chap:introduction}.
Programs in our DSL consist of sequences of method calls on a |Context|, which tracks the memory regions that are currently allocated in the program, using a |HList| of names.
We use the |NotIn| operation to constrain the allocation method, |malloc|, to prevent allocating regions with already used names:

\memImplicitContextMalloc

Scala~3 introduced an alternative to the ambiguous implicit pattern in the form of the |scala.util.NotGiven| type~\citep[Given Instances]{odersky2013scala}.
The compiler will synthesize an implicit parameter of type |NotGiven[T]| if and only if there is no implicit value of type |T| in scope.
We can use a "negative" implicit evidence to simplify the definition of |NotIn| by removing the ambiguous implicits and changing |casecons| to the following:

\memImplicitNotGiven

\subsection{Implicit Priorities}

When looking for implicits, the Scala compiler visits scopes sequentially, in a precisely defined order.
At any point in the search, if the implicits defined in the subset of scopes considered so far can fulfill the implicit query, the search succeeds and immediately terminates.
This incremental process has two consequences.
First, it allows the compiler to efficiently look for implicits by naturally pruning some of the search space.
Second, it provides an ad-hoc mechanism to disambiguate implicits.
Scala programmer can artificially partition their implicit definitions into multiple scopes to implement a priority system.

Let us consider an example of operation on |HList| whose implementation uses implicit priorities.
|RemoveAll| is a generalization of |Remove| that removes every occurrence of the given element instead of the first occurrence.
The implicit-based implementation takes the form of a trait with 3 type parameters: 2 inputs, |V| (the element to remove) and |Ps| (the list), and one output, |Out| (the list, with the elements removed):

\removeAllDefinition

From an algorithmic standpoint, we can implement |RemoveAll| as a recursive function with 3 cases, a base case for the empty list (|casenil|), a recursive case for when the head of the list matches the element to remove (|casematch|), and another recursive case for when the head doesn't match (|casedoesnt|):

\removeAllBroken

This direct implementation of |RemoveAll| using implicit definitions is, unfortunately, incorrect.
The issue is that the |casematch| and |casedoesnt| definitions are ambiguous when the head of the list matches the element to remove.
To work around that ambiguity, we split those definitions into two different scopes, so that the compiler always tries to apply the more specialized case (|casematch|) before considering the less specialized case (|casedoesnt|).
Concretely, we define low priority implicits in a separate trait, and have |RemoveAll|'s companion object extend that trait:

\removeAllPrioritized

\section{Conclusion}

In this chapter, we presented several techniques for type-level programming with implicits.
This style of programming is, unfortunately, quite cumbersome.
In addition to the heavy syntax, type-level programming with implicits also requires a deep understanding of the implicit resolution algorithm.
As we will see in \Cref{chap:performance-evaluation}, those techniques come at a high cost in terms of compilation time, which hinders their usability on a large scale.
Yet, despite those shortcomings, Scala programmers have shown a persistent interest in this style of programming, as demonstrated by its popularity in the open-source community~\citep{sabin2011shapeless, pilquist2013scodecs, blanvillain2016frameless}.

\chapter{Generalizing Scala's Singleton Types}
\label{chap:generalizing-singleton-types}

Type-level programming is an increasingly popular way to obtain additional type safety.
Unfortunately, it remains a second-class citizen in the majority of industrially-used programming languages.
We propose a new dependently-typed system with subtyping and singleton types whose goal is enabling type-level programming in an accessible style.
To this end, we have prototyped our system as an extension of the Scala programming language.
We demonstrate the practicality of our system with a case study in which we develop a strongly-typed interface for Spark datasets.
Through our formalization and implementation in the context of an industrial-strength compiler, we hope to provide valuable insights for language designers interested in dependent types.

\subsection*{Attribution}

This chapter is based on~\citep{schmid2020coming}, which was written in collaboration with Georg Schmid, Jad Hamza, and Viktor Kuncak.
Georg and I worked hand-in-hand on this project: we collaborated through countless whiteboard discussions and pair programming sessions, which resulted in shared first authorship of the implementation and most of the text.
The said report covers more material than what is included in this chapter, in particular, our formalization and the associated metatheory are out of the scope of this dissertation.

\input{preprint.tex}

\chapter{Match Types}
\label{chap:match-types}

Type-level programming is becoming more and more popular in the realm of functional programming.
However, the combination of type-level programming and subtyping remains largely unexplored in practical programming languages.
This chapter presents \emph{match types}, a type-level equivalent of pattern matching.
Match types integrate seamlessly into programming languages with subtyping and, despite their simplicity, offer significant additional expressiveness.
We formalize the feature of match types in a calculus based on \SystemFsub and prove its soundness.
We demonstrate the practicality of our system by implementing match types in the Scala~3 reference compiler, thus making type-level programming readily available to a broad audience of programmers.

\subsection*{Attribution}

This chapter is based on~\citep{blanvillain2022type}, which was written in collaboration with Jonathan Brachthäuser, Maxime Kjaer, and Martin Odersky, and published in POPL'22.
\Cref{sec:case-study-shape-safe-num-py} is based on Maxime's semester project entitled "Shape-safe TensorFlow in Dotty", where he designed a strongly typed TensorFlow interface, which checks tensor shapes at compile-time in order to prevent runtime errors in machine learning models.
Maxime also contributed to the Scala~3 compiler by adding support for arithmetic computations at the type level.
Jonathan mechanized the soundness proof for \SystemFm, and provided invaluable help and guidance with redaction.
Martin and I collaborated on the implementation of match types in the Scala 3 compiler, which is now actively used by the Scala community.

\input{article.tex}

\chapter{Type-Safe Regular Expressions}
\label{chap:type-safe-regular-expressions}

Regular expressions can easily go wrong.
Capturing groups, in particular, require meticulous care to avoid running into off-by-one errors and null pointer exceptions.
In this chapter, we propose a new design for Scala's regular expressions which completely eliminates this class of errors.
Our design makes extensive use of match types, Scala's new feature for type-level programming, to statically analyze regular expressions during type checking.
We show that our approach has a minor impact on compilation times, which makes it suitable for practical use.

\subsection*{Attribution}

This chapter is inspired by Andrea Veneziano's semester project entitled "Strongly-typed regular expressions in Dotty".
In his project, Andrea implemented a regex library in two different flavors, one based on match types and the other based on generalized singleton types, with the objective to compare the two approaches in terms of expressivity and ease of use.
The present chapter is a reboot of the match-type side of that project.

\input{symposium.tex}

\chapter{Performance Evaluation}
\label{chap:performance-evaluation}

In this chapter, we evaluate the performance of the various type-level programming techniques introduced in prior chapters.
Specifically, we are interested in the scalability of each technique when confronted with large type-level programs, in terms of compilation times (\Cref{sec:compilation-time}) and binary sizes (\Cref{sec:binary-size}).
In \Cref{sec:the-timing-of-match-type-reductions}, we discuss how the timing of match type reduction impacts compilation times.

\section{Method}

We evaluate the performance of the 3 type-level programming techniques presented in this dissertation by comparing them on a set of 4 benchmarks.
Each benchmark consists of a simple type-level operation, which we implemented once using implicits (\Cref{chap:ab-using-implicits}), once using generalized singletons (\Cref{chap:generalizing-singleton-types}), and once using match types (\Cref{chap:match-types}).

These benchmarks are inspired by code examples used throughout this dissertation. They all follow a similar pattern of a single function call whose result type is computed at the type level. We code generated variations of each benchmark with increasing input size, from 1 to 256, in increments of 8.
The source code of our benchmarks is available in the supplementary material of this dissertation\footnote{\texttt{https://olivierblanvillain.github.io/thesis/benchmarks.zip}}, alongside instructions on how to reproduce our experiments.

Here is a brief description of our benchmark suite:

\begin{enumerate}
  \item \textbf{Concat}\quad
  Concatenates a heterogeneous list with itself.

  \item \textbf{Remove}\quad
  Removes the last element of a heterogeneous list (\Cref{sec:hlists-remove}).

  \item \textbf{Join}\quad
  Computes the resulting schema after performing a self-join along the last column of a database table (\Cref{sec:motivating-example}).

  \item \textbf{Reduce}\quad
  Computes the dimension of a multidimensional array after dimensional reduction along all axes (\Cref{subsubsec:reduction}).
\end{enumerate}

We obtained the measurements presented in this chapter by averaging 60 independent compilations on a warm compiler, which we executed on an i7-7700K Processor running Oracle JVM~1.8.0\_212 and Linux.
For the implicit and match type benchmarks, we used the latest version of Scala~3 at the time of writing (3.1.2).
For the generalized singleton benchmarks, we used our prototype implementation, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.

\section{Compilation time}
\label{sec:compilation-time}

\timingGraphsZoomedOut{A comparison of the compilation time of implicits, generalized singletons and match types for our benchmark suite (lower is better).}

\Cref{fig:timingGraphsZoomedOut} compares the compilation time of implicits, generalized singletons and match types on our benchmark suite.
The compilation time of implicit-based benchmarks dominates the graphs on each figures, and suggests a quadratic complexity.
These results imply that implicits are ill-suited for large scale type-level computations: spending tens of seconds on a single operation is clearly out of the question for practical applications.

\timingGraphsZoomedIn{A comparison of the compilation time of generalized singletons and match types for our benchmark suite (lower is better).}

The same measurements are shown in \Cref{fig:timingGraphsZoomedIn}, but with millisecond y-axes, to focus on the comparison of non-implicit curves.
Each graph follows a similar trend, with singleton-based implementations outperforming their match-type-based counterparts by about a factor of two.
We speculate that this difference is due to the overhead of subtyping, which in inherent to the design of match type.
Our prototype for generalized singletons uses a dedicated type evaluation procedure that is specialized for the task and thus faster.

We obtained the data points of \Cref{fig:timingGraphsZoomedOut} and~\ref{fig:timingGraphsZoomedIn} using Java Microbenchmark Harness (JMH) with 60 seconds of warmup, 60 seconds of measurement and a single JVM fork.
As a result, the resulting measurements are precise (99.9\% confidence intervals vary between $\pm$ 0.5 and $\pm$ 2 milliseconds), but subject to variations caused by the unpredictability of the JVM's just in time compiler, which explains the bumps visible in \Cref{fig:timingGraphsZoomedIn}.

\section{Binary size}
\label{sec:binary-size}

\bytecodeSizeTable{Additional JVM Bytecode generated per input size increment (in bytes).}

In \Cref{fig:bytecodeSizeTable}, we show the amount of additional JVM Bytecode generated per benchmark size increment.
The binary sizes of singleton- and match-type-based implementations are stable with respect to benchmark input size, which is expected given that our benchmarks consist of type-level-only operations.

The binary sizes of implicit-based implementations grow linearly with the input size, by 6 bytes per input size increment for smaller benchmarks (Concat and Remove), and by 12 and 18 bytes for the larger benchmarks (Reduce and Join, respectively).
To put those numbers in perspective, they represent an overall binary size increase of 9\%, 9\%, 12\%, and 20\%, when comparing the size-1 instantiation to the size-256 instantiation of Concat, Remove, Reduce, and Join, respectively.
These increases in binary size are caused by the implicit evidences synthesized by the compiler during implicit resolution.
In our benchmarks, those evidences have no runtime purpose and are simply an artifact of the mechanism in play.
Other applications, such as those supported by the Shapeless library~\citep{sabin2011shapeless}, take advantage of implicit evidence synthesis to provide additional runtime functionalities.

\section{The Timing of Match Type Reductions}
\label{sec:the-timing-of-match-type-reductions}

In \Cref{chap:match-types}, we describe the match type reduction algorithm by adding new subtyping rules to \SystemFsub (\Cref{subsec:matches}).
The resulting system is entirely declarative: it doesn't specify \emph{when} to apply those rules.
In this section, we discuss how the timing of match type reduction impacts compilation times.

A correct type system implementation must attempt to reduce match types \emph{as late as possible}, during subtyping.
Indeed, failing to do so could lead to incorrect results in cases where a match type is reducible in the context of the subtyping query but is not reducible in any prior context.

In addition to the reductions attempted during subtyping, the compiler can try to reduce match types earlier in the pipeline, essentially at any point in time.
Doing so is always correct, since the types before and after reduction are mutually subtypes of each other.
Each early attempt of match type reduction is a trade-off: it can either be a small overhead, if the reduction fails, a no-op, if the reduction succeeds but the result is used exactly once, or an optimization, if the reduction succeeds and the result is used more than once.
Note that reducing match types in a lazy fashion would not be a replacement for early reduction attempts, because the subtyping and disjointness checks are context-dependent (\Cref{subsec:caching}).

Let us consider an example that illustrates the benefits of early reduction attempts:

\geqDefinition

\noindent
Without any early reductions, this definition will result in a quadratic number of additions. For example, here is the sequence of subtyping queries issued while compiling |GEQ[5,1]|:

\geqExample

\noindent
The first instance of |1+1| comes from the right-hand-side of the first |GEQ| call (with |B=1|), but that type is used in all the subsequent subtyping queries. As a result, reducing |1+1| to |2| once and for all leads to a speed-up in compilation.

Our Scala~3 implementation reduces match types at multiple points during compilation, including right after creating types, during subtyping, and after instantiating parameters of type applications.
This last point is essential to avoid the quadratic behavior of the |GEQ| example.
In \Cref{fig:timingGraphsTooEarlyTooLate}, we show the impact of this optimization on our Reduce benchmark.
In this figure, we compare the compilation times of the match type Reduce benchmark with (S1) and without (S2) the reduction of match types after type parameter instantiation.

\timingGraphsTooEarlyTooLate{A comparison of the compilation time of the match type Reduce benchmark with two different match type reduction strategies. S1 is the reduction strategy we implemented in the Scala~3 compiler. S2 is a variation of that strategy where we disabled the reduction of match types after type parameter instantiation.}

\chapter{Conclusion}
\label{chap:conclusion}

In this thesis, we presented three techniques for type-level programming in Scala: implicits, generalized singleton types, and match types.
We showed that programming with implicits can be relatively cumbersome, both for programmers and compilers, which led us to the following question: \emph{can we do better?}
Our main contribution, match types, certainly answers that question positively.
They do provide a simple, first-class solution to type-level programming which is both easy to use and fast to compile.
We formalized match types in a calculus based on \SystemFsub and proved its soundness, which gives us confidence that our approach is correct and sensible.
Our implementation of match types in the Scala~3 compiler makes them accessible to a large audience of programmers and is already in active use.
Beyond the scope of Scala, it will be interesting to see if our design can inspire other general-purpose languages interested in moving their type system towards dependent types.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%  Bibliography, Appendix, CV  %%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix
\chapter{Type Soundness for System FM}
\label{chap:type-soundness-for-system-fm}

\renewenvironment{proof}{{\it Proof: }}{\qed} % Show proofs
\input{proofs/permutation.tex}
\input{proofs/weakening.tex}
\input{proofs/strengthening.tex}
\input{proofs/substitution.tex}
\input{proofs/disjointness-subtyping-exclusivity.tex}

\harpoonx{Definition of the auxiliary relation $\protect \⇌$, used to state inversion of subtyping.}

\begin{definition*}
  $\Γ \⊢ \S \⇌ \T$ (defined in \Cref{fig:harpoon}) represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \emph{\SMatch1/2}, \emph{\SMatch3/4}, and \emph{\STrans} in both directions.
\end{definition*}

\input{proofs/inversion-of-subtyping.tex}
\input{proofs/canonical-forms.tex}
\input{proofs/inversion-of-typing.tex}
\input{proofs/minimum-types.tex}
\input{proofs/preservation.tex}
\input{proofs/progress.tex}

\backmatter
\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\bibliographystyle{apalik}
\bibliography{bibliography}

\cleardoublepage
\thispagestyle{empty}
\phantomsection
\addcontentsline{toc}{chapter}{Curriculum Vitae}
\includepdf{olivier-blanvillain-resume.pdf}
\thispagestyle{empty}~

\end{document}

### 2022-02-14 ack.tex
Typos

@@
-First and foremost, I would like to thank my advisor, Martin Odersky, for his help and support throughout my PhD study, and for providing me with this unique opportunity to work on my favorite language, Scala.
+I would like to thank my advisor, Martin Odersky, for his help and support throughout my PhD studies, and for providing me with this unique opportunity to work on my favorite language.
 
@@
 Dmitry Petrashko,
+Dragana Milovancevic,
 Fabien Salvi,
@@
 Fengyun Liu,
+Georg Schmid,
 Guillaume Martres,
@@
 Manohar Jonnalagedda,
-Matt Bovel,
+Matthieu Bovel,
+Maxime Kjaer,
 Natascha Fontana,
-Nicolas Alexander Stucki,
-Ólafur Páll Geirsson,
+Nicolas Stucki,
+Ólafur Geirsson,
 Paolo Giarrusso,
@@
 A special thanks to Georg Schmid for being such an amazing friend, colleague and co-author over the past five years.
-I will never forget what we went through together, the paper deadlines, the hacking sessions, the bike rides, all games...
-It's crazy how much we did and how much we learned together.
+I will never forget what we went through together, the paper deadlines, the bike rides, the hacking sessions, the games...
+It's crazy how much we did and learned together.
 And of course, thank you Mia Primorac for being there for all the fun parts!
@@
 I would like to thank my parents, Soledad and Christian, and my little sister, Alicia, for being there for me all these years.
-It was truly a pleasure to have you close by throughout my PhD.
-I must also thank the latest arrival in family, Leila, for being such an endless source of joy and cuteness.
+It was truly a chance and a pleasure to have you close by throughout my PhD.
+I must also thank the latest arrival in family, Leila, for being such an endless source of joy, cuteness and distraction.
 
-Last but not the least, I would like to thank Dragana Milovancevic for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
+Last but not the least, I would like to thank Dragana for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
 You are the best.

### 2022-02-14 article.tex
Type-Safe Regular Expressions: related work
Typos

@@
 \subsubsection*{Preliminary Lemmas}
-Our proof begins with preliminary technical lemmas: \prelemma{Permutation}, \prelemma{Weakening}, \prelemma{Strengthening}, \prelemma{Substitution}.
-We omit stating these lemmas as they are entirely standard yet relatively lengthy given that they span the three relations of our system: typing, subtyping, and disjointness.
-As usual, their proofs follow by mutual inductions on derivations.
+Our proof begins with preliminary technical lemmas:
+%
+\input{proofs/permutation.tex}
+\input{proofs/weakening.tex}
+\input{proofs/strengthening.tex}
+\input{proofs/substitution.tex}
+%
+These lemmas are entirely standard, and as usual, their proofs follow by mutual inductions on derivations.
 
@@
 
-\section{Related work}
+\section{Related Work}
 \label{sec:related-work}

### 2022-02-14 symposium.tex
Typos

@@
 Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the given regular expression, and to extract the content of the capturing groups when the matching succeeds.
-These operations operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.
+These operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.
 
@@
 \noindent
-Here, we instantiates |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
+Here, we instantiate |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
 The resulting code is type-safe: if we change the pattern to omit the option unpacking (replacing |Some(f)| by |f|), the compiler would raise a type error on the call to |.size| (size is defined on strings but not options).
 
-In the following section, we explain our technique to analysis regular expressions and statically compute a type-level representation of the regex's capturing groups (|P| in the example above).
+In the following section, we explain our technique to analyze regular expressions and statically compute a type-level representation of the regex's capturing groups (|P| in the example above).
 
@@
 \noindent
-The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending of the nullability analysis.
+The result of that type-level computation takes the form of a Scala tuple with |String| and |Option[String]| elements, depending on the nullability analysis.
 
@@
 Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
-If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that error at runtime.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than error at run-time.
 
@@
 Our first implementation of |Compile| treats every opening parenthesis as the start of a capturing group, which does not honor the syntax of Java's regular expression.
-Indeed, Java also supports several other special constructs which start with an opening parenthesis, for instance:
+Indeed, Java also supports several other special constructs that start with an opening parenthesis, for instance:
 
@@
 
-We handle nested groups by updated our algorithm to operate in two different modes while iterating through the regex's characters.
+We handle nested groups by updating our algorithm to operate in two different modes while iterating through the regex's characters.
 Our algorithm can either be outside of an optional group (|Opt=0|), in which case it computes nullability of newly encountered groups via |IsNullable|, or inside an optional group (|Opt>0|), in which case it treats every group as nullable.
@@
 The second version of our runtime component takes a completely different approach.
-Instead of analyzing regular expressions at runtime, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
-That synthesis is type-directed: it generates a single-purpose program that sanitized the output of the regex engine by following the shape of the tuple type computed at the type level.
+Instead of analyzing regular expressions at run-time, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
+That synthesis is type-directed: it generates a single-purpose program that sanitizes the output of the regex engine by following the shape of the tuple type computed at the type level.
 
-Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the results of our type-level analysis).
-The |Sanitizer| type class defines a single method that mutates an array in place to warp nullable elements in options:
+Our implementation consists of one type class, |Sanitizer|, and three implicit definitions, which will guide the compiler into generating the correct |Sanitizer[T]|, for any tuple |T| with |String| and |Option[String]| elements (here, |T| is the result of our type-level analysis).
+The |Sanitizer| type class defines a single method that mutates an array in place to put nullable elements in options:
 
@@
 To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
-We select the subset of those tests that are positive (they match successfully) and exercises capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
+We select the subset of those tests that are positive (they match successfully) and exercise capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
 We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.
@@
 
-We obtained these results by averaging the measurements obtained over 2 hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
-With these settings, we obtains margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
+We obtained these results by averaging the measurements obtained over 2-hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
+With these settings, we obtain margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
 We run our benchmarks on the latest version of Scala~3 at the time of writing (3.1.2).
@@
 This corresponds to the cost of match type reduction.
-The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost implicit resolution.
+The implicit-based variant adds another 21~ms (+0.1~ms per regex), which corresponds to the cost of implicit resolution.
 Both variants are thus relatively cheap, which validates their utility for practical uses.
@@
 \paragraph{Execution Times} At runtime, our code-duplicated variant adds 51~\textmu s over the baseline.
-Although this difference is three orders of magnitudes smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
+Although this difference is three orders of magnitude smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
 The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation time).
@@
 The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
-The main drawback of this solution lies in the library's learning curves: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are constrained to learn a new syntax, which is arguably more complicated.
+The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
 
@@
 In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala~3.
-An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at runtime), whereas her library also contains a complete regular expression engine.
+An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.
 
@@
 We propose to extend our approach with a dedicated data type for the regex alternative operator.
-In our current design, we map every optional regex operator to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
+In our current design, we map every capturing group to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
 For instance, in \lstinline!"(A)(B)|(C)"!
-(which matches either |AB| or |C|), our system represents capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
-This representation fails to account for the fact that the first and second groups have identical nullability status, and that this status is opposite to the nullability of the third group.
-Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the regular expression.
+(which matches either |AB| or |C|), our system represents the capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
+This representation fails to account for the fact that the first two groups have identical nullability status, and that this status is opposite to the nullability of the third group.
+Instead, we could represent those capturing groups as |Either[(String,| |String),| |String]|, which is isomorphic to the structure of the regular expression.

### 2022-02-14 thesis.tex
Typos

@@
 
-In this chapter, we evaluate the performance of the various type-level programming techniques introduced in prior chapters.
+In this chapter, we evaluate the performance of the various type-level programming techniques presented in prior chapters.
 Specifically, we are interested in the scalability of each technique when confronted with large type-level programs, in terms of compilation times (\Cref{sec:compilation-time}) and binary sizes (\Cref{sec:binary-size}).
@@
 
-We evaluate the performance of the 3 type-level programming techniques presented in this dissertation by comparing them on a set of 4 benchmarks.
-Each benchmark consists of a simple type-level operation, which we implemented once using implicits (\Cref{chap:ab-using-implicits}), once using generalized singletons (\Cref{chap:generalizing-singleton-types}), and once using match types (\Cref{chap:match-types}).
+We evaluate the performance of the three type-level programming techniques presented in this thesis by comparing them on a set of four benchmarks.
+Each benchmark consists of a simple type-level operation, which we implemented once with implicits (\Cref{chap:ab-using-implicits}), once with generalized singletons (\Cref{chap:generalizing-singleton-types}), and once with match types (\Cref{chap:match-types}).
 
@@
   \item \textbf{Remove}\quad
-  Removes the last element of a heterogeneous list (\Cref{sec:hlists-remove}).
+  Removes the last element of a heterogeneous list (inspired by the example presented in \Cref{sec:hlists-remove}).
 
   \item \textbf{Join}\quad
-  Computes the resulting schema after performing a self-join along the last column of a database table (\Cref{sec:motivating-example}).
+  Computes the resulting schema after performing a self-join along the last column of a database table (inspired by the example presented in \Cref{sec:motivating-example}).
 
   \item \textbf{Reduce}\quad
-  Computes the dimension of a multidimensional array after dimensional reduction along all axes (\Cref{subsubsec:reduction}).
+  Computes the dimension of a multidimensional array after dimensional reduction along all axes (inspired by the example presented in \Cref{subsubsec:reduction}).
 \end{enumerate}
 
-We obtained the measurements presented in this chapter by averaging 60 independent compilations on a warm compiler, which we executed on an i7-7700K Processor running Oracle JVM~1.8.0\_212 and Linux.
-For the implicit and match type benchmarks, we used the latest version of Scala~3 at the time of writing (3.1.2).
-For the generalized singleton benchmarks, we used our prototype implementation, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
+We obtained the measurements presented in this chapter by averaging independent compilations on warm compilers, which we executed on an i7-7700K processor running Oracle JVM~1.8.0\_212 and Linux.
+For the implicit and match type benchmarks, we used the latest version of Scala at the time of writing (3.1.2).
+For the generalized singleton benchmarks, we used our prototype, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
 
@@
 \Cref{fig:timingGraphsZoomedOut} compares the compilation time of implicits, generalized singletons and match types on our benchmark suite.
-The compilation time of implicit-based benchmarks dominates the graphs on each figures, and suggests a quadratic complexity.
+The compilation time of implicit-based benchmarks dominates the graphs on each figure, and suggests a quadratic complexity.
 These results imply that implicits are ill-suited for large scale type-level computations: spending tens of seconds on a single operation is clearly out of the question for practical applications.
@@
 
-The same measurements are shown in \Cref{fig:timingGraphsZoomedIn}, but with millisecond y-axes, to focus on the comparison of non-implicit curves.
+The same measurements are shown in \Cref{fig:timingGraphsZoomedIn} with millisecond y-axes, to focus on the comparison of non-implicit curves.
 Each graph follows a similar trend, with singleton-based implementations outperforming their match-type-based counterparts by about a factor of two.
 We speculate that this difference is due to the overhead of subtyping, which in inherent to the design of match type.
-Our prototype for generalized singletons uses a dedicated type evaluation procedure that is specialized for the task and thus faster.
+Our prototype for generalized singletons uses a dedicated type evaluation procedure that is specialized for the task and therefore faster.
 
-We obtained the data points of \Cref{fig:timingGraphsZoomedOut} and~\ref{fig:timingGraphsZoomedIn} using Java Microbenchmark Harness (JMH) with 60 seconds of warmup, 60 seconds of measurement and a single JVM fork.
-As a result, the resulting measurements are precise (99.9\% confidence intervals vary between $\pm$ 0.5 and $\pm$ 2 milliseconds), but subject to variations caused by the unpredictability of the JVM's just in time compiler, which explains the bumps visible in \Cref{fig:timingGraphsZoomedIn}.
+We obtained the data points of \Cref{fig:timingGraphsZoomedOut} and~\ref{fig:timingGraphsZoomedIn} using the Java Microbenchmark Harness library with 60 seconds of warmup, 60 seconds of measurement and a single JVM.
+As a result, the resulting measurements are precise (99.9\% confidence intervals vary between $\pm$ 0.5 and $\pm$ 2 milliseconds), but subject to variations caused by the unpredictability of the JVM's just-in-time compiler, which explains the bumps visible in \Cref{fig:timingGraphsZoomedIn}.
 
@@
 In \Cref{fig:bytecodeSizeTable}, we show the amount of additional JVM Bytecode generated per benchmark size increment.
-The binary sizes of singleton- and match-type-based implementations are stable with respect to benchmark input size, which is expected given that our benchmarks consist of type-level-only operations.
+The binary sizes of singleton- and match-type-based implementations are stable with respect to the benchmark input size, which is expected given that our benchmarks consist of type-level-only operations.
 
-The binary sizes of implicit-based implementations grow linearly with the input size, by 6 bytes per input size increment for smaller benchmarks (Concat and Remove), and by 12 and 18 bytes for the larger benchmarks (Reduce and Join, respectively).
-To put those numbers in perspective, they represent an overall binary size increase of 9\%, 9\%, 12\%, and 20\%, when comparing the size-1 instantiation to the size-256 instantiation of Concat, Remove, Reduce, and Join, respectively.
-These increases in binary size are caused by the implicit evidences synthesized by the compiler during implicit resolution.
-In our benchmarks, those evidences have no runtime purpose and are simply an artifact of the mechanism in play.
+The binary sizes of implicit-based implementations grow linearly with the input size, by 6 bytes per increment for smaller benchmarks (Concat and Remove), and by 12 and 18 bytes for the larger benchmarks (Reduce and Join, respectively).
+To put those numbers in perspective, they represent an overall binary size increase of 9\%, 9\%, 12\%, and 20\%, when comparing the size-1 instantiation to the size-256 instantiation of our Concat, Remove, Reduce, and Join benchmarks, respectively.
+These binary size increases are caused by the implicit evidences synthesized by the compiler after implicit resolution.
+In our benchmarks, those evidences have no runtime purpose and are simply artifacts of the mechanism in play.
 Other applications, such as those supported by the Shapeless library~\citep{sabin2011shapeless}, take advantage of implicit evidence synthesis to provide additional runtime functionalities.
@@
 
-In \Cref{chap:match-types}, we describe the match type reduction algorithm by adding new subtyping rules to \SystemFsub (\Cref{subsec:matches}).
+In \Cref{chap:match-types}, we described the match type reduction algorithm by adding new subtyping rules to \SystemFsub (\Cref{subsec:matches}).
 The resulting system is entirely declarative: it doesn't specify \emph{when} to apply those rules.
@@
 
-A correct type system implementation must attempt to reduce match types \emph{as late as possible}, during subtyping.
-Indeed, failing to do so could lead to incorrect results in cases where a match type is reducible in the context of the subtyping query but is not reducible in any prior context.
+A correct type system implementation must attempt to reduce match types \emph{as late as possible}, that is, during subtyping.
+Indeed, failing to do so could lead to incorrect results, for instance, if a match type is reducible in the context of the subtyping query, but is not reducible in any prior context.
 
-In addition to the reductions attempted during subtyping, the compiler can try to reduce match types earlier in the pipeline, essentially at any point in time.
-Doing so is always correct, since the types before and after reduction are mutually subtypes of each other.
-Each early attempt of match type reduction is a trade-off: it can either be a small overhead, if the reduction fails, a no-op, if the reduction succeeds but the result is used exactly once, or an optimization, if the reduction succeeds and the result is used more than once.
-Note that reducing match types in a lazy fashion would not be a replacement for early reduction attempts, because the subtyping and disjointness checks are context-dependent (\Cref{subsec:caching}).
+In addition to reducing match types during subtyping, the compiler can attempt to reduce match types earlier in the pipeline, essentially at any point in time.
+Doing so is always correct: the types before and after reduction are mutually subtypes.
+Each early attempt of match type reduction can be viewed as a performance trade-off: it can either introduce a small overhead, if the reduction fails, be a no-op, if the reduction succeeds but the result is used exactly once, or be an optimization, if the reduction succeeds and the result is used more than once.
 
@@
 \noindent
-Without any early reductions, this definition will result in a quadratic number of additions. For example, here is the sequence of subtyping queries issued while compiling |GEQ[5,1]|:
+Without any early reductions, this definition results in a quadratic number of additions. For example, here is the sequence of subtyping queries that would be issued while compiling |GEQ[5,1]|:
 
@@
 
-Our Scala~3 implementation reduces match types at multiple points during compilation, including right after creating types, during subtyping, and after instantiating parameters of type applications.
+Our implementation reduces match types at multiple points during compilation, including directly after creating types, during subtyping, and after instantiating parameters of type applications.
 This last point is essential to avoid the quadratic behavior of the |GEQ| example.
@@
 We formalized match types in a calculus based on \SystemFsub and proved its soundness, which gives us confidence that our approach is correct and sensible.
-Our implementation of match types in the Scala~3 compiler makes them accessible to a large audience of programmers and is already in active use.
+Our implementation of match types in the Scala~3 compiler is already in active use and makes type-level programming accessible to a large audience of programmers.
 Beyond the scope of Scala, it will be interesting to see if our design can inspire other general-purpose languages interested in moving their type system towards dependent types.
@@
 \input{proofs/minimum-types.tex}
+\input{proofs/progress.tex}
 \input{proofs/preservation.tex}
-\input{proofs/progress.tex}
 

### 2022-02-14 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -144,7 +144,7 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
       case "{" => CharAt[R, At + 1] match
         case "0" => true
         case _ => false
-      case "+" => IsMarked[R, At + 1, Hi]
+      case "+" => false
       case _ => false
 
 type IsPiped[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Int =

### 2022-02-14 scala/src/main/scala/regex-examples.scala

@@ -18,7 +18,7 @@ val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
 // end section regexDocumentation
 
 // start section regexRational
-val rational = Regex("(\\d+)\\.?(\\d+)?")
+val rational = Regex("(\\d+)(?:\\.(\\d+))?")
 "3.1415" match
   case rational(i, Some(f)) =>
     val n = i.size + f.size
@@ -61,7 +61,7 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
     case _ =>
       CharAt[R, At] match
         case "?" | "*" => true
-        case "+" => IsMarked[R, At + 1, Hi]
+        case "+" => false
         case _ => false
 // end section regexNaiveIsNullable
 }

### 2022-02-14 ack.tex
Update regex implementation

@@
 Guillaume Masse,
+Heather Miller,
 Jad Hamza,
@@
 Julien Richard-Foy,
-Manohar Jonnalagedda,
 Matthieu Bovel,
@@
 Sébastien Doeraene,
-Simon Guilloud, and
 Travis Lee---thank you!

### 2022-02-14 scala/regex/src/main/scala/bench/RegexBench2.scala

@@ -5,10 +5,10 @@ import annotation.experimental
 import org.openjdk.jmh.annotations._
 
 @experimental
-class RegexBench2 {
+object RegexBench2 {
 
-@Benchmark
-def run(): Unit = {
+// @Benchmark
+def main(args: Array[String]): Unit = {
 
 val r1 = Regex2("((((((((((a))))))))))"); "a" match { case r1(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
 val r2 = Regex2("((((((((((a))))))))))\\10"); "aa" match { case r2(g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) => assert((g0, g1, g2, g3, g4, g5, g6, g7, g8, g9) == ("a", "a", "a", "a", "a", "a", "a", "a", "a", "a")) }
@@ -193,5 +193,9 @@ val r180 = Regex2("foo(aA|bB)?+b"); "foobBb" match { case r180(Some(g0)) => asse
 val r181 = Regex2("foo(aA|bB){1,5}+b"); "foobBaAaAaAaAb" match { case r181(g0) => assert((g0) == ("aA")) }
 val r182 = Regex2("X(\\w+)(?=\\s)|X(\\w+)"); "Xab" match { case r182(None, Some(g1)) => assert((null, g1) == (null, "ab")) }
 val r183 = Regex2("x(~~)*(?:(?:F)?)?"); "x~~" match { case r183(Some(g0)) => assert((g0) == ("~~")) }
+
+val r184 = Regex2("(\\Qxxx\\E)"); "xxx" match { case r184(g0) => assert((g0) == ("xxx")) }
+val r185 = Regex2("([(][)])"); "()" match { case r185(g0) => assert((g0) == ("()")) }
+// val r186 = Regex2("(\\Q))((())((\\\\...\\E)"); "))((())((\\\\..." match { case r186(g0) => assert((g0) == ("))((())((\\\\...")) }
 
 }}

### 2022-02-14 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -90,28 +90,22 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] <: Tuple 
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
-      case ")" =>
-        Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
-      case "(" =>
-        Opt match
-          case 0 =>
-            IsNullable[R, Lo + 1, Hi, 0] match
-              case true =>
-                IsCapturing[R, Lo + 1] match {
-                  case false => Loop[R, Lo + 1, Hi, Acc, 1]
-                  case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
-                }
-              case false =>
-                IsCapturing[R, Lo + 1] match {
-                  case false => Loop[R, Lo + 1, Hi, Acc, IsPiped[R, Lo + 1, Hi, 0]]
-                  case true  => Loop[R, Lo + 1, Hi, String *: Acc, IsPiped[R, Lo + 1, Hi, 0]]
-                }
-          case _ =>
-            IsCapturing[R, Lo + 1] match {
-              case false => Loop[R, Lo + 1, Hi, Acc, Opt + 1]
-              case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
-            }
-      case "\\" => Loop[R, Lo + 2, Hi, Acc, Opt]
+      case "\\" => CharAt[R, Lo + 1] match
+        case "Q" => Loop[R, ToClosingQE[R, Lo + 2], Hi, Acc, Opt]
+        case _ => Loop[R, Lo + 2, Hi, Acc, Opt]
+      case "[" => Loop[R, ToClosingBracket[R, Lo + 1, 0], Hi, Acc, Opt]
+      case ")" => Loop[R, Lo + 1, Hi, Acc, Max[0, Opt - 1]]
+      case "(" => Opt match
+        case 0 => IsMarked[R, ToClosingParenthesis[R, Lo + 1, 0], Hi] match
+          case true => IsCapturing[R, Lo + 1] match
+            case false => Loop[R, Lo + 1, Hi, Acc, 1]
+            case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, 1]
+          case false => IsCapturing[R, Lo + 1] match
[…]

### 2022-02-14 scala/src/main/scala/regex-examples.scala

@@ -61,13 +61,31 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
     case _ =>
       CharAt[R, At] match
         case "?" | "*" => true
-        case "+" => false
         case _ => false
 // end section regexNaiveIsNullable
+
 }
 
+type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+  At match
+    case Hi => false
+    case _ =>
+      CharAt[R, At] match
+        case "?" | "*" => true
+        case _ => false
+
+// start section regexIsNullable
+type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
+  CharAt[R, At] match
+    case ")" => Lvl match
+      case 0 => IsMarked[R, At + 1, Hi]
+      case _ => IsNullable[R, At + 1, Hi, Lvl - 1]
+    case "(" => IsNullable[R, At + 1, Hi, Lvl + 1]
+    case "\\" => IsNullable[R, At + 2, Hi, Lvl]
+    case _    => IsNullable[R, At + 1, Hi, Lvl]
+// end section regexIsNullable
+
 object Last {
-import Lib.IsNullable
 
 // start section regexLastIteration
 type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =

### 2022-02-14 ack.tex
Add footnote and QE and character class

@@
 
-I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and taking part in the committee.
+I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and for taking part in the committee.
 
@@
 Georg Schmid,
+Guillaume Masse,
 Guillaume Martres,
-Guillaume Masse,
 Heather Miller,
@@
 A special thanks to Georg Schmid for being such an amazing friend, colleague and co-author over the past five years.
-I will never forget what we went through together, the paper deadlines, the bike rides, the hacking sessions, the games...
+I will never forget what we went through together, the paper deadlines, the bike rides, the hacking sessions...
 It's crazy how much we did and learned together.
@@
 I would like to thank my parents, Soledad and Christian, and my little sister, Alicia, for being there for me all these years.
-It was truly a chance and a pleasure to have you close by throughout my PhD.
+It was truly a bliss to have you close by throughout my PhD.
 I must also thank the latest arrival in family, Leila, for being such an endless source of joy, cuteness and distraction.
 
-Last but not the least, I would like to thank Dragana for being always there on my side, for being so kind and patient, and for her tremendous help in putting the ideas of this thesis into text.
+Last but not the least, I would like to thank Dragana for being always there on my side, for being so kind and patient, and for her help in putting the ideas of this thesis into text.
 You are the best.

### 2022-02-14 article.tex
Add footnote and QE and character class

@@
 %
-In this example, the definition of |f| in |C| type-checks because |X & String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X & String]| reduces to |Int| (|M|'s second case applies).
+In this example, the definition of |f| in |C| type-checks because |X| |&| |String| and |Int| are disjoint (since |String| and |Int| are disjoint) and |M[X| |&| |String]| reduces to |Int| (|M|'s second case applies).
 Class |D| refines the definition of |C| by giving concrete definition of |X|.
-The unsoundness manifests itself in the body of class |D|, where |X & String| is a subtype of |Int| and |M[X & String]| reduces to |String| (|M|'s first case applies).
+The unsoundness manifests itself in the body of class |D|, where |X| |&| |String| is a subtype of |Int| and |M[X| |&| |String]| reduces to |String| (|M|'s first case applies).
 There, it is possible to call the function~|f| with a string argument, which would result in a runtime error.
-Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int & String]| from reducing given that |Int & String| is not inhabited.
+Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int| |&| |String]| from reducing given that |Int| |&| |String| is not inhabited.
 
@@
 These annotations allow programmers to specify how the subtyping of annotated parameters influences the subtyping of the higher-kinded type.
-For instance, |type F[+T]| defines a type |F| that is covariant in its first type parameter, meaning that |T1 <: T2| implies |F[T1] <: F[T2]|.
-Contravariance, written |type G[-T]|, has the opposite meaning: |T1 <: T2| implies |G[T2] <: G[T1]|.
+For instance, |type F[+T]| defines a type |F| that is covariant in its first type parameter, meaning that |T1| |<:| |T2| implies |F[T1]| |<:| |F[T2]|.
+Contravariance, written |type G[-T]|, has the opposite meaning: |T1| |<:| |T2| implies |G[T2]| |<:| |G[T1]|.
 
-It would appear that co- and contravariant types are always overlapping, given that, for all types |X|, |F[Nothing] <: F[X]| and |G[Any] <: G[X]| (where |Nothing| and |Any| are Scala's bottom and top types).
+It would appear that co- and contravariant types are always overlapping, given that, for all types |X|, |F[Nothing]| |<:| |F[X]| and |G[Any]| |<:| |G[X]| (where |Nothing| and |Any| are Scala's bottom and top types).
 However, in the case of covariant parameters, an exception can be made when the said type parameter corresponds to a field or a constructor parameter: the |Nothing| instantiation can be ruled out because no runtime program can produce a value of that type.

### 2022-02-14 symposium.tex
Add footnote and QE and character class

@@
 Establishing the nullability of capturing groups is more complicated than it sounds.
-At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position\footnote{
+At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position
+\footnote{
 For brevity, our presentation does not account for the regex alternative operator, which also influences the nullability of capturing groups and can appear in both prefix and suffix positions.
@@
 For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
-To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups:
+To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups\footnote{
+For simplicity, our presentation treats all non-escaped parentheses as capturing group delimiters.
+Our implementation also takes \lstinline!\\Q...\\E! quotations into account and ignores parentheses that appear in character classes.
+}:
 

### 2022-02-14 thesis.tex
Add footnote and QE and character class

@@
 This definition is parametrized by the list's element type and by an ordering of that type, passed as an implicit parameter.
-Since |listOrdering| is itself implicit, it defines a \emph{rule}: it allows the compiler to materialize an implicit |Ordering[List[T]]|, given an implicit |Ordering[T]| (for any type |T|).
+Since |listOrdering| is itself implicit, it defines a \emph{rule}: it allows the compiler to materialize an implicit of type |Ordering[List[T]]| given an implicit type |Ordering[T]|, for any type |T|.
 
@@
 \begin{definition*}
-  $\Γ \⊢ \S \⇌ \T$ (defined in \Cref{fig:harpoon}) represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \emph{\SMatch1/2}, \emph{\SMatch3/4}, and \emph{\STrans} in both directions.
+  $\Γ \⊢ \S \⇌ \T$ (defined in \Cref{fig:harpoonx}) represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \emph{\SMatch1/2}, \emph{\SMatch3/4}, and \emph{\STrans} in both directions.
 \end{definition*}

### 2022-02-14 scala/src/main/scala/regex-examples.scala

@@ -52,8 +52,9 @@ object NaiveIsNullable {
 // start section regexNaiveIsNullable
 type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
   CharAt[R, At] match
-    case ")" => IsMarked[R, At + 1, Hi]
-    case _   => IsNullable[R, At + 1, Hi]
+    case ")"  => IsMarked[R, At + 1, Hi]
+    case "\\" => IsNullable[R, At + 2, Hi]
+    case _    => IsNullable[R, At + 1, Hi]
 
 type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
   At match
@@ -86,6 +87,7 @@ type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
 // end section regexIsNullable
 
 object Last {
+import Lib.IsNullable
 
 // start section regexLastIteration
 type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =

### 2022-02-15 ack.tex
typos

@@
 
-Last but not the least, I would like to thank Dragana for being always there on my side, for being so kind and patient, and for her help in putting the ideas of this thesis into text.
+Last but not the least, I would like to thank Dragana for being always there by my side, for being so kind and patient, and for her help in putting the ideas of this thesis into text.
 You are the best.

### 2022-02-15 scala/src/main/scala/regex-examples.scala

@@ -18,7 +18,7 @@ val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
 // end section regexDocumentation
 
 // start section regexRational
-val rational = Regex("(\\d+)(?:\\.(\\d+))?")
+val rational = Regex("(\\d+)\\.?(\\d+)?")
 "3.1415" match
   case rational(i, Some(f)) =>
     val n = i.size + f.size

### 2022-02-17 symposium.tex
Update symposium.tex

@@
 
-While the syntax of this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
+While the syntax used in this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
 First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
@@
 We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
+\thesisOnly{
 The source code of our implementation is available online\footnote{\texttt{https://gist.github.com/OlivierBlanvillain/418dcb74cbd2e12eb4a90b7336f40b17}}.
+}
+\paperOnly{
+The source code of our implementation is available in the supplementary material of this submission.
+}
 
@@
 Similar to its standard library counterpart, that class uses an instance of |java.util.regex.Pattern| to match the given input against the given regular expression, and to extract the content of the capturing groups when the matching succeeds.
-These operations take place within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.
+These operations happen within the |unapply| method, which is the way to define custom pattern matching extractors in Scala \citep{emir2007matching}.
 
-The main difference between Scala's original |Regex| implementation and ours lies in the signature of the |unapply| method, which determines the nature of extraction performed by that method.
-In our implementation, we instantiate the type |P| to a tuple and use that type in the result type of the |unapply| method to tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity.
+The main difference between Scala's original |Regex| implementation and ours lies in the signature of |unapply|, which determines the nature of pattern matching extraction performed by that method.
+In our implementation, we instantiate the type |P| to a tuple and use that type in the result type of |unapply| to tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity.
 Furthermore, the type of those groups corresponds to the type of the tuple elements.
@@
 \noindent
-Here, we instantiate |Regex|'s type parameter to |P = (String, Option[String])|, which we then use in |unapply|'s return type to obtain a precise representation of the regex's capturing groups.
+Here, we instantiate |Regex|'s type parameter to |P = (String,| |Option[String])|, which we then use in |unapply|'s result type to obtain a precise representation of the regex's capturing groups.
 The resulting code is type-safe: if we change the pattern to omit the option unpacking (replacing |Some(f)| by |f|), the compiler would raise a type error on the call to |.size| (size is defined on strings but not options).
@@
 Be reassured, this is not an oversight!
-Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as "unexpected internal error").
+Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error").
 Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
@@
 The runtime component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
-Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple, with nullable elements wrapped in options.
+Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple and warpping the nullable elements in options.
 That transformation must conform to the representation computed at the type level.
@@
 \noindent
-After having computed those functions, we simply do a pointwise application with the raw output of the regex engine, package the result into a tuple, \emph{et voilà!}
+After having computed those functions, we do a pointwise application with the raw output of the regex engine, package the result into a tuple, \emph{et voilà!}
 
@@
 The first alternative that comes to mind is perhaps to turn ourselves to a dependently typed language with support for type- and term-level polymorphism.
-In a language with shared term and type syntax, we could write our analysis generically, and derive two programs from it, one for each level.
+Using a language with shared term and type syntax would allow us to write our analysis generically, and derive two programs from it, one for each level.
 
-In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions.
+In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions\paperOnly{\citep[§ 2.1]{blanvillain2022type}}.
+\thesisOnly{
 In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
+}
 In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
 In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
-For instance, the compiler does not recognize |CharAt["foo", 0]| as a valid type for |"foo".charAt(0)|.
+For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)|.
 
 In the long run, it would be interesting to investigate extending Scala with a "precise" mode of type inference, that would automatically infer match types and singleton types, whenever possible.
+\thesisOnly{
 In \Cref{chap:generalizing-singleton-types}, we demonstrated the feasibility of this approach by generalizing Scala's singleton types.
 A promising avenue would be to revisit that line of work with the less ambitious goal of improving Scala's type inference.
+}
 
@@
 The second version of our runtime component takes a completely different approach.
-Instead of analyzing regular expressions at run-time, to reconstitute type-level analysis results, we use implicit resolution to \emph{synthesize} a type-specialized runtime.
+Instead of analyzing regular expressions at run-time, we use implicit resolution to \emph{synthesize} a type-specialized runtime, based on the results of our type-level analysis.
 That synthesis is type-directed: it generates a single-purpose program that sanitizes the output of the regex engine by following the shape of the tuple type computed at the type level.
@@
 From a design standpoint, this solution is appealing as it leads to no duplicated code or computation: we perform our capturing group analysis at the type-level, once and for all, and synthesize code accordingly.
-This combination of type-level programming and implicit-based code synthesis is reminiscent of multi-stage metaprogramming, in its compile-time variant \citep{stucki2018practical}.
+This combination of type-level programming and implicit-based code synthesis is reminiscent of multi-stage metaprogramming, in its com\-pile-time variant \citep{stucki2018practical}.
 In those terms, our approach corresponds to a multi-staged program whose stage 0 happens entirely within Scala's type checker.
@@
 To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
-We select the subset of those tests that are positive (they match successfully) and exercise capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
+We select the subset of those tests that are positive and exercise capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
 We use these benchmarks to evaluate the performance of our library in terms of compilation times and execution times.
 
-In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code-duplicated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.
+In \Cref{fig:regexBars}, we show the results of our experiments, which compare Scala's standard regular expression library (Std) against two flavors of our library, one using our code-dupli\-cated runtime (Dup), presented in \Cref{subsec:we-ain-t-need-no-dependent-types}, and the other using our implicit-based runtime (Impl), presented in \Cref{subsec:implicit-based-extractor-synthesis}.
 
@@
 
-We obtained these results by averaging the measurements obtained over 2-hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0\_212 and Linux.
+We obtained these results by averaging the measurements obtained over 2-hour runs, for each data point, which we executed on an i7-7700K Processor running Oracle JVM 1.8.0 and Linux.
 With these settings, we obtain margins of errors below \textpm1\%, with 99.9\% confidence (omitted in \Cref{fig:regexBars}).
@@
 
-There is surprisingly little literature concerned with the safety of regular expression capturing groups.
-In a recent survey paper on the correctness of regular expressions, \citet{li2021ensuring} classified the publications in this area into several categories, ranging from empirical studies to regular expression synthesis and repair.
+We found surprisingly little literature concerned with the safety of regular expression capturing groups.
+In a recent survey paper on regular expression correctness, \citet{li2021ensuring} classified the publications in this area into several categories, ranging from empirical studies to regular expression synthesis and repair.
 However, they only listed a single paper about static checking: the type system developed by \citet{spishak2012type} to validate regular expression syntax and capturing group usage in Java programs.
@@
 
-Stephanie Weirich presented, on several occasions \citep{weirich2014examples}, a regular expression library as an example of a dependently-typed program written in Haskell.
+Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
 To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
-In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala~3.
+In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala.
 An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.
@@
 In our current design, we map every capturing group to an |Option[String]|, which does not always accurately reflect the structure of regular expressions.
-For instance, in \lstinline!"(A)(B)|(C)"!
-(which matches either |AB| or |C|), our system represents the capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
+For instance, when confronted with \lstinline!"(A)(B)|(C)"!, our system represents this expression's capturing groups as a 3-tuple, |(Option[String],| |Option[String],| |Option[String])|.
 This representation fails to account for the fact that the first two groups have identical nullability status, and that this status is opposite to the nullability of the third group.

### 2022-02-17 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -11,15 +11,16 @@ import Lib._
 
 // start section regexUserLevel
 object Regex:
-  def apply[R <: String & Singleton](regex: R): Regex[Compile[R]] =
-    new Regex(regex)
+  def apply[R <: String & Singleton](regex: R) =
+    new Regex[Compile[R]](regex)
 
 class Regex[P] private (val regex: String):
   val pattern = Pattern.compile(regex)
   def unapply(s: String): Option[P] =
     val m = pattern.matcher(s)
     if (m.matches())
-      val a = Array.tabulate(m.groupCount)(i => m.group(i + 1))
+      val a = Array.tabulate(m.groupCount)(
+        i => m.group(i + 1))
       Some(transform[P](regex, a))
     else
       None
@@ -114,9 +115,9 @@ type IsCapturing[R <: String, At <: Int] <: Boolean =
     case "?" => CharAt[R, At + 1] match
       case "<" => CharAt[R, At + 2] match
         case "=" | "!" => false // lookbehinds
-        case _ => true          // named-capturing group
-      case _ => false           // other special constructs
-    case _ => true              // unnamed-capturing group
+        case _ => true // named-capturing group
+      case _ => false  // other special constructs
+    case _ => true     // unnamed-capturing group
 // end section regexIsCapturing
 
 type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =

### 2022-02-17 scala/src/main/scala/regex-examples.scala

@@ -32,7 +32,7 @@ import compiletime.ops.int.+
 type Compile[R <: String] =
   Reverse[Loop[R, 0, Length[R], EmptyTuple]]
 
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple] <: Tuple =
+type Loop[R, Lo, Hi, Acc <: Tuple] <: Tuple =
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
@@ -50,13 +50,13 @@ check["((B(C)))", (S, S, S)]
 
 object NaiveIsNullable {
 // start section regexNaiveIsNullable
-type IsNullable[R <: String, At <: Int, Hi <: Int] <: Boolean =
+type IsNullable[R <: String, At, Hi] =
   CharAt[R, At] match
     case ")"  => IsMarked[R, At + 1, Hi]
     case "\\" => IsNullable[R, At + 2, Hi]
     case _    => IsNullable[R, At + 1, Hi]
 
-type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+type IsMarked[R <: String, At, Hi] <: Boolean =
   At match
     case Hi => false
     case _ =>
@@ -67,7 +67,7 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
 
 }
 
-type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
+type IsMarked[R <: String, At, Hi] <: Boolean =
   At match
     case Hi => false
     case _ =>
@@ -76,7 +76,7 @@ type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
         case _ => false
 
 // start section regexIsNullable
-type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
+type IsNullable[R <: String, At, Hi, Lvl <: Int] <: Boolean =
   CharAt[R, At] match
     case ")" => Lvl match
       case 0 => IsMarked[R, At + 1, Hi]
@@ -87,10 +87,9 @@ type IsNullable[R <: String, At <: Int, Hi <: Int, Lvl <: Int] <: Boolean =
 // end section regexIsNullable
 
 object Last {
-import Lib.IsNullable
 
 // start section regexLastIteration
-type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] =
[…]

### 2022-02-17 scala/regex/src/main/scala/bench/RegexPackage.scala
Tweak code examples

@@ -109,16 +109,14 @@ type Loop[R <: String, Lo <: Int, Hi <: Int, Acc <: Tuple, Opt <: Int] <: Tuple 
           case true  => Loop[R, Lo + 1, Hi, Option[String] *: Acc, Opt + 1]
       case _ => Loop[R, Lo + 1, Hi, Acc, Opt]
 
-// start section regexIsCapturing
 type IsCapturing[R <: String, At <: Int] <: Boolean =
   CharAt[R, At] match
     case "?" => CharAt[R, At + 1] match
       case "<" => CharAt[R, At + 2] match
         case "=" | "!" => false // lookbehinds
         case _ => true // named-capturing group
-      case _ => false  // other special constructs
-    case _ => true     // unnamed-capturing group
-// end section regexIsCapturing
+      case _ => false // other special constructs
+    case _ => true // unnamed-capturing group
 
 type IsMarked[R <: String, At <: Int, Hi <: Int] <: Boolean =
   At match
@@ -261,7 +259,7 @@ def isPiped(r: String, at: Int, hi: Int, lvl: Int): Int =
 // those types. Furthermore, the code responsible for generating _n
 // extractors for tuples does not know about generic tuples, so we also
 // manually flattern those types here...
-type Reverse[L <: Tuple] = L match
+type Reverse[L] = L match
   case EmptyTuple => Unit
   case x1 *: EmptyTuple => x1
   case x1 *: x2 *: EmptyTuple => (x2, x1)

### 2022-02-17 scala/src/main/scala/regex-examples.scala

@@ -3,12 +3,37 @@ package bench
 import annotation.experimental
 import bench._
 import RegexPackage._
-import Lib.{CharAt, Reverse}
+import Lib.Reverse
+import compiletime.ops.string.Substring
 import compiletime.ops.int.Max
-import compiletime.ops.int.-
+import compiletime.ops.int.{-, +}
 
 @experimental
 object RegexExamples {
+
+type CharAt[R <: String, At <: Int] =
+  Substring[R, At, At + 1] match {
+    case ")" => ')'
+    case "(" => '('
+    case "\\" => '\\'
+    case "?" => '?'
+    case "*" => '?'
+    case "<" => '<'
+    case "=" => '='
+    case "!" => '!'
+    case _ => '_'
+  }
+
+// start section regexIsCapturing
+type IsCapturing[R <: String, At <: Int] =
+  CharAt[R, At] match
+    case "?" => CharAt[R, At + 1] match
+      case "<" => CharAt[R, At + 2] match
+        case "=" | "!" => false // lookbehinds
+        case _ => true // named-capturing group
+      case _ => false // other special constructs
+    case _ => true // unnamed-capturing group
+// end section regexIsCapturing
 
 // start section regexDocumentation
 val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
@@ -32,12 +57,12 @@ import compiletime.ops.int.+
 type Compile[R <: String] =
   Reverse[Loop[R, 0, Length[R], EmptyTuple]]
 
-type Loop[R, Lo, Hi, Acc <: Tuple] <: Tuple =
+type Loop[R, Lo, Hi, Acc <: Tuple] =
   Lo match
     case Hi => Acc
     case _  => CharAt[R, Lo] match
-      case "("  => Loop[R, Lo + 1, Hi, String *: Acc]
-      case "\\" => Loop[R, Lo + 2, Hi, Acc]
+      case '('  => Loop[R, Lo + 1, Hi, String *: Acc]
+      case '\\' => Loop[R, Lo + 2, Hi, Acc]
       case _ => Loop[R, Lo + 1, Hi, Acc]
 // end section regexFirstIteration
 
@@ -52,16 +77,16 @@ object NaiveIsNullable {
 // start section regexNaiveIsNullable
 type IsNullable[R <: String, At, Hi] =
   CharAt[R, At] match
[…]

### 2022-02-17 symposium.tex
More tweaks

@@
 In this chapter, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
-Our design makes extensive use of match types, Scala's new feature for type-level programming, to statically analyze regular expressions during type checking.
+Our design makes extensive use of match types \paperOnly{\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
 We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
@@
 
-  \item The runtime capturing group processing, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation computed.
+  \item The runtime capturing group processing component, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation.
   This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.
@@
 
-To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep[perl-tests.xml]{w3c1994qt3ts}.
+To assess the correctness of our library, we run our implementation against the QT3TS test suite \citep{w3c1994xquery}, which consists of about 1700 test cases for regular expressions \citep{w3c1994qt3ts}.
 We select the subset of those tests that are positive and exercise capturing groups, totaling 183 test cases, which we then collated to form our benchmark suite.
@@
 Both variants are thus relatively cheap, which validates their utility for practical uses.
-When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[2.11.3]{petrashko2017design}.
+When it comes to compilation times, it would be misleading to compare increments relative to the overall time, given that the compiler spends a large portion of its time outside of type checking \citep[§ 2.11.3]{petrashko2017design}.
 

### 2022-02-17 scala/regex/src/main/scala/bench/RegexPackage.scala

@@ -35,18 +35,16 @@ object Sanitizer:
     new Sanitizer[EmptyTuple](0):
       def mutate(arr: Array[Any]): Unit = ()
 
-  implicit def stringcase[T <: Tuple]
-    (implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
-      new Sanitizer[String *: T](ev.i + 1):
+  implicit def stringcase[T <: Tuple](implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
+      new Sanitizer[String *: T](ev.i+1):
         def mutate(arr: Array[Any]): Unit =
-          assert(arr(arr.size - i) != null)
+          assert(arr(arr.size-i) != null)
           ev.mutate(arr)
 
-  implicit def optioncase[T <: Tuple]
-    (implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
-      new Sanitizer[Option[String] *: T](ev.i + 1):
+  implicit def optioncase[T <: Tuple](implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
+      new Sanitizer[Option[String] *: T](ev.i+1):
         def mutate(arr: Array[Any]): Unit =
-          arr(arr.size - i) = Option(arr(arr.size - i))
+          arr(arr.size-i) = Option(arr(arr.size-i))
           ev.mutate(arr)
 // end section regexSanitizerTypeClass
 

### 2022-02-17 scala/src/main/scala/regex-examples.scala

@@ -84,10 +84,9 @@ type IsNullable[R <: String, At, Hi] =
 type IsMarked[R <: String, At, Hi] =
   At match
     case Hi => false
-    case _ =>
-      CharAt[R, At] match
-        case '?' | '*' => true
-        case _ => false
+    case _ => CharAt[R, At] match
+      case '?' | '*' => true
+      case _ => false
 // end section regexNaiveIsNullable
 
 }
@@ -135,55 +134,34 @@ type Loop[R, Lo, Hi, Opt <: Int, Acc <: Tuple] =
 def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean = false
 
 // start section regexTermLvlLoop
-val identity: String => Any = { x => assert(x != null); x }
+val id: String => Any = { x => assert(x != null); x }
+val opt: String => Any = { x => Option(x) }
 
-def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int)
-    : List[String => Any] =
+def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[String => Any] =
   lo match
     case `hi` => acc
     case _  => r.charAt(lo) match
-      case ')' => loop(r, lo + 1, hi, acc, 0.max(opt - 1))
+      case ')' =>
+        loop(r, lo+1, hi, acc, 0.max(opt-1))
       case '(' => opt match
-        case 0 => isNullable(r, lo + 1, hi, 0) match
-          case true  => loop(r, lo + 1, hi, Option.apply :: acc, 1)
-          case false => loop(r, lo + 1, hi, identity :: acc, 0)
-        case _ => loop(r, lo + 1, hi, Option.apply :: acc, opt + 1)
-      case '\\' => loop(r, lo + 2, hi, acc, opt)
-      case _ => loop(r, lo + 1, hi, acc, opt)
+        case 0 => isNullable(r, lo+1, hi, 0) match
+          case true =>
+            loop(r, lo+1, hi, opt::acc, 1)
+          case false =>
+            loop(r, lo+1, hi, id::acc, 0)
+        case _ => loop(r, lo+1, hi, opt::acc, opt+1)
+      case '\\' => loop(r, lo+2, hi, acc, opt)
+      case _ => loop(r, lo+1, hi, acc, opt)
 // end section regexTermLvlLoop
 
 // start section regexTransform
[…]

### 2022-03-01 symposium.tex
Incorporate Georg's comments

@@
 This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
-The Scala implementation innovates in its presentation: it uses the flexibility of Scala's pattern matching to offer a syntax that is both elegant and concise.
+The Scala implementation innovates in its presentation: it improves upon Java's solution by providing a more ergonomic API based on pattern matching that is both elegant and concise.
 The package's documentation starts with the following example:
@@
 
-In this chapter, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
+In this \thesisOnly{chapter} \paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
 Our design makes extensive use of match types \paperOnly{\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
@@
 
-Be reassured, this is not an oversight!
+Rest assured, this is not an oversight!
 Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error").
 Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
-If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than error at run-time.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than same error at run-time.
 
@@
 Establishing the nullability of capturing groups is more complicated than it sounds.
-At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position
+At first glance, it seems that this is simply a matter of looking for regex quantifiers in suffix position%
 \footnote{
@@
 \noindent
-Unfortunately, this naive solution is incorrect for regular expressions with nested capturing groups.
+Unfortunately, this naive solution does not handle regular expressions with nested capturing groups.
 For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
-To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups\footnote{
+To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups%
+\footnote{
 For simplicity, our presentation treats all non-escaped parentheses as capturing group delimiters.
@@
 The runtime component of our library is in charge of sanitizing and packaging the results of regular expression matchings.
-Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple and warpping the nullable elements in options.
+Concretely, this component's job consists of transforming the string array that comes out of the regex engine into an appropriately sized tuple and wrapping the nullable elements in options.
 That transformation must conform to the representation computed at the type level.
@@
 \paragraph{Execution Times} At runtime, our code-duplicated variant adds 51~\textmu s over the baseline.
-Although this difference is three orders of magnitude smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis.
+Although this difference is three orders of magnitude smaller than at compile-time, it corresponds to a 25\% increase, which is a rather substantial price to pay for redoing an already-performed analysis (|transform| redoes the analysis performed by |Compile|).
 The implicit-based variant gets rid of this cost by synthesizing a type-specialized runtime (at the cost of increased compilation time).

### 2022-03-02 symposium.tex
Background

@@
 \section{Introduction}
+
+Capturing groups allow programmers to extract substrings matched by parts of a regular expression.
+For example, the regex |"A(B)?"| matches both |"A"| and |"AB"|.
+In the first case, the capturing group is empty, and in the second case, the capturing group contains |"B"|.
 
@@
 This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
-The Scala implementation innovates in its presentation: it improves upon Java's solution by providing a more ergonomic API based on pattern matching that is both elegant and concise.
+The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
 The package's documentation starts with the following example:
@@
 
-In this \thesisOnly{chapter} \paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
-Our design makes extensive use of match types \paperOnly{\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
+In this \thesisOnly{chapter}\paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
+Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
 We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
+
+Our presentation is structured as follows:
+
+\begin{itemize}
+  \item
+\Cref{sec:background}
+  \item
+\Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}
+  \item
+\Cref{sec:evaluation}
+  \item
+\Cref{sec:related}
+\Cref{sec:conclusion}
+
+\end{itemize}
+
 \thesisOnly{
@@
 
+\section{Background}
+\label{sec:background}
+
+\subection{Match Types}
+
+In this section, we give a brief introduction to match types by inspecting the following example:
+
+\elemExample
+% type Elem[X] = X match
+%   case String => Char
+%   case Iterable[t] => Elem[t]
+%   case Any => X
+
+\noindent
+This example defines a type |Elem| parametrized by one type parameter |X|.
+The right-hand side is defined in terms of a match on that type parameter---a match type.
+A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
+For example, the above type reduces as follows:
+
+\begin{lstlisting}
+Elem[String] ~> Char
+Elem[Int] ~> Int
+Elem[Iterable[Int]] ~> Int
+\end{lstlisting}
+
+\noindent
+To reduce a match type, the scrutinee is compared to each pattern, one after the other, using subtyping.
+For example, although |String| is a subtype of both |String| and |Any|, the |Elem[String]| type reduces to |Char| because the corresponding case appears first.
+
+\subection{The Generic View of Tuples}
+
+% https://www.scala-lang.org/2021/02/26/tuples-bring-generic-programming-to-scala-3.html
+
 \section{Architecture}
+\label{sec:architecture}
 
@@
 Unfortunately, this naive solution does not handle regular expressions with nested capturing groups.
-For example, in |(A(B)?)|, that solution incorrectly labels the first group as optional.
+For example, in |"(A(B)?)"|, that solution incorrectly labels the first group as optional.
 To overcome this problem, we update our solution to keep track of the number of opening and closing parentheses (|Lvl|), which allows us to differentiate closing parentheses of inner and outer groups%
@@
 When an outer group is deemed optional, this overrides the nullability of all its inner groups, which also become optional.
-For instance, in |(A(B))?|, both the first (|A(B)|) and the second (|B|) capturing group are optional.
+For instance, in |"(A(B))?"|, both the first (|A(B)|) and the second (|B|) capturing group are optional.
 
@@
 
-In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions\paperOnly{\citep[§ 2.1]{blanvillain2022type}}.
+In principle, Scala~3's type inference should be able to solve half of that problem with its ability to correlate match types and match expressions\paperOnly{~\citep[§ 2.1]{blanvillain2022type}}.%
 \thesisOnly{
 In \Cref{subsec:a-lightweight-form-of-dependent-typing}, we show an example of that ability, where the compiler type checks a small program using a match type.
-}
+}%
 In our case, this mechanism should allow us to get rid of the unsafe cast but does not address the problem of code duplication.
@@
 \section{Evaluation}
+\label{sec:evaluation}
 
@@
 \section{Related Work}
+\label{sec:related}
 
@@
 \section{Conclusion}
+\label{sec:conclusion}
 
-In this chapter, we introduced a new design for type-safe regular expressions in Scala.
+In this \thesisOnly{chapter}\paperOnly{paper}, we introduced a new design for type-safe regular expressions in Scala.
 We presented our type-level analysis, which identifies capturing groups and computes their nullability.

### 2022-03-02 scala/src/main/scala/regex-examples.scala

@@ -24,6 +24,16 @@ type CharAt[R <: String, At <: Int] =
     case _ => '_'
   }
 
+// start section elemExample
+type Elem[X] = X match
+  case String => Char
+  case Iterable[t] => Elem[t]
+  case Any => X
+// end section elemExample
+
+val rev0test: Reverse0[String *: Int *: EmptyTuple, EmptyTuple]
+  = (1, "")
+
 // start section regexIsCapturing
 type IsCapturing[R <: String, At <: Int] =
   CharAt[R, At] match
@@ -135,7 +145,7 @@ def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean = false
 
 // start section regexTermLvlLoop
 val id: String => Any = { x => assert(x != null); x }
-val opt: String => Any = { x => Option(x) }
+val option: String => Any = { x => Option(x) }
 
 def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[String => Any] =
   lo match
@@ -146,10 +156,10 @@ def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[
       case '(' => opt match
         case 0 => isNullable(r, lo+1, hi, 0) match
           case true =>
-            loop(r, lo+1, hi, opt::acc, 1)
+            loop(r, lo+1, hi, option::acc, 1)
           case false =>
             loop(r, lo+1, hi, id::acc, 0)
-        case _ => loop(r, lo+1, hi, opt::acc, opt+1)
+        case _ => loop(r, lo+1, hi, option::acc, opt+1)
       case '\\' => loop(r, lo+2, hi, acc, opt)
       case _ => loop(r, lo+1, hi, acc, opt)
 // end section regexTermLvlLoop

### 2022-03-08 symposium.tex
Related work WIP

@@
 
-Our presentation is structured as follows:
-
-\begin{itemize}
-  \item
-\Cref{sec:background}
-  \item
-\Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}
-  \item
-\Cref{sec:evaluation}
-  \item
-\Cref{sec:related}
-\Cref{sec:conclusion}
-
-\end{itemize}
+This paper is structured as follows.
+In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
+In \Cref{sec:architecture},~\ref{sec:type-level} and~\ref{sec:term-level}, we present our library for type-safe regular expressions.
+In \Cref{sec:evaluation}, we evaluate the performance of our implementation in terms of compilation times and execution times.
+In \Cref{sec:related}, we discuss related work.
+We conclude the paper in \Cref{sec:conclusion}.
 
@@
 
-\subection{Match Types}
+In this section, we give a brief introduction to two recent additions to Scala's type system: match types and generic tuples.
+Our regular expression library makes extensive use those two features to enable type-safe and null-safe capturing group extraction.
 
-In this section, we give a brief introduction to match types by inspecting the following example:
+\subsection{Match Types}
+
+Match types provide first-class support for type-level computations by allowing programmers to perform pattern matching on types:
 
@@
 %   case String => Char
-%   case Iterable[t] => Elem[t]
+%   case Array[t] => Elem[t]
 %   case Any => X
@@
 This example defines a type |Elem| parametrized by one type parameter |X|.
-The right-hand side is defined in terms of a match on that type parameter---a match type.
+The right-hand side is defined in terms of a match on that type parameter -- a match type.
 A match type reduces to one of its right-hand sides, depending on the type of its scrutinee.
@@
 \begin{lstlisting}
-Elem[String] ~> Char
-Elem[Int] ~> Int
-Elem[Iterable[Int]] ~> Int
+Elem[String] =:= Char
+Elem[Int] =:= Int
+Elem[Array[Int]] =:= Int
 \end{lstlisting}
@@
 
-\subection{The Generic View of Tuples}
+\subsection{Generic Tuples}
+\label{subsec:generic-tuples}
 
-% https://www.scala-lang.org/2021/02/26/tuples-bring-generic-programming-to-scala-3.html
+Scala's tuples were originally defined as plain old data types.
+In Scala~3, tuples got enhanced with a generic representation~\citep{bazzucchi2021tuples}, similar to heterogeneous lists~\citep{kiselyov2004strongly}.
+This representation uses two types, |EmptyTuple| and |*:|, to describe tuples in a list-like fashion.
+The compiler treats those new types and the traditional class-based representation of tuples interchangeably.
+A 2-tuple of integers, for example, has two equivalent representations:
+
+\begin{lstlisting}
+(Int, Int) =:= Int *: Int *: EmptyTuple
+\end{lstlisting}
+
+\noindent
+This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples in a recursive manner.
+For example, the following match type reverses the order of a tuple's elements:
+
+\tupleReverseA
+\vspace{-3pt}
+\tupleReverseB
 
@@
 
-We present our library in terms of 3 components:
+We present our library for type-safe regular expressions in terms of 3 components:
 
@@
 Similarly, the |+| type (from |ops.int|) allows us to manipulate integers just like we would at the term level.
-The |EmptyTuple| and |*:| types are Scala~3's new generic representation of tuples.
-Those types remove the need for heterogeneous lists and let us manipulate tuples directly.
+The |EmptyTuple| and |*:| types are Scala~3's new generic representation of tuples, presented in \Cref{subsec:generic-tuples}.
 
@@
 
+The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
+The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
+Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
+At the time of writing, the library is still in an early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.
+
+The safe-regex-rs~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to code generate arity-specific extractors.
+Unlike the
+Their implementation is
+
+Rust \url{https://gitlab.com/leonhard-llc/safe-regex-rs}
+
 \section{Conclusion}
@@
 
-In this \thesisOnly{chapter}\paperOnly{paper}, we introduced a new design for type-safe regular expressions in Scala.
+In this \thesisOnly{chapter}\paperOnly{paper}, we introduce a new design for type-safe regular expressions in Scala.
 We presented our type-level analysis, which identifies capturing groups and computes their nullability.

### 2022-03-08 scala/src/main/scala/regex-examples.scala

@@ -27,12 +27,27 @@ type CharAt[R <: String, At <: Int] =
 // start section elemExample
 type Elem[X] = X match
   case String => Char
-  case Iterable[t] => Elem[t]
+  case Array[t] => Elem[t]
   case Any => X
 // end section elemExample
 
-val rev0test: Reverse0[String *: Int *: EmptyTuple, EmptyTuple]
+object Rev {
+
+// start section tupleReverseA
+type Reverse[T <: Tuple] = Rev[T, EmptyTuple]
+// end section tupleReverseA
+
+// start section tupleReverseB
+type Rev[T <: Tuple, Acc <: Tuple] <: Tuple =
+  T match
+    case x *: xs => Rev[xs, x *: Acc]
+    case EmptyTuple => Acc
+// end section tupleReverseB
+
+val rev0test: Rev[String *: Int *: EmptyTuple, EmptyTuple]
   = (1, "")
+
+}
 
 // start section regexIsCapturing
 type IsCapturing[R <: String, At <: Int] =

### 2022-03-10 symposium.tex
Submission ready

@@
 Furthermore, the values that come out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
-Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale code base.
+Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale codebase.
 
@@
 In this section, we give a brief introduction to two recent additions to Scala's type system: match types and generic tuples.
-Our regular expression library makes extensive use those two features to enable type-safe and null-safe capturing group extraction.
+Our regular expression library makes extensive use of those two features to enable type-safe and null-safe capturing group extraction.
 
@@
 
-Match types provide first-class support for type-level computations by allowing programmers to perform pattern matching on types:
+Match types provide first-class support for type-level computations in the form of pattern matching on types:
 
@@
 \noindent
-This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples in a recursive manner.
+This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples recursively.
 For example, the following match type reverses the order of a tuple's elements:
@@
 
-  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe experience.
+  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing a type-safe and null-safe experience.
 \end{enumerate}
@@
 Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
-If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than same error at run-time.
+If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
 
@@
 \noindent
-Similar to our handling of backslash characters, this implementation intentionally does not guard against out-of-bounds errors, since these errors correspond to ill-formed regular expressions.
+Similar to our handling of backslash characters, this implementation intentionally does not prevent out-of-bounds errors, since these errors correspond to ill-formed regular expressions.
 
@@
 Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
-At the time of writing, the library is still in an early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.
+At the time of writing, the library is still in the early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.
 
-The safe-regex-rs~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to code generate arity-specific extractors.
-Unlike the
-Their implementation is
-
-Rust \url{https://gitlab.com/leonhard-llc/safe-regex-rs}
+The safe-regex~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to statically compile regular expressions and generate arity-specific extractors.
+Performance-wise, macros improve over the status quo since they remove the need to compile regular expressions at runtime.
+Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.
 
@@
 We presented our type-level analysis, which identifies capturing groups and computes their nullability.
-We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
+We built our library following the design of the original Scala regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
 We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.

### 2022-03-10 scala/src/main/scala/regex-examples.scala

@@ -61,14 +61,14 @@ type IsCapturing[R <: String, At <: Int] =
 // end section regexIsCapturing
 
 // start section regexDocumentation
-val date = Regex("(\\d{4})-(\\d{2})-(\\d{2})")
+val date = Regex("""(\d{4})-(\d{2})-(\d{2})""")
 "2004-01-20" match
   case date(y, m, d) =>
     s"$y was a good year for PLs."
 // end section regexDocumentation
 
 // start section regexRational
-val rational = Regex("(\\d+)\\.?(\\d+)?")
+val rational = Regex("""(\d+)(?:\.(\d+))?""")
 "3.1415" match
   case rational(i, Some(f)) =>
     val n = i.size + f.size
@@ -160,7 +160,7 @@ def isNullable(r: String, at: Int, hi: Int, lvl: Int): Boolean = false
 
 // start section regexTermLvlLoop
 val id: String => Any = { x => assert(x != null); x }
-val option: String => Any = { x => Option(x) }
+val nu: String => Any = { x => Option(x) }
 
 def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[String => Any] =
   lo match
@@ -171,10 +171,10 @@ def loop(r: String, lo: Int, hi: Int, acc: List[String => Any], opt: Int): List[
       case '(' => opt match
         case 0 => isNullable(r, lo+1, hi, 0) match
           case true =>
-            loop(r, lo+1, hi, option::acc, 1)
+            loop(r, lo+1, hi, nu::acc, 1)
           case false =>
             loop(r, lo+1, hi, id::acc, 0)
-        case _ => loop(r, lo+1, hi, option::acc, opt+1)
+        case _ => loop(r, lo+1, hi, nu::acc, opt+1)
       case '\\' => loop(r, lo+2, hi, acc, opt)
       case _ => loop(r, lo+1, hi, acc, opt)
 // end section regexTermLvlLoop

### 2022-03-10 slides.tex
Slides

\documentclass[10pt]{beamer}

\usetheme[progressbar=frametitle]{metropolis}
\usepackage{appendixnumberbeamer}

\usepackage{booktabs}
\usepackage[scale=2]{ccicons}

\usepackage{pgfplots}
\usepgfplotslibrary{dateplot}

\usepackage{xspace}
\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}

\title{Abstractions for Type-Level Programming}
\date{\today}
\author{Olivier Blanvillain}
\institute{EPFL}
% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}

\begin{document}

\maketitle

\begin{frame}{Table of contents}
  \setbeamertemplate{section in toc}[sections numbered]
  \tableofcontents%[hideallsubsections]
\end{frame}

\section[Intro]{Introduction}

\begin{frame}[fragile]{Metropolis}

  The \themename theme is a Beamer theme with minimal visual noise
  inspired by the \href{https://github.com/hsrmbeamertheme/hsrmbeamertheme}{\textsc{hsrm} Beamer
  Theme} by Benjamin Weiss.

  Enable the theme by loading

  \begin{verbatim}    \documentclass{beamer}
    \usetheme{metropolis}\end{verbatim}

  Note, that you have to have Mozilla's \emph{Fira Sans} font and XeTeX
  installed to enjoy this wonderful typography.
\end{frame}
\begin{frame}[fragile]{Sections}
  Sections group slides of the same topic

  \begin{verbatim}    \section{Elements}\end{verbatim}

  for which \themename provides a nice progress indicator \ldots

\end{frame}

\section{Titleformats}

\begin{frame}{Metropolis titleformats}
  \themename supports 4 different titleformats:
  \begin{itemize}
    \item Regular
    \item \textsc{Smallcaps}
    \item \textsc{allsmallcaps}
    \item ALLCAPS
  \end{itemize}
  They can either be set at once for every title type or individually.
\end{frame}

\subsection{Tricks}

{
    \metroset{titleformat frame=smallcaps}
\begin{frame}{Small caps}
  This frame uses the \texttt{smallcaps} titleformat.

  \begin{alertblock}{Potential Problems}
    Be aware, that not every font supports small caps. If for example you typeset your presentation with pdfTeX and the Computer Modern Sans Serif font, every text in smallcaps will be typeset with the Computer Modern Serif font instead.
  \end{alertblock}
\end{frame}
}

{
\metroset{titleformat frame=allsmallcaps}
\begin{frame}{All small caps}
  This frame uses the \texttt{allsmallcaps} titleformat.

  \begin{alertblock}{Potential problems}
    As this titleformat also uses smallcaps you face the same problems as with the \texttt{smallcaps} titleformat. Additionally this format can cause some other problems. Please refer to the documentation if you consider using it.

    As a rule of thumb: Just use it for plaintext-only titles.
  \end{alertblock}
\end{frame}
}

{
\metroset{titleformat frame=allcaps}
\begin{frame}{All caps}
  This frame uses the \texttt{allcaps} titleformat.

  \begin{alertblock}{Potential Problems}
    This titleformat is not as problematic as the \texttt{allsmallcaps} format, but basically suffers from the same deficiencies. So please have a look at the documentation if you want to use it.
  \end{alertblock}
\end{frame}
}

\section{Elements}

\begin{frame}[fragile]{Typography}
      \begin{verbatim}The theme provides sensible defaults to
\emph{emphasize} text, \alert{accent} parts
or show \textbf{bold} results.\end{verbatim}

  \begin{center}becomes\end{center}

  The theme provides sensible defaults to \emph{emphasize} text,
  \alert{accent} parts or show \textbf{bold} results.
\end{frame}

\begin{frame}{Font feature test}
  \begin{itemize}
    \item Regular
    \item \textit{Italic}
    \item \textsc{SmallCaps}
    \item \textbf{Bold}
    \item \textbf{\textit{Bold Italic}}
    \item \textbf{\textsc{Bold SmallCaps}}
    \item \texttt{Monospace}
    \item \texttt{\textit{Monospace Italic}}
    \item \texttt{\textbf{Monospace Bold}}
    \item \texttt{\textbf{\textit{Monospace Bold Italic}}}
  \end{itemize}
\end{frame}

\begin{frame}{Lists}
  \begin{columns}[T,onlytextwidth]
    \column{0.33\textwidth}
      Items
      \begin{itemize}
        \item Milk \item Eggs \item Potatos
      \end{itemize}

    \column{0.33\textwidth}
      Enumerations
      \begin{enumerate}
        \item First, \item Second and \item Last.
      \end{enumerate}

    \column{0.33\textwidth}
      Descriptions
      \begin{description}
        \item[PowerPoint] Meeh. \item[Beamer] Yeeeha.
      \end{description}
  \end{columns}
\end{frame}
\begin{frame}{Animation}
  \begin{itemize}[<+- | alert@+>]
    \item \alert<4>{This is\only<4>{ really} important}
    \item Now this
    \item And now this
  \end{itemize}
\end{frame}
\begin{frame}{Figures}
  \begin{figure}
    \newcounter{density}
    \setcounter{density}{20}
    \begin{tikzpicture}
      \def\couleur{alerted text.fg}
      \path[coordinate] (0,0)  coordinate(A)
                  ++( 90:5cm) coordinate(B)
                  ++(0:5cm) coordinate(C)
                  ++(-90:5cm) coordinate(D);
      \draw[fill=\couleur!\thedensity] (A) -- (B) -- (C) --(D) -- cycle;
      \foreach \x in {1,...,40}{%
          \pgfmathsetcounter{density}{\thedensity+20}
          \setcounter{density}{\thedensity}
          \path[coordinate] coordinate(X) at (A){};
          \path[coordinate] (A) -- (B) coordinate[pos=.10](A)
                              -- (C) coordinate[pos=.10](B)
                              -- (D) coordinate[pos=.10](C)
                              -- (X) coordinate[pos=.10](D);
          \draw[fill=\couleur!\thedensity] (A)--(B)--(C)-- (D) -- cycle;
      }
    \end{tikzpicture}
    \caption{Rotated square from
    \href{http://www.texample.net/tikz/examples/rotated-polygons/}{texample.net}.}
  \end{figure}
\end{frame}
\begin{frame}{Tables}
  \begin{table}
    \caption{Largest cities in the world (source: Wikipedia)}
    \begin{tabular}{lr}
      \toprule
      City & Population\\
      \midrule
      Mexico City & 20,116,842\\
      Shanghai & 19,210,000\\
      Peking & 15,796,450\\
      Istanbul & 14,160,467\\
      \bottomrule
    \end{tabular}
  \end{table}
\end{frame}
\begin{frame}{Blocks}
  Three different block environments are pre-defined and may be styled with an
  optional background color.

  \begin{columns}[T,onlytextwidth]
    \column{0.5\textwidth}
      \begin{block}{Default}
        Block content.
      \end{block}

      \begin{alertblock}{Alert}
        Block content.
      \end{alertblock}

      \begin{exampleblock}{Example}
        Block content.
      \end{exampleblock}

    \column{0.5\textwidth}

      \metroset{block=fill}

      \begin{block}{Default}
        Block content.
      \end{block}

      \begin{alertblock}{Alert}
        Block content.
      \end{alertblock}

      \begin{exampleblock}{Example}
        Block content.
      \end{exampleblock}

  \end{columns}
\end{frame}
\begin{frame}{Math}
  \begin{equation*}
    e = \lim_{n\to \infty} \left(1 + \frac{1}{n}\right)^n
  \end{equation*}
\end{frame}
\begin{frame}{Line plots}
  \begin{figure}
    \begin{tikzpicture}
      \begin{axis}[
        mlineplot,
        width=0.9\textwidth,
        height=6cm,
      ]

        \addplot {sin(deg(x))};
        \addplot+[samples=100] {sin(deg(2*x))};

      \end{axis}
    \end{tikzpicture}
  \end{figure}
\end{frame}
\begin{frame}{Bar charts}
  \begin{figure}
    \begin{tikzpicture}
      \begin{axis}[
        mbarplot,
        xlabel={Foo},
        ylabel={Bar},
        width=0.9\textwidth,
        height=6cm,
      ]

      \addplot plot coordinates {(1, 20) (2, 25) (3, 22.4) (4, 12.4)};
      \addplot plot coordinates {(1, 18) (2, 24) (3, 23.5) (4, 13.2)};
      \addplot plot coordinates {(1, 10) (2, 19) (3, 25) (4, 15.2)};

      \legend{lorem, ipsum, dolor}

      \end{axis}
    \end{tikzpicture}
  \end{figure}
\end{frame}
\begin{frame}{Quotes}
  \begin{quote}
    Veni, Vidi, Vici
  \end{quote}
\end{frame}

{%
\setbeamertemplate{frame footer}{My custom footer}
\begin{frame}[fragile]{Frame footer}
    \themename defines a custom beamer template to add a text to the footer. It can be set via
    \begin{verbatim}\setbeamertemplate{frame footer}{My custom footer}\end{verbatim}
\end{frame}
}

\begin{frame}{References}
  Some references to showcase [allowframebreaks] \cite{knuth92,ConcreteMath,Simpson,Er01,greenwade93}
\end{frame}

\section{Conclusion}

\begin{frame}{Summary}

  Get the source of this theme and the demo presentation from

  \begin{center}\url{github.com/matze/mtheme}\end{center}

  The theme \emph{itself} is licensed under a
  \href{http://creativecommons.org/licenses/by-sa/4.0/}{Creative Commons
  Attribution-ShareAlike 4.0 International License}.

  \begin{center}\ccbysa\end{center}

\end{frame}

\appendix

\begin{frame}[fragile]{Backup slides}
  Sometimes, it is useful to add slides at the end of your presentation to
  refer to during audience questions.

  The best way to do this is to include the \verb|appendixnumberbeamer|
  package in your preamble and call \verb|\appendix| before your backup slides.

  \themename will automatically turn off slide numbering and progress bars for
  slides in the appendix.
\end{frame}
\end{document}

### 2022-03-15 slides.tex
scala2022.hotcrp

@@
 
-\usetheme[progressbar=frametitle]{metropolis}
-\usepackage{appendixnumberbeamer}
+\usepackage[outputdir=latex.out,]{minted}
+\usepackage{fontspec}
 
-\usepackage{booktabs}
-\usepackage[scale=2]{ccicons}
+%% Minted setup (syntax highlighting)
+\usemintedstyle{vs}
+\setminted{fontsize=\footnotesize}
+\newminted[code]{scala}{gobble=0}
+\newenvironment{slide}[2][]
+  {\begin{frame}[fragile,environment=slide,#1]{#2}}
+  {\end{frame}}
+\newcommand\makemintedshortinline[2]{%
+  \catcode`#2=13
+  \begingroup%
+  \lccode`\~=`#2
+  \lowercase{\endgroup\protected\def~{\mintinline{#1}~}}%
+}
 
-\usepackage{pgfplots}
-\usepgfplotslibrary{dateplot}
-
-\usepackage{xspace}
-\newcommand{\themename}{\textbf{\textsc{metropolis}}\xspace}
+%% Beamer theme setup
+\usetheme[block=fill]{metropolis}
+\setbeamercolor{progress bar}{bg=black, fg=black}
+\setbeamercolor{title separator}{bg=black, fg=black}
+\setbeamercolor{progress bar in head/foot}{bg=black, fg=black}
+\setbeamercolor{progress bar in section page}{bg=black, fg=black}
+\setbeamercovered{transparent}
 
 \title{Abstractions for Type-Level Programming}
-\date{\today}
+\date{Tuesday, 22 March 2022}
 \author{Olivier Blanvillain}
-\institute{EPFL}
-% \titlegraphic{\hfill\includegraphics[height=1.5cm]{logo.pdf}}
+
+\makemintedshortinline{scala}\|
 
@@
   \setbeamertemplate{section in toc}[sections numbered]
-  \tableofcontents%[hideallsubsections]
+  \tableofcontents
 \end{frame}
 
-\section[Intro]{Introduction}
+\section[Type-Level Programming\\What? Why?]{What? Why?}
 
-\begin{frame}[fragile]{Metropolis}
+\begin{slide}{Example 1: Database queries (2016)}
+Spark API, practically untyped:
+\begin{code}
+phoneBook                                           : DataFrame
+  .innerJoin(cityMap, col("address"))               : DataFrame
+  .select(col("name"), col("neighborhood"))         : DataFrame
+  .filter(x => uniqueName(x.getAs[String]("name"))) : DataFrame
+  .groupBy(col("neighborhood")).count()             : DataFrame
+  .sortDesc(col("count"))                           : DataFrame
+  .head.getAs[String](0)                            : String
+\end{code}
+\pause
+The same API, revisited with type-level programming:
+\begin{code}
+phoneBook                          : DF[PhoneBookEntry]
+  .innerJoin(cityMap, 'address)    : DF[(Address, String, String, ... )]
+  .select('_2, '_3).as[Family]()   : DF[Family]
+  .filter(x => uniqueName(x.name)) : DF[Family]
+  .groupBy('neighborhood).count()  : DF[(String, Long)]
+  .sortDesc('_2)                   : DF[(String, Long)]
+  .head._1                         : String
+\end{code}
+\end{slide}
 
-  The \themename theme is a Beamer theme with minimal visual noise
-  inspired by the \href{https://github.com/hsrmbeamertheme/hsrmbeamertheme}{\textsc{hsrm} Beamer
-  Theme} by Benjamin Weiss.
+\begin{slide}{Example 2: Regular expressions (2022)}
+Regular expressions in Scala's standard library:
+\begin{code}
+val rational = new Regex("""(\d+)(?:\.(\d+))?""")
+"3.1415" match
+  case rational(captures: Seq[String]) =>
+    val i = captures(0)
+    val j = captures(1)
+\end{code}
+\pause
+The same API, revisited with type-level programming:
+\begin{code}
+val rational = new R("""(\d+)(?:\.(\d+))?""")
+"3.1415" match
+  case rational(captures: (String, Option[String])) =>
+    val i = captures._1
+    val j = captures._2
+\end{code}
+\end{slide}
 
-  Enable the theme by loading
+\begin{slide}{What changed between 2016 and 2022?}
+\begin{itemize}
+  \item Example 1 uses a hack, implicits:
+  \begin{itemize}
+    \item convoluted to use
+    \item slow to compile
+    \item popular nonetheless
+  \end{itemize}
+  \pause
+  \item Example 2 uses a first-class language construct, match types:
+  \begin{itemize}
+    \item easier to use
+    \item faster to compile
+    \item solid theoretical foundation
+  \end{itemize}
+\end{itemize}
+\end{slide}
 
-  \begin{verbatim}    \documentclass{beamer}
-    \usetheme{metropolis}\end{verbatim}
+\begin{slide}{Compilation times for the type-level join operation}
+\begin{center}
+\input{figures/join.gex}
+\end{center}
+\end{slide}
 
-  Note, that you have to have Mozilla's \emph{Fira Sans} font and XeTeX
-  installed to enjoy this wonderful typography.
-\end{frame}
-\begin{frame}[fragile]{Sections}
-  Sections group slides of the same topic
+\begin{slide}{Compilation times for the type-level join operation, zoomed in}
+\begin{center}
+\input{figures/join-2.gex}
+\end{center}
+\end{slide}
 
-  \begin{verbatim}    \section{Elements}\end{verbatim}
+\section{Generalizing Scala's Singleton~Types}
 
-  for which \themename provides a nice progress indicator \ldots
+\begin{slide}{Singleton types in Scala}
+Scala supports a few singleton types:
+\begin{itemize}
+  \item |x.type|, the type of the variable |x| (since forever)
+  \item |42|, the type of the integer literal |42| (since 2016)
+  \item |+|, the type of the primitive addition (since 2020)
+\end{itemize}
+\pause
+How much of the language can we lift to the type level?
+\end{slide}
 
-\end{frame}
+\begin{slide}{Generalized singleton types: a proposal}
+\end{slide}
 
-\section{Titleformats}
+\section{Type-Level Programming with Match Types}
 
-\begin{frame}{Metropolis titleformats}
-  \themename supports 4 different titleformats:
-  \begin{itemize}
-    \item Regular
-    \item \textsc{Smallcaps}
-    \item \textsc{allsmallcaps}
-    \item ALLCAPS
-  \end{itemize}
-  They can either be set at once for every title type or individually.
-\end{frame}
-
-\subsection{Tricks}
-
-{
-    \metroset{titleformat frame=smallcaps}
-\begin{frame}{Small caps}
-  This frame uses the \texttt{smallcaps} titleformat.
-
-  \begin{alertblock}{Potential Problems}
-    Be aware, that not every font supports small caps.
-If for example you typeset your presentation with pdfTeX and the Computer Modern Sans Serif font, every text in smallcaps will be typeset with the Computer Modern Serif font instead.
-  \end{alertblock}
-\end{frame}
-}
-
-{
-\metroset{titleformat frame=allsmallcaps}
-\begin{frame}{All small caps}
-  This frame uses the \texttt{allsmallcaps} titleformat.
-
-  \begin{alertblock}{Potential problems}
-    As this titleformat also uses smallcaps you face the same problems as with the \texttt{smallcaps} titleformat.
-Additionally this format can cause some other problems.
-Please refer to the documentation if you consider using it.
-
-    As a rule of thumb: Just use it for plaintext-only titles.
-  \end{alertblock}
-\end{frame}
-}
-
-{
-\metroset{titleformat frame=allcaps}
-\begin{frame}{All caps}
-  This frame uses the \texttt{allcaps} titleformat.
-
-  \begin{alertblock}{Potential Problems}
-    This titleformat is not as problematic as the \texttt{allsmallcaps} format, but basically suffers from the same deficiencies. So please have a look at the documentation if you want to use it.
-  \end{alertblock}
-\end{frame}
-}
-
-\section{Elements}
-
-\begin{frame}[fragile]{Typography}
-      \begin{verbatim}The theme provides sensible defaults to
-\emph{emphasize} text, \alert{accent} parts
-or show \textbf{bold} results.\end{verbatim}
-
-  \begin{center}becomes\end{center}
-
-  The theme provides sensible defaults to \emph{emphasize} text,
-  \alert{accent} parts or show \textbf{bold} results.
-\end{frame}
-
-\begin{frame}{Font feature test}
-  \begin{itemize}
-    \item Regular
-    \item \textit{Italic}
-    \item \textsc{SmallCaps}
-    \item \textbf{Bold}
-    \item \textbf{\textit{Bold Italic}}
-    \item \textbf{\textsc{Bold SmallCaps}}
-    \item \texttt{Monospace}
-    \item \texttt{\textit{Monospace Italic}}
-    \item \texttt{\textbf{Monospace Bold}}
-    \item \texttt{\textbf{\textit{Monospace Bold Italic}}}
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Lists}
-  \begin{columns}[T,onlytextwidth]
-    \column{0.33\textwidth}
-      Items
-      \begin{itemize}
-        \item Milk \item Eggs \item Potatos
-      \end{itemize}
-
-    \column{0.33\textwidth}
-      Enumerations
-      \begin{enumerate}
-        \item First, \item Second and \item Last.
-      \end{enumerate}
-
-    \column{0.33\textwidth}
-      Descriptions
-      \begin{description}
-        \item[PowerPoint] Meeh. \item[Beamer] Yeeeha.
-      \end{description}
-  \end{columns}
-\end{frame}
-\begin{frame}{Animation}
-  \begin{itemize}[<+- | alert@+>]
-    \item \alert<4>{This is\only<4>{ really} important}
-    \item Now this
-    \item And now this
-  \end{itemize}
-\end{frame}
-\begin{frame}{Figures}
-  \begin{figure}
-    \newcounter{density}
-    \setcounter{density}{20}
-    \begin{tikzpicture}
-      \def\couleur{alerted text.fg}
-      \path[coordinate] (0,0)  coordinate(A)
-                  ++( 90:5cm) coordinate(B)
-                  ++(0:5cm) coordinate(C)
-                  ++(-90:5cm) coordinate(D);
-      \draw[fill=\couleur!\thedensity] (A) -- (B) -- (C) --(D) -- cycle;
-      \foreach \x in {1,...,40}{%
-          \pgfmathsetcounter{density}{\thedensity+20}
-          \setcounter{density}{\thedensity}
-          \path[coordinate] coordinate(X) at (A){};
-          \path[coordinate] (A) -- (B) coordinate[pos=.10](A)
-                              -- (C) coordinate[pos=.10](B)
-                              -- (D) coordinate[pos=.10](C)
-                              -- (X) coordinate[pos=.10](D);
-          \draw[fill=\couleur!\thedensity] (A)--(B)--(C)-- (D) -- cycle;
-      }
-    \end{tikzpicture}
-    \caption{Rotated square from
-    \href{http://www.texample.net/tikz/examples/rotated-polygons/}{texample.net}.}
-  \end{figure}
-\end{frame}
-\begin{frame}{Tables}
-  \begin{table}
-    \caption{Largest cities in the world (source: Wikipedia)}
-    \begin{tabular}{lr}
-      \toprule
-      City & Population\\
-      \midrule
-      Mexico City & 20,116,842\\
-      Shanghai & 19,210,000\\
-      Peking & 15,796,450\\
-      Istanbul & 14,160,467\\
-      \bottomrule
-    \end{tabular}
-  \end{table}
-\end{frame}
-\begin{frame}{Blocks}
-  Three different block environments are pre-defined and may be styled with an
-  optional background color.
-
-  \begin{columns}[T,onlytextwidth]
-    \column{0.5\textwidth}
-      \begin{block}{Default}
-        Block content.
-      \end{block}
-
-      \begin{alertblock}{Alert}
-        Block content.
-      \end{alertblock}
-
-      \begin{exampleblock}{Example}
-        Block content.
-      \end{exampleblock}
-
-    \column{0.5\textwidth}
-
-      \metroset{block=fill}
-
-      \begin{block}{Default}
-        Block content.
-      \end{block}
-
-      \begin{alertblock}{Alert}
-        Block content.
-      \end{alertblock}
-
-      \begin{exampleblock}{Example}
-        Block content.
-      \end{exampleblock}
-
-  \end{columns}
-\end{frame}
-\begin{frame}{Math}
-  \begin{equation*}
-    e = \lim_{n\to \infty} \left(1 + \frac{1}{n}\right)^n
-  \end{equation*}
-\end{frame}
-\begin{frame}{Line plots}
-  \begin{figure}
-    \begin{tikzpicture}
-      \begin{axis}[
-        mlineplot,
-        width=0.9\textwidth,
-        height=6cm,
-      ]
-
-        \addplot {sin(deg(x))};
-        \addplot+[samples=100] {sin(deg(2*x))};
-
-      \end{axis}
-    \end{tikzpicture}
-  \end{figure}
-\end{frame}
-\begin{frame}{Bar charts}
-  \begin{figure}
-    \begin{tikzpicture}
-      \begin{axis}[
-        mbarplot,
-        xlabel={Foo},
-        ylabel={Bar},
-        width=0.9\textwidth,
-        height=6cm,
-      ]
-
-      \addplot plot coordinates {(1, 20) (2, 25) (3, 22.4) (4, 12.4)};
-      \addplot plot coordinates {(1, 18) (2, 24) (3, 23.5) (4, 13.2)};
-      \addplot plot coordinates {(1, 10) (2, 19) (3, 25) (4, 15.2)};
-
-      \legend{lorem, ipsum, dolor}
-
-      \end{axis}
-    \end{tikzpicture}
-  \end{figure}
-\end{frame}
-\begin{frame}{Quotes}
-  \begin{quote}
-    Veni, Vidi, Vici
-  \end{quote}
-\end{frame}
-
-{%
-\setbeamertemplate{frame footer}{My custom footer}
-\begin{frame}[fragile]{Frame footer}
-    \themename defines a custom beamer template to add a text to the footer. It can be set via
-    \begin{verbatim}\setbeamertemplate{frame footer}{My custom footer}\end{verbatim}
-\end{frame}
-}
-
-\begin{frame}{References}
-  Some references to showcase [allowframebreaks] \cite{knuth92,ConcreteMath,Simpson,Er01,greenwade93}
-\end{frame}
-
-\section{Conclusion}
-
-\begin{frame}{Summary}
-
-  Get the source of this theme and the demo presentation from
-
-  \begin{center}\url{github.com/matze/mtheme}\end{center}
-
-  The theme \emph{itself} is licensed under a
-  \href{http://creativecommons.org/licenses/by-sa/4.0/}{Creative Commons
-  Attribution-ShareAlike 4.0 International License}.
-
-  \begin{center}\ccbysa\end{center}
-
-\end{frame}
-
-\appendix
-
-\begin{frame}[fragile]{Backup slides}
-  Sometimes, it is useful to add slides at the end of your presentation to
-  refer to during audience questions.
-
-  The best way to do this is to include the \verb|appendixnumberbeamer|
-  package in your preamble and call \verb|\appendix| before your backup slides.
-
-  \themename will automatically turn off slide numbering and progress bars for
-  slides in the appendix.
-\end{frame}
 \end{document}

### 2022-03-15 symposium.tex
scala2022.hotcrp

@@
+\def\paper{\thesisOnly{chapter}\paperOnly{paper}}
 \section{Introduction}
@@
 
-In this \thesisOnly{chapter}\paperOnly{paper}, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
+In this \paper, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
 Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
@@
 
-This paper is structured as follows.
+This \paper is structured as follows.
 In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
@@
 In \Cref{sec:related}, we discuss related work.
-We conclude the paper in \Cref{sec:conclusion}.
+We conclude the \paper in \Cref{sec:conclusion}.
 
 \thesisOnly{
-The source code of our implementation is available online\footnote{\texttt{https://gist.github.com/OlivierBlanvillain/418dcb74cbd2e12eb4a90b7336f40b17}}.
+The source code of our implementation is available online\footnote{\texttt{https://github.com/OlivierBlanvillain/regsafe}} under the MIT licence.
 }
 \paperOnly{
-The source code of our implementation is available in the supplementary material of this submission.
+The source code of our implementation is available online\footnote{This URL is hidden for the purpose of the double-blind review process.} under the MIT licence.
 }
@@
 
-In this \thesisOnly{chapter}\paperOnly{paper}, we introduce a new design for type-safe regular expressions in Scala.
+In this \paper, we introduce a new design for type-safe regular expressions in Scala.
 We presented our type-level analysis, which identifies capturing groups and computes their nullability.

### 2022-03-15 slides.scala

@@ -1,0 +1,54 @@
+def bestNeighborhood(phoneBook: TypedDataFrame[PhoneBookEntry], cityMap: TypedDataFrame[CityMapEntry]) =
+  phoneBook
+    .innerJoin(cityMap).using('address)     :TypedDataFrame[(Address, String, Double, String)]
+    .select('_2, '_4)                       :TypedDataFrame[(String, String)]
+    .as[Family]()                           :TypedDataFrame[Family]
+    .flatMap(f => f.residents.split(' ').map(r => Person(r, f.neighborhood)))
+                                            :TypedDataFrame[Person]
+    .filter(p => NLPLib.uniqueName(p.name)) :TypedDataFrame[Person]
+    .groupBy('neighborhood).count()         :TypedDataFrame[(String, Long)]
+    .as[NeighborhoodCount]()                :TypedDataFrame[NeighborhoodCount]
+    .sortDesc('count)                       :TypedDataFrame[NeighborhoodCount]
+    .select('neighborhood)                  :TypedDataFrame[Tuple1[String]]
+    .head._1
+
+
+
+phoneBook                             : DF[PhoneBookEntry]
+  .innerJoin(cityMap).using('address) : DF[(Address, String, String, ...)]
+  .select('_2, '_3).as[Family]()      : DF[Family]
+  .filter(x => uniqueName(x.name))    : DF[Family]
+  .groupBy('neighborhood).count()     : DF[(String, Long)]
+  .sortDesc('_2)                      : DF[(String, Long)]
+  .head._1                            : String
+
+
+phoneBook                                           : DataFrame
+  .innerJoin(cityMap, col("address"))               : DataFrame
+  .select(col("name"), col("neighborhood"))         : DataFrame
+  .filter(x => uniqueName(x.getAs[String]("name"))) : DataFrame
+  .groupBy(col("neighborhood")).count()             : DataFrame
+  .sortDesc(col("count"))                           : DataFrame
+  .head.getAs[String](0)                            : String
+
+
+
+val captures: Option[Seq[String]] = rational.unapply("3.1415")
+captures match
+  case Some(seq) if seq(1) != null =>
+    var n = seq(0).length + seq(1).length
+    println(s"This number is $n digits long")
+  case _ =>
+    println("Not a number")
+
+val rational = Regex("""(\d+)(?:\.(\d+))?""")
+"3.1415" match
+  case rational(captures: Seq[String]) =>
[…]

### 2022-03-18 article.tex
Slides

@@
 \begin{lstlisting}
-type Elem[X] = X match {
+type Elem[X] = X match
   case String => Char
@@
   case Any => X
-}
 \end{lstlisting}
@@
 \begin{lstlisting}
-def elem[X](x: X): Elem[X] = x match {
+def elem[X <: Singleton](x: X): Elem[X] = x match
   case x: String => x.charAt(0)
@@
   case x: Any => x
-}
 \end{lstlisting}
@@
 \begin{lstlisting}
-type M[X] = X match {
+type M[X] = X match
   case Int => String
   case String => Int
-}
-class C {
+class C:
   type X
   def f(bad: M[X & String]): Int = bad
-}
-class D extends C {
+class D extends C:
   type X = Int
-}
 \end{lstlisting}
@@
 \begin{lstlisting}
-trait AgeClass {
+trait AgeClass:
   type Age
   def addAge(a: Age, i: Int): Int
-}
-object AgeObject extends AgeClass {
+object AgeObject extends AgeClass:
   type Age = Int
   def addAge(a: Age, i: Int): Int = a + i
-}
 \end{lstlisting}
@@
 \begin{lstlisting}
-type M[X] = X match {
+type M[X] = X match
   case Age => Char
   case Int => Bool
-}
 \end{lstlisting}

### 2022-03-18 preprint.tex
Add footnote and QE and character class
Draft submission
Slides

@@
 Instead, the result type of a method call |t.m(|\ldots{}|)| is always the result type of |m| in |t|'s static type.
-Consequently, |dependent| methods effectively become |final|, given that only a provably-equivalent implementation could be used to override it.
+Consequently, |dependent| methods effectively become |final|, given that only a provably-equivalent implementation could be used to override it, modulo side effects.
 
@@
 
-\section{Use Case}
+\section{Case Study: A Type-Safe Database Interface}
 \label{sec:use-case}
@@
 
-\subsection{A Type-Safe Database Interface}
+\subsection{Type-Safe Datasets}
 
@@
 dependent case class Column[T](k: String):
-  def ===(that: Column[T]): Column[Boolean] = Column(s"(${this.k} === ${that.k})")
+  def ===(that: Column[T]): Column[Boolean] =
+    Column(s"(${this.k} === ${that.k})")
 \end{lstlisting}
@@
 Safety is obtained by inserting dynamic checks, similarly to gradual typing.
+
+% TODO: Compare to pure subtype systems

### 2022-03-18 slides.tex
Slides

@@
 
-\usepackage[outputdir=latex.out,]{minted}
 \usepackage{fontspec}
@@
 %% Minted setup (syntax highlighting)
+\usepackage[outputdir=latex.out]{minted}
 \usemintedstyle{vs}
 \setminted{fontsize=\footnotesize}
-\newminted[code]{scala}{gobble=0}
+\newminted[code]{scala}{gobble=0, escapeinside=@@}
 \newenvironment{slide}[2][]
@@
 }
+\renewcommand{\fcolorbox}[4][]{#4}
 
@@
 
+\usepackage{xspace}
+\usepackage{mathtools}
+\def\C{\normalfont\text{C}}
+\def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
+\def\Fsub{\normalfont{F}\ensuremath{_{\<:}}\xspace}
+
+\usepackage[autostyle=false, style=english]{csquotes}
+\MakeOuterQuote{"}
+
+\usepackage{xcolor}
+\definecolor{typeColor}{HTML}{2B91B0}
+\definecolor{implColor}{HTML}{57B5E8}
+\definecolor{singColor}{HTML}{009E73}
+\definecolor{mtpeColor}{HTML}{9400D4}
+\newcommand{\tp}{\color{typeColor}}
+
+\usepackage{tikz}
+\usetikzlibrary{decorations.pathreplacing,positioning, arrows.meta}
+
 \title{Abstractions for Type-Level Programming}
@@
 
+\newcommand{\timeline}[5] {
+  \begin{tikzpicture}
+    \draw[lightgray!#5!black, thick, -Triangle] (0,0) -- (10cm,0);
+    \foreach \x in {0,1,...,6}
+    \draw[lightgray!#5!black] (\x*1.4 cm,3pt) -- (\x*1.4 cm,-3pt);
+
+    \foreach \x/\descr in {0/2016,1/2017,2/2018,3/2019,4/2020,5/2021,6/2022}
+    \node[lightgray!#5!black, font=\scriptsize, text height=1.75ex,
+    text depth=.5ex] at (\x*1.4,-.3) {$\descr$};
+
+    \fill[lightgray!#1!implColor] (0.25, 0.5) circle (4pt) node[lightgray!#1!black,align=center,above=7pt,font=\small]{DataFrame\\Library};
+
+    \draw[lightgray!#2!singColor, line width=4pt]
+    (2.25*1.4,.5) -- (4.417*1.4,.5) node[lightgray!#2!black,midway,above=5pt,font=\small]{Generalized Singletons};
+
+    \draw[lightgray!#3!mtpeColor, line width=4pt]
+    (2.5*1.4,-.7) -- (6.15*1.4,-.7) node[lightgray!#3!black,midway,below=5pt,font=\small]{Match Types};
+    \draw[thick,lightgray!#3!mtpeColor,-Triangle,line width=1.2pt] (6.15*1.4,-.7) -- (6.25*1.4,-.7);
+
+    \fill[lightgray!#4!mtpeColor] (6.1*1.4, 0.5) circle (4pt) node[lightgray!#4!black,align=center,above=7pt,font=\small]{Regex\\Library};
+  \end{tikzpicture}
+}
+% Jun 12 2018  Add TypeOf  #4671
+% Sep 20 2018  Add match type #4964
+% ECOOP  Jan 12 → Apr 1  2019, rejected
+% OOPSLA Mai 15 → Aug 1  2020, rejected
+% PLDI   Nov 20 → Feb 25 2021, rejected
+% OOPSLA Apr 16 → Jul 2  2021, rejected
+% POPL   Jul 7  → Sep 28 2021, accepted
+
 \begin{document}
@@
 
-\begin{frame}{Table of contents}
-  \setbeamertemplate{section in toc}[sections numbered]
-  \tableofcontents
-\end{frame}
-
-\section[Type-Level Programming\\What? Why?]{What? Why?}
+\begin{slide}{Example 1: Database queries (2016)}
+\begin{center}
+\includegraphics[width=0.3\textwidth]{figures/spark.pdf}
+\end{center}
+\end{slide}
 
 \begin{slide}{Example 1: Database queries (2016)}
-Spark API, practically untyped:
+Spark APIs are practically untyped:
 \begin{code}
-phoneBook                                           : DataFrame
-  .innerJoin(cityMap, col("address"))               : DataFrame
-  .select(col("name"), col("neighborhood"))         : DataFrame
-  .filter(x => uniqueName(x.getAs[String]("name"))) : DataFrame
-  .groupBy(col("neighborhood")).count()             : DataFrame
-  .sortDesc(col("count"))                           : DataFrame
-  .head.getAs[String](0)                            : String
+class DataFrame {
+  /** Inner equi-join with another DataFrame on the given column.
+   *  The join column will only appear once in the output. */
+  def join(right: DataFrame, column: String): DataFrame
+}
 \end{code}
 \pause
-The same API, revisited with type-level programming:
+Revisited with type-level programming:
 \begin{code}
-phoneBook                          : DF[PhoneBookEntry]
-  .innerJoin(cityMap, 'address)    : DF[(Address, String, String, ... )]
-  .select('_2, '_3).as[Family]()   : DF[Family]
-  .filter(x => uniqueName(x.name)) : DF[Family]
-  .groupBy('neighborhood).count()  : DF[(String, Long)]
-  .sortDesc('_2)                   : DF[(String, Long)]
-  .head._1                         : String
+class DF[X] {
+  def join[Y](df: DF[Y], col: String): DF[col+(X-col)++(Y-col)]
+}
 \end{code}
@@
 \begin{code}
-val rational = new Regex("""(\d+)(?:\.(\d+))?""")
-"3.1415" match
-  case rational(captures: Seq[String]) =>
-    val i = captures(0)
-    val j = captures(1)
+val rational = new Regex("(\\d+)\\.?(\\d+)?")
+rational.unapply("3.1415"): Option[Seq[String]]
 \end{code}
 \pause
-The same API, revisited with type-level programming:
+Revisited with type-level programming:
 \begin{code}
-val rational = new R("""(\d+)(?:\.(\d+))?""")
-"3.1415" match
-  case rational(captures: (String, Option[String])) =>
-    val i = captures._1
-    val j = captures._2
+rational.unapply("3.1415"): Option[(String, Option[String])]
+\end{code}
+\pause
+\begin{code}
+class Regex(pattern: String) {
+  def unapply(s: String): Option[GroupsOf[@pattern@]]
+}
 \end{code}
@@
 \begin{slide}{What changed between 2016 and 2022?}
-\begin{itemize}
-  \item Example 1 uses a hack, implicits:
+\begin{tikzpicture}
+  \draw[black, thick, -Triangle] (0,0) -- (10cm,0);
+  \foreach \x in {0,1,...,6}
+  \draw[black] (\x*1.4 cm,3pt) -- (\x*1.4 cm,-3pt);
+
+  \foreach \x/\descr in {0/2016,1/2017,2/2018,3/2019,4/2020,5/2021,6/2022}
+  \node[black, font=\scriptsize, text height=1.75ex,
+  text depth=.5ex] at (\x*1.4,-.3) {$\descr$};
+
+  \fill[implColor] (0.25, 0.5) circle (4pt) node[black,align=center,above=7pt,font=\small]{DataFrame\\Library};
+
+  \fill[mtpeColor] (6.1*1.4, 0.5) circle (4pt) node[black,align=center,above=7pt,font=\small]{Regex\\Library};
+\end{tikzpicture}
+\end{slide}
+
+\begin{slide}{What changed between 2016 and 2022?}
+Example 1 uses a hack, implicits:
 \begin{itemize}
@@
   \item slow to compile
-    \item popular nonetheless
 \end{itemize}
-  \pause
-  \item Example 2 uses a first-class language construct, match types:
-  \begin{itemize}
-    \item easier to use
-    \item faster to compile
-    \item solid theoretical foundation
-  \end{itemize}
-\end{itemize}
+
+Example 2 uses a new language construct, match types, one of the main contributions of this thesis.
 \end{slide}
@@
 
-\begin{slide}{Compilation times for the type-level join operation, zoomed in}
-\begin{center}
-\input{figures/join-2.gex}
-\end{center}
+\begin{slide}{Timeline of my PhD}
+\timeline{0}{0}{0}{0}{0}
 \end{slide}
 
-\section{Generalizing Scala's Singleton~Types}
+\begin{slide}{Part I: Generalized Singleton Types}
+\timeline{80}{0}{80}{80}{80}
+\end{slide}
 
-\begin{slide}{Singleton types in Scala}
+\begin{slide}{Singleton types}
 Scala supports a few singleton types:
@@
   \item |42|, the type of the integer literal |42| (since 2016)
-  \item |+|, the type of the primitive addition (since 2020)
+  \item |+|, the type of integer addition (since 2020)
 \end{itemize}
 \pause
-How much of the language can we lift to the type level?
+Question: How much of the language can we represent in types?
 \end{slide}
 
-\begin{slide}{Generalized singleton types: a proposal}
+\begin{slide}{Generalized singleton types: A proposal}
+Our proposal consists of 3 changes:
+\begin{enumerate}
+  \item new types for if-then-else, pattern matching, constructors, functions calls
+  \item "precise mode" of type inference
+  \item type evaluation, during subtyping
+\end{enumerate}
 \end{slide}
 
-\section{Type-Level Programming with Match Types}
+\begin{slide}{New types}
+What's the type of | if (x == 0) "zero" else "one"|?
+\begin{itemize}
+  \pause\item |String| ~(Scala 2)
+  \pause\item |"zero"|~{\footnotesize\texttt\textbar}~|"one"| ~(Scala 3)
+  \pause\item |If[x.type == 0,|| "zero", "one"]| ~(proposed)
+  \pause\item |{ if (x == 0) "zero" else "one" }| ~(proposed, syntactic sugar)
+\end{itemize}
+\pause
+Similarly, we add new types for other constructs: |Match[]|, |New[]| and |Call[]|, and the corresponding syntactic sugar.
+\end{slide}
 
+\begin{slide}{"Precise mode" of type inference}
+We need two modes of type inference, for backwards compatibility.
+
+\vspace{5pt}
+\begin{overprint}
+\onslide<1>
+\begin{code}
+@\phantom{dependent}@
+def int2str(x: Int) =
+  if (x == 0) "zero" else "one"
+\end{code}
+
+\onslide<2>
+\begin{code}
+@\phantom{dependent}@
+def int2str(x: Int): String =
+  if (x == 0) "zero" else "one"
+\end{code}
+
+\onslide<3>
+\begin{code}
+dependent
+def int2str(x: Int)
+  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ "one" }
+  =   if (x == 0) "zero" else "one"
+\end{code}
+
+\onslide<4-5>
+\begin{code}
+dependent
+def int2str(x: Int)
+  : If[x.type @\tp{}== 0@, "zero", "one"]
+  =   if (x == 0) "zero" else "one"
+\end{code}
+
+\onslide<6>
+\begin{code}
+dependent
+def int2str(x: Int)
+  : If[x.type @\tp{}== 0@, "zero", String]
+  =   if (x == 0) "zero" else readString()
+\end{code}
+
+\onslide<7>
+\begin{code}
+dependent
+def int2str(x: Int)
+  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ (_: String) }
+  =   if (x == 0) "zero" else readString()
+\end{code}
+\end{overprint}
+\medskip
+\pause\pause\pause\pause
+
+The "precise mode" is best effort: it lifts whatever possible to the type level and approximates the rest.
+\end{slide}
+
+\begin{slide}{List concatenation with generalized singleton types}
+\begin{code}
+dependent def concat(xs: List, ys: List) <: List =
+  xs match
+    case x :: xs => x :: concat(xs, ys)
+    case Nil => ys
+\end{code}
+\pause
+\begin{overprint}
+\onslide<1-2>
+\begin{code}
+dependent val l1 = "A" :: Nil
+dependent val l2 = "B" :: Nil
+dependent val l3 = concat(l1, l2)
+l3: { "A" :: "B" :: Nil }
+\end{code}
+
+\onslide<3>
+\begin{code}
+val l1: List     = "A" :: Nil
+dependent val l2 = "B" :: Nil
+dependent val l3 = concat(l1, l2)
+l3: { concat(l1, l2) }
+\end{code}
+
+\onslide<4>
+\begin{code}
+dependent val l1 = "A" :: Nil
+val l2: List     = "B" :: Nil
+dependent val l3 = concat(l1, l2)
+l3: { "A" :: l2 }
+\end{code}
+
+\end{overprint}
+\end{slide}
+
+\begin{slide}{Type evaluation}
+During subtyping, we evaluate both sides:
+\begin{code}
+A <: B  iff  eval(A) <: eval(B)
+\end{code}
+\pause
+
+Straightforward for if-then-else:
+\begin{code}
+eval(If[true, A, B]) = A
+eval(If[false, A, B]) = B
+\end{code}
+\pause
+
+More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
+\pause
+\smallskip
+
+We evaluate type tests using subtyping and type disjointness:
+\begin{code}
+eval(@\tp{}x.asInstanceOf@[Foo]) = true   iff  x.type <: Foo
+eval(@\tp{}x.asInstanceOf@[Foo]) = false  iff  disj(x.type, Foo)
+\end{code}
+\end{slide}
+
+\begin{slide}{Type evaluation: pattern matching example}
+\begin{overprint}
+\onslide<1>
+\begin{code}
+dependent val foo(x: Any) =
+  x match
+    case s: String => s
+    case i: Int => i+1
+\end{code}
+\onslide<2>
+\begin{code}
+dependent val foo(x: Any): {
+  x match
+    case s: String => @\tp{}s@
+    @\tp{}case i: Int@ => @\tp{}i+1@
+} =
+  x match
+    case s: String => s
+    case i: Int => i+1
+\end{code}
+\onslide<3->
+\begin{code}
+dependent val foo(x: Any): {
+  if (x.isInstanceOf[String]) x.asInstanceOf[String]
+  @\tp{}else if @(@\tp{}x.isInstanceOf@[@\tp{}Int@]) @\tp{}x.asInstanceOf@[Int] @\tp{}+ 1@
+  @\tp{}else throw new@ MatchError()
+} =
+  x match
+    case s: String => s
+    case i: Int => i+1
+\end{code}
+\end{overprint}
+\bigskip
+\begin{overprint}
+\onslide<4>
+|foo(readString())|
+\onslide<5>
+|foo(readString()): { foo(_: String) }|
+\onslide<6>
+|foo(readString()): { (_: String).asInstanceOf[String] }|
+\onslide<7>
+|foo(readString()): { (_: String) }|
+\onslide<8>
+|foo(readString()): String|
+\end{overprint}
+\end{slide}
+
+\begin{slide}{Generalized Singletons: Conclusion}
+TODO
+\end{slide}
+
+% Termination
+% Virtual dispatch
+
+\begin{slide}{Part II: Match Types}
+\timeline{80}{80}{0}{80}{80}
+\end{slide}
+
+\begin{slide}{Overview}
+\begin{code}
+type Elem[X] = X match
+  case String => Char
+  case List[t] => Elem[t]
+  case Any => X
+\end{code}
+\end{slide}
+
+\begin{slide}{Disjointness}
+|Elem[Seq[Int]]| is stuck
+\end{slide}
+
+\begin{slide}{Relating match terms and types: a defining property}
+|Elem[Seq[Int]]| is stuck
+\end{slide}
+
+\begin{slide}{Formalizing Match Types}
+System FM is an extension of System \Fsub with
+\begin{itemize}
+  \item pattern matching
+  \item opaque classes
+  \item match types
+\end{itemize}
+Design goals:
+\begin{itemize}
+  \item explain the essence of match types
+  \item give us confidence in our design
+  \item be as simple as possible
+\end{itemize}
+\end{slide}
+
+\begin{slide}{System FM is parametric}
+System FM is parametrized by a set of externally defined classes:
+
+FM(C, $\Psi$, $\Xi$)
+\begin{itemize}
+  \item $\C \hspace{2.3pt} \coloneqq$ set of classes
+  \item $\Psi \coloneqq$ class inheritance relation
+  \item $\Xi \hspace{1.1pt} \coloneqq$ class disjointness relation
+\end{itemize}
+
+For example:
+
+|class A|                    -> C={A}, $\Psi$={}, $\Xi$={}
+|class A; class B extends A| -> C={A, B}, $\Psi$={(A, B)}, $\Xi$={}
+|class A; class D|           -> C={A, D}, $\Psi$={}, $\Xi$={(A, D)}
+|class A; trait T|           -> C={A, T}, $\Psi$={}, $\Xi$={}
+\end{slide}
+
+\begin{slide}{System FM's syntax}
+\includegraphics[width=\textwidth]{figures/FMSyntax.pdf}
+\end{slide}
+
+\begin{slide}{System FM's disjointness relation}
+\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
+\end{slide}
+
+\begin{slide}{System FM's new subtyping rules}
+\includegraphics[width=\textwidth]{figures/FMSubtyping.pdf}
+\end{slide}
+
+\begin{slide}{System FM's type safety}
+theorems
+
+two versions of the proof, pen and paper and Coq
+\end{slide}
+
+\begin{slide}{Structure of the type safety proof}
+\includegraphics[width=\textwidth]{figures/FMProofStructure.pdf}
+\end{slide}
+
+\begin{slide}{Limitations}
+- type binders
+- higher-kinded types
+- recursion
+\end{slide}
+
+\begin{slide}{Empty Types}
+\begin{code}
+type M[X] = X match
+  case Int => String
+  case String => Int
+class C:
+  type X
+  def f(bad: M[X & String]): Int = bad
+class D extends C:
+  type X = Int
+\end{code}
+\end{slide}
+
+\begin{slide}{Variance}
+\begin{code}
+F[+T]: T1 <: T2 => F[T1] <: F[T2]
+G[-T]: T1 <: T2 => G[T1] >: G[T2]
+
+case class Tuple2[+T1, +T2](_1: T1, _2: T2)
+\end{code}
+\end{slide}
+
+\begin{slide}{Match Types, Conclusion}
+TODO
+\end{slide}
+
+\begin{slide}{Type-Safe Regular Expressions}
+\timeline{80}{80}{80}{0}{80}
+\end{slide}
+
+\begin{slide}{Type-Safe Regular Expressions (2022)}
+\begin{code}
+val rational = new Regex("(\\d+)\\.?(\\d+)?")
+rational.unapply("3.1415"): Option[(String, Option[String])]
+\end{code}
+\pause
+\begin{code}
+type GroupsOf[P <: String] = ...
+
+\end{code}
+\end{slide}
+
+% \appendix
 \end{document}

### 2022-03-18 symposium.tex
Slides

@@
 
+% TODO: consider inline functions as a 3rd alternative
+
 \begin{enumerate}
@@
 
-\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline (an average of +0.3~ms per regex).
+\paragraph{Compilation Times} Our library in its code-duplicated variant adds a total of 53~ms over the baseline's total compilation time (an average of +0.3~ms per regex).
 This corresponds to the cost of match type reduction.

### 2022-03-18 thesis.tex
Draft submission
Update symposium.tex
Slides

@@
 We develop the example presented in \Cref{chap:introduction} by looking into the remove operation on |HList|s.
-Remove operation takes as argument an element and a list, and returns that list with the first occurrence of the element removed.
+The remove operation takes as argument an element and a list, and returns that list with the first occurrence of the element removed.
 This operation should yield an error if the element is not part of the list.
@@
 
+% TODO: Add a paragraph about the evolution of the "rule for picking a most specific alternative" during implicit resolution, see https://dotty.epfl.ch/docs/reference/changed-features/implicit-resolution.html
+
 \section{Conclusion}
@@
 The present chapter is a reboot of the match-type side of that project.
+We plan to submit this work to the upcoming Scala Symposium.
 
+\newcommand\thesisOnly[1]{#1}
+\newcommand\paperOnly[1]{}
 \input{symposium.tex}

### 2022-03-18 scala/src/main/scala/reduce-match-types.scala

@@ -7,10 +7,9 @@ case object Ø extends Shape
 // cheating a little bit here, but there's a footnote to explain it!
 object SimplifiedShape {
 // start section shapeEnum
-enum Shape {
+enum Shape:
   case #:[H <: Int, T <: Shape](head: H, tail: T)
   case Ø
-}
 // end section shapeEnum
 }
 
@@ -36,10 +35,9 @@ def multiply[T, S <: Shape](x: NDArray[T, S], y: NDArray[T, S]): NDArray[T, S] =
 import scala.compiletime.ops.int.*
 // start section numelementsType
 type NumElements[X <: Shape] <: Int =
-  X match {
+  X match
     case Ø => 1
     case head #: tail => head * NumElements[tail]
-  }
 // end section numelementsType
 
 // start section reshapeDef
@@ -50,24 +48,20 @@ def reshape[T, From <: Shape, To <: Shape](arr: NDArray[T, From], newshape: To)
 import scala.compiletime.ops.int.+
 // start section reduceType
 type ReduceAxes[S <: Shape, Axes <: None | Shape] <: Shape =
-  Axes match {
+  Axes match
     case None => Ø
     case Shape => Loop[S, Axes, 0]
-  }
 // end section reduceType
 
 // start section reduceLoop
 type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
-  S match {
-    case head #: tail => Contains[Axes, I] match {
+  S match
+    case head #: tail => Contains[Axes, I] match
       case true => Loop[tail, Remove[Axes, I], I + 1]
       case false => head #: Loop[tail, Axes, I + 1]
-    }
-    case Ø => Axes match {
+    case Ø => Axes match
       case Ø => Ø
       // otherwise, do not reduce further
-    }
-  }
 // end section reduceLoop
 
 type Contains[Haystack <: Shape, Needle <: Int] <: Boolean = Haystack match {

### 2022-03-19 slides.tex
Slides

@@
   \lccode`\~=`#2
-  \lowercase{\endgroup\protected\def~{\mintinline{#1}~}}%
+  \lowercase{\endgroup\protected\def~{\mintinline[escapeinside=@@]{#1}~}}%
 }
@@
 \usepackage{mathtools}
+\def\A{\normalfont\text{A}}
+\def\B{\normalfont\text{B}}
 \def\C{\normalfont\text{C}}
+\def\D{\normalfont\text{D}}
+\def\T{\normalfont\text{T}}
 \def\<:{\allowbreak\operatorname{\normalfont\text{<}:}\allowbreak}
@@
 \definecolor{mtpeColor}{HTML}{9400D4}
-\newcommand{\tp}{\color{typeColor}}
+\newcommand{\tp}[1]{\color{typeColor}#1}
 
@@
 def int2str(x: Int)
-  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ "one" }
+  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ "one" }
   =   if (x == 0) "zero" else "one"
@@
 def int2str(x: Int)
-  : If[x.type @\tp{}== 0@, "zero", "one"]
+  : If[x.type @\tp{== 0}@, "zero", "one"]
   =   if (x == 0) "zero" else "one"
@@
 def int2str(x: Int)
-  : If[x.type @\tp{}== 0@, "zero", String]
+  : If[x.type @\tp{== 0}@, "zero", String]
   =   if (x == 0) "zero" else readString()
@@
 def int2str(x: Int)
-  : { if (x @\tp{}== 0@) "zero" @\tp{}else@ (_: String) }
+  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ (_: String) }
   =   if (x == 0) "zero" else readString()
@@
 
-\begin{slide}{List concatenation with generalized singleton types}
+\begin{slide}{"Precise mode" of type inference: list concatenation example}
 \begin{code}
@@
 \begin{code}
-eval(@\tp{}x.asInstanceOf@[Foo]) = true   iff  x.type <: Foo
-eval(@\tp{}x.asInstanceOf@[Foo]) = false  iff  disj(x.type, Foo)
+eval(@\tp{x.asInstanceOf}@[Foo]) = true   iff  x.type <: Foo
+eval(@\tp{x.asInstanceOf}@[Foo]) = false  iff  disj(x.type, Foo)
 \end{code}
@@
   x match
-    case s: String => @\tp{}s@
-    @\tp{}case i: Int@ => @\tp{}i+1@
+    case s: String => @\tp{s}@
+    @\tp{case i: Int}@ => @\tp{i+1}@
 } =
@@
   if (x.isInstanceOf[String]) x.asInstanceOf[String]
-  @\tp{}else if @(@\tp{}x.isInstanceOf@[@\tp{}Int@]) @\tp{}x.asInstanceOf@[Int] @\tp{}+ 1@
-  @\tp{}else throw new@ MatchError()
+  @\tp{else if }@(@\tp{x.isInstanceOf}@[@\tp{Int}@]) @\tp{x.asInstanceOf}@[Int] @\tp{+ 1}@
+  @\tp{else throw new}@ MatchError()
 } =
@@
 \bigskip
+
 \begin{overprint}
 \onslide<4>
+|foo(42): { foo(42) }|
+\onslide<5>
+|foo(42): { 42.asInstanceOf[Int] + 1 }|
+\onslide<6>
+|foo(42): { 42 + 1 }|
+\onslide<7>
+|foo(42): { 43 }|
+\onslide<8->
+|foo(42): 43|
+\end{overprint}
+
+\begin{overprint}
+\onslide<9>
 |foo(readString())|
-\onslide<5>
+\onslide<10>
 |foo(readString()): { foo(_: String) }|
-\onslide<6>
+\onslide<11>
 |foo(readString()): { (_: String).asInstanceOf[String] }|
-\onslide<7>
+\onslide<12>
 |foo(readString()): { (_: String) }|
-\onslide<8>
+\onslide<13>
 |foo(readString()): String|
@@
 
-\begin{slide}{Generalized Singletons: Conclusion}
+\begin{slide}{Generalized Singletons: Recap}
+
+We proposed the following 3 changes:
+\begin{enumerate}
+  \item lift a subset of Scala's language constructs to the type level
+  \item add a "precise mode" of type inference
+  \item evaluate those types during subtyping
+\end{enumerate}
+\end{slide}
+
 TODO
@@
 \end{itemize}
-
+\pause
 For example:
 
-|class A|                    -> C={A}, $\Psi$={}, $\Xi$={}
-|class A; class B extends A| -> C={A, B}, $\Psi$={(A, B)}, $\Xi$={}
-|class A; class D|           -> C={A, D}, $\Psi$={}, $\Xi$={(A, D)}
-|class A; trait T|           -> C={A, T}, $\Psi$={}, $\Xi$={}
+\begin{tabular}{@{}llll@{}}
+|class A|          &$\C = \{\A\}$     & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
+|class A; class B extends @\tp{A}@| &$\C = \{\A, \B\}$ & $\Psi = \{(\B, \A)\}$ & $\Xi = \varnothing$\\
+|class A; class D| &$\C = \{\A, \D\}$ & $\Psi = \varnothing$  & $\Xi = \{(\A, \D)\}$\\
+|class A; trait T| &$\C = \{\A, \T\}$ & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
+\end{tabular}
 \end{slide}
@@
 
-\begin{slide}{System FM's disjointness relation}
-\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
+\begin{slide}{System FM's pattern matching evaluation}
+Given a concrete scrutinee, we evaluate match expressions with a few lookups in the inheritance relation:
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/FMEMatch.pdf}
+\end{center}
 \end{slide}
 
-\begin{slide}{System FM's new subtyping rules}
-\includegraphics[width=\textwidth]{figures/FMSubtyping.pdf}
+\begin{slide}{System FM's disjointness relation}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/FMDisjointness.pdf}
+\end{center}
+\end{slide}
+
+\begin{slide}{System FM's match types evaluation rule (subtyping)}
+\includegraphics[width=\textwidth]{figures/FMSMatch.pdf}
 \end{slide}
@@
 \begin{slide}{System FM's type safety}
-theorems
+We show type safety through the progress and preservation.
 
-two versions of the proof, pen and paper and Coq
+Our proof comes in two versions:
+\begin{enumerate}
+  \item Pen and paper (\textasciitilde30 pages)
+  \item Coq mechanization (\textasciitilde6000 LOC)
+\end{enumerate}
 \end{slide}
@@
 \begin{slide}{Structure of the type safety proof}
-\includegraphics[width=\textwidth]{figures/FMProofStructure.pdf}
-\end{slide}
-
-\begin{slide}{Limitations}
-- type binders
-- higher-kinded types
-- recursion
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/FMProofStructure.pdf}
+\end{center}
 \end{slide}
@@
 \begin{slide}{Empty Types}
+Our system does not like empty types.
+
+|Nothing|, in particular, is both subtype and disjoint from every type.
+
+Intersections require special care.
+
+\pause
+
 \begin{code}
@@
   case String => Int
+
 class C:
@@
   def f(bad: M[X & String]): Int = bad
+
 class D extends C:
@@
 \begin{slide}{Variance}
+At first glance, we can't prove disjointness when variance is involved.
+
+|F[+T]:| \quad |T1 <: T2| \quad implies \quad |F[T1] <: F[T2]|\\
+|G[-T]:| \quad |T1 <: T2| \quad implies \quad |G[T1] >: G[T2]|
+\pause
+
+We make an exception for covariant type that are used as class fields:
 \begin{code}
-F[+T]: T1 <: T2 => F[T1] <: F[T2]
-G[-T]: T1 <: T2 => G[T1] >: G[T2]
+case class Some[+A](value: A)
+\end{code}
 
-case class Tuple2[+T1, +T2](_1: T1, _2: T2)
-\end{code}
+|Some[String]| and |Some[Int]| are disjoint, since there is no runtime value of type |Some[Nothing]|.
+
 \end{slide}
@@
 
-\begin{slide}{Type-Safe Regular Expressions}
-\timeline{80}{80}{80}{0}{80}
+\begin{slide}{Thank you!}
+\timeline{0}{0}{0}{0}{0}
 \end{slide}
 
-\begin{slide}{Type-Safe Regular Expressions (2022)}
+\appendix
+
+\begin{slide}{Type-Safe Regular Expressions (simplified)}
 \begin{code}
-val rational = new Regex("(\\d+)\\.?(\\d+)?")
-rational.unapply("3.1415"): Option[(String, Option[String])]
-\end{code}
-\pause
-\begin{code}
-type GroupsOf[P <: String] = ...
+val rational = new Regex("(\\d+)\\.(\\d+)")
+rational.unapply("3.1415"): Option[(String, String)]
 
+type GroupsOf[P <: String] =
+  Reverse[Loop[P, 0, Length[P], EmptyTuple]]
+
+type Loop[P, Lo, Hi, Acc <: Tuple] =
+  Lo match
+    case Hi => Acc
+    case _ => CharAt[P, Lo] match
+      case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
+      case '\\' => Loop[P, Lo + 2, Hi, Acc]
+      case _ => Loop[P, Lo + 1, Hi, Acc]
 \end{code}

### 2022-03-19 slides.tex
m
Update slides

@@
 
-\begin{slide}{Example 1: Database queries (2016)}
+\begin{slide}{Example 1: database queries (2016)}
 \begin{center}
@@
 
-\begin{slide}{Example 1: Database queries (2016)}
+\begin{slide}{Example 1: database queries (2016)}
 Spark APIs are practically untyped:
@@
 
-\begin{slide}{Example 2: Regular expressions (2022)}
+\begin{slide}{Example 2: regular expressions (2022)}
 Regular expressions in Scala's standard library:
@@
 \begin{slide}{What changed between 2016 and 2022?}
-Example 1 uses a hack, implicits:
+The Spark library uses a hack, implicits:
 \begin{itemize}
@@
 
-Example 2 uses a new language construct, match types, one of the main contributions of this thesis.
+The regex library uses match types, a new language construct introduced in this thesis.
 \end{slide}
@@
 
-\begin{slide}{Generalized singleton types: A proposal}
+\begin{slide}{Generalized singleton types: a proposal}
 Our proposal consists of 3 changes:
@@
 \pause
-Similarly, we add new types for other constructs: |Match[]|, |New[]| and |Call[]|, and the corresponding syntactic sugar.
+Similarly, we add new types for other constructs: |Match[]|, |New[]|, |Call[]|, |TypeTest[]|, and the corresponding syntactic sugar.
 \end{slide}
@@
 \end{code}
-\pause
 \begin{overprint}
@@
 More interesting for pattern matching: we "desugar" pattern matching expressions into if-then-else \& type tests.
-\pause
-\smallskip
-
+\end{slide}
+\begin{slide}{Type evaluation: type tests}
 We evaluate type tests using subtyping and type disjointness:
-\begin{code}
-eval(@\tp{x.asInstanceOf}@[Foo]) = true   iff  x.type <: Foo
-eval(@\tp{x.asInstanceOf}@[Foo]) = false  iff  disj(x.type, Foo)
-\end{code}
+\begin{itemize}
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
+\end{itemize}
 \end{slide}
@@
     case s: String => @\tp{s}@
-    @\tp{case i: Int}@ => @\tp{i+1}@
+    @\tp{case i: Int}@ @=>@ @\tp{i+1}@
 } =
@@
 \bigskip
+\pause\pause\pause
 
-\begin{overprint}
-\onslide<4>
-|foo(42): { foo(42) }|
-\onslide<5>
-|foo(42): { 42.asInstanceOf[Int] + 1 }|
-\onslide<6>
-|foo(42): { 42 + 1 }|
-\onslide<7>
 |foo(42): { 43 }|
-\onslide<8->
-|foo(42): 43|
-\end{overprint}
+\pause
 
-\begin{overprint}
-\onslide<9>
-|foo(readString())|
-\onslide<10>
-|foo(readString()): { foo(_: String) }|
-\onslide<11>
-|foo(readString()): { (_: String).asInstanceOf[String] }|
-\onslide<12>
-|foo(readString()): { (_: String) }|
-\onslide<13>
-|foo(readString()): String|
-\end{overprint}
+|foo(readInt(): { (_: Int) + 1 })|
 \end{slide}
 
-\begin{slide}{Generalized Singletons: Recap}
+\begin{slide}{Generalized singletons: recap}
 
-We proposed the following 3 changes:
+We proposed a generalization of Scala's singleton types:
 \begin{enumerate}
@@
   \item add a "precise mode" of type inference
-  \item evaluate those types during subtyping
+  \item evaluate types during subtyping
 \end{enumerate}
-\end{slide}
-
-TODO
 \end{slide}
@@
 
-\begin{slide}{Overview}
+\begin{slide}{Pattern matching on types}
 \begin{code}
@@
 \end{code}
+
+\medskip
+
+\begin{overprint}
+\onslide<2-5>
+\begin{code}
+Elem[String] =:= Char
+\end{code}
+
+\onslide<3-5>
+\begin{code}
+Elem[String] =:= Char
+Elem[Int] =:= Int
+\end{code}
+
+\onslide<4-5>
+\begin{code}
+Elem[String] =:= Char
+Elem[Int] =:= Int
+Elem[List[Int]] =:= Int
+\end{code}
+
+\onslide<5>
+\begin{code}
+Elem[String] =:= Char
+Elem[Int] =:= Int
+Elem[List[Int]] =:= Int
+Elem[Seq[Int]] =:= Elem[Seq[Int]]
+\end{code}
+
+\onslide<6>
+We evaluate type tests using subtyping and type disjointness:
+\begin{itemize}
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
+  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
+\end{itemize}
+
+\onslide<7>
+Match types can be explained in terms of generalized singletons:
+
+\medskip
+\begin{code}
+type Elem[X] = { (_: X) match
+  case _: String => (_: Char)
+  case _: List[t] @=>@ (_: Elem[t])
+  case _: Any => (_: X)
+}
+\end{code}
+
+\onslide<8>
+The type/term correspondence dictates match type semantics:
+
+\medskip
+\begin{code}
+def elem[X <: Singleton](x: X): Elem[X] = x match
+  case x: String => x.charAt(0)
+  case x: List[t] @=>@ elem(x.head)
+  case x: Any => x
+\end{code}
+
+\onslide<9>
+The type/term correspondence dictates match type semantics:
+
+\medskip
+\begin{code}
+def elem[X <: Singleton](x: X): Elem[X] = x match
+  case x: String => x.charAt(0)
+  case x: List[t] => elem(x.head)
+  case x: Any => x
+
+elem[Any]("foo")
+\end{code}
+\vspace{-20pt}
+\end{overprint}
 \end{slide}
 
-\begin{slide}{Disjointness}
-|Elem[Seq[Int]]| is stuck
-\end{slide}
-
-\begin{slide}{Relating match terms and types: a defining property}
-|Elem[Seq[Int]]| is stuck
-\end{slide}
-
-\begin{slide}{Formalizing Match Types}
+\begin{slide}{Formalization}
 System FM is an extension of System \Fsub with
@@
 \end{itemize}
+\pause
 Design goals:
@@
 \begin{tabular}{@{}llll@{}}
-|class A|          &$\C = \{\A\}$     & $\Psi = \varnothing$  & $\Xi = \varnothing$\\
 |class A; class B extends @\tp{A}@| &$\C = \{\A, \B\}$ & $\Psi = \{(\B, \A)\}$ & $\Xi = \varnothing$\\
@@
 
-\begin{slide}{Empty Types}
+\begin{slide}{Empty types}
 Our system does not like empty types.
@@
 
-\begin{slide}{Match Types, Conclusion}
-TODO
+\begin{slide}{Match types: recap}
+We presented match types, a new language construct for type-level programming:
+\begin{itemize}
+  \item implemented in the Scala 3 compiler
+  \item formalized in an extension of System \Fsub
+  \item already in active use
+\end{itemize}
 \end{slide}
@@
 
-\begin{slide}{Type-Safe Regular Expressions (simplified)}
+\begin{slide}{Type-safe regular expressions (simplified)}
 \begin{code}
@@
 
-% \appendix
 \end{document}

### 2022-03-21 slides.tex
slides

@@
 
+\usepackage{ulem}
 \usepackage{xspace}
@@
 
-The regex library uses match types, a new language construct introduced in this thesis.
+The regex library uses technology developed in this thesis.
 \end{slide}
@@
 \begin{center}
-\input{figures/join.gex}
+\input{figures/join-slides.gex}
 \end{center}
@@
 \begin{slide}{Singleton types}
-Scala supports a few singleton types:
+\begin{block}{Definition}
+A singleton type is a type that contains exactly one value.
+\end{block}
+
+\pause
+
+Scala has a few of those:
 \begin{itemize}
@@
 \pause
-Question: How much of the language can we represent in types?
+Question: How much more of the language can we represent as singleton types?
 \end{slide}
@@
 \end{code}
+
 \begin{overprint}
-\onslide<1-2>
+\onslide<1>
 \begin{code}
-dependent val l1 = "A" :: Nil
-dependent val l2 = "B" :: Nil
-dependent val l3 = concat(l1, l2)
+dependent val l1 @=@ "A" :: Nil
+dependent val l2 @=@ "B" :: Nil
+dependent val l3 @=@ concat(l1, l2)
 l3: { "A" :: "B" :: Nil }
+\end{code}
+
+\onslide<2>
+\begin{code}
+val l1: List     @=@ "A" :: Nil
+dependent val l2 @=@ "B" :: Nil
+dependent val l3 @=@ concat(l1, l2)
+l3: { concat(l1, l2) }
 \end{code}
@@
 \begin{code}
-val l1: List     = "A" :: Nil
-dependent val l2 = "B" :: Nil
-dependent val l3 = concat(l1, l2)
-l3: { concat(l1, l2) }
-\end{code}
-
-\onslide<4>
-\begin{code}
-dependent val l1 = "A" :: Nil
-val l2: List     = "B" :: Nil
-dependent val l3 = concat(l1, l2)
+dependent val l1 @=@ "A" :: Nil
+val l2: List     @=@ "B" :: Nil
+dependent val l3 @=@ concat(l1, l2)
 l3: { "A" :: l2 }
@@
 During subtyping, we evaluate both sides:
-\begin{code}
-A <: B  iff  eval(A) <: eval(B)
-\end{code}
+\begin{itemize}
+  \item |A <: B|  \quad if \quad eval(|A|) |<:| eval(|B|)
+\end{itemize}
 \pause
@@
 Straightforward for if-then-else:
-\begin{code}
-eval(If[true,  A, B]) = A
-eval(If[false, A, B]) = B
-\end{code}
+\begin{itemize}
+  \item eval(|If[true, A, B]|) = |A|
+  \item eval(|If[false, A, B]|) = |B|
+\end{itemize}
 \pause
@@
 \begin{itemize}
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
+  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
+  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |false| \quad if |x.type| and |@\tp{T}@| are disjoint
+  \item eval(|@\tp{x.isInstanceOf}@[T]|) = |@\tp{x.isInstanceOf}@[T]| \quad otherwise
 \end{itemize}
@@
 
-|foo(readInt(): { (_: Int) + 1 })|
+|foo(readInt()): { (_: Int) + 1 }|
+\end{slide}
+
+\begin{slide}{Generalized, or not singleton at all?}
+\begin{code}
+dependent
+def int2str(x: Int)
+  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ (_: String) }
+  =   if (x == 0) "zero" else readString()
+\end{code}
+\begin{overprint}
+\onslide<1>
+\begin{block}{Definition}
+A singleton type is a type that contains exactly one value.
+\end{block}
+
+\onslide<2>
+\begin{block}{Definition}
+\sout{A singleton type is a type that contains exactly one value.}
+\end{block}
+\end{overprint}
 \end{slide}
@@
 \begin{slide}{Generalized singletons: recap}
-
 We proposed a generalization of Scala's singleton types:
@@
 
-% Termination
-% Virtual dispatch
+% % Termination
+% % Virtual dispatch
 
@@
 \begin{overprint}
-\onslide<2-5>
-\begin{code}
-Elem[String] =:= Char
-\end{code}
-
-\onslide<3-5>
-\begin{code}
-Elem[String] =:= Char
-Elem[Int] =:= Int
-\end{code}
-
-\onslide<4-5>
-\begin{code}
-Elem[String] =:= Char
-Elem[Int] =:= Int
-Elem[List[Int]] =:= Int
-\end{code}
-
-\onslide<5>
-\begin{code}
-Elem[String] =:= Char
-Elem[Int] =:= Int
-Elem[List[Int]] =:= Int
-Elem[Seq[Int]] =:= Elem[Seq[Int]]
-\end{code}
-
-\onslide<6>
-We evaluate type tests using subtyping and type disjointness:
-\begin{itemize}
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = true@ @| \quad if |x.type| is a subtype of |@\tp{T}@|
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = false| \quad if |x.type| and |@\tp{T}@| are disjoint
-  \item |eval(@\tp{x.isInstanceOf}@[T]) = @\tp{x.isInstanceOf}@[T]| \quad otherwise
-\end{itemize}
-
-\onslide<7>
+\onslide<2>
 Match types can be explained in terms of generalized singletons:
@@
 
-\onslide<8>
+\onslide<3>
+We evaluate type tests using subtyping and type disjointness:
+\begin{itemize}
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |true@ @| \quad if |@\tp{X}@| is a subtype of |@\tp{T}@|
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |false| \quad if |@\tp{X}@| and |@\tp{T}@| are disjoint
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |@\tp{X.isInstanceOf}@[T]| \quad otherwise
+\end{itemize}
+
+\onslide<4>
 The type/term correspondence dictates match type semantics:
@@
 \begin{code}
-def elem[X <: Singleton](x: X): Elem[X] = x match
+def elem[X](x: X): Elem[X] = x match
   case x: String => x.charAt(0)
@@
 
-\onslide<9>
+\onslide<5>
 The type/term correspondence dictates match type semantics:
@@
 \begin{code}
-def elem[X <: Singleton](x: X): Elem[X] = x match
+def elem[X](x: X): Elem[X] = x match
   case x: String => x.charAt(0)
-  case x: List[t] => elem(x.head)
+  case x: List[t] @=>@ elem(x.head)
   case x: Any => x
@@
 \end{code}
-\vspace{-20pt}
+\vspace{-10pt}
 \end{overprint}
@@
 \begin{itemize}
+  \item classes (defined externally)
   \item pattern matching
-  \item opaque classes
   \item match types
@@
 
-\begin{slide}{System FM's pattern matching evaluation}
-Given a concrete scrutinee, we evaluate match expressions with a few lookups in the inheritance relation:
+\begin{slide}{System FM's disjointness relation}
 \begin{center}
-\includegraphics[width=0.8\textwidth]{figures/FMEMatch.pdf}
+\includegraphics[width=\textwidth]{figures/FMDisjointness.pdf}
 \end{center}
@@
 
-\begin{slide}{System FM's disjointness relation}
+\begin{slide}{System FM's evaluation rules}
+We evaluate term-level matches by querying the class inheritance relation:
 \begin{center}
-\includegraphics[width=0.9\textwidth]{figures/FMDisjointness.pdf}
+\includegraphics{figures/FMEMatch.pdf}
 \end{center}
-\end{slide}
 
-\begin{slide}{System FM's match types evaluation rule (subtyping)}
-\includegraphics[width=\textwidth]{figures/FMSMatch.pdf}
+\visible<2>{We evaluate type-level matches using subtyping and disjointness:}
+
+\begin{center}
+\visible<2>{\includegraphics{figures/FMSMatch.pdf}}
+\end{center}
 \end{slide}
@@
 \begin{slide}{Empty types}
+
+\begin{block}{Definition}
+An empty type is a type that contains no values.
+\end{block}
+
+\pause
+
 Our system does not like empty types.
@@
 |Nothing|, in particular, is both subtype and disjoint from every type.
-
-Intersections require special care.
 
@@
 \begin{slide}{Variance}
-At first glance, we can't prove disjointness when variance is involved.
+\begin{block}{Definitions}
+|F| is covariant in |@\tp{T}@|, written |F[+T]|, if |@\tp{T1}@| |<:| |@\tp{T2}@| implies |F[T1]| |<:| |F[T2]|.\\
+|G| is contravariant in |@\tp{T}@|, |G[-T]|, if |@\tp{T1}@| |<:| |@\tp{T2}@| implies |G[T1]| |>:| |G[T2]|.
+\end{block}
 
-|F[+T]:| \quad |T1 <: T2| \quad implies \quad |F[T1] <: F[T2]|\\
-|G[-T]:| \quad |T1 <: T2| \quad implies \quad |G[T1] >: G[T2]|
+\pause
+
+We cannot prove disjointness when variance is involved:
+
+$\forall$ |T1|, |T2|. \quad  |F[Nothing] <: F[T1]| \quad  and  \quad |F[Nothing] <: F[T2]|\\
+$\forall$ |T1|, |T2|. \quad  |    G[Any] <: G[T1]|  \quad and  \quad |    G[Any] <: G[T2]|
+
 \pause
@@
 
-\appendix
+% \appendix
 
-\begin{slide}{Type-safe regular expressions (simplified)}
-\begin{code}
-val rational = new Regex("(\\d+)\\.(\\d+)")
-rational.unapply("3.1415"): Option[(String, String)]
+% \begin{slide}{Type-safe regular expressions (simplified)}
+% \begin{code}
+% val rational = new Regex("(\\d+)\\.(\\d+)")
+% rational.unapply("3.1415"): Option[(String, String)]
 
-type GroupsOf[P <: String] =
-  Reverse[Loop[P, 0, Length[P], EmptyTuple]]
+% type GroupsOf[P <: String] =
+%   Reverse[Loop[P, 0, Length[P], EmptyTuple]]
 
-type Loop[P, Lo, Hi, Acc <: Tuple] =
-  Lo match
-    case Hi => Acc
-    case _ => CharAt[P, Lo] match
-      case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
-      case '\\' => Loop[P, Lo + 2, Hi, Acc]
-      case _ => Loop[P, Lo + 1, Hi, Acc]
-\end{code}
-\end{slide}
+% type Loop[P, Lo, Hi, Acc <: Tuple] =
+%   Lo match
+%     case Hi => Acc
+%     case _ => CharAt[P, Lo] match
+%       case '('  => Loop[P, Lo + 1, Hi, String *: Acc]
+%       case '\\' => Loop[P, Lo + 2, Hi, Acc]
+%       case _ => Loop[P, Lo + 1, Hi, Acc]
+% \end{code}
+% \end{slide}
 

### 2022-03-22 slides.tex
Slides

@@
 
+\usepackage[sorting=none, backend=biber, maxbibnames=9]{biblatex}
+\settowidth\labelnumberwidth{\usebeamertemplate*{bibliography item}}
+\AtEveryBibitem{\clearfield{note}}
+\AtEveryBibitem{\clearfield{volume}}
+\AtBeginBibliography{\small}
+\bibliography{bibliography}
+
+\usepackage{textpos}
 \usepackage{ulem}
@@
 
-The regex library uses technology developed in this thesis.
+The regex library uses a first-class language construct:
+\begin{itemize}
+  \item easier to use
+  \item faster to compile
+\end{itemize}
+
 \end{slide}
@@
 \begin{slide}{Timeline of my PhD}
+\begin{overprint}
+\onslide<1>
 \timeline{0}{0}{0}{0}{0}
-\end{slide}
-
-\begin{slide}{Part I: Generalized Singleton Types}
+\onslide<2>
+\timeline{80}{80}{80}{0}{80}
+\vspace{-10pt}
+\begin{refsection}%
+\raggedright
+\phantom{\cite{blanvillain2022regular}}%
+\printbibliography[heading=subbibliography]%
+\end{refsection}
+\vspace{-40pt}
+\onslide<3>
 \timeline{80}{0}{80}{80}{80}
+\onslide<4>
+\timeline{80}{0}{80}{80}{80}
+\vspace{-10pt}
+\begin{refsection}%
+\phantom{\cite{schmid2020coming}}%
+\printbibliography[heading=subbibliography]%
+\end{refsection}
+\vspace{-40pt}
+\end{overprint}
 \end{slide}
@@
 
-\begin{slide}{Generalized, or not singleton at all?}
-\begin{code}
-dependent
-def int2str(x: Int)
-  : { if (x @\tp{== 0}@) "zero" @\tp{else}@ (_: String) }
-  =   if (x == 0) "zero" else readString()
-\end{code}
-\begin{overprint}
-\onslide<1>
-\begin{block}{Definition}
-A singleton type is a type that contains exactly one value.
-\end{block}
-
-\onslide<2>
-\begin{block}{Definition}
-\sout{A singleton type is a type that contains exactly one value.}
-\end{block}
-\end{overprint}
-\end{slide}
-
 \begin{slide}{Generalized singletons: recap}
@@
 \begin{slide}{Part II: Match Types}
+\begin{overprint}
+\onslide<1>
 \timeline{80}{80}{0}{80}{80}
+\onslide<2>
+\timeline{80}{80}{0}{80}{80}
+\vspace{-10pt}
+\begin{refsection}%
+\phantom{\cite{blanvillain2022type}}%
+\printbibliography[heading=subbibliography]%
+\end{refsection}
+\vspace{-40pt}
+\end{overprint}
 \end{slide}
@@
 \onslide<2>
+Some examples of reduction:
+\begin{itemize}
+  \item |Elem[String] =:= Char|
+  \item |Elem[Int] =:= Int|
+  \item |Elem[List[Int]] =:= Int|
+  \item |Elem[Any] =:= Elem[Any]|
+\end{itemize}
+
+\onslide<3>
+We evaluate type tests using subtyping and type disjointness:
+\begin{itemize}
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |true@ @| \quad if |@\tp{X}@| is a subtype of |@\tp{T}@|
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |false| \quad if |@\tp{X}@| and |@\tp{T}@| are disjoint
+  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |@\tp{X.isInstanceOf}@[T]| \quad otherwise
+\end{itemize}
+
+\onslide<4>
 Match types can be explained in terms of generalized singletons:
@@
 
-\onslide<3>
-We evaluate type tests using subtyping and type disjointness:
-\begin{itemize}
-  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |true@ @| \quad if |@\tp{X}@| is a subtype of |@\tp{T}@|
-  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |false| \quad if |@\tp{X}@| and |@\tp{T}@| are disjoint
-  \item eval(|@\tp{X.isInstanceOf}@[T]|) = |@\tp{X.isInstanceOf}@[T]| \quad otherwise
-\end{itemize}
-
-\onslide<4>
+\onslide<5>
 The type/term correspondence dictates match type semantics:
@@
 
-\onslide<5>
+\onslide<6>
 The type/term correspondence dictates match type semantics:
@@
 \end{code}
-\vspace{-10pt}
 \end{overprint}
@@
 
+\begin{slide}{Formalization vs implementation}
+Our implementation goes much further than System FM:
+\begin{itemize}
+  \item recursion / non-termination
+  \item type parameter bindings (|case List[t]|)
+  \item empty types
+  \item variance
+\end{itemize}
+\end{slide}
+
 \begin{slide}{Empty types}
@@
 
-We cannot prove disjointness when variance is involved:
+We cannot prove disjointness when variance is involved, |F[T1]| and |F[T2]| always overlap:
 

### 2022-03-28 preprint.tex
Partially address Viktor's comments

@@
 During type normalization, our system evaluates pattern matching expressions according to Scala's runtime semantics.
-Patterns are checked top-to-bottom, and type-tests are evaluated using runtime type information available after type erasure.
+Patterns are checked top-to-bottom, and type-tests are evaluated using subtyping checks.
 
@@
 We could have equivalently defined |getName| by omitting the |dependent| qualifier and writing its result type explicitly.
+The difference between the two would only be matter of syntax.
 
@@
 Exception handlers are typed imprecisely using the default mode of type-inference.
-Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{e1; e2}| is simply |{ e2 }|.
+Exceptions thrown in statement positions are not reflected in singleton types, since the type of |{ e1; e2 }| is simply |{ e2 }|.
 However, exceptions thrown in tail positions (such as in remove from \Cref{sec:motivating-example}) can lead to types normalizing to |Error(e)|.

### 2022-03-28 thesis.tex
Partially address Viktor's comments

@@
 The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
+The right-hand side of those definitions is devoid of meaning, instances of the |Remove| trait merely act as placeholders (the type parameters of those constructor calls are inferred from their expected type).
 The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.
@@
 \noindent
-The |Singleton| type bound is a marker that instructs type inference to preserve literal singleton types (by default, the compiler widens those types).
+The |Singleton| type bound is a marker that instructs type inference to preserve union types and literal singleton types (by default, the compiler widens those types).
 The "|?|" type is Scala~3's new syntax for wildcard types~\citep[Wildcard Arguments in Types]{odersky2013scala}.
@@
 \label{chap:match-types}
+
+% TODO: Compare to singleton types
 

### 2022-03-30 article.tex
Partially address Viktor's comments
Comments

@@
 %
-We have implemented match types in the latest version of Scala, an industry-grade, production-ready compiler.
+We have implemented match types in the latest version of Scala.
 Match types have received a great interest from the Scala community, and are already in active use.
@@
 
+% TODO: tell the truth about empty types
+
 \subsection{Null Values}
@@
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
+
+% TODO make it clear that contravariant types are never considered disjoint.
+% TODO Incorporate email discussion on match types co- and contravariance.
 
@@
 The outer loop, |Loop|, traverses the shape and counts the current index.
-The inner loops are implemented using |Contains| and |Remove|, standard operations on HLists (omitted).
+The inner loops are implemented using the standard |Contains| and |Remove| operations.
 When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.

### 2022-03-30 preprint.tex
Comments

@@
 \begin{lstlisting}
-def addRes(a: Any, b: Any) =
+dependent def addRes(a: Any, b: Any) =
   (a, b) match

### 2022-04-05 preprint.tex
Copy paste more of the dependent scala  related work

@@
 Consequently, the lack of totality checks does not affect Scala's present notion of safety.
-Exceptions or infinite loops in the evaluation of a `dependent` method would prevent the completion of type-checking.
+Exceptions or infinite loops in the evaluation of a |dependent| method would prevent the completion of type-checking.
 
@@
 
-% TODO: Compare to pure subtype systems
+Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
+His type system is much more expressive than ours and allows bounded quantification over both types and terms using the notion of a |Type| type and power types.
+Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.
+
+In~\citep{aspinall1994subtyping} Aspinall introduces a dependently-typed system with subtyping \emph{and} singleton types that resembles ours in its type language.
+His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
+Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
+Aspinall's system also has primitives types and allows for atomic subtyping among them, but no congruence rules, so partially-widened forms like |(_: Any) :: Nil| cannot be represented.
+
+Pure subtype systems~\citep{hutchins2010pure} are different in that they only contain a single syntactic category for both terms and types, and a single relation, subtyping, that subsumes typing, subtyping and type evaluation of our system.
+However, their system allows for partially-widened types similar to ours and also enables the computation with different levels of precision.
+For instance, it is able to conclude that |Int + 5| can be approximated as |Int|.
+
+Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
+Though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their goals largely orthogonal to ours.

### 2022-04-07 thesis.tex
Comments
Address Martin's comments on abmiguities

@@
 \subsection{Implicit Ambiguities}
+\label{subsec:implicit-ambiguities}
 
@@
 \subsection{Implicit Priorities}
+\label{subsec:implicit-priorities}
 
@@
 This incremental process has two consequences.
-First, it allows the compiler to efficiently look for implicits by naturally pruning some of the search space.
+First, it allows the compiler to efficiently look for implicits by naturally pruning the search space.
 Second, it provides an ad-hoc mechanism to disambiguate implicits.
@@
 
-% TODO: Add a paragraph about the evolution of the "rule for picking a most specific alternative" during implicit resolution, see https://dotty.epfl.ch/docs/reference/changed-features/implicit-resolution.html
+In Scala~3, the rules for implicit ambiguities changed to take implicit parameters into account, thus introducing another, more direct, disambiguation mechanism.
+All else being equal, an implicit definition that takes some implicit parameters is considered to be less specific than an implicit definition that takes none~\citep[Changes in Implicit Resolution]{odersky2013scala}.
+As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).
+TODO
 
@@
 
+% TODO you could write a bit more. Maybe move some of the text of the conclusions of individual chapters there?
+
 In this thesis, we presented three techniques for type-level programming in Scala: implicits, generalized singleton types, and match types.

### 2022-04-07 preprint.tex
Add Richard's comments

@@
 This effectively allows users to execute programs involving functions and pattern matching at the type level.
+
+% TODO: Add a sentence about the state of adoptions of these ideas
 
@@
 The difference between the two would only be matter of syntax.
+% TODO: say something about the fact that generalized singleton types are not singletons...
 
 Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
-However, in the case of a |dependent| method, an explicit result type is only used as an upper bound for the actual precise result type and will only be used to type-check the method's body.
+However, in the case of a |dependent| method, an explicit result type only serves as an upper bound used to type-check the method's body.
 At other call sites, the (precise) inferred result type is used.
@@
 However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
+% TODO: More generally, when dependently typed functions call into a non-dependently typed one, ...
 
@@
 It would be interesting to explore the addition of an effect system to our Scala extension.
+
+% TODO: Add an example
 

### 2022-04-07 article.tex
Add Richard's comments
Define Λ and V

@@
 
+% TODO: define Ψ and Ξ
+% TODO: cite the precise figure in Pierce 2002
 \FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$.
@@
 We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
-Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$.
+Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representative for abstractions and type abstractions, respectively.
 Elements of $\P$ can be understood as equivalence classes for \Fm's types.
@@
 Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
+% TODO: clarify that harpoon is asymmetric (and so it disj)
 It represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
@@
 In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.
+% TODO: say that our approach would not work in a system with full type erasure
 
@@
 
-Unlike our calculus, the Scala implementation also allows match types to be defined recursively.
+Unlike our calculus, the Scala implementation also allows match type definitions to be recursive.
 Recursive match types can cause subtyping checks to loop indefinitely.
@@
 Instead, explicit type annotations must be provided.
+% TODO: Add examples
 

### 2022-04-08 article.tex
Define Ψ and Ξ in figure caption

@@
 
-% TODO: define Ψ and Ξ
-% TODO: cite the precise figure in Pierce 2002
-\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$.
-\hl{\text{Highlights}} correspond to additions to \SystemFsub.}
+\FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$.
+The $\Ψ$ relation is a partial order on $\C$ that describes the inheritance between classes.
+\hl{\text{Highlights}} correspond to additions to \SystemFsub, as per \citep[Figure 26-1]{pierce2002types}.}
 
-\FMTypingRules{\SystemFm type system for a given set of classes $\C$ with class inheritance $\Ψ$ \\and class disjointness $\Ξ$. \hl{\text{Highlights}} correspond to additions to \SystemFsub.}
+\FMTypingRules{\SystemFm type system for a given set of classes $\C$ with class inheritance $\Ψ$ and class disjointness $\Ξ$.
+$\Ψ$ is a partial order on $\C$ that describes the inheritance between classes.
+$\Ξ$ is symmetric relation over $\C$ that relates classes which share no inhabitants.
+\hl{\text{Highlights}} correspond to additions to \SystemFsub, as per \citep[Figure 26-1]{pierce2002types}.}
 

### 2022-04-08 article.tex
Talk a bit more about type erasure

@@
 
-% \olivier{
-% Type erasure is the process by which type arguments of generic types (type constructors) are removed from programs during compilation.
-% Scala's erasure affects pattern matching because it prevents certain patterns from being checked at runtime.
-% For instance, both `List[String]` and `List[Int]` erase to `List[Object]`, which makes them indistinguishable at runtime.
-% This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
-% Match types reduce during type checking (before erasure) and are not affected by this limitation.
-% Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.
-% }
+Scala's primary platform is the Java virtual machine (JVM).
+On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
+Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
+For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
+This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
+Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
+Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.
 
-Scala's primary platform is the Java virtual machine (JVM).
-On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased~\citep{schinz2005compiling}.
-For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.
-
-Erasure directly affects the type tests that can be performed at runtime.
-For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it will lead to a compiler warning since the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
+In practice, the Scala compiler copes with this limitation by emitting warnings for unchecked patterns.
+For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
 
@@
 In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.
-% TODO: say that our approach would not work in a system with full type erasure
+
+In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.
 

### 2022-04-08 symposium.tex
Address Martin's comments on abmiguities
Add Richard's comments
Talk a bit more about type erasure

@@
 
-% TODO: consider inline functions as a 3rd alternative
-
 \begin{enumerate}
@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
-In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala.
+% TODO: Rework that sentence
 An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.
@@
 The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
-The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
+The implementation uses conditional types \citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
 Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).

### 2022-04-19 article.tex
Do TODOs

@@
 Instead, explicit type annotations must be provided.
-% TODO: Add examples
+
+For example, consider the following function definition:
+
+\begin{lstlisting}
+def foo[X](x: X) = x match
+  case x: Int => x.toString
+  case x: String => x.toInt
+\end{lstlisting}
+
+In the absence of result type annotation, Scala's type inference will first assign \lstinline!String | Int!
+for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
+To obtain a precise result type for this method, users should annotate it explicitly:
+
+\begin{lstlisting}
+type Foo[X] = X match
+  case Int => String
+  case String => Int
+
+def foo[X](x: X): Foo[X] = x match
+  case x: Int => x.toString
+  case x: String => x.toInt
+\end{lstlisting}
 

### 2022-04-19 symposium.tex
Do TODOs

@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-% TODO: Rework that sentence
-An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.
+An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
 

### 2022-04-19 preprint.tex
Do TODOs
Generalized singleton type

@@
 This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
-When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|.
+When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
+\footnotetext{%
+When a generalized singleton type contains a component of form |_: T|, then that type is not singleton: it contains more than one value.
+}
 
@@
 The difference between the two would only be matter of syntax.
-% TODO: say something about the fact that generalized singleton types are not singletons...
 
@@
 However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
-% TODO: More generally, when dependently typed functions call into a non-dependently typed one, ...
+More generally, when a dependently-typed function call into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
 

### 2022-04-19 article.tex
Do TODOs

@@
 Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
-% TODO: clarify that harpoon is asymmetric (and so it disj)
-It represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
+This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
 Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
+Although $\Γ \⊢ \S \⇌ \T$ witness mutual subtyping between $\S$ and $\T$, that relation is asymmetric since its left-hand side is always a match type.
 %

### 2022-04-19 preprint.tex
Do TODOs

@@
 \footnotetext{%
-When a generalized singleton type contains a component of form |_: T|, then that type is not singleton: it contains more than one value.
-}
+When a generalized singleton type contains a component of form \texttt{(\_: T)}, then that type is not singleton: it might contains more than one value.}
 

### 2022-04-21 preprint.tex
Do TODO

@@
 
-% TODO: Add a sentence about the state of adoptions of these ideas
-
 The remaining of the chapter is organized as follows:
@@
 Our original presentation of generalized singleton types contains a formalization \citep[Section 3 and 4]{schmid2020coming}, which is out of the scope of this chapter.
+
+A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
+Although the ideas presented in this did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala 3 language.
 
@@
 
-% TODO: Add an example
-
 \subsection{Virtual Dispatch}

### 2022-04-21 thesis.tex
Add Richard's comments
Define Λ and V
Do TODOs
Do TODO

@@
 As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).
-TODO
 
@@
 The present chapter is a reboot of the match-type side of that project.
-We plan to submit this work to the upcoming Scala Symposium.
+This work is under submission at the Scala Symposium (2022).
 
@@
 For the implicit and match type benchmarks, we used the latest version of Scala at the time of writing (3.1.2).
-For the generalized singleton benchmarks, we used our prototype, which lives in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
+For the generalized singleton benchmarks, we used our prototype, which is available in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
 

### 2022-04-21 article.tex
Tell the truth about empty types

@@
 
-% TODO: tell the truth about empty types
+In practice, it would be quite restrictive to only reduce match types with provably inhabited scrutinees since it would disallow matching on abstract types.
+Instead, our implementation incorporates non-emptyness in the
+definition of the disjointness relation:
+
+$disj(T1, T2) = T1 nonempty and \forall e \in T1. e \notin T2$
+
+This updated definition allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} in the presence of empty types (disjointness/subtyping exclusivity).
+
+% TODO: tell the truth about empty types?
 
@@
 
-In practice, the Scala compiler copes with this limitation by emitting warnings for unchecked patterns.
+In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
 For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
@@
 
-In the absence of result type annotation, Scala's type inference will first assign \lstinline!String | Int!
+In the absence of result type annotation, Scala's type inference will first assign \lstinline!String|Int!
 for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.

### 2022-04-21 ack.tex
ack.tex

@@
 
-I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and for taking part in the committee.
+I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for reviewing my dissertation and for providing valuable feedback for improvement.
 

### 2022-04-21 article.tex
Do TODO

--- 2022-04-19 article.tex
@@
 Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
-To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting any reduction.
+To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
+This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} in the presence of empty types (disjointness/subtyping exclusivity).
 
@@
 Checking for scrutinee inhabitance prevents this class of errors. In this example, it would prevent |M[Int| |&| |String]| from reducing given that |Int| |&| |String| is not inhabited.
-
-% TODO: tell the truth about empty types
 

### 2022-04-22 article.tex
Make it clear
Talk about co- and contravariance of MT

@@
 
-\subsection{Variance}
+\subsection{Disjointness of Variant Types}
+\label{subsec:disjointness-of-variant-types}
 
@@
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
+There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
 
-% TODO make it clear that contravariant types are never considered disjoint.
-% TODO Incorporate email discussion on match types co- and contravariance.
+\subsection{Match Type Variance}
+
+In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotation of higher-kinded types.
+In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.
+
+At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
+However, it would be conceivably to lift this restriction in the future, as shown by the analysis presented in this section.
+
+Consider the following definition of a match type |M|:
+%
+\begin{lstlisting}
+type M[X] = X match
+  case P1 => E1
+  case P2 => E2
+  case P3 => E3
+\end{lstlisting}
+%
+In \Cref{sec:formalization}, we define the semantics of this type in terms of subtyping rules.
+Given this particular definition of |M|, we can write a direct semantic definition of |M| with equations:
+%
+\begin{itemize}
+ \item |M[X] = E1| \quad if |X <: P1|
+ \item |M[X] = E2| \quad if |disj(X, P1)| and |X <: P2|
+ \item |M[X] = E3| \quad if |disj(X, P1)| and |disj(X, P2)| and |X <: P3|
+\end{itemize}
+%
+This equational definition is consistent under the assumptions of \Cref{lem:disjointness-subtyping-exclusivity}.
+
+We show that |M[X]| is covariant if |E1|, |E2|, and |E3| are covariant with respect to |X|.
+Given |T1 <: T2|, we proceed by case analysis on |M[T2]|:
+%
+\begin{itemize}
+  \item If |M[T2] = E1|, then |M[T1] = E1| by subtyping transitivity and the result follows from the covariance of |E1|.
+
+  \item If |M[T2] = E2|, then (|disj(T2, P1)| and |T1 <: T2|) implies |disj(T1, P1)| (using D-SUB), and (|T2 <: P2| and |T1 <: T2|) implies |T1 <: P2|, which yields |M[T1] = E2|.
+  The result follows from the covariance of |E2|.
+
+  \item |M[T2] = E3| is analogous.
+\end{itemize}
+%
+Similarly, we can show that |M[X]| is contravariant if |E1|, |E2|, and |E3| are contravariant with respect to |X|.
+
+In practice, it would be hard to show that |E1|, |E2|, and |E3| are co-/contravariant when type bindings get into the picture.
+For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.
 

### 2022-04-22 ack.tex
Annotate diffs

--- 2022-02-15 ack.tex
@@
 
-I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and for taking part in the committee.
+\begin{diff}
+I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for reviewing my dissertation and for providing valuable feedback for improvement.
+\end{diff}
 

### 2022-04-22 article.tex
Annotate diffs

@@
 We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
+\begin{diff}
 Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representative for abstractions and type abstractions, respectively.
+\end{diff}
 Elements of $\P$ can be understood as equivalence classes for \Fm's types.
@@
 Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
+\begin{diff}
 This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
@@
 Although $\Γ \⊢ \S \⇌ \T$ witness mutual subtyping between $\S$ and $\T$, that relation is asymmetric since its left-hand side is always a match type.
+\end{diff}
 %
@@
 Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
+\begin{diff}
 To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
 This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} in the presence of empty types (disjointness/subtyping exclusivity).
+\end{diff}
 
@@
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
+\begin{diff}
 There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
+\end{diff}
 
@@
 
+\begin{diff}
 In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotation of higher-kinded types.
@@
 For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.
+\end{diff}
 
@@
 
+\begin{diff}
 Scala's primary platform is the Java virtual machine (JVM).
@@
 For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
+\end{diff}
 
@@
 
+\begin{diff}
 In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.
+\end{diff}
 
@@
 
+\begin{diff}
 For example, consider the following function definition:
-
+%
 \begin{lstlisting}
@@
 \end{lstlisting}
-
+%
 In the absence of result type annotation, Scala's type inference will first assign \lstinline!String|Int!
@@
 To obtain a precise result type for this method, users should annotate it explicitly:
-
+%
 \begin{lstlisting}
@@
 \end{lstlisting}
+\end{diff}
 

### 2022-04-22 preprint.tex
Annotate diffs

@@
 
+\begin{diff}
 A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
 Although the ideas presented in this did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala 3 language.
+\end{diff}
 
@@
 This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
+\begin{diff}
 When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
@@
 When a generalized singleton type contains a component of form \texttt{(\_: T)}, then that type is not singleton: it might contains more than one value.}
+\end{diff}
 
@@
 Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
+\begin{diff}
 However, in the case of a |dependent| method, an explicit result type only serves as an upper bound used to type-check the method's body.
+\end{diff}
 At other call sites, the (precise) inferred result type is used.
@@
 However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
+\begin{diff}
 More generally, when a dependently-typed function call into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
+\end{diff}
 
@@
 
+\begin{diff}
 Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
@@
 Though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their goals largely orthogonal to ours.
+\end{diff}

### 2022-04-22 symposium.tex
Annotate diffs

@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
+\begin{diff}
 An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+\end{diff}
 

### 2022-04-22 thesis.tex
Annotate diffs

@@
 
+\begin{diff}
 In Scala~3, the rules for implicit ambiguities changed to take implicit parameters into account, thus introducing another, more direct, disambiguation mechanism.
@@
 As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).
+\end{diff}
 
@@
 The present chapter is a reboot of the match-type side of that project.
+\begin{diff}
 This work is under submission at the Scala Symposium (2022).
+\end{diff}
 

### 2022-04-22 preprint.tex
Address Dragana's comments

@@
 A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
-Although the ideas presented in this did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala 3 language.
+Although the ideas presented in this chapter did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala language.
 \end{diff}
@@
 \footnotetext{%
-When a generalized singleton type contains a component of form \texttt{(\_: T)}, then that type is not singleton: it might contains more than one value.}
+When a generalized singleton type contains a component of form \texttt{(\_: T)}, that type is not singleton: it might contains more than one value.}
 \end{diff}
@@
 \begin{diff}
-More generally, when a dependently-typed function call into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
+More generally, when a dependently-typed function calls into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
 \end{diff}
@@
 Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
-His type system is much more expressive than ours and allows bounded quantification over both types and terms using the notion of a |Type| type and power types.
+\citeauthor{cardelli1988structural}'s type system is much more expressive than ours and allows bounded quantification over both types and terms, using the notion of a |Type| type and power types.
 Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.
 
-In~\citep{aspinall1994subtyping} Aspinall introduces a dependently-typed system with subtyping \emph{and} singleton types that resembles ours in its type language.
+In~\citep{aspinall1994subtyping} Aspinall introduces a dependently-typed system with subtyping \emph{and} singleton types which resembles ours in its type language.
 His equivalence relation on types is more powerful and is not syntax-directed, unlike our type evaluation relation.
 Furthermore, singleton types in his work are indexed by the type through which equality is "viewed", thereby enabling a form of polymorphism beyond ours.
-Aspinall's system also has primitives types and allows for atomic subtyping among them, but no congruence rules, so partially-widened forms like |(_: Any) :: Nil| cannot be represented.
+Aspinall's system also has primitive types and allows for atomic subtyping among them, but no congruence rules, so partially-widened forms like |(_: Any) :: Nil| cannot be represented.
 
 Pure subtype systems~\citep{hutchins2010pure} are different in that they only contain a single syntactic category for both terms and types, and a single relation, subtyping, that subsumes typing, subtyping and type evaluation of our system.
-However, their system allows for partially-widened types similar to ours and also enables the computation with different levels of precision.
+However, \citeauthor{hutchins2010pure}'s system allows for partially-widened types similar to ours and also enables computations with different levels of precision.
 For instance, it is able to conclude that |Int + 5| can be approximated as |Int|.
@@
 Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
-Though they introduce a form of dependency, path-dependent types were not designed for type-level computation, rendering their goals largely orthogonal to ours.
+Though they introduce a form of dependency, path-dependent types were not designed for type-level computations, rendering their goals largely orthogonal to ours.
 \end{diff}

### 2022-04-22 article.tex
Address Dragana's comments
Do TODO

@@
 \hl{\text{Highlights}} correspond to additions to \SystemFsub, as per \citep[Figure 26-1]{pierce2002types}.}
+
+\subsection{Comparison to Generalized Singleton Types}
+
+In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
+Generalized singletons and match types share some similarities, such as the notion of disjointness which plays an important role in both features.
+The two features relate on deeper level: we can explain match types in terms of generalized singletons.
+For example, the we can give an alternative definition of |Elem| match type using a generalized singleton:
+%
+\begin{lstlisting}
+type Elem[X] = { (_: X) match
+  case _: String => (_: Char)
+  case _: List[t] => (_: Elem[t])
+  case _: Any => (_: X)
+}
+\end{lstlisting}
+%
+This definition makes the term/type correspondence evident: the |Elem[X]| type reduces in the way than the corresponding match expression.
 

### 2022-04-22 slides.scala
Remove unused file

@@ -1,54 +1,0 @@
-def bestNeighborhood(phoneBook: TypedDataFrame[PhoneBookEntry], cityMap: TypedDataFrame[CityMapEntry]) =
-  phoneBook
-    .innerJoin(cityMap).using('address)     :TypedDataFrame[(Address, String, Double, String)]
-    .select('_2, '_4)                       :TypedDataFrame[(String, String)]
-    .as[Family]()                           :TypedDataFrame[Family]
-    .flatMap(f => f.residents.split(' ').map(r => Person(r, f.neighborhood)))
-                                            :TypedDataFrame[Person]
-    .filter(p => NLPLib.uniqueName(p.name)) :TypedDataFrame[Person]
-    .groupBy('neighborhood).count()         :TypedDataFrame[(String, Long)]
-    .as[NeighborhoodCount]()                :TypedDataFrame[NeighborhoodCount]
-    .sortDesc('count)                       :TypedDataFrame[NeighborhoodCount]
-    .select('neighborhood)                  :TypedDataFrame[Tuple1[String]]
-    .head._1
-
-
-
-phoneBook                             : DF[PhoneBookEntry]
-  .innerJoin(cityMap).using('address) : DF[(Address, String, String, ...)]
-  .select('_2, '_3).as[Family]()      : DF[Family]
-  .filter(x => uniqueName(x.name))    : DF[Family]
-  .groupBy('neighborhood).count()     : DF[(String, Long)]
-  .sortDesc('_2)                      : DF[(String, Long)]
-  .head._1                            : String
-
-
-phoneBook                                           : DataFrame
-  .innerJoin(cityMap, col("address"))               : DataFrame
-  .select(col("name"), col("neighborhood"))         : DataFrame
-  .filter(x => uniqueName(x.getAs[String]("name"))) : DataFrame
-  .groupBy(col("neighborhood")).count()             : DataFrame
-  .sortDesc(col("count"))                           : DataFrame
-  .head.getAs[String](0)                            : String
-
-
-
-val captures: Option[Seq[String]] = rational.unapply("3.1415")
-captures match
-  case Some(seq) if seq(1) != null =>
-    var n = seq(0).length + seq(1).length
-    println(s"This number is $n digits long")
-  case _ =>
-    println("Not a number")
-
-val rational = Regex("""(\d+)(?:\.(\d+))?""")
-"3.1415" match
-  case rational(captures: Seq[String]) =>
[…]

### 2022-04-22 article.tex
Address Dragana's comments

@@
 \begin{diff}
-Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representative for abstractions and type abstractions, respectively.
+Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representatives for abstractions and type abstractions, respectively.
 \end{diff}
@@
 \begin{diff}
-This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$ with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
+This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$, with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
 Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
-Although $\Γ \⊢ \S \⇌ \T$ witness mutual subtyping between $\S$ and $\T$, that relation is asymmetric since its left-hand side is always a match type.
+Although $\Γ \⊢ \S \⇌ \T$ witnesses mutual subtyping between $\S$ and $\T$, that relation is asymmetric, since its left-hand side is always a match type.
 \end{diff}
@@
 To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
-This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} in the presence of empty types (disjointness/subtyping exclusivity).
+This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} (disjointness/subtyping exclusivity) in the presence of empty types.
 \end{diff}
@@
 \begin{diff}
-In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotation of higher-kinded types.
+In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotations of higher-kinded types.
 In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.
-
-At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
-However, it would be conceivably to lift this restriction in the future, as shown by the analysis presented in this section.
 
@@
 In \Cref{sec:formalization}, we define the semantics of this type in terms of subtyping rules.
-Given this particular definition of |M|, we can write a direct semantic definition of |M| with equations:
+Given this particular definition of |M|, we can write a direct semantic definition of |M| using the following equations:
 %
@@
 For non-binding cases, that analysis is straightforward and analogous to the variance checks that are already in place in the compiler.
+
+At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
+However, it would be conceivable to lift this restriction in the future, as shown by the analysis presented in this section.
 \end{diff}
@@
 Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
-Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.
+Erasure hence creates an expressivity gap between pattern matching at the term and at the type level.
 
 In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
-For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
+For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning that the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
 \end{diff}
@@
 %
-In the absence of result type annotation, Scala's type inference will first assign \lstinline!String|Int!
+In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int!
 for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.

### 2022-04-22 ack.tex
Remove diff

--- 2022-02-15 ack.tex
@@
 
-I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and for taking part in the committee.
+I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for reviewing my dissertation and for providing valuable feedback.
 

### 2022-04-22 article.tex
Remove diff

@@
 We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
-\begin{diff}
 Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representatives for abstractions and type abstractions, respectively.
-\end{diff}
 Elements of $\P$ can be understood as equivalence classes for \Fm's types.
@@
 Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
-\begin{diff}
 This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$, with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
@@
 Although $\Γ \⊢ \S \⇌ \T$ witnesses mutual subtyping between $\S$ and $\T$, that relation is asymmetric, since its left-hand side is always a match type.
-\end{diff}
 %
@@
 Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
-\begin{diff}
 To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
 This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} (disjointness/subtyping exclusivity) in the presence of empty types.
-\end{diff}
 
@@
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
-\begin{diff}
 There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
-\end{diff}
 
@@
 
-\begin{diff}
 In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotations of higher-kinded types.
@@
 However, it would be conceivable to lift this restriction in the future, as shown by the analysis presented in this section.
-\end{diff}
 
@@
 
-\begin{diff}
 Scala's primary platform is the Java virtual machine (JVM).
@@
 For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning that the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
-\end{diff}
 
@@
 
-\begin{diff}
 In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.
-\end{diff}
 
@@
 
-\begin{diff}
 For example, consider the following function definition:
@@
 \end{lstlisting}
-\end{diff}
 

### 2022-04-22 preprint.tex
Remove diff

@@
 
-\begin{diff}
 A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
 Although the ideas presented in this chapter did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala language.
-\end{diff}
 
@@
 This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
-\begin{diff}
 When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
@@
 When a generalized singleton type contains a component of form \texttt{(\_: T)}, that type is not singleton: it might contains more than one value.}
-\end{diff}
 
@@
 Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
-\begin{diff}
 However, in the case of a |dependent| method, an explicit result type only serves as an upper bound used to type-check the method's body.
-\end{diff}
 At other call sites, the (precise) inferred result type is used.
@@
 However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
-\begin{diff}
 More generally, when a dependently-typed function calls into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
-\end{diff}
 
@@
 
-\begin{diff}
 Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
@@
 Though they introduce a form of dependency, path-dependent types were not designed for type-level computations, rendering their goals largely orthogonal to ours.
-\end{diff}

### 2022-04-22 thesis.tex
Remove diff

--- 2022-04-21 thesis.tex
@@
 In Scala~3, the rules for implicit ambiguities changed to take implicit parameters into account, thus introducing another, more direct, disambiguation mechanism.
-All else being equal, an implicit definition that takes some implicit parameters is considered to be less specific than an implicit definition that takes none~\citep[Changes in Implicit Resolution]{odersky2013scala}.
+All else being equal, an implicit definition that takes implicit parameters is considered less specific than an implicit definition does not take implicit parameters~\citep[Changes in Implicit Resolution]{odersky2013scala}.
 As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).
@@
 \label{chap:match-types}
-
-% TODO: Compare to singleton types
 
@@
 
-% TODO you could write a bit more. Maybe move some of the text of the conclusions of individual chapters there?
-
 In this thesis, we presented three techniques for type-level programming in Scala: implicits, generalized singleton types, and match types.

### 2022-04-22 article.tex
Spacing
Typos

@@
 In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
-Generalized singletons and match types share some similarities, such as the notion of disjointness which plays an important role in both features.
-The two features relate on deeper level: we can explain match types in terms of generalized singletons.
-For example, the we can give an alternative definition of |Elem| match type using a generalized singleton:
+Generalized singletons and match types share some similarities, such as the notion of disjointness, which plays an important role in both features.
+The two features relate on a deeper level: we can explain match types in terms of generalized singletons.
+For example, we can give an alternative definition of the |Elem| match type using a generalized singleton:
 %
@@
 %
-This definition makes the term/type correspondence evident: the |Elem[X]| type reduces in the way than the corresponding match expression.
+This definition makes the term/type correspondence evident: the |Elem[X]| type reduces analogously to the corresponding match expression.
 
@@
 \end{lstlisting}%
-\vspace{-3pt}%
+\vspace{-1pt}%
 \begin{lstlisting}
@@
 \begin{minipage}[t]{.52\linewidth}%
+\vspace{-12pt}%
 \hfuzz=3pt

### 2022-04-22 scala/regex/src/main/scala/bench/RegexPackage.scala
Spacing

@@ -31,17 +31,15 @@ abstract class Sanitizer[T](val i: Int):
   def mutate(arr: Array[Any]): Unit
 
 object Sanitizer:
-  implicit val basecase: Sanitizer[EmptyTuple] =
+  implicit val basecase =
     new Sanitizer[EmptyTuple](0):
       def mutate(arr: Array[Any]): Unit = ()
-
-  implicit def stringcase[T <: Tuple](implicit ev: Sanitizer[T]): Sanitizer[String *: T] =
+  implicit def stringcase[T <: Tuple](implicit ev: Sanitizer[T]) =
       new Sanitizer[String *: T](ev.i+1):
         def mutate(arr: Array[Any]): Unit =
           assert(arr(arr.size-i) != null)
           ev.mutate(arr)
-
-  implicit def optioncase[T <: Tuple](implicit ev: Sanitizer[T]): Sanitizer[Option[String] *: T] =
+  implicit def optioncase[T <: Tuple](implicit ev: Sanitizer[T]) =
       new Sanitizer[Option[String] *: T](ev.i+1):
         def mutate(arr: Array[Any]): Unit =
           arr(arr.size-i) = Option(arr(arr.size-i))

### 2022-04-22 scala/src/main/scala/reduce-match-types.scala

@@ -60,8 +60,7 @@ type Loop[S <: Shape, Axes <: Shape, I <: Int] <: Shape =
       case true => Loop[tail, Remove[Axes, I], I + 1]
       case false => head #: Loop[tail, Axes, I + 1]
     case Ø => Axes match
-      case Ø => Ø
-      // otherwise, do not reduce further
+      case Ø => Ø // otherwise, do not reduce further
 // end section reduceLoop
 
 type Contains[Haystack <: Shape, Needle <: Int] <: Boolean = Haystack match {

### 2022-04-22 scala/src/main/scala/safe-implicits.scala

@@ -187,8 +187,7 @@ addEventListener("mouseover", myHandler) // implicitly
 // start section addEventListenerImplicitCall
 addEventListener("mouseover", myHandler)(
   // Evidence that "mouseover" is a valid event type, infered automatically.
-  Remove.casetail(Remove.casetail(Remove.casehead))
-)
+  Remove.casetail(Remove.casetail(Remove.casehead)))
 // end section addEventListenerImplicitCall
 
 }

### 2022-04-23 ack.tex
Revert "Remove diff"
This reverts commit baca8b24302dc02b00fcc2f8fee3e4173838d1a4.

@@
 
+\begin{diff}
 I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for reviewing my dissertation and for providing valuable feedback.
+\end{diff}
 

### 2022-04-23 article.tex
More diff

--- 2022-04-22 article.tex
@@
 
+\subsection{Comparison to Generalized Singleton Types}
+\begin{diff}
+In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
+Generalized singletons and match types share some similarities, such as the notion of disjointness, which plays an important role in both features.
+The two features relate on a deeper level: we can explain match types in terms of generalized singletons.
+For example, we can give an alternative definition of the |Elem| match type using a generalized singleton:
+%
+\begin{lstlisting}
+type Elem[X] = { (_: X) match
+  case _: String => (_: Char)
+  case _: List[t] => (_: Elem[t])
+  case _: Any => (_: X) }
+\end{lstlisting}
+%
+This definition makes the term/type correspondence evident: the |Elem[X]| type reduces analogously to the corresponding match expression.
+\end{diff}
+
 \FMSyntaxEvaluation{\SystemFm syntax and evaluation rules for a given set of classes $\C$ with class inheritance $\Ψ$.
@@
 \hl{\text{Highlights}} correspond to additions to \SystemFsub, as per \citep[Figure 26-1]{pierce2002types}.}
-
-\subsection{Comparison to Generalized Singleton Types}
-
-In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
-Generalized singletons and match types share some similarities, such as the notion of disjointness which plays an important role in both features.
-The two features relate on deeper level: we can explain match types in terms of generalized singletons.
-For example, the we can give an alternative definition of |Elem| match type using a generalized singleton:
-%
-\begin{lstlisting}
-type Elem[X] = { (_: X) match
-  case _: String => (_: Char)
-  case _: List[t] => (_: Elem[t])
-  case _: Any => (_: X)
-}
-\end{lstlisting}
-%
-This definition makes the term/type correspondence evident: the |Elem[X]| type reduces in the way than the corresponding match expression.
 
@@
 Scala tuples, for example, fall into this category.
-|Tuple2|, the class for pairs, is defined as follows:
-%
-\begin{lstlisting}
-case class Tuple2[+T1, +T2](_1: T1, _2: T2)
-\end{lstlisting}
-%
+|Tuple2|, the class for pairs, is defined as follows: |case class Tuple2[+T1, +T2](_1: T1, _2: T2)|.
+\begin{diff}
 Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
-\begin{diff}
 There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
@@
 For example, consider the following function definition:
-%
+
+\begin{minipage}{\linewidth}
 \begin{lstlisting}
@@
 \end{lstlisting}
+\end{minipage}
 %
@@
 When |Loop| reaches the end of the list, if there are still axes to remove, these are out of bounds for the initial shape: the match type intentionally gets stuck in such cases.
-%
+
+\begin{minipage}{\linewidth}
 \reduceType
 \reduceLoop
+\end{minipage}
 

### 2022-04-23 symposium.tex
More diff

--- 2022-04-19 symposium.tex
@@
 
+\begin{diff}
 Capturing groups allow programmers to extract substrings matched by parts of a regular expression.
@@
 In the first case, the capturing group is empty, and in the second case, the capturing group contains |"B"|.
+\end{diff}
 
@@
 This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
+\begin{diff}
 The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
 The package's documentation starts with the following example:
+\end{diff}
 
@@
 
+\begin{diff}
 This \paper is structured as follows.
@@
 We conclude the \paper in \Cref{sec:conclusion}.
+\end{diff}
 
@@
 
+\begin{diff}
 \section{Background}
@@
 \tupleReverseB
+\end{diff}
 
@@
 
+\begin{diff}
 We present our library for type-safe regular expressions in terms of 3 components:
+\end{diff}
 
@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+\begin{diff}
+An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
 
@@
 Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.
+\end{diff}
 

### 2022-04-23 thesis.tex
More diff

@@
 The two implicit definitions, |casehead| and |casetail|, correspond to the base case and the recursive case of the list removal operation, respectively.
+\begin{diff}
 The right-hand side of those definitions is devoid of meaning, instances of the |Remove| trait merely act as placeholders (the type parameters of those constructor calls are inferred from their expected type).
+\end{diff}
 The |Remove| trait is intended to be used as an implicit parameter, with constrained input types, and an unconstrained output type.

### 2022-05-10 article.tex
Revert "More diff"
This reverts commit 0cde87495c11c430f41f6e19a3c53dc7d20732d7.

@@
 \subsection{Comparison to Generalized Singleton Types}
-\begin{diff}
+
 In \Cref{chap:generalizing-singleton-types}, we presented a generalization of singleton types for type-level programing.
@@
 This definition makes the term/type correspondence evident: the |Elem[X]| type reduces analogously to the corresponding match expression.
-\end{diff}
 
@@
 |Tuple2|, the class for pairs, is defined as follows: |case class Tuple2[+T1, +T2](_1: T1, _2: T2)|.
-\begin{diff}
 Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
 Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
+\begin{diff}
 There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.

### 2022-05-10 symposium.tex
Revert "More diff"
This reverts commit 0cde87495c11c430f41f6e19a3c53dc7d20732d7.

--- 2022-04-19 symposium.tex
@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+\begin{diff}
+An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+\end{diff}
 

### 2022-05-10 symposium.tex
Revert "Revert "Remove diff""
This reverts commit 0c51f09a5f4e75f93b96f298c1be08460983d13a.

--- 2022-04-22 symposium.tex
@@
 The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-\begin{diff}
-An important conceptual difference between both approaches is ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
-\end{diff}
+An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
 

### 2022-05-10 symposium.tex
Incorporate symposium's feedback

@@
-\def\paper{\thesisOnly{chapter}\paperOnly{paper}}
+\def\paper{\thesisOnly{chapter}\paperOnly{paper}\xspace}
+
 \section{Introduction}
@@
 
-This \paper is structured as follows.
+This \begin{diff}\paper is\end{diff} structured as follows.
 In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
@@
 In \Cref{sec:related}, we discuss related work.
-We conclude the \paper in \Cref{sec:conclusion}.
+We conclude the \begin{diff}\paper in\end{diff} \Cref{sec:conclusion}.
 
@@
 
-  \item The runtime capturing group processing component, which extracts and sanitizes the output of the regex engine, in accordance with the previously computed type-level representation.
+  \item The runtime capturing group processing component, which extracts and sanitizes the output of the \begin{diff}JVM's\end{diff} regex engine, in accordance with the previously computed type-level representation.
   This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.
@@
 Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string (the |Singleton| bound is necessary to guide Scala's type inference).
+% TODO: could you explain in more detail why "the Singleton bound is necessary to guide Scala’s type inference"?
 We use that type-level string to instantiate the |Compile| type, which is our type-level analysis component.
@@
 Furthermore, the type of those groups corresponds to the type of the tuple elements.
+% TODO L196: "Furthermore, the type of those groups corresponds to the type of the tuple elements." This part is a bit hard to understand at a first reading, because only later on L236, the reader learns that matched values can only be of type String or Option String, and there's a singular/plural mismatch ("the type" is singular but applies to each of "those groups").
+It might help to rephrase this sentence and already mention "String or Option String".
 This allows us to enrich the results of Java's regex engine by wrapping nullable values into options.
@@
 Rest assured, this is not an oversight!
-Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error").
-Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
+Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
+\begin{diff}
+Even though systematic detecting of invalid regular expressions is out of the scope of our library, doing so when possible is still desirable.
+\end{diff}
 If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
+\begin{diff}
+At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
+% TODO
+\end{diff}
 
@@
 
+% TODO: How understandable are the compiler's error messages if the programmer makes a regex-related mistake?
+In particular, what's the error message if the programmer provides an incorrect number of arguments to an extractor?
+What's the error message in case of unbalanced parentheses in the regex?
+What about other forms of invalid regexes?
+Could you list the kinds of regex-library-related error messages to be expected?
+
 \section{Term-Level}
@@
 In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
-For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)|.
+For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)| \begin{diff}expression\end{diff}.
 
@@
 Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
-The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
-An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+The library provides safety guarantees similar to ours, but from the slightly different \begin{diff}angle of named-capturing groups.
+An important conceptual\end{diff} difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
 
@@
 Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.
+
+% Another type-safe regex library that is not discussed in related work is Tyre: https://github.com/Drup/tyre
+% TODO \citep{radanne2020tyre}
 

### 2022-05-11 symposium.tex
Explain the Singleton bound in more details

@@
 It takes a regular expression and returns an instance of |Regex|.
-Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string (the |Singleton| bound is necessary to guide Scala's type inference).
-% TODO: could you explain in more detail why "the Singleton bound is necessary to guide Scala’s type inference"?
+Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string\footnote{
+\begin{diff}
+The \texttt{Singleton} bound is a marker that instructs type inference to preserve unions and literal singleton types.
+By default, the compiler widens those types, notably in method applications.
+For instance, given \texttt{1} \texttt{::} \texttt{Nil}, Scala's default type inference yield \texttt{List[Int]} instead of the more precise \texttt{List[1]}.
+To alter this behavior, one could mark \texttt{::}'s type argument with the \texttt{Singleton} bound.
+\end{diff}
+}.
 We use that type-level string to instantiate the |Compile| type, which is our type-level analysis component.
@@
 For simplicity, our presentation treats all non-escaped parentheses as capturing group delimiters.
-Our implementation also takes \lstinline!\\Q...\\E! quotations into account and ignores parentheses that appear in character classes.
+Our implementation also takes \texttt{\textbackslash{}Q...\textbackslash{}E} quotations into account and ignores parentheses that appear in character classes.
 }:

### 2022-05-11 thesis.tex
Explain the Singleton bound in more details

--- 2022-04-22 thesis.tex
@@
 \noindent
-The |Singleton| type bound is a marker that instructs type inference to preserve union types and literal singleton types (by default, the compiler widens those types).
+The |Singleton| type bound is a marker that instructs type inference to preserve unions and literal singleton types (by default, the compiler widens those types).
 The "|?|" type is Scala~3's new syntax for wildcard types~\citep[Wildcard Arguments in Types]{odersky2013scala}.

### 2022-05-11 symposium.tex
Rephrase,"the type of those groups..."

@@
 In our implementation, we instantiate the type |P| to a tuple and use that type in the result type of |unapply| to tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity.
-Furthermore, the type of those groups corresponds to the type of the tuple elements.
-% TODO L196: "Furthermore, the type of those groups corresponds to the type of the tuple elements." This part is a bit hard to understand at a first reading, because only later on L236, the reader learns that matched values can only be of type String or Option String, and there's a singular/plural mismatch ("the type" is singular but applies to each of "those groups").
-It might help to rephrase this sentence and already mention "String or Option String".
-This allows us to enrich the results of Java's regex engine by wrapping nullable values into options.
-Our implementation performs this wrapping via the |transform| method, which is our runtime processing component.
+\begin{diff}
+The |P| type is composed of |String| and |Option[String]| types, depending on the nullability of the correspond capturing group.
+Our implementation enriches the results of Java's regex engine by wrapping nullable values into options, and packaging the overall result into a tuple of type P.
+We perform this operation in the |transform| method, which is the runtime component of our library.
+\end{diff}
 As a result, our implementation never returns null values and removes the null-checking burden from the user.

### 2022-05-12 symposium.tex
Cite Tyre

@@
 The Scala standard library contains a package to manipulate regular expressions.
-This package is based on Java's implementation, and thus benefits from the high performance of the JVM's regex engine.
+This package is based on Java's implementation, and thus benefits from the high performance of \begin{diff}Java's\end{diff} regex engine.
 The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
@@
 
-In this \paper, we propose a new design for Scala's regular expression library which provides a type-safe and null-safe mechanism for capturing group extraction.
+In this \paper, we propose a new design for Scala's regular expression library, which provides a type-safe and null-safe mechanism for capturing group extraction.
 Our design makes extensive use of match types\paperOnly{~\citep{blanvillain2022type}, Scala~3's new feature for type-level programming,} to statically analyze regular expressions during type checking.
-We build our interface to mimic Scala's original regular expression API so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
+We build our interface to mimic Scala's original regular expression API, so that Scala programmers can use it as a drop-in replacement and enjoy the additional safety with minimal migration costs.
 
@@
 
-\thesisOnly{
-The source code of our implementation is available online\footnote{\texttt{https://github.com/OlivierBlanvillain/regsafe}} under the MIT licence.
-}
-\paperOnly{
-The source code of our implementation is available online\footnote{This URL is hidden for the purpose of the double-blind review process.} under the MIT licence.
-}
+The source code of our implementation is available online\footnote{\url{https://github.com/OlivierBlanvillain/regsafe}}, under the MIT license.
 
@@
 
-  \item The runtime capturing group processing component, which extracts and sanitizes the output of the \begin{diff}JVM's\end{diff} regex engine, in accordance with the previously computed type-level representation.
+  \item The runtime capturing group processing component, which extracts and sanitizes the output of \begin{diff}Java's\end{diff} regex engine, in accordance with the previously computed type-level representation.
   This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.
@@
 At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
-% TODO
 \end{diff}
@@
 In its final iteration, our capturing group analysis is, to the best of our knowledge, on par with Java's implementation of regular expression.
-
-% TODO: How understandable are the compiler's error messages if the programmer makes a regex-related mistake?
-In particular, what's the error message if the programmer provides an incorrect number of arguments to an extractor?
-What's the error message in case of unbalanced parentheses in the regex?
-What about other forms of invalid regexes?
-Could you list the kinds of regex-library-related error messages to be expected?
 
@@
 The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
+\begin{diff}
+Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
+\end{diff}
 
@@
 Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.
-
-% Another type-safe regex library that is not discussed in related work is Tyre: https://github.com/Drup/tyre
-% TODO \citep{radanne2020tyre}
 

### 2022-05-14 symposium.tex
Minor

@@
 \noindent
-The extractor pattern, |case date(y,m,d)|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.
+The extractor pattern, |case| |date(y,m,d)|, replaces the need for manually indexing into the regular expression's capturing groups, and shields users from off-by-one error.
 

### 2022-05-15 symposium.tex
Incorporate Dragana's feedback

@@
 It takes a regular expression and returns an instance of |Regex|.
-Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's string\footnote{
+Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's |String|\footnote{
 \begin{diff}
@@
 By default, the compiler widens those types, notably in method applications.
-For instance, given \texttt{1} \texttt{::} \texttt{Nil}, Scala's default type inference yield \texttt{List[Int]} instead of the more precise \texttt{List[1]}.
+For instance, given \texttt{1} \texttt{::} \texttt{Nil}, Scala's default type inference yields \texttt{List[Int]} instead of the more precise \texttt{List[1]}.
 To alter this behavior, one could mark \texttt{::}'s type argument with the \texttt{Singleton} bound.
@@
 Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error", on the JVM).
-\begin{diff}
-Even though systematic detecting of invalid regular expressions is out of the scope of our library, doing so when possible is still desirable.
-\end{diff}
 If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
@@
 At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
+The systematic detection of invalid regular expressions is out of the scope of this paper.
 \end{diff}
@@
 The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
-The main drawback of this solution lies in the library's learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
 \begin{diff}
 Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
+The main drawback of these solutions is the learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
 \end{diff}

### 2022-05-15 symposium.tex
Remove diff

@@
 The Scala standard library contains a package to manipulate regular expressions.
-This package is based on Java's implementation, and thus benefits from the high performance of \begin{diff}Java's\end{diff} regex engine.
+This package is based on Java's implementation, and thus benefits from the high performance of Java's regex engine.
 The Scala implementation innovates in its presentation: it improves upon Java's solution by providing an ergonomic API based on pattern matching that is both elegant and concise.
@@
 
-This \begin{diff}\paper is\end{diff} structured as follows.
+This \paper is structured as follows.
 In \Cref{sec:background}, we give an introduction to match types and generic tuples, two recent additions to Scala's type system which we rely on in our implementation.
@@
 In \Cref{sec:related}, we discuss related work.
-We conclude the \begin{diff}\paper in\end{diff} \Cref{sec:conclusion}.
+We conclude the \paper in \Cref{sec:conclusion}.
 
@@
 
-  \item The runtime capturing group processing component, which extracts and sanitizes the output of \begin{diff}Java's\end{diff} regex engine, in accordance with the previously computed type-level representation.
+  \item The runtime capturing group processing component, which extracts and sanitizes the output of Java's regex engine, in accordance with the previously computed type-level representation.
   This component takes the form of a function called |transform|, which we present in \Cref{sec:term-level}.
@@
 Instead of directly accepting a |String| argument, that method takes a type parameter, |R|, which will be inferred by the compiler, and allows us to get our hands on a type-level instance of the regular expression's |String|\footnote{
-\begin{diff}
 The \texttt{Singleton} bound is a marker that instructs type inference to preserve unions and literal singleton types.
@@
 To alter this behavior, one could mark \texttt{::}'s type argument with the \texttt{Singleton} bound.
-\end{diff}
 }.
@@
 In our implementation, we instantiate the type |P| to a tuple and use that type in the result type of |unapply| to tell the compiler that a successful matching consists of exactly $n$ capturing groups, where $n$ is the tuple's arity.
-\begin{diff}
 The |P| type is composed of |String| and |Option[String]| types, depending on the nullability of the correspond capturing group.
@@
 We perform this operation in the |transform| method, which is the runtime component of our library.
-\end{diff}
 As a result, our implementation never returns null values and removes the null-checking burden from the user.
@@
 If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.
-\begin{diff}
 At the time of writing, the Scala~3 compiler does not support the customization of match type errors.
 The systematic detection of invalid regular expressions is out of the scope of this paper.
-\end{diff}
 
@@
 In practice, at the time of writing, this solution does not play well with predefined types from the standard library: the compiler lacks the knowledge necessary to correlate predefined types with their term-level counterpart.
-For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)| \begin{diff}expression\end{diff}.
+For instance, the compiler does not recognize |CharAt["hello", 0]| as a valid type for the |"hello".charAt(0)| expression.
 
@@
 The regex-applicative package stands on the safe side of that spectrum: it allows programmers to write regular expressions using a parser combinator library, which entirely removes the need to reference capturing groups \citep{cheplyaka2011regex}.
-\begin{diff}
 Tyre~\citep{radanne2020tyre} is an OCaml library that also uses combinators to define type-safe regular expressions.
 The main drawback of these solutions is the learning curve: programmers already familiar with POSIX regular expressions~\citep{ieee2018the} are forced to learn a new syntax, which is arguably more complicated.
-\end{diff}
 
 Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
-The library provides safety guarantees similar to ours, but from the slightly different \begin{diff}angle of named-capturing groups.
-An important conceptual\end{diff} difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
+The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
+An important conceptual difference between the two approaches is that ours only focuses on type checking (we rely on Java's regular expressions at run-time), whereas hers also includes a fully-fledged regular expression engine.
 

### 2022-05-17 public.tex
update

\documentclass[10pt]{beamer}

\usepackage{fontspec}

%% Minted setup (syntax highlighting)
\usepackage[outputdir=latex.out]{minted}
\usemintedstyle{vs}
\setminted{fontsize=\footnotesize}
\newminted[code]{scala}{gobble=0, escapeinside=@@}
\newenvironment{slide}[2][]
  {\begin{frame}[fragile,environment=slide,#1]{#2}}
  {\end{frame}}
\newcommand\makemintedshortinline[2]{%
  \catcode`#2=13
  \begingroup%
  \lccode`\~=`#2
  \lowercase{\endgroup\protected\def~{\mintinline[escapeinside=@@]{#1}~}}%
}
\renewcommand{\fcolorbox}[4][]{#4}

%% Beamer theme setup
\usetheme[block=fill]{metropolis}
\setbeamercolor{progress bar}{bg=black, fg=black}
\setbeamercolor{title separator}{bg=black, fg=black}
\setbeamercolor{progress bar in head/foot}{bg=black, fg=black}
\setbeamercolor{progress bar in section page}{bg=black, fg=black}
\setbeamercovered{transparent}

\usepackage[sorting=none, backend=biber, maxbibnames=9]{biblatex}
% \settowidth\labelnumberwidth{\usebeamertemplate*{bibliography item}}
\AtEveryBibitem{\clearfield{note}}
\AtEveryBibitem{\clearfield{volume}}
\AtBeginBibliography{\small}
\bibliography{bibliography}

\usepackage{textpos}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{ulem}

\usepackage[autostyle=false, style=english]{csquotes}
\MakeOuterQuote{"}

\usepackage{xcolor}
\definecolor{typeColor}{HTML}{2B91B0}
\definecolor{implColor}{HTML}{57B5E8}
\definecolor{singColor}{HTML}{009E73}
\definecolor{mtpeColor}{HTML}{9400D4}
\definecolor{FAFAFA}{HTML}{FAFAFA}
\newcommand{\tp}[1]{\color{typeColor}#1}

\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,positioning, arrows.meta}

\def\checkmark{\textcolor{green}{\tikz\fill[scale=0.6](0,.35) -- (.25,0) -- (1,.7) -- (.25,.15) -- cycle;}}
\def\cross{$\mathbin{\tikz [x=1.4ex,y=1.4ex,line width=.2ex, red] \draw (0,0) -- (1,1) (0,1) -- (1,0);}$}

\title{Abstractions for Type-Level Programming}
\date{Wednesday, 25 May 2022}
\author{Olivier Blanvillain}

\makemintedshortinline{scala}\|

\newcommand{\timeline}[5] {
  \begin{tikzpicture}
    \draw[lightgray!#5!black, thick, -Triangle] (0,0) -- (10cm,0);
    \foreach \x in {0,1,...,6}
    \draw[lightgray!#5!black] (\x*1.4 cm,3pt) -- (\x*1.4 cm,-3pt);

    \foreach \x/\descr in {0/2016,1/2017,2/2018,3/2019,4/2020,5/2021,6/2022}
    \node[lightgray!#5!black, font=\scriptsize, text height=1.75ex,
    text depth=.5ex] at (\x*1.4,-.3) {$\descr$};

    \fill[lightgray!#1!implColor] (0.25, 0.5) circle (4pt) node[lightgray!#1!black,align=center,above=7pt,font=\small]{DataFrame\\Library};

    \draw[lightgray!#2!singColor, line width=4pt]
    (2.25*1.4,.5) -- (4.417*1.4,.5) node[lightgray!#2!black,midway,above=5pt,font=\small]{Generalized Singletons};

    \draw[lightgray!#3!mtpeColor, line width=4pt]
    (2.5*1.4,-.7) -- (6.15*1.4,-.7) node[lightgray!#3!black,midway,below=5pt,font=\small]{Match Types};
    \draw[thick,lightgray!#3!mtpeColor,-Triangle,line width=1.2pt] (6.15*1.4,-.7) -- (6.25*1.4,-.7);

    \fill[lightgray!#4!mtpeColor] (6.1*1.4, 0.5) circle (4pt) node[lightgray!#4!black,align=center,above=7pt,font=\small]{Regex\\Library};
  \end{tikzpicture}
}

\begin{document}

\maketitle

\begin{slide}{Welcome!}
\Large
\begin{center}
Bienvenue à ma déf\textcolor{typeColor}{ense} de thèse !\\
\phantom{Bienvenue à ma} déf\textcolor{implColor}{anse} \phantom{de thèse !}\\
\phantom{Bienvenue à ma} déf\textcolor{singColor}{ence} \phantom{de thèse !}\\
\phantom{Bienvenue à ma} déf\textcolor{mtpeColor}{ance} \phantom{de thèse !}\\
\end{center}
\end{slide}

\begin{slide}{Welcome!}
\Large
\begin{center}
Bienvenue à ma souten\textcolor{typeColor}{ense} de thèse !\\
\phantom{Bienvenue à ma} souten\textcolor{implColor}{anse} \phantom{de thèse !}\\
\phantom{Bienvenue à ma} souten\textcolor{singColor}{ence} \phantom{de thèse !}\\
\phantom{Bienvenue à ma} souten\textcolor{mtpeColor}{ance} \phantom{de thèse !}\\
\end{center}
\end{slide}

\begin{slide}{Spell checkers to the rescue}
\begin{center}
\includegraphics[width=\textwidth]{figures/scribens.png}
\end{center}
\end{slide}

\section{Programming Methods Laboratory\hspace{-2pt}}

\begin{slide}{An analogy...}
\begin{center}
\Large
Programming $\longleftrightarrow$ Poetry
\vspace{10pt}

\includegraphics[width=0.5\textwidth]{figures/poetry.jpg}
\end{center}
\end{slide}

\begin{slide}{An analogy...}
\begin{center}
\Large
Scala $\longleftrightarrow$ La Langue Française
\vspace{10pt}

\includegraphics[width=0.6\textwidth]{figures/fr.jpg}
\end{center}
\end{slide}

\begin{slide}{An analogy...}
\begin{center}
\Large
Compiler $\longleftrightarrow$ Spell checker
\vspace{10pt}

\includegraphics[width=\textwidth]{figures/nil.png}
\end{center}
\end{slide}

\begin{slide}{An analogy...}
\begin{center}
\begin{minipage}{0.45\textwidth}
\Large
Standard library $\longleftrightarrow$
\end{minipage}
\begin{minipage}{0.45\textwidth}
\vspace{10pt}
\includegraphics[width=\textwidth]{figures/larousse.jpg}
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{An analogy...}
\begin{center}
\begin{minipage}{0.37\textwidth}
\Large
Type system $\longleftrightarrow$
\end{minipage}
\begin{minipage}{0.45\textwidth}
\vspace{10pt}
\includegraphics[width=\textwidth]{figures/bescherelle.jpg}
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{LAMP -- Poetry Methods Laboratory}
\large
\begin{itemize}
  \item Scala $\longleftrightarrow$ La Langue Française
  \item Compiler $\longleftrightarrow$ Spell checker
  \item Standard library $\longleftrightarrow$ Le Petit Larousse Illustré
  \item Type system $\longleftrightarrow$ Bescherelle (rêgles de grammaires)
\end{itemize}
\end{slide}

\section{My Contribution}

\begin{slide}{My contribution}
\Large
My contribution:

A new Scala feature for type-level programming.

I will give you an intuition using examples:

\begin{itemize}
  \item A poetry example
  \item A math example
  \item A programming example
\end{itemize}
\end{slide}

\begin{slide}{Type-level programming: poetry example}
\Large
What do I mean by Type-Level Programming?

A way to \emph{customize} the spell checker.

\begin{center}
\hspace{20pt}
\begin{minipage}{0.45\textwidth}%
\phantom{-- en alexandrin}\\
\phantom{Roses are red,}\\
\phantom{violets are blue,}\\
\phantom{so are you.}
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: poetry example}
\Large
What do I mean by Type-Level Programming?

A way to \emph{customize} the spell checker.

\begin{center}
\hspace{20pt}
\begin{minipage}{0.45\textwidth}%
\phantom{-- en alexandrin}\\
Roses are red,\\
violets are blue,\\
so are you.
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: poetry example}
\Large
What do I mean by Type-Level Programming?

A way to \emph{customize} the spell checker.

\begin{center}
\hspace{20pt}
\begin{minipage}{0.45\textwidth}%
\textcolor{blue}{"En alexandrin"}\\
Roses are red,\\
violets are blue,\\
so are you. \textcolor{red}{\cross}
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: poetry example}
\Large
What do I mean by Type-Level Programming?

A way to \emph{customize} the spell checker.

\begin{center}
\hspace{20pt}
\begin{minipage}{0.45\textwidth}%
\textcolor{blue}{"En alexandrin"}\\
Roses are red,\\
violets are blue,\\
and so are you. \checkmark
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
How many ways are there to pick 13 cards from a 52-card deck?
\begin{center}
\begin{minipage}{0.6\textwidth}
\includegraphics[width=\textwidth]{figures/cards.png}
\end{minipage}
\begin{minipage}{0.2\textwidth}
$$
\phantom{binom{52}{13}}
$$
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
How many ways are there to pick 13 cards from a 52-card deck?
\begin{center}
\begin{minipage}{0.6\textwidth}
\includegraphics[width=\textwidth]{figures/cards.png}
\end{minipage}
\begin{minipage}{0.2\textwidth}
$$
\binom{52}{13}
$$
\end{minipage}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
The binomial coefficients are defined as follows:

$$
\binom{\textcolor{typeColor}{n}}{\textcolor{mtpeColor}{k}} = \frac{\textcolor{typeColor}{n}!}{\textcolor{mtpeColor}{k}!(\textcolor{typeColor}{n} - \textcolor{mtpeColor}{k})!}
$$
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
The $\binom{\textcolor{typeColor}{n}}{\textcolor{mtpeColor}{k}}$ symbol usually read as:

\begin{itemize}
  \item "\textcolor{typeColor}{n} choose \textcolor{mtpeColor}{k}" in English
  \item "\textcolor{mtpeColor}{k} parmis \textcolor{typeColor}{n}" in French
\end{itemize}
\pause

Programmers might confuse \textcolor{typeColor}{n} \& \textcolor{mtpeColor}{k}? Should it be

\begin{itemize}
\item binomialCoefﬁcient(\textcolor{typeColor}{n}, \textcolor{mtpeColor}{k}) or
\item binomialCoefﬁcient(\textcolor{mtpeColor}{k}, \textcolor{typeColor}{n})?
\end{itemize}
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
\textbf{def} binomialCoefﬁcient(\textcolor{typeColor}{n}: Int, \textcolor{mtpeColor}{k}: Int): Int\\
\phantom{def }

binomialCoefﬁcient(52, 13) \checkmark

binomialCoefﬁcient(13, 52) \checkmark
\end{slide}

\begin{slide}{Type-level programming: math example}
\Large
\textbf{def} binomialCoefﬁcient(\textcolor{typeColor}{n}: Int, \textcolor{mtpeColor}{k}: Int)\\
\phantom{def }(\textbf{implicit} ev: \textcolor{typeColor}{n}.type >= \textcolor{mtpeColor}{k}.type \raisebox{-.06\baselineskip}{=}:\raisebox{-.06\baselineskip}{=} true): Int

binomialCoefﬁcient(52, 13) \checkmark

binomialCoefﬁcient(13, 52) \cross
\end{slide}

\begin{slide}{Type-level programming: regular expression example}
\Large
A regular expression is a simple tool to validate and extract data from text.
\vspace{10pt}

\includegraphics[width=\textwidth]{figures/cff.png}
\end{slide}

\begin{slide}{Type-level programming: regular expression example}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}llrrr@{}}
\toprule
|input| & |pattern| & |out[0]| & |out[1]| & |out[2]| \\
\midrule
|"17:00"| & |"(\d+)[.:h](\d+)"| & |"17"| & |"00"|  & \\
\pause
|"25.05.2022"| & |"(\d+).(\d+).(\d+)"| & |"25"| & |"05"| & |"2022"| \\
\bottomrule
\end{tabular}
\end{center}
\end{slide}

\begin{slide}{Type-level programming: regular expression example}
\begin{center}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}llrrr@{}}
\toprule
|input| & |pattern| & |out[0]| & |out[1]| & |out[2]| \\
\midrule
|"17:00"| & |"(\d+)[.:h](\d+)"| & |"17"| & |"00"|  & \cross \\
|"25.05.2022"| & |"(\d+).(\d+).(\d+)"| & |"25"| & |"05"| & |"2022"| \\
\bottomrule
\end{tabular}
\end{center}
\end{slide}

\begin{slide}{Timeline of my PhD}
\timeline{0}{0}{0}{0}{0}
\end{slide}

\begin{slide}{Conclusion}
\Large
In this talk, I presented an analogy to explain what we do in the \sout{Programming} Poetry Methods Laboratory.

I also presented a few example to illustate my contribution:

\begin{center}
  \includegraphics[width=0.25\textwidth]{figures/poetry.jpg}
  \hspace{10pt}
  \includegraphics[width=0.32\textwidth]{figures/cards.png}
  \hspace{10pt}
  \includegraphics[width=0.25\textwidth]{figures/cff-logo.jpg}
\end{center}

\end{slide}

\begin{slide}{Thank you!}
\begin{center}
\Huge Thank you!
\end{center}
\end{slide}

\end{document}

### 2022-05-27 public.tex
Add pictures
update slides
Fix typo

@@
 \usepackage{booktabs}
-\usepackage{ulem}
 
@@
 \begin{minipage}{0.45\textwidth}%
+\phantom{-- en alexandrin}\\
+Roses are red,\\
+violets are blue,\\
+so are you. \checkmark
+\end{minipage}
+\end{center}
+\end{slide}
+
+\begin{slide}{Type-level programming: poetry example}
+\Large
+What do I mean by Type-Level Programming?
+
+A way to \emph{customize} the spell checker.
+
+\begin{center}
+\hspace{20pt}
+\begin{minipage}{0.45\textwidth}%
 \textcolor{blue}{"En alexandrin"}\\
@@
 \end{tabular}
+
+\phantom{\includegraphics[width=0.92\textwidth]{figures/error2.png}}
 \end{center}
@@
 \end{tabular}
+
+\phantom{\includegraphics[width=0.92\textwidth]{figures/error2.png}}
+\end{center}
+\end{slide}
+
+\begin{slide}{Type-level programming: regular expression example}
+\begin{center}
+\renewcommand{\arraystretch}{1.2}
+\begin{tabular}{@{}llrrr@{}}
+\toprule
+|input| & |pattern| & |out[0]| & |out[1]| & |out[2]| \\
+\midrule
+|"17:00"| & |"(\d+)[.:h](\d+)"| & |"17"| & |"00"|  & \cross \\
+|"25.05.2022"| & |"(\d+).(\d+).(\d+)"| & |"25"| & |"05"| & |"2022"| \\
+\bottomrule
+\end{tabular}
+
+\includegraphics[width=0.92\textwidth]{figures/error2.png}
 \end{center}
@@
 \Large
-In this talk, I presented an analogy to explain what we do in the \sout{Programming} Poetry Methods Laboratory.
+In this talk, I presented an analogy to explain what we do in the Poetry Methods Laboratory.
 
-I also presented a few example to illustate my contribution:
+I also presented a few examples to illustate my contribution:
 
@@
 \end{center}
+\end{slide}
 
+\begin{slide}{Thank you!}
+\begin{center}
+\Huge Thank you!
+\end{center}
+\end{slide}
+
+\section{Acknowledgments}
+
+\begin{slide}{Martin}
+\begin{center}
+\includegraphics[width=0.8\textwidth]{figures/martin.jpg}
+\end{center}
+\end{slide}
+
+\begin{slide}{LAMP, LARA \& Scala Center}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/ski-trip.jpg}
+\end{center}
+\end{slide}
+
+\begin{slide}{Georg \& Mia}
+\begin{center}
+\includegraphics[width=\textwidth]{figures/bike-trip.jpg}
+\end{center}
+\end{slide}
+
+\begin{slide}{Alicia \& Leila}
+\begin{center}
+\includegraphics[width=0.7\textwidth]{figures/alicia-2.png}
+\end{center}
+\end{slide}
+
+\begin{slide}{Sol, Christian \& Leila}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{figures/parents.jpg}
+\end{center}
+\end{slide}
+
+\begin{slide}{Dragana \& Leila}
+\begin{center}
+\includegraphics[width=0.75\textwidth]{figures/dragana.jpg}
+\end{center}
 \end{slide}
